### 2023-12-27 README.Rmd
- First commit. - Results (draft) - Methodology (draft)

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# sourcearticle

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/jbelmiro/sourcearticle/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): Quarto source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2023-12-27 analysis/paper/paper.qmd
- First commit. - Results (draft) - Methodology (draft)

---
title: "Title Goes Here"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
institute:
  - fosg:
      name: Formatting Open Science Group
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

```{r packages, echo=FALSE}
# Call packages
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(scales)
library(janitor)

```

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

The samples chosen for this study consist of the chert and chalcedony lithic artifacts with individual IDs (excluding buckets) from the Upper Paleolithic levels of the Terrace and Shelter areas of Vale Boi. The sample from the Terrace area (VB-T) consists of 3653 from levels 8 to 4, while the sample from the Shelter area (VBS) consists of 1652 from levels C to A. These two areas were chosen since both have detailed spatial information (as mentioned in section *IN*), several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal and geoarchaeological data.

## Geo-archaeological surveys

The geological data and comparative collection from local and regional cherts used in this study were obtained through previous prospections and petrographic works (\*\*), located at the Lusolit lithotheque (hosted at ICArEHB, University of Algarve). These samples also included cherts collected from previous prospections from other areas of Portugal (such as central Portugal). Hand samples were used throughout the macroscopic characterization for comparison with the archaeological materials, and the archaeological thin sections were compared with the previously studied geological chert thin sections. The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz was also visited, to understand the macroscopic variability from the Cadiz local and regional cherts, to discern between local and regional chert types and possible long-distance cherts.

## Macroscopic characterization

We applied a macroscopic analysis to both Terrace and Shelter samples, which was divided into two steps. The first step included a preliminary characterization using a hand lens of 10x magnification. This step allowed us to create groups and subgroups based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these include characteristics such as color, translucency, and feel; a data dictionary for the recorded variables can be found in the Supplementary Materials (\*\*).

In the second step, we used a stereomicroscope () to observe each sample in further detail, which allowed us to better characterize the artifacts (especially regarding inclusions and alterations), and the previously established groups. The data dictionary used for the individual samples can also be found in Supplementary Materials (\*\*). Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis is simplified and focused on collecting data related to weight, cortex, and alterations.

## Petrographic characterization

To better characterize the different chert groups identified in the macroscopic analysis, we analyzed ++ thin sections. The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels. For groups which showed macroscopic variability, several thin sections were analysed to better characterize these cherts. @tbl-TS-table Petrographic thin sections were analyzed by polarized light microscopy () to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations. A total of 20 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory. The archaeological samples chosen for the petrographic study are from the Terrace area, from several levels, and focused on characterizing the macroscopic variability, reflected on the several groups and subgroups (table \*\*).

```{r}
#| label: tbl-TS-table
#| tbl-cap: Thin section list.
#| echo: false

TS_short_list <- read_csv(here::here('analysis/data/raw_data/TS_short_list.csv'))

knitr::kable(TS_short_list)

```

# Results

```{r}
#| label: db-setup
#| echo: false

# Terrace DB preparation
# Read DBs
T_RM_data <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_TERRACE.csv')) # Terrace RM analysis
context <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_context.csv')) # Terrace Field DB - lithic artifact info
xyz <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_xyz.csv')) # Terrace Field DB - just coordinates

# Df cleaning and uniformizing
cols_to_concat <- c("UNIT", "ID") # Vector to concatenate Unit and ID in one single variable (e.g. A6-100) - to use with both Field databases
context$ID <- str_squish(context$ID) # Remove empty spaces from ID field
xyz$ID <- str_squish(xyz$ID) # Remove empty spaces from UNIT field

# Concatenate UNIT and ID variables in one single variable (e.g. A6-100)
context <- context %>%
  # dplyr::rename("Unit" = "j") %>% # Renaming first column
  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
  select(ID, Level, Spit, Code) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

xyz <- xyz %>%
  #dplyr::rename(Unit = UNIT) %>% # Rename to homogenize
  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artifact
  select(ID, X, Y, Z) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

field_data <- full_join(xyz, context, by = "ID") # Join field tables
T_dataset <- left_join(T_RM_data, field_data, by = "ID") # Join field and RM type tables

T_dataset <- T_dataset %>% 
  #filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
  filter(Level !="4B" & Level !="7C" & Level!="8") # Filtering out levels with barely any artifacts

cols_to_concat_spit <- c("Level", "Spit")
RM_by_spit <- T_dataset %>%
  filter(!Spit %in% c("limp_corte", "profile", " ", NA)) %>%
  select(ID, Z, Level, Spit, RMTYPE) %>%
  unite(col='depth', cols_to_concat_spit, sep=".", remove=FALSE) %>% 
  dplyr::group_by(depth) %>%
  dplyr::count(RMTYPE) %>%
  mutate(freq = n / sum(n)*100)

# Mutate level 5 to add phases: top of 5 (4E/5) and bottom of 5 (5), following Belmiro et al. (2020).
phase_dataset <- T_dataset %>% 
  mutate(Level = case_when(
    Level == 5 & Z > 24.1 ~ "4E/5",
    Level == 5 & Z < 24.1 ~ "5",
    TRUE ~ Level))

# Shelter DB preparation
S_RM_data <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
S_field_data <- read_csv(here::here('analysis/data/raw_data/VB_Shelter_xyz.csv')) # Shelter coordinates

S_RM_data <- S_RM_data %>%
  mutate_at("ID", str_replace, "VB", "") %>% # Remove "VB" from ID
  mutate(RMTYPE = case_when(
    RMTYPE == "Type 2" ~ "TYPE 2",
    .default = RMTYPE))

cols_to_concat_shelter <- c("YEAR", "ID") # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)

S_field_data <- S_field_data %>%
  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% # Unite YEAR with Id s«using cols_to_concat
  select(ID, X, Y, Z) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)
S_dataset <- left_join(S_RM_data, S_field_data, by = "ID")

```

```{r}
#| label: t-table

T_table <- phase_dataset %>%
  filter(RMTYPE !="INDET") %>% 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
                                        "TYPE 12", "TYPE 13","TYPE 13B", 
                                        "TYPE 11", "TYPE 11F", "Oolitic", "GB", "TL"))) %>%
  tabyl(RMTYPE, Level) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(2) %>%
  adorn_ns("front")

```

The three main used raw materials at the archaeological site of Vale Boi are quartz, greywacke, and chert. Using the field database with all individually coordinated lithic artifacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. Since some lithic assemblages of the Terrace area are currently under study, the data regarding the percentage of raw materials in the “Other” category was not available for all archaeological levels. @fig-rm-comparison shows that for the selected levels, and using single coordinated artifacts (excluding buckets), chert represents only \~20-15% of the lithic assemblage. Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke. Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site. Frequently, the quartz and greywacke assemblage includes large amounts of shattered and/or unknapped chunks or slabs. Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz. Based on this, despite the low amounts of individually plotted chert artifacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.

```{r}
#| label: tbl-indet-freq
#| tbl-cap: N and percentage of grouped and indeterminate artifacts by level.

indet_table <- phase_dataset %>% 
  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
  tabyl(Group,Level) %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(2) %>%
  adorn_ns("front")

```

As previously mentioned, whenever too altered to confidently attribute a group, the artifacts were instead classified as indeterminable (INDET). As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artifacts, corresponding to 40% of the total sample (n=1027). This is due to the frequent alterations of the sample, especially related with patinas and alteration rinds, pits and fire/heat alterations. For all other levels, indeterminate represents in mean \~28% of the chert artifacts.

```{r}
#| label: fig-rm-comparison
#| fig-cap: "Comparison of chert, chalcedony and other raw materials coordinate in the terrace area of the archaeological site of Vale Boi."
#| echo: false

context_short <- context %>% 
  filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
dataset_short <- T_dataset %>% 
  select(ID, RMTYPE)
T_short <- full_join(context_short, dataset_short, by = "ID")
T_short["RMTYPE"][is.na(T_short["RMTYPE"])] <- "Other"

T_short <- T_short %>% 
 mutate(Rawmaterial = case_when(
    RMTYPE == "Other" ~ "Other",
    RMTYPE == "Chalcedony" ~ "Chalcedony",
    RMTYPE != "Other" | RMTYPE != "Chalcedony" ~ "Chert"))

#T_short_table <- T_short %>% 
#   tabyl(Rawmaterial, Level) %>%
# adorn_totals(c("row", "col")) %>%
#  adorn_percentages("col") %>%
#  adorn_pct_formatting(2) %>%
#  adorn_ns("front")

T_short %>%
  dplyr::group_by(Level, Rawmaterial) %>% 
  dplyr::summarise(n = n()) %>%
  mutate(percent = n/sum(n)) %>% 
  ggplot(aes(x=Level, y = percent, fill = Rawmaterial)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() + #type of theme
  labs(x = "Level", y = NULL) +# change labels
  guides(fill=guide_legend(title=NULL)) + # remove legend title
  geom_text(aes(label=ifelse(percent >= 0.07, paste0(sprintf("%.1f", percent*100),"%"),"")),
                position=position_stack(vjust=0.5), colour="white", size=3) +
  theme(axis.title.x = element_text(size = 10, vjust = -1))

```

Through the macroscopic and petrographic analysis we identified 24 types of chert and three types of chalcedony. Of these, one type of chalcedony and 9 types of chert are composed of several samples (\>10) and present across both areas (Terrace and Shelter) and throughout the Upper Paleolithic occupations of the Terrace area. Thin sections were produced for the lithics of the latter groups. One type had several samples (\>10; Type GB) but was only identified in the Shelter area, while 14 varieties of chert and 2 chalcedony are composed of one to three samples and restricted to specific levels. The latter varieties were collapsed into a category named TL (Trace lithotypes).

We present here a short description of each identified macroscopic and petrographic type. A detailed macroscopic and petrographic description and figures for each type and sub-type of chert and chalcedony can be found in the SOM (++). The petrographic analysis PDF files for all thin sections studied in this paper (including macroscopic and microscopic figures) can also be found in our GitHub page (++) as well as the OSF repository (++).

## Type description

The chalcedony is a massive, fibrous chalcedony mudstone. It is characterized by a colorless and translucent appearance, often partially or completely white patinated. It has good to medium quality, due to frequent large crystallizations, irregularities or fractures which impact the knapping. In thin section, the chalcedony is composed by fibrous chalcedony, without allochems or bioclasts, and no porosity (\<1%). Radiating spherulitic textures are very frequent.

The first chert groups identified at Vale Boi are massive micro-cryptocrystalline quartzes from a marine environment. Petrographically, these cherts are homogeneous, except for a small amount of artifacts with a banded variety within type 12. These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite. Group 12 shows higher percentages of micrite/sparite. In all samples, fossils range from common to very frequent, albeit frequently unidentifiable. When identifiable in thin-section, they are sponge spicules, with the rare occurrence of possible foraminifera and ostracod in a sample from type 12. Porosity varies between 10% and 1%, of vug type. Macroscopically however, these groups show variability, which resulted in their separation in 4 different macroscopic groups; this separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.

Group 2 is characterized by single-color and multiple-color samples, ranging between yellow, grayish brown, pale red, and white, and with variable patterns such as splotched, broad mottling, speckling, and laminations. They are opaque, with sub-translucent areas mainly focused in the white areas of the sample. Whenever present, the cortex can be rounded or irregular, varying between thin to thick. Inclusions include black speckles and red speckles, which are probably organics and iron oxides. Bioclasts are frequently unidentifiable, but also include monaxon sponge spicules, bivalve shells replaced by quartz or chalcedony, and rarely triaxon sponge spicules. In the shelter area, the variety of identified fossils also includes possible uncommon ostracods, a possible small foraminifer, and echinoderm spines.

Group 15 is characterized by light gray and reddish gray colors and varies between opaque to sub-translucent. It is characterized by the presence of splotches (often coarser and sub-translucent), speckling, and fine lamination. Whenever present, the cortex is white, rounded, and of medium thickness. Allochems such as iron oxides or opaques are also visible macroscopically as black speckling or rare red speckling. The bioclasts are frequently unidentifiable, due to poor preservation and are replaced by microcrystalline quartz or fibrous chalcedony, except for uncommon sponge spicules.

Type 7 shows a heterogeneous structure with two types of fabrics. The interior of the samples and the nodules is gray, and sub-translucent with a single mica mineral, flecks, and speckles. The outer edge and within the grey fabric as splotches or broad mottling are white, opaque, and with yellow speckles. The cortex is white, with a sharp transition and with a medium to thick width. The morphology of the cortex in fully cortical samples is originally available as irregular (slightly oval) nodules within the parent rock. Whenever identifiable, bioclasts are limited to sponge spicules.

Type 12 chert has a conchoidal fracture (albeit uneven at times) and acceptable to low quality due to the lack of homogeneity and frequent fractures, as well as a coarser fabric in certain areas. It has gray colors and ranges from opaque to sub-translucent. The group displays frequent patterns: splotched, banded, broad mottling, and speckling. The cortex is from an outcrop, frequently rounded, but also irregular. Based on reference samples, the chert seems to be enclosed within the parent rock as irregular or tabular nodules). The cortex is thin, with a sharp transition, and some artifacts show chunks of the parent rock attached to the cortex. Bioclasts are uncommon, and when identifiable are sponge spicules.

Group 11 corresponds to a massive cryptocrystalline quartz with a mudstone texture. The lack of local similar cherts, as well as identifiable fossils does not allow us to attribute a depositional environment or epoch/stage. However, rare spicules were identified; given the exclusivity of the sponge’s (Porifera) living environment to be marine (Flügel, 2010), group 11 may be considered a chert formed in a marine environment. This chert is translucent and shows a variety of colors, ranging from brown, red, and gray, which seem to vary in function of the artifact’s thickness. Patterns are rare, however, whenever present they are laminated, with flecks, and splotches. In unaltered samples, bioclasts are rare. The cortex is rounded, and rarely pebble-like, ranging from thin to medium thickness. Two sub-groups were individualized based on macroscopic characteristics such as color, translucency variation, and patterns, although they may simply reflect the use of specific nodules from within Type 11’s variability. Five thin sections were produced to better characterize the macroscopic variability of this group and sub-groups. The petrographic analysis shows the samples are characterized by a mudstone texture and massive microstructure. It is mainly composed of cryptocrystalline quartz (between 95-97% in the samples), with accessory fibrous chalcedony found replacing fossils or filling fractures, and in some samples macrocrystalline quartz, microcrystalline quartz, and micrite/sparite. Allochems are opaques (u), iron oxides (u), and bioclasts (r/u). Bioclasts are unidentifiable and poorly preserved, replaced by fibrous chalcedony and rarely by microcrystalline quartz. Sponge spicules were identified in a single sample. Porosity is present between \<1-1%, of vug-type when identifiable.

Group 11F corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment. Identified and well-preserved fossils are foraminifera from the Pfenderenidae family, which are larger benthic foraminifera from marine environments. The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous. Macroscopically, this group is characterized by gray colors and a white area, frequently at the center of the sample, which seems to be a weathering rind and where the presence of fossils is highlighted. The translucency ranges from sub-translucent to opaque and visible patterns are splotches, broad mottling, spots, and speckling. Whenever present, the cortex is rounded and thin. Petrographically, this group is a packstone (mudstone texture close to the edges) with a massive microstructure. It is composed of 97% micro-cryptocrystalline quartz, macrocrystalline quartz, and fibrous chalcedony, found mostly replacing fossils. Other minerals include uncommon iron oxides. The porosity is of vug type (3%). Fossil content is very frequent, albeit mostly unidentifiable and poorly preserved. Identified fossils are rare sponge spicules , rare ostracods with some degree of preservation, and common small, unidentified foraminifera (100 μm) of differing preservation degrees. Identifiable foraminifera are rare and from the Pfenderenidae family (500 μm).

Group 13 is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids. It is characterized by two colors, white (opaque) and grey (sub-translucent), with splotches, speckles and horizontally dotted laminations. The cortex is from an outcrop, rounded, and varying between medium thickness to thick. This group shows several facies, with different types and concentrations of allochems and bioclasts, of which identifiable are peloids, ooids, and sponge spicules. A detailed scheme of the facies can be found in the SOM (++). The petrography results show this chert has a massive microstructure and packstone texture. It is composed mainly of micro-cryptocrystalline quartz (80%), fibrous chalcedony (10%), and micrite (10%). Allochems are opaques, iron oxides, ooids and peloids. The ooids are poorly preserved and only uncommonly show concentric lamellae in plane polarized light. Unlike macroscopical observations, bioclasts under the thin section are all unidentifiable, although rare fossils may be bivalve shells. Porosity is vug (10%).

Group 13B corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone. It is opaque and has a heterogeneous structure and is characterized by a variable color distribution: it shows a horizontal laminated and finely laminated pattern, with gray dark gray bands intercalated with light gray bands. Whenever present, the cortex is rounded (from an outcrop source) and is generally thin. In specific samples, the cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers. Petrographically this chert is a banded packstone (banding of sedimentary origin), composed mostly of micro-cryptocrystalline quartz (48%), micrite (40%), dolomite (10%), and accessory fibrous chalcedony and mica (muscovite). The allochems are peloids, unidentifiable bioclasts, iron oxides, and opaque minerals. Porosity is low (1%).

Group Oolitic is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids. Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: brown (opaque) and black (sub-translucent) fabric. The quality is acceptable since it has fine granularity and a smooth feel, but shows natural fractures and occasional inclusions, such as macrocrystalline quartz or chalcedony veins that may impact the knapping. The cortex is difficult to identify since it is similar to the brown fabric, highly altered, and with a gradual transition. Petrographically this group is a packstone with a massive microstructure. The chosen sample for the thin section is composed of 35% micro-cryptocrystalline quartz, 53% micrite, 9% macrocrystalline quartz, and accessory fibrous chalcedony (filling fractures) and mica; the percentages of quartz and micrite might be related to the fabrics present in the artifacts. Allochems are common iron oxides and very frequent ooids. Bioclasts are rare, poorly preserved, and replaced by quartz. The porosity is intraparticle and vug (10%). The ooids are distributed homogeneously across the sample (although clearly macroscopically visible in the brown fabric) and are highly abundant and concentrated. They are poorly sorted, varying between 500 and 20 μm. They show round-to-elliptical shapes, and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010], pp. 144\].

Finally, group TL is characterized by several samples with differing characteristics from the established larger groups. Based on their different characteristics they have been interpreted as exotic or long-distance raw materials. A detailed description of these types can also be found in the SOM (++). TL01 is a jasper, with a massive microstructure and red coloration, with frequent fractures filled with colorless, translucent fibrous chalcedony. TL02 is an orange chalcedony of good quality, differing from the group of white/translucent chalcedony due to its color. TL03 is a massive chert with uncommon spots. TL04 is a yellow massive chert, with inclusions and fractures filled with fibrous chalcedony. TL05 is a grey, banded/massive chert, where the banding seems to be of ++ origin. TL06 is a sub-translucent, oolitic packstone chert. Unlike the oolitic cherts in the previous groups, this chert shows well-sorted and small (200-220 μm), round ooids with differentiated walls and cores. TL07 is characterized by a laminated dendritic pattern, with uncommon inclusions. TL08 is a translucent and colorless, massive chert, without inclusions. TL09 is a translucent chalcedony with a massive microstructure and packstone texture. It is characterized by unidentified, moss-like orange/red inclusions, and no other allochems or bioclasts. TL10 is a massive chert with frequent iron oxide speckling or banding. TL11 is a banded, translucent chert with no inclusions. TL12 is a massive oolitic packstone chert, white and translucent. The ooids are not visible under the naked eyes, and can only be identified under the stereomicroscope. TL13 is a translucent, banded chert without inclusions. TL14 is a massive chert with large benthic foraminifera (1-2 mm). TL15 is a massive brown chert with common iron oxides and unidentifiable fossils. And finally, TL 16 is a radiolarite with unidentifiable fossils.

## Source attribution

Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 15, 7 and 12 to be local, as they are congruent with the described characteristics. Groups 2 and 15 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils. The macroscopic similarities of Groups 7 and 12 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, as mostly located in the Ferrel area. As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape. This is especially relevant, since the Ferrel outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiro2023].

The similarities between chert artifacts at the UP occupations of Vale Boi and the chert from the Rio Maior region (Estremadura Portuguesa) have been previously suggested although without systematic studies to ascertain the attribution.

These Cretaceous cherts, found in secondary deposition settings, are described as frequently translucent, with geodes, with colors ranging from yellow to red or grey, and mineralogically homogeneous between them. Petrographic studies highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matias2016] . When identified, sponge spicules and possible rare foraminifera have been identified [@matias2012]. These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from central Portugal hosted at LusoLit.

To aid on the comparison, we compared the lithics attributed to Type 11 to hand-samples from the LusoLit lithotheque, previously collected from the Rio Maior area. Similarly, three thin sections were produced of geological samples with different macroscopic characteristics from the Central Portugal area to compare with the archaeological thin sections. Alike the archaeological thin sections from Type 11 and congruent with other studies [@matias2016], these thin sections show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%). From these cherts, only the dark grey nodules (RT231) showed identifiable fossils which were rare ostracods.

Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Group 11 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts from the Estremaduran area in central Portugal.

The peloidal/oolitic packstone cherts are not congruent with the cherts from the Algarve area, making them not local or regional. The consultation of the lithotheque hosted at UCA allowed us to identify a sub-set of cherts with similarities to those from Type 13. These cherts are massive, sub-translucent, peloidal/oolitic packstones. Allochems include very frequent peloids and ooids. The peloids are densely arranged in the samples, more or less visible depending on the alterations to the surface or the thicknes. The ooids are poorly sorted, with oval or round shapes and replaced by quartz. Uncommonly the ooids show a preserved nucleus. The presence of bioclasts varies between samples, randing between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon) and echinoderm spines (rare). @fig-type13-source exemplifies several of the facies identified in Type 13 cherts from the Terrace and Shelter areas (left) and matching samples from the UCA lithotheque (right). Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allows us to interpret Type 13 as coming from the Andalusian region of southern Spain.

Group 13B is a small group but present throughout the whole stratigraphy of the Terrace and in the Solutrean occupation of the Shelter, and with significant different characteristics from the local cherts of the Algarve region. Although its origin is unknown, when visiting the Cadiz lithotheque, we identified a small group of banded black cherts without noticeable fossils or inclusions. Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Group 13B belongs to Andalusian chert formations in southern Spain.

From the identified groups, only group 11F and Oolitic remain unknown. Despite the macroscopic similarities between type 11F and cherts from the Andalusian chert formations, the samples inspected with a stereomicroscope showed no visible foraminifera, especially from the ++ family. However, the archaeological thin section revealed that despite no foraminifera being visible macroscopically, they are visible under the microscope in thin section. The fact that no foraminifera are visible in the the small number of analyzed samples from the Cadiz lithotheque may simply be a result of their uncommon visibility. Despite this, given the lack of this specific fossil in the reference samples, the probable source of Group 11F remains unknown. Similarly, the black oolitic chert shows no similarities to any consulted reference material. Despite the existence of oolitic limestones in the Algarve region, no chert nodules are mentioned (++) and the visit of outcrops mentioned in the geological cartography did not allow their identification; the trade of oolitic cherts in recent prehistory may indicate however that these types of cherts are available in the landscape in long-distance (++), and their presence in Vale Boi may simply reflect similar mobility patterns.

Apart from the groups common throughout the stratigraphy of the Terrace and Shelter, group GB and samples from the TL group also show no identifiable source. The exceptions are TL01, TL10-11 and TL15-16. Jasper is known in the SPZ area, from the Alentejo region to Spain, above the Algarve. Although no petrographic data is available, the macroscopic similarities between TL01 and a hand-sample hosted at Luso-lit seem to indicate that the source of this jasper is the SPZ formations. Similarly, despite different from the frequent Cretaceous nodules, the macroscopic similarities between TL10 and TL11 and specific nodules from Central Portugal in the Luso-lit lithotheque may indicate these cherts are from the Estremadura area. This difference in appearance may be due to the already mentioned variability of the nodule's macroscopic characteristics. Finally, TL15 and TL16 show macroscopic similarities to radiolarite reference samples located at the Cadiz lithotheque, which may indicate the source of these cherts is from Andalusian geological formations.

## Stratigraphical distribution

Chert groups are present across the stratigraphy (except for GB which appears only in level 4 and in a small number of artifacts; n=2). Different chert types were consistently used in the Terrace area throughout time, with different intensities of use. @fig-type-per-level shows the percentage of chalcedony and chert types used throughout the stratigraphy in the Terrace area of Vale Boi; this data excludes the unidentifiable and altered samples, categorized in a group named INDET. In the Solutrean and Proto-solutrean levels, local cherts make up more than \~70% of the chert. Despite this homogeneous pattern, the use of group 2 and group 7 marks significant differences between the Solutrean occupations (levels 4, 4C, and 4D) and the Proto-solutrean occupations (4E and 4E/5); in the first group the yellow/brown cherts (type 2) represent \~30% of local chert use, while in the second group, the use of type 2 diminishes drastically (\~10-15%), while Type 7 becomes the most used chert with \~38% in level 4E and \~23% in the top of level 5. For both occupations, chert non-congruent with the known local sources represent percentages lower than 30%. This pattern seems to shift in the bottom of 5 and throughout the Gravettian occupation; the use of chert local cherts drops to 45-60%, while the percentages of specific non-local chert types increase significantly, such as Type 11 and Type 13. Other non-local cherts such as the Oolitic cherts and TL categories also show smaller increases in the Gravettian occupation. As seen in previous studies, chalcedony seems to be present but always in small frequencies. Despite this, there is a small increase in the use of chalcedony in level 5 (both top and bottom), from lower than 5% in level 4E to 13.5% in top of 5, with a tendency to decrease in the older levels.

```{r}
#| label: fig-type-per-level
#| fig-cap: "Chalcedony and chert types per level, without unidentifiable samples. Only percentages superior to the 5% threshold are presented in the figure."
#| echo: false

phase_dataset %>%
  filter(RMTYPE != "INDET") %>% 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
                                        "TYPE 12", "TYPE 13","TYPE 13B", 
                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
  dplyr::group_by(Level, RMTYPE) %>% 
  dplyr::summarise(n = n()) %>%
  mutate(percent = n/sum(n)) %>% 
  ggplot(aes(x=Level, y = percent, fill = RMTYPE)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() + #type of theme
  labs(x = "Level", y = NULL) +# change labels
  guides(fill=guide_legend(title=NULL)) + # remove legend title
  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
                position=position_stack(vjust=0.5), colour="white", size=3) +
  theme(axis.title.x = element_text(size = 10, vjust = -1)) %>% 
  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))

T_dataset %>%
  filter(Z != "0") %>% 
  filter(!Z >= "25") %>% 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
                                        "TYPE 12", "TYPE 13","TYPE 13B", 
                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
  filter(RMTYPE != "INDET") %>% 
  ggplot(aes(RMTYPE, Z, color = RMTYPE)) +
  geom_violin(trim = FALSE, scale="width", bw=0.1) + 
  labs(title="Violin plot", 
       subtitle="City Mileage vs Class of vehicle",
       caption="Source: mpg",
       x="Class of Vehicle",
       y="City Mileage") +
  stat_summary(fun.data=mean_sdl, geom="errorbar", color="black", width=0.2)
  ggsave("violin_plot.png")

T_dataset %>% 
  filter(Z != "0") %>% 
  filter(!Z >= "25") %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
                                        "TYPE 12", "TYPE 13","TYPE 13B", 
                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
  dplyr::group_by(Z, RMTYPE) %>% 
  dplyr::summarise(n = n()) %>%
  ggplot(aes(x = n, y = Z)) +
  geom_point(alpha = 0.1, color = "coral") +
  facet_grid(cols = vars(RMTYPE)) +
  labs(x = "Count", y = "Depth (cm)") +
   theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

The same raw materials found in the Terrace area are also found in the Solutrean occupation of the Shelter, albeit with some differences. Although with similar percentages to the local cherts, the use of local chert in the Shelter area is heavily dominated by the brown/yellow cherts that frequently characterize the Lower Jurassic outcrops of the Algarve; although the other macroscopic types are present but not significant. The presence of cherts not congruent with the local sources is, alike level 4, close to 25%, although dominated mostly by Type 11. Type GB is present in higher numbers than in level 4, however still representing less than 5% of the assemblage.

```{r}
#| label: fig-type-per-level_shelter
#| fig-cap: "Chalcedony and chert types per level, without unidentifiable samples. Only percentages superior to the 5% threshold are presented in the figure."
#| echo: false

S_dataset %>%
  filter(RMTYPE != "INDET") %>% 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
                                        "TYPE 12", "TYPE 13","TYPE 13B", 
                                        "TYPE 11", "TYPE 11F","TYPE GB", "Oolitic", "TL"))) %>%
  dplyr::group_by(AREA, RMTYPE) %>% 
  dplyr::summarise(n = n()) %>%
  mutate(percent = n/sum(n)) %>% 
  ggplot(aes(x = AREA, y = percent, fill = RMTYPE)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() + #type of theme
  labs(x = "Level", y = NULL) +# change labels
  guides(fill=guide_legend(title=NULL)) + # remove legend title
  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
                position=position_stack(vjust=0.5), colour="white", size=3) +
  theme(axis.title.x = element_text(size = 10, vjust = -1)) %>% 
  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))

```

```{r}
#| label: fig-type13-source
#| fig-cap: Comparison between chert samples of different facies from Type 13 of the Terrace area (left) and samples from the Universidad de Cádiz lithotheque.

knitr::include_graphics(here::here("analysis/figures/Terrace_UCA_comparison.png"))

```

```{r}
#| label: tbl-TL
#| tbl-cap: A table of TL

TL_table_T <- phase_dataset %>% 
  filter(RMTYPE=="TL") %>% 
  select("Level", "SUBTYPE")
TL_table_S <- S_dataset %>% 
  filter(RMTYPE=="TL") %>% 
  mutate_at("LAYER", ~replace_na(.,"Shelter")) %>% 
  select("LAYER", "SUBTYPE") %>% 
  rename(Level = LAYER)

TL_table <- rbind(TL_table_T, TL_table_S)
TL_table <- TL_table %>% 
  tabyl(SUBTYPE, Level) %>% 
  adorn_totals(c("row","col"))

```

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2023-12-27 analysis/supplementary-materials/SOM_lithotypes.qmd
- First commit. - Results (draft) - Methodology (draft)

---
title: "Terrace and shelter lithotype complete description"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
title-block-published: "Last updated"  
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: ../paper/references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  out.width = "100%",
  fig.align = "center"
)
```

# Chert group description

## Group 2

Group 2 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment. The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Pliensbachian.

It is a fine-grained chert, with a conchoidal fracture and good quality (see the Data dictionary for macroscopic analysis document for details on the attribution of quality to raw materials used in this analysis). It is macroscopically heterogeneous in color, patterns, and textural characteristics; due to this, initially, two sub-groups were created: single-color @fig-type2-variability and multiple-color () groups.

The single-color group's main colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), grayish brown (10YR 5/2) and white (2.5Y 8/1). The color variability and presence of allochems and bioclasts produce macroscopic patterns such as splotched, speckling, and laminations. The structure is homogeneous. The artifacts have medium to dull luster; some samples show shiny luster, however, this may be related to alterations to the surface. Translucency is opaque and the feel is smooth to semi-smooth. The multiple-color group shows a variable color distribution: colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), pale red (2.5YR 6/2), and white (2.5Y 8/1). The color variability creates macroscopic patterns such as splotched, broad mottling, speckling, and laminations (Fig. \@ref(fig:type2Bvariability)). The structure is similar to group single-color, although slightly heterogeneous as translucency varies from opaque to sub-translucent, with light passing through at around 3.5 mm, limited to the white areas.

For both sub-groups, whenever present, the cortex can be rounded or irregular and from an outcrop, varying between ++ to ++ mm in thickness, frequently with a sharp transition.

Inclusions include black speckles, which may be organics or iron oxides, and red speckles which may be iron oxides. Bioclasts include white, round speckles, spicule-shaped skeletal remains, monaxon sponge spicules, bivalve shells (rare, hereon r) replaced by quartz or chalcedony, and rarely triaxon sponge spicules. In the single-color group fossil remains range from uncommon to frequent, while in the multiple-color group fossils range from uncommon to very frequent; in all cases, the bioclasts are frequently unidentifiable.

In the shelter area, the variety of identified fossils also includes possible ostracods (uncommon, hereon u), a possible small foraminifer (Fig. \@ref(fig:shelter2fossils) d), and echinoderm spines (Fig. \@ref(fig:shelter2fossils) a-c).

Five thin sections were produced to account for the macroscopic variability of this group. All thin sections show a wackestone texture and massive microstructure. Micro-cryptocrystalline quartz represents more than 90% of the orthochem composition (between 93 to 99%), while other orthochems are present in lower frequencies: macrocrystalline quartz (1-5%) as grains scattered in the sample or replacing fossils and fibrous chalcedony (1-5%) replacing fossils. Accessory minerals are also present in some samples: dolomite, micrite, and mica (muscovite). Allochems present are oxide patina and grains (common, hereon c), unidentifiable opaques (common to very frequent, hereon vf), and bioclasts (vf). All samples have very frequent unidentifiable fossils. In samples from multiple-color group, although in lesser frequencies, monaxon sponge spicules (H21-3095) and a possible triaxone sponge spicule (H21-4234) are identifiable. All samples show vug-type porosity between 1-5%.

```{r}
#| label: fig-type2-variability
#| fig-cap: Chert macroscopic variability within group 2 (single color).
#| fig-width: 8

knitr::include_graphics("../figures/SOM/Group2_macromeso_b.png")

```

```{r type2Bvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 2 (multiple color).", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group2B_macromeso.png")

```

```{r shelter2fossils, echo=FALSE, fig.cap="Identified fossil variability from group 2 from the Shelter area. Fossils present in the figures are: a-b, crinoid stem section with 4 series; c, cross-section of crinoid columnal with pentagonal stem plate; d, possible foraminifer (black arrow).", out.width = '80%'}

knitr::include_graphics("../figures/SOM/SHELTER2_FOSSILS.png")

```

A group of samples was individualized from group 2 due to color and textural differences (named 2R during the analysis process). These are interpreted as alterations to group 2 cherts, possibly due to heat, although further experimentation is needed. The color distribution is variable and it has a heterogeneous structure, possibly due to the alterations. The colors are red (10R 5/6), light yellowish brown (10YR 6/4), and pinkish-gray (5YR 7/2). It shows shiny to dull luster and semi-smooth to rough feel.

Whenever present, cortex seems to be from an outcrop, rounded, with thicknesses ranging from 3.8 mm to 0.5 mm, with a sharp transition.

Inclusions seem to be iron oxides, which create speckling patterns. Bioclasts include 1) frequently poorly-preserved spicules, mostly monaxon, although some tetraxon spicules may be present; 2) white, round fossils; 3) uncommon, unidentifiable fossils replaced by chalcedony or quartz. In general, the alterations seem to exacerbate the visibility of the iron oxides and skeletal remains.

```{r type2Rvariabilitymeso, echo=FALSE, fig.cap="Archaeological samples attributed to Type 2R. ", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group2R_macromeso-01.png")

```

A thin section was made to confirm whether the color and textural differences are solely macroscopic, and do not represent a different petrographic type. The sample (H19-2426 from Level 4D) shows no outstanding differences from the samples from group 2.

## Group 15

Group 15 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment. The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Pliensbachian.

Group 15 is a fine to medium chert, with a conchoidal fracture and good quality. The color distribution is uniform, despite subtle changes to gray. The colors are light gray (10R 7/1) and reddish gray (10YR 6/1). It has medium to dull luster and is sub-translucent to opaque, with light passing through only in the very fine parts (\~1.4 mm). The splotches and lamination adjacent to the cortex seem to be sub-translucent as well, showing more translucency than the chert itself. The feel is semi-smooth and the patterns are splotches, speckling, concentric lamination (limited to a small number of samples), and fine lamination (found only between the cortex and the chert).

Whenever present, the cortex seems to be from an outcrop and rounded. Despite the brown coloration possibly due to post-depositional alterations, the original color is white. Cortex thickness ranges from 3.8 mm to 1.2 mm, and has a sharp transition. One single sample from Level 6 shows a gradual cortex transition. Although the original morphology of this chert is unknown, the presence of cortex on two opposed extremities in a sample (H19-3196, \@ref(fig:type15variability) a) indicates the chert can be found in irregularly rounded nodules within the parent rock.

Petrographically, this chert is a massive wackestone, composed mostly of micro-cryptocrystalline quartz (96%) with accessory macrocrystalline quartz and fibrous chalcedony. Allochems are iron oxides (vf), unidentified opaques (u) and bioclasts. Porosity is low (\<1%).

Allochems such as iron oxides or opaques are also visible macroscopically as black speckling or rare red speckling. The bioclasts are unidentifiable, due to poor preservation and are replaced by microcrystalline quartz or fibrous chalcedony. The common round unidentified fossils can be seen only under the stereomicroscope and in the thin section. The uncommon long, spicule-like fossils can be seen both macroscopically and in the thin section. The presence of sponge spicules, however, is more obvious in artifacts from Level 4 and 4D.

```{r type15variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 15.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group15_macromeso.png")
```

## Group 7

Group 7 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment. The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin locate their epoch and stage in the Lower Jurassic, Pliensbachian.

This chert is fine-grained, with a conchoidal fracture and good quality. It is macroscopically homogeneous throughout the archaeological layers, although layer 4 shows less homogeneity between the artifacts, possibly due to post-depositional alterations.

Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: grey (G) and white (W) fabric. The grey fabric corresponds to the interior of the nodule. It is light gray (5Y 7/1, 7/2), light olive gray (5Y 6/2), and gray (5Y 6/1, 5/1), sub-translucent (light passing through in the thin areas, around 3.2 mm thickness) and has medium luster. The visible inclusions are limited to a single mica mineral (\~50 μm), white flecks (concentrated near the outer edges of white fabric), and speckles, which may be circular, unidentifiable fossil remains, and rare spicule-like remains. In layer 6 spicule-shaped remains are uncommon. The white fabric appears in two areas in the nodules: 1) in the outer edge, surrounding the grey fabric; 2) within the grey fabric, where it creates splotched and broad mottling patterns. This fabric is white (5Y 8/1, 10YR 8/1) and very pale brown (10YR 8/2, 8/4), opaque and dull. In the pale brown splotches, inclusions are yellow speckles (possibly iron oxides, Fig. \@ref(fig:type7variability) a). The white fabric generally has barely visible fossils and uncommon unidentifiable fossils replaced by chalcedony or quartz.

Whenever present, the cortex seems to be from an outcrop, mostly rounded, although a sample from layer 5 shows a thick (30 mm) irregular cortex. The cortex is white or light brown with a sharp transition between 1.4 mm to 10 mm thick. The morphology of the cortex in fully cortical samples (i.e., I18-2099; \@ref(fig:type7variability) d) may indicate that this chert is originally available as irregular (slightly oval) nodules within the parent rock.

Petrography results show this group is a massive wackestone composed mostly of micro-cryptocrystalline quartz (98%), with accessory macrocrystalline quartz and fibrous chalcedony. Allochems are iron oxides (c), opaques (vf), and bioclasts (c). Despite the macroscopically visible spicule-like white shapes, these fossils are not apparent in thin section. The porosity is low (1%).

```{r type7variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 7.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group7_macromeso.png")

```

## Group 12

Group 12 corresponds to a massive and banded micro-cryptocrystalline quartz from a marine environment and possibly a shelf depositional environment. The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin, as well as the large size of the nodules and presence of parent rock, allow us to interpret this chert as local, possibly the Lower Jurassic although the exact outcrop is not identifiable.

This chert is fine to medium-grained, with a conchoidal fracture (albeit uneven at times). The quality is acceptable to low due to the lack of homogeneity and frequent fractures, as well as a coarser fabric in certain areas. It has a heterogeneous structure, frequently showing different characteristics within one single artifact. The colors are gray (5Y 6/1), light brownish gray (10YR 6/2), and pale brown (10YR 6/3). Fresh surfaces display slightly different colors: light gray (2.5Y 7/1) and gray (2.5Y 6/1). It has medium to dull luster, coarse to medium feel, and is opaque to sub-translucent; in the sub-translucent areas, light passes through to max. 4 mm of thickness. The group displays frequent patterns, mostly related to this heterogeneity, but also due to patination; these are splotched, banded, broad mottling, and speckling.

The cortex is from an outcrop, frequently rounded, but also irregular. Based on reference samples, the chert seems to be enclosed within the parent rock as irregular or tabular nodules (Fig. \@ref(fig:type12variability) b). The cortex is thin (\~0.5 mm), with a sharp transition, and some artifacts show chunks of the parent rock attached to the cortex (Fig. \@ref(fig:type12variability) a).

Based on the presence of bands, two sub-groups were created: 12 (Fig. \@ref(fig:type12variability) a-c) and 12B (Fig. \@ref(fig:type12variability) d), the latter characterized by banding. Two thin sections were produced to characterize the macroscopic variability. The petrographic results show that group 12 is a wackestone, with a massive (group 12) or banded (group 12B) microstructure. They are mainly composed of micro-cryptocrystalline quartz (75% and 72%, respectively) and micrite (20% in both), and in smaller percentages macrocrystalline quartz and fibrous chalcedony. Allochems are unidentified opaques (vf), iron oxides (u), and bioclasts (c). The sub-groups show larger differences in the porosity, with 12 showing 1% vug porosity while 12B shows 10% vug and moldic porosity, concentrated in the banded areas with more opaques.

Bioclasts are uncommon both macroscopically and in thin section: sponge spicules and round, unidentifiable fossils can be seen by the naked eye or under the microscope. In thin section however, several other unidentified fossils are present, some of which may be an ostracod (both in 12 and 12B) and a foraminifer (12).

```{r type12variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 12.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group12_macromeso.png")

```

## Group 11

Group 11 corresponds to a massive cryptocrystalline quartz with mudstone texture. The lack of local similar cherts, as well as identifiable fossils does not allow us to attribute a depositional environment or epoch/stage. However, rare spicules were identified; given the exclusivity of sponge's (Porifera) living environment to be marine [@flugelMicrofaciesCarbonateRocks2010], group 11 may be considered a chert formed in a marine environment.

This chert is fine-grained and translucent (light passing through at 17 mm of thickness), with a conchoidal fracture and good quality. The structure is heterogeneous, with shiny to medium luster and semi-smooth to smooth feel. This group of translucent cherts shows a variety of colors: grayish brown (2.5Y 5/2), light brownish gray (10YR 6/2), brown (7.5 YR 4/2), reddish brown (2.5YR 4/3) and dark gray (10YR 4/1). The color distribution is variable; in some cases, the color seems to be related to the thickness of the sample; thicker parts show darker colors, while thinner parts show lighter colors. Alterations and inclusions also impact the color distribution and the discernment of natural versus the altered color. Alterations such as patina also seem to impact the translucency, which becomes opaque, and the luster, which becomes medium to dull.

Patterns are rare, however, whenever present they are laminated, with flecks, and splotches. The flecks and splotches are characterized by a semi-smooth feel, sub-translucency, and lighter color than the chert.

In unaltered samples, bioclasts are rare.

The cortex is rounded, from an outcrop source, and rarely rounded and smooth like a pebble. It is frequently around 0.5 to ++ mm thick and the transition is gradual.

```{r typeCPTvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 11.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/GroupCPT_macromeso.png")

```

### Sub-group LLB

A sub-group named LLB (corresponding to the light brownish gray colors) was created during the macroscopic analysis process due to the presence of a group of artifacts with homogeneous characteristics (and refits) within the previous group.

This sub-group is characterized by white (10YR 8/1), light brownish gray (10YR 6/2), and brown (7.5YR 5/2) colors. It has a heterogeneous structure, with two fabrics: white and brown. The white fabric is located by the cortex and is opaque. It gradually turns to a light brownish gray which is sub-translucent (light passing through until \~10 mm), and to a brown which is translucent. The luster ranges from medium to dull, the feel is smooth and the grain is fine. The fracture is conchoidal and the quality is good. Patterns are laminated, splotched, and speckled. The patterns occur on the brown fabric, in a light brownish-gray color.

The cortex is rounded, from an outcrop source. It is around 0.5 to 1 mm thick and the transition is gradual.

Uncommon oval skeletal remains are visible but are replaced by quartz. No other allochems or bioclasts are visible.

```{r typeLLB11Evariability, echo=FALSE, fig.cap="Chert macroscopic variability within the sub-group LLB (a-b) and 11E (c-d).", out.width = '100%'}

knitr::include_graphics("../figures/SOM/groupLLB_11E_variability.png")

```

### Sub-group 11E

Similarly to the previous sub-group, a sub-group named 11E was created during the macroscopic analysis process due to the presence of a group of artifacts with homogeneous characteristics (and refits).

This sub-group is characterized by a heterogeneous structure. The colors are brown (10YR 5/3), gray (2.5Y 5/1), light gray (10YR 7/2) and white (10YR 8/1) The translucency is variable, some areas are translucent (light passing through at 8 mm of thickness), while the splotches are sub-translucent. The luster ranges from shiny to medium, the feel is smooth and the grain is fine. The fracture is conchoidal and the quality is acceptable, due to the presence of fractures. Patterns are splotches, broad mottling, spots, speckling, and laminated.

Whenever present, cortex is rounded, from an outcrop source. Its thickness ranges from 0.5 to 0.7 mm and the transition is sharp.

Identifiable fossils are rare, and limited to a possible bivalve fragment. Uncommon, oval-shaped bioclasts are also visible, however, these are always replaced by quartz. The white splotches show a concentration of spicule-like shapes, however, these may be due to patina.

Five thin sections were produced to better characterize the macroscopic variability of this group and sub-groups.The results show no outstanding differences between the samples and sub-groups. The petrographic analysis shows the samples are characterized by a mudstone texture and massive microstructure. It is mainly composed of cryptocrystalline quartz (between 95-97% in the samples, with one exception where it represents 91%), with accessory fibrous chalcedony found replacing fossils or filling fractures, and in some samples macrocrystalline quartz, microcrystalline quartz and micrite/sparite. Allochems are opaques (u), iron oxides (u) and bioclasts (r/u). Oxides and opaques are often concentrated in the edges of the sample, and may be related to post-depositional processes. Bioclasts are unidentifiable and poorly preserved, replaced by fibrous chalcedony and rarely by microcrystalline quartz. In a single sample (I20-3689) sponge spicules were identified. However, in sample H19-4074 (sub-type 11E), the absence of spicules or spicule-like shapes in the thin section suggests that they are an effect of white patina on the surface of the samples. Porosity is present between \<1-1%, of vug-type when identifiable.

## Group 11F

Group 11F corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment. Identified and well-preserved fossils are foraminifera from the Pfenderenidae family, which are larger benthic foraminifera from marine environments. The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous.

Macroscopically, this group is characterized by gray colors (Gley 1 6/ and 5/) and a white (10YR 8/1) area, frequently at the center of the sample, which seems to be a weathering rind. However, the white broad mottling highlights the presence of very frequent bioclasts (\@ref(fig:type11Fvariability)). The translucency is variable: the grey areas are sub-translucent (light passing through at 2.5/3 mm of thickness), while the broad mottling is opaque; due to the presence of this sort of alteration in all artifacts from this group, a correct measurement of translucency was not possible. The luster ranges from shiny to medium, the feel is smooth and the grain is fine. The fracture is conchoidal and the quality is good. Patterns are splotches, broad mottling, spots, and speckling.

Whenever present, cortex is rounded, from an outcrop source. Its thickness ranges from 0.5 to 0.7 mm and the transition is sharp.

Petrographically, this group is a packstone (mudstone texture close to the edges) with a massive microstructure. It is composed of 97% micro-cryptocrystalline quartz. Other orthochems are macrocrystalline quartz and fibrous chalcedony, found mostly replacing fossils. Allochems are bioclasts. Other minerals include uncommon iron oxides. The porosity is of vug type (3%).

Fossil content is very frequent, albeit mostly unidentifiable and poorly preserved. Petrographic results show these fossils are replaced by macrocrystalline quartz, microcrystalline quartz and fibrous chalcedony. Larger unidentifiable fossils (\~200 μm; r) are frequently replaced by several generations of microcrystalline quartz, macrocrystalline quartz, and fibrous chalcedony. Identified fossils are sponge spicules (r; visible only as ghosts and poorly preserved), ostracods (r) with some degree of preservation, and common small foraminifera (100 μm) of differing preservation degrees (Fig. \@ref(fig:11Ffossils)). Identifiable foraminifera are rare and from the Pfenderenidae family (Fig. \@ref(fig:11Ffossils), black arrows); these have been identified both through thin section (Fig. \@ref(fig:11Ffossils) d) and macroscopically (Fig. \@ref(fig:11Ffossils) a, c). They are well-preserved, replaced by microcrystalline quartz, and measure around 500 μm.

```{r type11Fvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 11F.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group11F_macromeso.png")

```

```{r 11Ffossils, echo=FALSE, fig.cap="Foraminifera from Group 11F from the Terrace and Shelter area. Figures a-c were obtained through a stereomicroscope; figure d is a thin section in plane polarized light. Black arrows: foraminifera from the Pfenderenidae family; White arrow: other unidentified foraminifera.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/group11F_FOSSILS.png")

```

## Group 13

Group 13 is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.

It is a fine-grained chert, with a conchoidal fracture. The quality is good. It has a heterogeneous structure, with two fabrics: 1) white fabric; 2) grey fabric. The colors are . The white fabric is characterized by dull luster, opaque translucency, and semi-smooth feel. The grey fabric varies between medium to dull luster and it is opaque to sub-translucent, with light passing through at around 1.9 mm of thickness in some areas. The feel is semi-smooth and the patterns are splotched, speckled, and laminated dotted horizontally.

The cortex is from an outcrop, rounded, and varying between a minimum of 3.4 mm of thickness and a maximum of 12 mm of thickness. The cortex transition is frequently sharp, although some samples show a gradual transition.

This group shows several facies, with different types and concentrations of allochems and bioclasts (Fig. \@ref(fig:type13facies)). Facies 01a shows a wackestone to packstone texture, with visible small fossils (very frequently below 50 μm). Fossils include round white fossils, with grey or non-colored walls; sponge spicules, monaxon pointed at one end or tetraxon. A possible bivalve shell replaced by quartz or chalcedony was identified on a single sample. Other fossils are unidentifiable, but very frequently they are round and light, scattered within the chert. Facies 01b is similar to 01a, but composed mainly of peloids or unpreserved ooids, with uncommon presence of spicules. Facies 01c shows a decrease in the visible fossils (wackestone texture), and identifiable fossils are spicules and small round fossils with walls. Facies 02 is characterized by a packstone texture, composed mostly of larger (\~200-500 μm, uncommonly 1 cm) round or oval fossils, white or grey), completely or partially replaced by quartz or chalcedony. The smaller round fossils with walls present in Facies 01 are uncommon. Due to poor preservation, these fossils are unidentifiable. Facies 03 is mostly white, with faint round shapes; this facies is interpreted as a white patinated area of both Facies 01 and 02. Facies 04 is characterized by a packstone texture, and composed of ooids (\~50-200 μm, uncommonly 500 μm), which are completely or partially replaced by quartz and chalcedony. In this facies, the chert is light (white or light grey) and it seems to be sub-translucent. Fossils seen in other facies are not visible.

The petrography results for a sample with Facies 01a and 01b (L19-64) show this chert has a massive microstructure and packstone texture. It is composed mainly of micro-cryptocrystalline quartz (80%), fibrous chalcedony (10%), and micrite (10%). Allochems are opaques (u), iron oxides (vf), ooids (vf) and peloids (vf). The ooids are poorly preserved and only uncommonly show concentric lamellae in plane polarized light. Unlike macroscopical observations, bioclasts under the thin section are all unidentifiable, although rare fossils may be bivalve shells. Porosity is vug (10%).

```{r type13variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 11F.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group13_macromeso.png")

```

```{r type13facies, echo=FALSE, fig.cap="Several fossil facies within Type 13.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/GROUP13_facies-01.png")
```

## Type 13B

Group 13B corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone.

It is fine-grained with a conchoidal fracture. The quality is good. It has a heterogeneous structure and is characterized by a variable color distribution: it shows a horizontal laminated and finely laminated pattern, with gray (5Y 5/1) and dark gray (5Y 4/1) bands intercalated with light gray (5Y 6/1) and light olive gray (5Y 6/2) bands. This chert varies between dull, medium and shiny luster, although this variation may be related to post-depositional alterations; samples with little to no alterations show mostly a medium luster. The translucency is opaque and the feel is semi-smooth.

Whenever present, the cortex is rounded (from an outcrop source), and is generally thin, with a min. thickness of 0.1 mm and a max. thickness of 2.5 mm. It is characterized by a sharp transition. In specific samples (i.e., I21-3546, \@ref(fig:type13variability) a) cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers. In this case, the sample indicates a bed thickness of at least \~19 mm in that particular area of the outcrop.

Petrographically this chert is a banded packstone (banding of sedimentary origin), composed mostly of micro-cryptocrystalline quartz (48%), micrite (40%), dolomite (10%), and accessory fibrous chalcedony and mica (muscovite). Allochems are peloids (vf), bioclasts (vf), iron oxides (u), and opaque minerals (vf). Porosity is low (1%). The bioclasts are all unidentifiable and only visible in thin section.

```{r type13Bvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 13B.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group13B_macromeso.png")

```

## Oolitic chert

This group is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.

Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: brown and black fabric. The brown fabric is pale brown (10YR 6/3) and light yellowish brown (10YR 6/4), opaque, and shows medium to dull luster. The black fabric is black (7.5YR 2.5/1), shiny, and sub-translucent (light passing through in the thin areas, around 1 to 2 mm thickness). It corresponds to the fabric which creates horizontal laminated patterns within the brown fabric (Fig. \@ref(fig:ooiliticvariability)c). This chert is fine-grained with a conchoidal fracture. The quality is acceptable since it has fine granularity and a smooth feel, but shows natural fractures and occasional inclusions, such as macrocrystalline quartz or chalcedony veins that may impact the knapping.

The cortex is difficult to identify since it is similar to the brown fabric, highly altered, and with a gradual transition. Whenever present, it is from an outcrop and shows a maximum thickness of 3.8 mm.

Petrographically this group is a packstone with a massive microstructure. The chosen sample for thin section is composed of 35% micro-cryptocrystalline quartz, 53% micrite, 9% macrocrystalline quartz, and accessory fibrous chalcedony (filling fractures) and mica; the percentages of quartz and micrite might be related to the amount of fabrics present in the artifacts. Allochems are common iron oxides and very frequent ooids. Bioclasts are rare, poorly preserved and replaced by quartz; they occur between the ooids and with long, thin shapes. The porosity is intraparticle and vug (10%).

The ooids are distributed homogeneously across the sample (although clearly macroscopically visible in the brown fabric) and are highly abundant and concentrated. They are poorly sorted, varying between 500 and 20 μm. They show round to elliptical shapes, and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010], pp. 144\]. When preserved, the nucleus represents half the width of the ooid and the cortex is normal.

```{r ooiliticvariability, echo=FALSE, fig.cap="Chert macroscopic variability within the Oolitic chert group.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/oolitic_macromeso.png")

```

## Trace lithotypes

### TL01

# References

### 2024-01-05 analysis/paper/paper.qmd
- First (first) draft of the article.

@@
   fig-dpi: 600
+number-sections: true
 filters:
@@
   keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
 ---
@@
 library(janitor)
+library(ggpubr)
+library(knitr)
 
@@
 
-Here is a citation [@Marwick2017]
+Knappable raw materials are a key element used to understand the lifeways of past hunter-gatherers throughout time, as lithics are ubiquitous throughout prehistory.
+More than mere stones with which stone tools are produced, these resources are intimately connected to the group’s technological, social, and cultural organization, possibly with an impact on the group's survivability [@oestmoFormalModelingApproach2017].
+Several key topics to understand hunter-gatherer behavior and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binford_organization_1979; @binford_righteous_1985; @gould_lithic_1985; @gould_lithic_1985; @kuhn_unpacking_1991; @mccall_behavioral_2007], occupation types and their duration [@kuhn_upper_2004; @surovell_toward_2009]) to the establishment and dimension of social networks [@whallon_social_2006] and exchange between groups or individuals [@wurzHowiesonsPoortBacked1999; @gamblePalaeolithicSocietiesEurope1999].
+This has frequently been done through the systematic characterization of raw materials (both geologically available and archaeological), identification of corresponding sources for a clear differentiation between local and long-distance raw materials, and the study of their proportions within a site, often throughout different occupations and chronologies.
+These approaches and concepts have been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula (++).
+Frequently coupled with technological analyses, the study of raw materials, especially chert, through a wide range of methods, such as macroscopic, petrographic, and geochemical, have recurrently aided in answering key questions to Iberian archaeology, such as the disappearance of Neanderthals and consequent replacement by Modern Humans [@aubry_far_2022].
+Raw material studies have also contributed to understanding how different groups during different chronological periods organized their technology, managed time, and risk (through the embedded/direct procurement approach), and chose resources [@matias2016; @matias_o_2012; @pereira_raw_2021; @pereiraLithicEconomyTerritory2016].
+Furthermore, these studies frequently highlight long-distance procurement even if in low proportions at times.
+In fact, raw material studies and the distinction between local/regional and exotic have also been successfully applied to identify long-distance contacts (\>150 km) during the Upper Paleolithic in the Côa Valley which has been interpreted as the result of reciprocal exchange [@aubry_raw_2004-1; @aubry_upper_2016].
+Although commonly known and identified during late prehistory [@nocete_circulation_2005; @rodriguezPrehistoricFlintExploitations2011; @costa_lithic_2022], the identification of these contacts has expanded our understanding of often large-scale social networks and contacts during the Upper Paleolithic in central Portugal, to paint a picture previously outlined of a wide interconnected territory.
+However, unlike other regions of the Iberian Peninsula such as central Portugal, and despite the potential of raw material studies to provide key data on the lifeways and organization of Paleolithic hunter-gatherers through time, as well as the mobility range and existence of social networks, the behaviors of management and procurement of raw materials are still in a preliminary phase in southern Portugal, especially in the Algarve region.
+Although much is known about the technological organization of Paleolithic hunter-gatherer groups in southwestern Portugal, especially due to the archaeological site of Vale Boi (see ++), raw material studies have been scarcer.
+Previous studies applied macroscopic methods to characterize the cherts from Vale Boi.
+These preliminary results have shown the presence of a large component of local cherts from the Jurassic limestone formations in conjunction with a less abundant and variable allochthonous chert component throughout the several Upper Paleolithic occupations of the site [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004].
+@pereira_territory_2016-1 suggest that the exogenous group may have been the result of embedded procurement within a 16 km radius, while the allochthonous group may have been the result of transportation during trips to other areas of the Algarve, Alentejo, or even south Spain.
+The presence of chert from the Cretaceous formations of central Portugal has also been suggested [@bicho_ecodynamics_2013].
+Furthermore, previous studies aiming to characterize the local and regional chert sources in the Algarve region have established a comparative macroscopic and petrographic database [@belmiro2023; @pereira_presenting_2016-1], a key element to raw material studies, especially to distinguish between local and non-local raw materials and understanding changes in the archaeological record [@pop_simulating_2015; @surovell_toward_2009].
+These preliminary results and the existence of regional reference collections show the potential for a systematic raw material study at the site, using different methods and approaches, applied to the lithic assemblages throughout the several Upper Paleolithic occupations, to understand how these hunter-gatherer groups behaved through time: how they explored their landscape and lithic resources, where they moved across a territory nestled between water and land, and with whom did they establish social networks.
+To answer these questions, we aim to characterize the raw material procurement and mobility patterns of hunter-gatherer groups throughout the Upper Paleolithic of southern Portugal.
+By studying the chalcedony and chert artifacts of the UP levels of Vale Boi we aim to characterize these raw materials macroscopically and petrographically, to identify different chert groups and their respective sources.
 
-# Background
+## Site description
 
-# Methods
+Vale Boi is located on the western coast of the Algarve region (south Portugal), in a small valley that connects south to the Atlantic current coast at around 2 km and is bordered by a limestone hill.
+This hill is characterized by limestone exposures that form rocks shelters facing west and southwest [@cascalheira_tecnologia_2010].
+The site extends for more than 10 000 sq m through the slope of the valley, through which three main occupation areas have been identified: Slope, Shelter and Terrace.
+The excavation methodology included the recording of all artifacts with dimensions superior to 2 cm or complete small artifacts (i.e., bladelets) with a total station, and was applied to all main areas.
+Several occupations have been identified in both the Terrace and the Shelter area.
+The Terrace includes occupations from the Holocene to the Upper Paleolithic, resulting in the identification of eight main lithostratigraphic units.
+The UP occupations start with the Gravettian (levels 8-6), followed by Proto-solutrean (bottom of 5 and 5/4E, sensu Belmiro et al., 2021), and Solutrean occupations (4D-4 and 3B).
+The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations.
+From these, the Solutrean levels show the largest occupation, with three, well-preserved archaeological horizons [@cascalheira_vale_2013].
 
-The samples chosen for this study consist of the chert and chalcedony lithic artifacts with individual IDs (excluding buckets) from the Upper Paleolithic levels of the Terrace and Shelter areas of Vale Boi.
-The sample from the Terrace area (VB-T) consists of 3653 from levels 8 to 4, while the sample from the Shelter area (VBS) consists of 1652 from levels C to A.
-These two areas were chosen since both have detailed spatial information (as mentioned in section *IN*), several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal and geoarchaeological data.
+## Geological context
 
-## Geo-archaeological surveys
+The Algarve region is a complex territory, characterized by several geomorphic sub-regions and geological units, which provide different lithological resources used throughout prehistory.
+There are two main geological units, the South Portuguese Zone (SPZ) in the north sector of Algarve and the Algarve basin to the south.
+The SPZ is characterized by schists, greywacke and quartzites, while the Algarve basin is characterized by limestones, dolomites, and marls.
+It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules and beds can be found.
+Chalcedony is located in a 15 km radius from Vale Boi, in the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013].
+The cherts can be found in different Jurassic formations throughout the Algarve, but are most frequently known for the Lower Jurassic beds and small nodules in the limestone cliffs at ++ km distance from the site, which have been previously studied and characterized [@ribeiro_evolucao_2005; @belmiro2023].
+Chert is also available in the less known Upper Jurassic limestone formation in western Algarve (++ km from the site), and in the center and eastern portion of the Algarve, in Middle and Upper Jurassic formations with chert nodules [@belmiro2023].
 
-The geological data and comparative collection from local and regional cherts used in this study were obtained through previous prospections and petrographic works (\*\*), located at the Lusolit lithotheque (hosted at ICArEHB, University of Algarve).
-These samples also included cherts collected from previous prospections from other areas of Portugal (such as central Portugal).
+# Materials and methods
+
+To characterize the siliceous raw materials in the Upper Paleolithic occupations of Vale Boi and identify their geological sources, we applied a macroscopic and petrographic approach to the study of the chalcedony and chert archaeological materials from the Terrace and Shelter area.
+We applied two different methods to obtain a more complete characterization of the materials and compensate for the inherent caveats of each methods [@luedtke_archaeologists_1992].
+Macroscopic analysis, for example, shows several caveats related to the lack of quantitative variables [@bustillo_is_2009].
+However, given the large sample size of chalcedony and chert lithics at the archaeological site, a macroscopic approach presents a non-destructive, easy and less costly of characterizing all the materials.
+As such, we apply a first step of macroscopic analysis, followed by a petrographic study to reduce the subjectivity of the analysis [@bustillo_is_2009] and complete the characterization with petrographic data.
+Similarly, the combination of these two methods was previously applied successfully in the characterization of the cherts from the Algarve region as the result of recent prospection works [@belmiro2023], and as such, a comprehensive macroscopic and petrographic comparative collection is available for the comparison with archaeological materials when using the same methods.
+
+This reference collection is located at the Lusolit lithotheque (hosted at ICArEHB, University of Algarve), and aided in the characterization and source attribution of cherts.
 Hand samples were used throughout the macroscopic characterization for comparison with the archaeological materials, and the archaeological thin sections were compared with the previously studied geological chert thin sections.
-The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz was also visited, to understand the macroscopic variability from the Cadiz local and regional cherts, to discern between local and regional chert types and possible long-distance cherts.
+The geological samples at the lithotheque also included cherts collected from previous prospections from other areas of Portugal (such as central Portugal) and chalcedony, of which 4 thin sections were produced for this study (3 of chert from the Estremadura and ++ of chalcedony from the Algarve).
+The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz (hereon UCA) was also visited, to understand the macroscopic variability of the Cadiz local and regional cherts, to aid in the source attribution of cherts not congruent with local sources and identify possible long-distance chert-use.
 
-## Macroscopic characterization
+As aforementioned, the study started with a macroscopic analysis of archaeological materials.
+The samples chosen for this study consist of the chert and chalcedony lithic artifacts with individual IDs (excluding buckets) from the Upper Paleolithic levels of the Terrace and Shelter areas of Vale Boi.This means that the smaller fraction of artifacts (lithics smaller than 2 cm) was not considered, and thus we may be missing patterns of differential raw material use and miniaturization.
+However, similar methodologies were applied to studies using formal models for raw material and mobility studies applied to archaeology, where only artifacts above 2 cm were considered to improve comparative studies between assemblages [@surovell_toward_2009].
 
-We applied a macroscopic analysis to both Terrace and Shelter samples, which was divided into two steps.
+The sample from the Terrace area (VB-T) consists of 3600 lithic artifacts from levels 4 to 7/8 (excluding levels with a small sample such as level 4B), while the sample from the Shelter area (VBS) consists of ++ lithic artifacts from levels C to A.
+These two areas were chosen since both have detailed spatial information (as mentioned in section *IN*), several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal, and geoarchaeological data.
+The archaeological materials are currently located at ICArEHB's laboratories (University of Algarve) and no permits were necessary for their study.
+The macroscopic analysis was divided into two steps.
 The first step included a preliminary characterization using a hand lens of 10x magnification.
-This step allowed us to create groups and subgroups based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these include characteristics such as color, translucency, and feel; a data dictionary for the recorded variables can be found in the Supplementary Materials (\*\*).
-
-In the second step, we used a stereomicroscope () to observe each sample in further detail, which allowed us to better characterize the artifacts (especially regarding inclusions and alterations), and the previously established groups.
+This step allowed us to create groups and subgroups based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these include characteristics such as color, translucency, and feel [@luedtke_archaeologists_1992]; a data dictionary for the recorded variables can be found in the Supplementary Materials (++).
+In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artifacts (especially regarding inclusions and alterations), and the previously established groups.
 The data dictionary used for the individual samples can also be found in Supplementary Materials (\*\*).
 Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis is simplified and focused on collecting data related to weight, cortex, and alterations.
-
-## Petrographic characterization
-
-To better characterize the different chert groups identified in the macroscopic analysis, we analyzed ++ thin sections.
-The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels.
-For groups which showed macroscopic variability, several thin sections were analysed to better characterize these cherts.
-@tbl-TS-table Petrographic thin sections were analyzed by polarized light microscopy () to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations.
-A total of 20 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory.
-The archaeological samples chosen for the petrographic study are from the Terrace area, from several levels, and focused on characterizing the macroscopic variability, reflected on the several groups and subgroups (table \*\*).
+A caveat of using this methodology of analysis is the inability to unequivocally characterize and group artifacts which have surface alterations such as patinas, pits or fire-related alterations.
+Whenever artifacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), they were analysed using the individual sample dataset, and instead grouped in a category of indeterminate cherts (INDET).
 
 ```{r}
-#| label: tbl-TS-table
-#| tbl-cap: Thin section list.
-#| echo: false
+#| label: fig-terrace-alterations
+#| fig-cap: Photo and close-up of samples with extensive alterations which impossibilitate the identification of chert type.
+a) artifact with cracks, white patina and reddening covering 100% of the surface; b) artifact with surface alterations, displaying a portion with patina (brown) and a portion without patina (gray), cleaned with HCL.
 
-TS_short_list <- read_csv(here::here('analysis/data/raw_data/TS_short_list.csv'))
-
-knitr::kable(TS_short_list)
+knitr::include_graphics(here::here("analysis/figures/terrace_alterations.png"))
 
 ```
+
+To better characterize the different chert groups identified in the macroscopic analysis, 22 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory.
+The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels (@tbl-thinsections).
+For groups that showed macroscopic variability, several thin sections were analyzed to ensure the collection of petrographic data that reflected this variability.
+Petrographic thin sections were analyzed by polarized light microscopy (Nikon LV100ND) to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations.
+
+| Type               | Sample ID | Level | TS \# |
+|--------------------|-----------|-------|-------|
+| Chalcedony         | I19-1615  | 4D    | 8     |
+| Type 2             | I19-2835  | 5     | 17    |
+| Type 2 (rmu_17)    | I19-2226  | 4E    | 18    |
+| Type 2 (altered)   | H18-1938  | 4D    | 20    |
+| Type 2 (reddening) | H19-2426  | 4D    | 3     |
+| Type 2B            | H21-3095  | 4C    | 6     |
+| Type 2B            | H21-4234  | 5     | 7     |
+| Type 15            | I20-3160  | 5     | 10    |
+| Type 7             | H20-2441  | 4E    | 19    |
+| Type 12            | I21-3252  | 4E    | 14    |
+| Type 12B           | H18-2708  | 5     | 2     |
+| CPT                | I19-3350  | 6     | 9     |
+| CPT                | J18-778   | 5     | 13    |
+| CPT                | I20-3689  | 6     | 16    |
+| Sub-type LLB       | I20-3951  | 6     | 11    |
+| Sub-type 11E       | H19-4074  | 6     | 4     |
+| Type 11F           | H19-2924  | 4E    | 1     |
+| Type 13 (01b; 01a) | L19-64    | 5     | 12    |
+| Type 13 (04)       | J18-1264  | 8     | 22    |
+| Type 13B           | I21-2966  | 4C    | 15    |
+| Oolitic            | H19-4216  | 6     | 5     |
+| White patinated    | H20-4166  | 6     | 21    |
+
+: A table with thin sections {#tbl-thinsections}
+
+The petrographic analysis PDF files for all thin sections studied in this paper (including macroscopic and microscopic figures) can be found in our GitHub page (++) as well as the OSF repository (++).
+The geological thin sections can also be consulted in the LusoLit online database at https://lusolit.icarehb.com/.
+A detailed macroscopic and petrographic description and figures for each type and sub-type of chert and chalcedony can be found in the SOM (++), as well as the datadictionary used for the petrographic analysis.
+The entirety of the R code used for the analysis and visual representations contained in this paper can be accessed through our online research compendium (++).
+We use the rrtools package by Ben Marwick to create a research compendium and write a reproducible journal article; rrtools is a free and open-source RStudio package accessible at <https://github.com/benmarwick/rrtools>.
+To produce these files, we followed the procedures outlined by @Marwick2017 for constructing research compendiums, aiming to enhance the reproducibility of our research.
+The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickhamPackagesOrganizeTest2015] containing the code required to generate all tables and figures.
+To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY [additional details can be found in @Marwick2017].
 
@@
   #filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
-  filter(Level !="4B" & Level !="7C" & Level!="8") # Filtering out levels with barely any artifacts
+  filter(Level !="4B") %>%  # Filtering out levels with barely any artifacts
+  mutate(Level = case_when(Level == "7C" ~ "7", # Collapsing samples from Level 7C into 7
+                           TRUE ~ Level)) %>% 
+  mutate(Level = case_when(Level == 8 ~ "7", # Collapsing two samples from Level 8 into 7, since 8 disappears in field_data
+                           TRUE ~ Level))
 
-cols_to_concat_spit <- c("Level", "Spit")
-RM_by_spit <- T_dataset %>%
-  filter(!Spit %in% c("limp_corte", "profile", " ", NA)) %>%
-  select(ID, Z, Level, Spit, RMTYPE) %>%
-  unite(col='depth', cols_to_concat_spit, sep=".", remove=FALSE) %>% 
-  dplyr::group_by(depth) %>%
-  dplyr::count(RMTYPE) %>%
-  mutate(freq = n / sum(n)*100)
+#cols_to_concat_spit <- c("Level", "Spit")
+#RM_by_spit <- T_dataset %>%
+  #filter(!Spit %in% c("limp_corte", "profile", " ", NA)) %>%
+  #select(ID, Z, Level, Spit, RMTYPE) %>%
+  #unite(col='depth', cols_to_concat_spit, sep=".", remove=FALSE) %>% 
+  #dplyr::group_by(depth) %>%
+  #dplyr::count(RMTYPE) %>%
+  #mutate(freq = n / sum(n)*100)
 
@@
 S_RM_data <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
-S_field_data <- read_csv(here::here('analysis/data/raw_data/VB_Shelter_xyz.csv')) # Shelter coordinates
+S_field_data <- read_csv(here::here('analysis/data/raw_data/VB_Shelter_xyz_codes.csv')) # Shelter coordinates
+S_layerdata <- read_csv(here::here('analysis/data/raw_data/shelter_layers_JC.csv'))
 
@@
 
+S_layerdata <- S_layerdata %>% 
+  mutate_at("Date", str_replace, "20", "") %>%
+  dplyr::rename("YEAR" = "Date") %>% 
+  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% 
+  select(ID, Layer)
+
 S_field_data <- S_field_data %>%
+  mutate_at("YEAR", str_replace, "20", "") %>%
   unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% # Unite YEAR with Id s«using cols_to_concat
-  select(ID, X, Y, Z) %>% # Selecting important variables
+  select(ID, X, Y, Z, CODE) %>% # Selecting important variables
   distinct(ID, .keep_all = TRUE)
-S_dataset <- left_join(S_RM_data, S_field_data, by = "ID")
+
+shelter_field_data <- full_join(S_field_data, S_layerdata, by = "ID")
+S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID", relationship = "many-to-many")
 
@@
 The three main used raw materials at the archaeological site of Vale Boi are quartz, greywacke, and chert.
+Other raw materials such as chalcedony, dolerite and schists have also been identified.
 Using the field database with all individually coordinated lithic artifacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials.
-Since some lithic assemblages of the Terrace area are currently under study, the data regarding the percentage of raw materials in the “Other” category was not available for all archaeological levels.
-@fig-rm-comparison shows that for the selected levels, and using single coordinated artifacts (excluding buckets), chert represents only \~20-15% of the lithic assemblage.
-Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke.
-Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site.
-Frequently, the quartz and greywacke assemblage includes large amounts of shattered and/or unknapped chunks or slabs.
-Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz.
+@fig-rm-comparison shows that for the selected levels, chert represents only \~20-15% of the lithic assemblage.
+Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke \[++\].
+Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site \[++\].
+Frequently, the quartz and greywacke assemblages include large amounts of shattered and/or unknapped chunks or slabs.
+Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021].
 Based on this, despite the low amounts of individually plotted chert artifacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.
-
-```{r}
-#| label: tbl-indet-freq
-#| tbl-cap: N and percentage of grouped and indeterminate artifacts by level.
-
-indet_table <- phase_dataset %>% 
-  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
-  tabyl(Group,Level) %>% 
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
-  adorn_ns("front")
-
-```
-
-As previously mentioned, whenever too altered to confidently attribute a group, the artifacts were instead classified as indeterminable (INDET).
-As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artifacts, corresponding to 40% of the total sample (n=1027).
-This is due to the frequent alterations of the sample, especially related with patinas and alteration rinds, pits and fire/heat alterations.
-For all other levels, indeterminate represents in mean \~28% of the chert artifacts.
+The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].
 
@@
 
-#T_short_table <- T_short %>% 
-#   tabyl(Rawmaterial, Level) %>%
-# adorn_totals(c("row", "col")) %>%
-#  adorn_percentages("col") %>%
-#  adorn_pct_formatting(2) %>%
-#  adorn_ns("front")
-
 T_short %>%
@@
 
-Through the macroscopic and petrographic analysis we identified 24 types of chert and three types of chalcedony.
-Of these, one type of chalcedony and 9 types of chert are composed of several samples (\>10) and present across both areas (Terrace and Shelter) and throughout the Upper Paleolithic occupations of the Terrace area.
-Thin sections were produced for the lithics of the latter groups.
-One type had several samples (\>10; Type GB) but was only identified in the Shelter area, while 14 varieties of chert and 2 chalcedony are composed of one to three samples and restricted to specific levels.
+As previously mentioned, whenever too altered to confidently attribute a group, the artifacts were instead classified as indeterminable (INDET).
+As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artifacts, corresponding to 40% of the total sample (n=1027).
+This is due to the frequent alterations of the sample, especially related to patinas and alteration rinds, pits and fire/heat alterations.
+For all other levels in the Terrace, indeterminate represents in mean \~28% of the chert artifacts, while in the Shelter the frequency of indetermiates is slightly lower (\~20-25%).
+
+```{r}
+#| label: tbl-indet-freq
+#| tbl-cap: N and percentage of grouped and indeterminate artifacts by level.
+
+terrace_indet_table <- phase_dataset %>% 
+  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
+  tabyl(Group,Level) %>% 
+  adorn_totals("row") %>%
+  adorn_percentages("col") %>%
+  adorn_pct_formatting(2) %>%
+  adorn_ns("front")
+
+shelter_indet_table <- S_dataset %>% 
+  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
+  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
+  filter(!Layer %in% c("D", "ND")) %>% 
+  tabyl(Group,Layer) %>% 
+  adorn_totals("row") %>%
+  adorn_percentages("col") %>%
+  adorn_pct_formatting(2) %>%
+  adorn_ns("front")
+
+indet_table <- full_join(terrace_indet_table, shelter_indet_table, by = "Group")
+
+kable(head(indet_table))
+
+```
+
+## Characterization and source attribution
+
+Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony groups, present throughout the stratigraphy in the Terrace and Shelter area.
+A single different group was identified in the Shelter area, with limited presence in the Terrace (n=2), and 14 varieties of chert and 2 chalcedony composed of one to three samples were identified in several levels of the Terrace and Shelter.
 The latter varieties were collapsed into a category named TL (Trace lithotypes).
+Through the comparison with the reference collections, we identified cherts congruent with the local sources and cherts which are not local.
 
-We present here a short description of each identified macroscopic and petrographic type.
-A detailed macroscopic and petrographic description and figures for each type and sub-type of chert and chalcedony can be found in the SOM (++).
-The petrographic analysis PDF files for all thin sections studied in this paper (including macroscopic and microscopic figures) can also be found in our GitHub page (++) as well as the OSF repository (++).
+The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts.
+The chalcedony is a massive, fibrous chalcedony without allochems.
+It is characterized by a colorless and translucent appearance, often partially or completely white patinated.
+It is characterized by frequent large crystallizations, irregularities and fractures which impact the knapping.
+It is macroscopically and petrographically congruent with the hydrothermal chalcedony found in lenses in Triassic formations, also frequently characterized by irregularities and fractures.
 
-## Type description
-
-The chalcedony is a massive, fibrous chalcedony mudstone.
-It is characterized by a colorless and translucent appearance, often partially or completely white patinated.
-It has good to medium quality, due to frequent large crystallizations, irregularities or fractures which impact the knapping.
-In thin section, the chalcedony is composed by fibrous chalcedony, without allochems or bioclasts, and no porosity (\<1%).
-Radiating spherulitic textures are very frequent.
-
-The first chert groups identified at Vale Boi are massive micro-cryptocrystalline quartzes from a marine environment.
-Petrographically, these cherts are homogeneous, except for a small amount of artifacts with a banded variety within type 12.
+The cherts are massive micro-cryptocrystalline quartzes from a marine environment.
+Macroscopically, these groups show variability which resulted in their separation in four different macroscopic groups (Group 2 to 5).
+In general, these are opaque to sub-translucent cherts, with colors varying brown, yellow, pale red and grey, with the frequent presence of spicules and iron oxides.
+The differences are in their colors, nodule morphology, frequency, and variety of bioclasts, something already identified in the geological samples [@belmiro2023].
+Petrographically, however, these cherts are homogeneous, except for a small number of artifacts with a banded variety within group 5.
+The macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.
 These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite.
@@
 Porosity varies between 10% and 1%, of vug type.
-Macroscopically however, these groups show variability, which resulted in their separation in 4 different macroscopic groups; this separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.
+Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 15, 7 and 12 to be local, as they are congruent with the described characteristics.
+Groups 2 and 15 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils.
+The macroscopic similarities of Groups 7 and 12 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the Ferrel area.
+As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape.
+This is especially relevant, since the Ferrel outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiro2023].
 
-Group 2 is characterized by single-color and multiple-color samples, ranging between yellow, grayish brown, pale red, and white, and with variable patterns such as splotched, broad mottling, speckling, and laminations.
-They are opaque, with sub-translucent areas mainly focused in the white areas of the sample.
-Whenever present, the cortex can be rounded or irregular, varying between thin to thick.
-Inclusions include black speckles and red speckles, which are probably organics and iron oxides.
-Bioclasts are frequently unidentifiable, but also include monaxon sponge spicules, bivalve shells replaced by quartz or chalcedony, and rarely triaxon sponge spicules.
-In the shelter area, the variety of identified fossils also includes possible uncommon ostracods, a possible small foraminifer, and echinoderm spines.
-
-Group 15 is characterized by light gray and reddish gray colors and varies between opaque to sub-translucent.
-It is characterized by the presence of splotches (often coarser and sub-translucent), speckling, and fine lamination.
-Whenever present, the cortex is white, rounded, and of medium thickness.
-Allochems such as iron oxides or opaques are also visible macroscopically as black speckling or rare red speckling.
-The bioclasts are frequently unidentifiable, due to poor preservation and are replaced by microcrystalline quartz or fibrous chalcedony, except for uncommon sponge spicules.
-
-Type 7 shows a heterogeneous structure with two types of fabrics.
-The interior of the samples and the nodules is gray, and sub-translucent with a single mica mineral, flecks, and speckles.
-The outer edge and within the grey fabric as splotches or broad mottling are white, opaque, and with yellow speckles.
-The cortex is white, with a sharp transition and with a medium to thick width.
-The morphology of the cortex in fully cortical samples is originally available as irregular (slightly oval) nodules within the parent rock.
-Whenever identifiable, bioclasts are limited to sponge spicules.
-
-Type 12 chert has a conchoidal fracture (albeit uneven at times) and acceptable to low quality due to the lack of homogeneity and frequent fractures, as well as a coarser fabric in certain areas.
-It has gray colors and ranges from opaque to sub-translucent.
-The group displays frequent patterns: splotched, banded, broad mottling, and speckling.
-The cortex is from an outcrop, frequently rounded, but also irregular.
-Based on reference samples, the chert seems to be enclosed within the parent rock as irregular or tabular nodules).
-The cortex is thin, with a sharp transition, and some artifacts show chunks of the parent rock attached to the cortex.
-Bioclasts are uncommon, and when identifiable are sponge spicules.
+Five chert types, not congruent with local sources, were identified (Groups 6 to 10), including the TL group.
 
@@
 
+The similarities between chert artifacts at the UP occupations of Vale Boi and the chert from the Rio Maior region (Estremadura Portuguesa) have been previously suggested although without systematic studies to ascertain the attribution.
+These Cretaceous cherts, found in secondary deposition settings, are described as frequently translucent, with geodes, with colors ranging from yellow to red or grey, and mineralogically homogeneous between them.
+Petrographic studies highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matias2016].
+When identified, sponge spicules and possible rare foraminifera have been identified [@matias2012].
+These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from central Portugal hosted at LusoLit.
+Alike the archaeological thin sections from Type 11 and congruent with other studies [@matias2016], the thin sections produced from the Cretaceous cherts show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
+From these cherts, only the dark grey nodules (RT231) showed identifiable fossils which were rare ostracods.
+Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Group 11 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts from the Estremadura area in central Portugal.
+
 Group 11F corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment.
@@
 Identifiable foraminifera are rare and from the Pfenderenidae family (500 μm).
+Despite the macroscopic similarities between type 11F and cherts from the Andalusian chert formations, the samples inspected with a stereomicroscope showed no visible foraminifera, especially from the ++ family.
+However, the archaeological thin section revealed that despite no foraminifera being visible macroscopically, they are visible under the microscope in thin section.
+The fact that no foraminifera are visible in the the small number of analyzed samples from the Cadiz lithotheque may simply be a result of their uncommon visibility.
+Despite this, given the lack of this specific fossil in the reference samples, the probable source of Group 11F remains unknown.
 
@@
 Porosity is vug (10%).
+The consultation of the lithotheque hosted at UCA allowed us to identify a sub-set of cherts with similarities to those from Type 13.
+These cherts are massive, sub-translucent, peloidal/oolitic packstones.
+Allochems include very frequent peloids and ooids.
+The peloids are densely arranged in the samples, more or less visible depending on the alterations to the surface or the thickness.
+The ooids are poorly sorted, with oval or round shapes and replaced by quartz.
+Uncommonly the ooids show a preserved nucleus.
+The presence of bioclasts varies between samples, ranging between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon) and echinoderm spines (rare).
+Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allows us to interpret Type 13 as coming from the Andalusian region of southern Spain.
 
@@
 Porosity is low (1%).
+Although its origin is unknown, when visiting the Cadiz lithotheque, we identified a small group of banded black cherts without noticeable fossils or inclusions.
+Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Group 13B belongs to Andalusian chert formations in southern Spain.
 
@@
 144\].
+This chert shows no similarities to any consulted reference material.
+Despite the existence of oolitic limestones in the Algarve region, no chert nodules are mentioned (++) and the visit of outcrops mentioned in the geological cartography did not allow their identification; the trade of oolitic cherts in recent prehistory may indicate however that these types of cherts are available in the landscape in long-distance (++), and their presence in Vale Boi may simply reflect similar mobility patterns.
 
-Finally, group TL is characterized by several samples with differing characteristics from the established larger groups.
-Based on their different characteristics they have been interpreted as exotic or long-distance raw materials.
-A detailed description of these types can also be found in the SOM (++).
-TL01 is a jasper, with a massive microstructure and red coloration, with frequent fractures filled with colorless, translucent fibrous chalcedony.
-TL02 is an orange chalcedony of good quality, differing from the group of white/translucent chalcedony due to its color.
-TL03 is a massive chert with uncommon spots.
-TL04 is a yellow massive chert, with inclusions and fractures filled with fibrous chalcedony.
-TL05 is a grey, banded/massive chert, where the banding seems to be of ++ origin.
-TL06 is a sub-translucent, oolitic packstone chert.
-Unlike the oolitic cherts in the previous groups, this chert shows well-sorted and small (200-220 μm), round ooids with differentiated walls and cores.
-TL07 is characterized by a laminated dendritic pattern, with uncommon inclusions.
-TL08 is a translucent and colorless, massive chert, without inclusions.
-TL09 is a translucent chalcedony with a massive microstructure and packstone texture.
-It is characterized by unidentified, moss-like orange/red inclusions, and no other allochems or bioclasts.
-TL10 is a massive chert with frequent iron oxide speckling or banding.
-TL11 is a banded, translucent chert with no inclusions.
-TL12 is a massive oolitic packstone chert, white and translucent.
-The ooids are not visible under the naked eyes, and can only be identified under the stereomicroscope.
-TL13 is a translucent, banded chert without inclusions.
-TL14 is a massive chert with large benthic foraminifera (1-2 mm).
-TL15 is a massive brown chert with common iron oxides and unidentifiable fossils.
-And finally, TL 16 is a radiolarite with unidentifiable fossils.
-
-## Source attribution
-
-Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 15, 7 and 12 to be local, as they are congruent with the described characteristics.
-Groups 2 and 15 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils.
-The macroscopic similarities of Groups 7 and 12 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, as mostly located in the Ferrel area.
-As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape.
-This is especially relevant, since the Ferrel outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiro2023].
-
-The similarities between chert artifacts at the UP occupations of Vale Boi and the chert from the Rio Maior region (Estremadura Portuguesa) have been previously suggested although without systematic studies to ascertain the attribution.
-
-These Cretaceous cherts, found in secondary deposition settings, are described as frequently translucent, with geodes, with colors ranging from yellow to red or grey, and mineralogically homogeneous between them.
-Petrographic studies highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matias2016] .
-When identified, sponge spicules and possible rare foraminifera have been identified [@matias2012].
-These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from central Portugal hosted at LusoLit.
-
-To aid on the comparison, we compared the lithics attributed to Type 11 to hand-samples from the LusoLit lithotheque, previously collected from the Rio Maior area.
-Similarly, three thin sections were produced of geological samples with different macroscopic characteristics from the Central Portugal area to compare with the archaeological thin sections.
-Alike the archaeological thin sections from Type 11 and congruent with other studies [@matias2016], these thin sections show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
-From these cherts, only the dark grey nodules (RT231) showed identifiable fossils which were rare ostracods.
-
-Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Group 11 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts from the Estremaduran area in central Portugal.
-
-The peloidal/oolitic packstone cherts are not congruent with the cherts from the Algarve area, making them not local or regional.
-The consultation of the lithotheque hosted at UCA allowed us to identify a sub-set of cherts with similarities to those from Type 13.
-These cherts are massive, sub-translucent, peloidal/oolitic packstones.
-Allochems include very frequent peloids and ooids.
-The peloids are densely arranged in the samples, more or less visible depending on the alterations to the surface or the thicknes.
-The ooids are poorly sorted, with oval or round shapes and replaced by quartz.
-Uncommonly the ooids show a preserved nucleus.
-The presence of bioclasts varies between samples, randing between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon) and echinoderm spines (rare).
-@fig-type13-source exemplifies several of the facies identified in Type 13 cherts from the Terrace and Shelter areas (left) and matching samples from the UCA lithotheque (right).
-Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allows us to interpret Type 13 as coming from the Andalusian region of southern Spain.
-
-Group 13B is a small group but present throughout the whole stratigraphy of the Terrace and in the Solutrean occupation of the Shelter, and with significant different characteristics from the local cherts of the Algarve region.
-Although its origin is unknown, when visiting the Cadiz lithotheque, we identified a small group of banded black cherts without noticeable fossils or inclusions.
-Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Group 13B belongs to Andalusian chert formations in southern Spain.
-
-From the identified groups, only group 11F and Oolitic remain unknown.
-Despite the macroscopic similarities between type 11F and cherts from the Andalusian chert formations, the samples inspected with a stereomicroscope showed no visible foraminifera, especially from the ++ family.
-However, the archaeological thin section revealed that despite no foraminifera being visible macroscopically, they are visible under the microscope in thin section.
-The fact that no foraminifera are visible in the the small number of analyzed samples from the Cadiz lithotheque may simply be a result of their uncommon visibility.
-Despite this, given the lack of this specific fossil in the reference samples, the probable source of Group 11F remains unknown.
-Similarly, the black oolitic chert shows no similarities to any consulted reference material.
-Despite the existence of oolitic limestones in the Algarve region, no chert nodules are mentioned (++) and the visit of outcrops mentioned in the geological cartography did not allow their identification; the trade of oolitic cherts in recent prehistory may indicate however that these types of cherts are available in the landscape in long-distance (++), and their presence in Vale Boi may simply reflect similar mobility patterns.
+Finally, the TL groups are characterized by several samples with differing characteristics from the previously established groups, and with without correspondence to the local cherts and chalcedony from the Algarve region.
+These are composed of blanks or retouched tools, rarely with cortex, and rarely the same TL group is found in different archaeological layers.
+A detailed description of these types can be found in the SOM (++).
 
@@
 The exceptions are TL01, TL10-11 and TL15-16.
-Jasper is known in the SPZ area, from the Alentejo region to Spain, above the Algarve.
-Although no petrographic data is available, the macroscopic similarities between TL01 and a hand-sample hosted at Luso-lit seem to indicate that the source of this jasper is the SPZ formations.
-Similarly, despite different from the frequent Cretaceous nodules, the macroscopic similarities between TL10 and TL11 and specific nodules from Central Portugal in the Luso-lit lithotheque may indicate these cherts are from the Estremadura area.
+Jasper is known in the SPZ area, from the Alentejo region to Spain, above the Algarve suggesting that the source of this jasper may be the SPZ formations.
+Similarly, despite different from the frequent Cretaceous nodules, the macroscopic similarities between TL10 and TL11 and specific nodules from Central Portugal in the Lusolit lithotheque may indicate these cherts are from the Estremadura area.
 This difference in appearance may be due to the already mentioned variability of the nodule's macroscopic characteristics.
@@
 Different chert types were consistently used in the Terrace area throughout time, with different intensities of use.
-@fig-type-per-level shows the percentage of chalcedony and chert types used throughout the stratigraphy in the Terrace area of Vale Boi; this data excludes the unidentifiable and altered samples, categorized in a group named INDET.
-In the Solutrean and Proto-solutrean levels, local cherts make up more than \~70% of the chert.
-Despite this homogeneous pattern, the use of group 2 and group 7 marks significant differences between the Solutrean occupations (levels 4, 4C, and 4D) and the Proto-solutrean occupations (4E and 4E/5); in the first group the yellow/brown cherts (type 2) represent \~30% of local chert use, while in the second group, the use of type 2 diminishes drastically (\~10-15%), while Type 7 becomes the most used chert with \~38% in level 4E and \~23% in the top of level 5.
+@fig-type-per-level A shows the percentage of chalcedony and chert types used throughout the stratigraphy in the Terrace area of Vale Boi, while @fig-type-per-level B shows the percentages of weight sum of these raw material types per level; this data excludes the unidentifiable and altered samples, categorized in a group named INDET.
+In the Solutrean and Proto-solutrean levels, local cherts make up more than \~70% of the chert (n) and 80% when considered the weight.
+Despite this homogeneous pattern, the use of group 2 and group 7 marks significant differences between the Solutrean occupations (levels 4, 4C, and 4D) and the Proto-Solutrean occupations (4E and 4E/5); in the first group the yellow/brown cherts (type 2) represent \~30% of local chert use, while in the second group, the use of type 2 diminishes drastically (\~10-15%).
+In contrast, Type 7 becomes the most used chert with percentages above 35% in level 4E and \~20% in the top of level 5.
 For both occupations, chert non-congruent with the known local sources represent percentages lower than 30%.
 This pattern seems to shift in the bottom of 5 and throughout the Gravettian occupation; the use of chert local cherts drops to 45-60%, while the percentages of specific non-local chert types increase significantly, such as Type 11 and Type 13.
+However, when considering the sum of the weight, the percentages of Type 11 diminish, instead representing around 20% of the chert use, and bringing for level 6 the use of local chert to 50% in comparison to chert types of non-local source.
 Other non-local cherts such as the Oolitic cherts and TL categories also show smaller increases in the Gravettian occupation.
-As seen in previous studies, chalcedony seems to be present but always in small frequencies.
+Chalcedony seems to be present but always in small frequencies.
 Despite this, there is a small increase in the use of chalcedony in level 5 (both top and bottom), from lower than 5% in level 4E to 13.5% in top of 5, with a tendency to decrease in the older levels.
@@
 
-phase_dataset %>%
+count_bar <- phase_dataset %>%
   filter(RMTYPE != "INDET") %>% 
@@
   mutate(percent = n/sum(n)) %>% 
-  ggplot(aes(x=Level, y = percent, fill = RMTYPE)) +
+  ggplot(aes(x=Level,y=percent, fill = RMTYPE)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  #theme_minimal() + #type of theme
+  labs(x = "Level", y = "Count %") +# change labels
+  guides(fill=guide_legend(title=NULL)) + # remove legend title
+  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
+                position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_text(size = 10, vjust = -1), legend.position = "none") +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+
+weight_bar <- phase_dataset %>%
+  filter(RMTYPE != "INDET" & WEIGHT != "0") %>% 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
+                                        "TYPE 12", "TYPE 13","TYPE 13B", 
+                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  dplyr::group_by(Level, RMTYPE) %>% 
+  dplyr::summarise(wsum = sum(WEIGHT)) %>% 
+  mutate(percent = wsum/sum(wsum)) %>% 
+  ggplot(aes(x = Level, y = percent, fill = RMTYPE)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  scale_y_continuous(labels = scales::percent_format(), position = "right") +
+  #theme_minimal() + #type of theme
+  labs(x = "Level", y = "Weight %") +# change labels
+  guides(fill=guide_legend(title=NULL)) + # remove legend title
+  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
+                position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_text(size = 10, vjust = -1)) + 
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+
+ggarrange(count_bar, weight_bar, labels = c("A", "B"), ncol=2, common.legend = TRUE, legend="bottom")
+
+```
+
+```{r}
+#| label: extra-graph
+#| include: false
+
+T_dataset %>%
+  filter(Z != "0") %>% 
+  filter(!Z >= "25") %>% 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
+                                        "TYPE 12", "TYPE 13","TYPE 13B", 
+                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  filter(RMTYPE != "INDET") %>% 
+  ggplot(aes(RMTYPE, Z, fill = RMTYPE)) +
+  geom_violin(trim = FALSE, scale="width", adjust = .5) + 
+  geom_point(position = position_jitter(seed = 1, width = 0.2), alpha = 0.1, color = "darkorange") +
+  geom_hline(yintercept = c(24.5,24.3,24.1,23.8,23.45), colour="#BB0000", linetype="dashed") +
+  labs(title="Groups per depth", 
+       subtitle="Chert groups in the Terrace area by depth",
+       caption="Source: mpg",
+       x="",
+       y="Depth") +
+  stat_summary(fun.data=mean_sdl, geom="errorbar", color="black", width=0.2) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank())
+
+```
+
+```{r}
+#| label: tbl-weights
+#| tbl-cap: A table with weights.
+#| include: false
+
+T_stats <- phase_dataset %>%
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET", "GB", "TL")) %>% 
+  select(WEIGHT, Level, RMTYPE) %>% 
+  group_by(Level, RMTYPE) %>%
+  summarize(mean=mean(WEIGHT, na.rm=TRUE),
+            sd= sd(WEIGHT, na.rm=TRUE),
+            min=min(WEIGHT, na.rm=TRUE),
+            max=max(WEIGHT,na.rm=TRUE)) %>% 
+  unite("min/max", min:max, sep= "/", remove = TRUE) %>% 
+  ungroup() %>% 
+  arrange(factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
+                                        "TYPE 12", "TYPE 13","TYPE 13B", 
+                                        "TYPE 11", "TYPE 11F", "Oolitic")))
+
+stats_4 <- T_stats %>%
+  filter(Level=="4") %>% 
+  select(!(Level)) %>% 
+  rename_at(c(2,3,4), ~ paste(., 4, sep = "."))
+
+stats_4C <- T_stats %>%
+  filter(Level=="4C") %>% 
+  select(!(Level)) %>% 
+  rename_at(c(2,3,4), ~ paste(., "4c", sep = "."))
+
+stats_4D <- T_stats %>%
+  filter(Level=="4D") %>%
+  select(!(Level)) %>% 
+  rename_at(c(2,3,4), ~ paste(., "4d", sep = "."))
+
+stats_4E <- T_stats %>%
+  filter(Level=="4E") %>%
+  select(!(Level)) %>% 
+  rename_at(c(2,3,4), ~ paste(., "4e", sep = "."))
+
+stats_4E5 <- T_stats %>%
+  filter(Level=="4E/5") %>% 
+  select(!(Level)) %>% 
+  rename_at(c(2,3,4), ~ paste(., "4e5", sep = "."))
+
+stats_5 <- T_stats %>%
+  filter(Level=="5") %>%
+  select(!(Level)) %>% 
+  rename_at(c(2,3,4), ~ paste(., "5", sep = "."))
+
+stats_6 <- T_stats %>%
+  filter(Level=="6") %>%
+  select(!(Level)) %>% 
+  rename_at(c(2,3,4), ~ paste(., "6", sep = "."))
+
+stats_7 <- T_stats %>%
+  filter(Level=="7") %>%
+  select(!(Level)) %>% 
+  rename_at(c(2,3,4), ~ paste(., "7", sep = "."))
+
+stats_table <- left_join (stats_4, stats_4C, by = "RMTYPE") %>% 
+  left_join(., stats_4D, by = "RMTYPE") %>% 
+  left_join(., stats_4E, by = "RMTYPE") %>% 
+  left_join(., stats_4E5, by = "RMTYPE") %>% 
+  left_join(., stats_5, by = "RMTYPE") %>% 
+  left_join(., stats_6, by = "RMTYPE") %>% 
+  left_join(., stats_7, by = "RMTYPE")
+```
+
+```{r}
+#| label: fig-weight
+#| fig-cap: Boxplot of weights.
+
+phase_dataset %>% 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
+                                        "TYPE 12", "TYPE 13","TYPE 13B", 
+                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  ggplot(aes(RMTYPE, WEIGHT, color = RMTYPE)) +
+  geom_boxplot() +
+  facet_wrap(vars(Level)) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  #geom_jitter(shape=16, position=position_jitter(0.2)) +
+  coord_cartesian(ylim =  c(0, 40)) +
+  scale_color_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank())
+
+```
+
+When plotting the weight, it is apparent that most of the samples are located under the 0.3 kg mark, with median and bottom quartiles often falling under the 0.5 kg line.
+Local types, such as Type 2, Type 7 and Type 12, although mostly represented by artifacts with a small mass (\~0.3 kg), show the heaviest and highest number of heavy artifacts.
+In comparison, other types, although showing some variation in the weight of artifacts, show lighter numbers of artifacts, and their means and medians are frequently closer to the bottom quartile.
+As seen in @fig-type-per-level B, non-local types show larger masses from level 5 to level 7.
+
+The same raw materials found in the Terrace area are also found in the Solutrean occupation of the Shelter, albeit with some differences.
+Although with similar percentages to the local cherts, the use of local chert in the Shelter area is heavily dominated by the brown/yellow cherts that frequently characterize the Lower Jurassic outcrops of the Algarve; although the other macroscopic types are present but not significant.
+The presence of cherts not congruent with the local sources is, alike level 4, close to 25%, although dominated mostly by Type 11.
+Type GB is present in higher numbers than in level 4, however still representing less than 5% of the assemblage.
+Similarly to the Solutrean occupations of the Terrace, although the artifacts are generally small (lower weights), the local types show a higher variability in mass, with several outliers and heavy samples when compared to the non-local types.
+
+```{r}
+#| label: fig-type-per-level_shelter
+#| fig-cap: "Chalcedony and chert types per level, without unidentifiable samples. Only percentages superior to the 5% threshold are presented in the figure."
+#| echo: false
+
+s_data_plots <- S_dataset %>% 
+  filter(RMTYPE != "INDET" & WEIGHT != "0") %>% 
+  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
+  filter(CODE != "B" & Layer != "D")
+
+shelter_count <- s_data_plots %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
+                                        "TYPE 12", "TYPE 13","TYPE 13B", 
+                                        "TYPE 11", "TYPE 11F","TYPE GB", "Oolitic", "TL"))) %>%
+  dplyr::group_by(Layer, RMTYPE) %>% 
+  dplyr::summarise(n = n()) %>%
+  mutate(percent = n/sum(n)) %>% 
+  ggplot(aes(x = Layer, y = percent, fill = RMTYPE)) +
   geom_bar(stat = "identity", width = 0.5) +
@@
 
-T_dataset %>%
-  filter(Z != "0") %>% 
-  filter(!Z >= "25") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
-  filter(RMTYPE != "INDET") %>% 
-  ggplot(aes(RMTYPE, Z, color = RMTYPE)) +
-  geom_violin(trim = FALSE, scale="width", bw=0.1) + 
-  labs(title="Violin plot", 
-       subtitle="City Mileage vs Class of vehicle",
-       caption="Source: mpg",
-       x="Class of Vehicle",
-       y="City Mileage") +
-  stat_summary(fun.data=mean_sdl, geom="errorbar", color="black", width=0.2)
-  ggsave("violin_plot.png")
-
-T_dataset %>% 
-  filter(Z != "0") %>% 
-  filter(!Z >= "25") %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
-  dplyr::group_by(Z, RMTYPE) %>% 
-  dplyr::summarise(n = n()) %>%
-  ggplot(aes(x = n, y = Z)) +
-  geom_point(alpha = 0.1, color = "coral") +
-  facet_grid(cols = vars(RMTYPE)) +
-  labs(x = "Count", y = "Depth (cm)") +
-   theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank())
-
-```
-
-The same raw materials found in the Terrace area are also found in the Solutrean occupation of the Shelter, albeit with some differences.
-Although with similar percentages to the local cherts, the use of local chert in the Shelter area is heavily dominated by the brown/yellow cherts that frequently characterize the Lower Jurassic outcrops of the Algarve; although the other macroscopic types are present but not significant.
-The presence of cherts not congruent with the local sources is, alike level 4, close to 25%, although dominated mostly by Type 11.
-Type GB is present in higher numbers than in level 4, however still representing less than 5% of the assemblage.
-
-```{r}
-#| label: fig-type-per-level_shelter
-#| fig-cap: "Chalcedony and chert types per level, without unidentifiable samples. Only percentages superior to the 5% threshold are presented in the figure."
-#| echo: false
-
-S_dataset %>%
-  filter(RMTYPE != "INDET") %>% 
+shelter_weight <- s_data_plots %>%
   mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
@@
                                         "TYPE 11", "TYPE 11F","TYPE GB", "Oolitic", "TL"))) %>%
-  dplyr::group_by(AREA, RMTYPE) %>% 
-  dplyr::summarise(n = n()) %>%
-  mutate(percent = n/sum(n)) %>% 
-  ggplot(aes(x = AREA, y = percent, fill = RMTYPE)) +
+  dplyr::group_by(Layer, RMTYPE) %>% 
+  dplyr::summarise(wsum = sum(WEIGHT)) %>% 
+  mutate(percent = wsum/sum(wsum)) %>% 
+  ggplot(aes(x = Layer, y = percent, fill = RMTYPE)) +
   geom_bar(stat = "identity", width = 0.5) +
-  scale_y_continuous(labels = scales::percent_format()) +
-  theme_minimal() + #type of theme
-  labs(x = "Level", y = NULL) +# change labels
+  scale_y_continuous(labels = scales::percent_format(), position = "right") +
+  #theme_minimal() + #type of theme
+  labs(x = "Level", y = "Weight %") +# change labels
   guides(fill=guide_legend(title=NULL)) + # remove legend title
@@
                 position=position_stack(vjust=0.5), colour="white", size=3) +
-  theme(axis.title.x = element_text(size = 10, vjust = -1)) %>% 
+  theme(axis.title.x = element_text(size = 10, vjust = -1)) + 
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
@@
 
+ggarrange(shelter_count, shelter_weight, labels = c("A", "B"), ncol=2, common.legend = TRUE, legend="bottom")
+
 ```
@@
 ```{r}
-#| label: fig-type13-source
-#| fig-cap: Comparison between chert samples of different facies from Type 13 of the Terrace area (left) and samples from the Universidad de Cádiz lithotheque.
+#| label: weight-shelter
+#| fig-cap: Boxplot of weights.
 
-knitr::include_graphics(here::here("analysis/figures/Terrace_UCA_comparison.png"))
+s_data_plots %>% 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","TL") & Layer != "ND") %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
+                                        "TYPE 12", "TYPE 13","TYPE 13B", 
+                                        "TYPE 11", "TYPE 11F","TYPE GB", "Oolitic"))) %>%
+  ggplot(aes(RMTYPE, WEIGHT, color = RMTYPE)) +
+  geom_boxplot() +
+  facet_wrap(vars(Layer)) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  #geom_jitter(shape=16, position=position_jitter(0.2)) +
+  coord_cartesian(ylim =  c(0, 40)) +
+  scale_color_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank())
 
@@
 #| tbl-cap: A table of TL
+#| include: false
 
@@
 
+kable(head(TL_table))
+
 ```
@@
 
-# Conclusion
+The macroscopic and petrographic analysis of chert and chalcedony lithic materials from the UP levels of the Terrace and Shelter areas of Vale Boi have provided crucial insights into the raw material procurement and mobility patterns of hunter-gatherers.
+These raw materials are useful proxies to identify mobility patterns since they show patterns of local and non-local procurement.
+The prevalence of local raw materials in the Solutrean and Proto-Solutrean assemblages show a preferential use of local resources, although with discrete differences in the used nodules/outcrops.
+The presence of non-local raw materials throughout the stratigraphy suggests a continuous pattern of long-distance sourcing, which was intense during the Gravettian and slowed down with time.
+
+Despite differences in proportion, there seems to be continuity in the raw materials used throughout the UP, including the allochthonous sources.
+This may be explained by the tradition of shared knowledge and culture regarding the location of raw materials and other resources in the landscapes, or the maintenance of social networks through time.
+In this sense, raw materials and the connection they suggest seem to present themselves as resilient components of UP occupations and organization, and less prone to cultural or technological change.
+This persistence however, may also be explained by cached raw materials that were accumulated at the site and were intensively used and reused throughout several re-occupations of the site, eventually disappearing and leading to a more intense use of the local sources, either through large nodules or large blanks that remained unused [@surovell_toward_2009; @kuhn_mousterian_1995; @hofmanSHIFTINGSANDSFOLSOMMIDLAND2024; @binford_organization_1979].
+
+The local chalcedony and chert show the groups were obtaining a large portion of their raw materials around the 20 km range of the site.
+The existence of direct procurement of raw materials may be considered since the known sources of chalcedony are located at 7-6 km of distance from Vale Boi, and chert can be found in the inland outcrop of Ferrel, \~8 km from the site.
+This hypothesis may be strengthened by the existence of UP occupations at Ferrel, which may correspond to quarrying and chert catchment during prehistory [@zambujo_jazidas_1998].
+The procurement of local cherts may have also occurred for larger distances, possibly through embedded activities.
+The other chert sources (Lower Jurassic and Upper Jurassic) of western Algarve are located by the beach and cliff areas as previously mentioned, between 18-15 km of Vale Boi.
+Although our current data does not allow us to identify which specific outcrops were visited based on macroscopic and petrographic observations, the presence of marine resources, such as shells, at the site shows these groups visited the coast, which means they could easily obtain raw materials in the vicinity without necessary detours [@binford_organization_1979; @surovell_toward_2009].
+We may see these patterns of close-distance procurement and possibly caching in the presence of heavy (large mass) nodules and chunks of chert from local types, as well as the considerable amount of outliers (representing larger blocks and chunks) in the weight boxplots throughout the stratigraphy (@fig-weight).
+
+However, the discussion involving local cherts seems further complicated by the existence of similar outcrops in eastern Algarve.
+Previous studies showed that macroscopic criteria is not enough to discern whether the used sources are Lower Jurassic by the site or Middle Jurassic since they are similar macroscopically and petrographically, and with similar quality [@belmiro2023].
+In this sense, and similar to other archaeological studies [@surovell_toward_2009], we cannot discredit that some portion of the local cherts we interpret as being procured within a 20 km radius of Vale Boi, may be the result of long-distance procurement or the result of regional mobility, between \~80-100 km where the Middle Jurassic cherts are located.
+In fact, previous studies have shown that the outcrops in this region were used, due to the presence of UP archaeological sites by the outcrops \[++\].
+
+Given the similar quality of the raw materials and lack of geochemical studies, and following raw material formal and informal models for raw material procurement, we interpret that most of these raw materials are probably from the local (under 20 km sources) while a small quantity may be the result of long distance displacements.
+Similarly, the Upper Jurassic cherts in proximity to the latter sources, around 90 km, despite having macroscopic characteristics not congruent with the local sources, have not been identified in the assemblages.
+This may be explained by the medium quality and generally small size of the nodules (frequently between 3-5 cm), or the location of the outcrops, since the known outcrops are located inland, and it may be possible that they were not accessible or visible.
+
+The non-local cherts seem to show two modalities: knapped at the site, and present in larger quantities, and individual samples of variable chert and chalcedony types, but present in a limited number of artifacts.
+The allochthonous cherts with possibly identified sources seem to come from two directions: 1) north, where the Cretaceous cherts from Central Portugal were obtained, along with the small quantity of jasper; 2) west, where the peloidal/oolitic chert from south Spain can be found.
+Despite both these sources being \~300 km away, these cherts may have been obtained at closer distances, through riverbeds with smaller transport distances, without pebble-like cortex.
+The near nonexistence of pebble cortex does shows they were likely not provisioned at the beach or riverbeds.
+Despite these high distance travel paths, studies have suggested the existence of low-cost paths to travel through the Iberian Peninsula.
+In fact, previous formal models designed to explain hunter-gatherer mobility have suggested that variables such as how topography influences mobility may be important for the distances traveled by groups or individuals [@brantinghamMeasuringForagerMobility2006].
+For example, Castillo [-@Quirós Castillo, 2020] describes how Copper age groups reduced subsistence risks by creating exchange networks and practicing regional mobility seen in the exchange of autochthonous raw materials.
+These exchanges would have come with excessive energy costs and risks.
+However, the author points out that there are important features about the Iberian Peninsula regarding mobility such as the low cost of moving through the plateaus and the connectivity of the Guadalquivir River valley with the Guadiana and Tagus rivers (Quirós Castillo, 2020).
+In this sense, albeit distant, the existence of the north/east directions of exchange at Vale Boi may simply reflect the use of these natural highways for hunter-gatherer mobility and explain the high percentages of allochthonous chert use.
+The identification of the smaller subset of unknown source allochthonous cherts and chalcedonies (TL) may reflect different types of long-distance exchanges of finished products and blanks or with different social networks, possibly even reflecting individual networks and mobility [sensu @gamblePalaeolithicSocietiesEurope1999].
+
+A higher percentage of these allochthonous cherts, especially as we see in the Gravettian occupation, may be interpreted then as a high degree of communication and mobility between the groups occupying Vale Boi and an extended social network elsewhere rather than secondary catchment.
+This is a significantly different interpretation from that obtained from the Gravettian technology from south Portugal.
+Using a technological approach to the Gravettian lithic assemblages of Vale Boi, Marreiros et al.
+[-@marreiros2015] concluded that there are significant differences between the assemblages of central Portugal and southern Portugal.
+In this sense, studies suggest the presence of Gravettian regional technological facies.
+These have been interpreted as a different exploitation of territories as well as the reflection of different ecological and ethnographic boundaries between hunter-gatherers [@marreiros2015].
+Based on these combined data, it may be possible that the technology data indeed reflects different coexistence boundaries, as the regional band is often no more than ++ km \[++\], as such not being expected to represent the full range of hunter-gatherer mobility \[++\].
+The existence of a high frequency of allochtonous raw materials may be the reflection of these contacts and exchanges between different groups, focusing on good quality raw materials in detriment of those found locally, while still maintaining boundaries between groups, expressed at the technological level.
+A similar scenario is expressed by Brantingham [@brantinghamMeasuringForagerMobility2006] when discussing social exchange; that exchange may occur between two bands occupying non-overlapping territories but with established regular meetings between, when they would exchange their raw materials, thus explaining how very long-distance sources might be identified in archaeological sites.
+
+A different interpretation for the high frequencies of non-local cherts during the Gravettian may be used following the "Mean per Capita Occupation Span" model by Surovell [-@surovell_toward_2009].
+According to this model, short occupations should display high proportions of transported artifacts (in this case, understood by allochthonous raw materials and artifacts produced with these raw materials), and as the duration of the occupation increases, local artifacts should become dominant.
+Similarly, the author also suggests long-term occupations may be a product of the environment, as patchy environments and colder temperatures present risks to increased mobility, and the settlement of camp in a locally abundant environment in a scarce regional setting could lead to long-term occupations which would present a less risky option [@surovell_toward_2009][Kelly 1995].
+In this theoretical paradigm, the Gravettian occupations could be the result of high mobility and short-term occupations of Vale Boi, with possible reoccupations [sensu @surovell_toward_2009], thus showing high amounts of used non-local raw materials.
+As the climatic conditions begin to change, with the onset of Heinrich Event 2 during the Proto-Solutrean and the consequent glacial maximum that characterized the Solutrean [@sanchez_goni_millennial-scale_2010, @schmidt_rapid_2012], which caused a dryer, harsher climate with patchy vegetation \[++\], the resource abundant characteristics of an ecological niche region such as southwestern Portugal \[++\], could have decreased the mobility of hunter-gatherers.
+With longer-term occupations to reduce risk, the amount of non-local raw materials from previous expeditions (obtained both through catchment at the source or through trade), would eventually reduce as local raw materials became more dominant.
+
+In general, our results build upon previous preliminary studies, where the majority of the chert used seemed to be the local Lower Jurassic cherts within a 20 km range of the archaeological site, while non-local materials composed a small percentage of used chert [@pereira_territory_2016-1].
+Previously suggested long-distance sourcing, such as the Cretaceous cherts from central Portugal are confirmed through visual and petrographic data [@pereira_territory_2016-1, ++], as well as the north and east mobility paths that connected the hunter-gatherer groups from the southwest of Portugal to the center of Portugal and the south of Spain [@pereira_territory_2016-1].
+These cherts and chalcedonies with different sources present a picture of hunter-gatherers with complex systems and modes of exploring the landscape, managing resources, and possibly managing risk.
+Allochthonous raw materials, even if a small percentage in the Proto-Solutrean and Solutrean occupations, the Vale Boi case study shows that hunter-gatherers occupying a marginal territory in the Iberian Peninsula and Europe, bordered south and west by sea, did not live isolated and instead demonstrate an extensive social network.
+Alike the results identified in central Portugal (++), we also identify a pattern of interconnection between the Upper Paleolithic hunter-gatherers of the Iberian Peninsula.
+
+This study showed the potential of macroscopic and petrographic analysis to identify different chert and chalcedony types and identify their possible sources.
+However, the macroscopic and petrographic observations have frequently been associated with geochemical analyses, through methods such as XRF or LA-ICP-MS to further strengthen the attribution of source.
+For future research, we aim to use of a geochemical approach in the studied collection and geological reference collection to further expand upon the discussion of mobility through the UP of Vale Boi.
+Testing with fire and the creation of an altered reference collection may also aid in strengthening the raw material types and group of currently unidentifiable samples as mentioned in previous studies [@pereira_territory_2016-1].
+Finally, for future research, we aim to correlate technological data from existing and in-progress datasets to further understand how raw materials (local and non-local) were being used and managed through time and include these in the broader spectrum of the changing environmental conditions occurring throughout the Upper Paleolithic.
 

### 2024-01-05 analysis/supplementary-materials/SOM_datadictionaries.qmd
- First (first) draft of the article.

---
title: "Data dictionaries"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
title-block-published: "Last updated"  
format: 
  pdf:
    classoption: landscape
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: ../paper/references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

# 1. Data dictionaries

## 1.1. Group characterization

## 1.2. Single characterization

## 1.3. Petrographic characterization

# 2. References

### 2024-01-08 analysis/paper/paper.qmd
First draft

@@
 ---
-title: "Title Goes Here"
+title: "Beyond the coastal periphery: chert use and procurement patterns in the Upper Paleolithic of Southwesternmost Portugal"
 author:
-  - Jane Doe:
+  - Joana Belmiro:
       correspondence: "yes"
-      email: janedoe@fosg.org
-      orcid: 0000-0003-1689-0557
+      email: jfbelmiro@ualg.pt
       institute:
-        - fosg
-        - fop
-  - John Q. Doe:
-      institute: fosg
-      orcid: 0000-0003-1689-0558
-  - Peder Ås:
-      institute: fosg
-      orcid: 0000-0003-1689-0559
-  - Juan Pérez:
-      orcid: 0000-0003-1689-0551
+        - ICArEHB
+  - Xavier Terradas:
       institute: 
-        - name: Acme Corporation
-  - Max Mustermann:
-      orcid: 0000-0003-1689-0552
+        - CSIC
+  - Salvador Dominguez-Bella:
       institute: 
-  - fosg:
-      name: Formatting Open Science Group
-      address: 23 Science Street, Eureka, Mississippi, USA
-  - fop: Federation of Planets
+        - UCA
+  - JOão Cascalheira:
+      institute:
+        - ICArEHB
 title-block-published: "Last updated" 
-date: now
-date-format: long
 format: 
@@
 
-Knappable raw materials are a key element used to understand the lifeways of past hunter-gatherers throughout time, as lithics are ubiquitous throughout prehistory.
-More than mere stones with which stone tools are produced, these resources are intimately connected to the group’s technological, social, and cultural organization, possibly with an impact on the group's survivability [@oestmoFormalModelingApproach2017].
-Several key topics to understand hunter-gatherer behavior and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binford_organization_1979; @binford_righteous_1985; @gould_lithic_1985; @gould_lithic_1985; @kuhn_unpacking_1991; @mccall_behavioral_2007], occupation types and their duration [@kuhn_upper_2004; @surovell_toward_2009]) to the establishment and dimension of social networks [@whallon_social_2006] and exchange between groups or individuals [@wurzHowiesonsPoortBacked1999; @gamblePalaeolithicSocietiesEurope1999].
-This has frequently been done through the systematic characterization of raw materials (both geologically available and archaeological), identification of corresponding sources for a clear differentiation between local and long-distance raw materials, and the study of their proportions within a site, often throughout different occupations and chronologies.
-These approaches and concepts have been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula (++).
-Frequently coupled with technological analyses, the study of raw materials, especially chert, through a wide range of methods, such as macroscopic, petrographic, and geochemical, have recurrently aided in answering key questions to Iberian archaeology, such as the disappearance of Neanderthals and consequent replacement by Modern Humans [@aubry_far_2022].
-Raw material studies have also contributed to understanding how different groups during different chronological periods organized their technology, managed time, and risk (through the embedded/direct procurement approach), and chose resources [@matias2016; @matias_o_2012; @pereira_raw_2021; @pereiraLithicEconomyTerritory2016].
-Furthermore, these studies frequently highlight long-distance procurement even if in low proportions at times.
-In fact, raw material studies and the distinction between local/regional and exotic have also been successfully applied to identify long-distance contacts (\>150 km) during the Upper Paleolithic in the Côa Valley which has been interpreted as the result of reciprocal exchange [@aubry_raw_2004-1; @aubry_upper_2016].
-Although commonly known and identified during late prehistory [@nocete_circulation_2005; @rodriguezPrehistoricFlintExploitations2011; @costa_lithic_2022], the identification of these contacts has expanded our understanding of often large-scale social networks and contacts during the Upper Paleolithic in central Portugal, to paint a picture previously outlined of a wide interconnected territory.
+Knappable raw materials play a crucial role in understanding the lifeways of past hunter-gatherers across time, given the ubiquitous presence of lithics throughout prehistory.
+More than mere stones with which stone tools are produced, these resources are intimately connected to a group’s technological, social, and cultural organization, potentially influencing the group's overall survivability [@oestmoFormalModelingApproach2017].
+Several key topics to the study of hunter-gatherer behavior and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binford_organization_1979; @binford_righteous_1985; @gould_lithic_1985; @gould_lithic_1985; @kuhn_unpacking_1991; @mccall_behavioral_2007], occupation types and their duration, [@kuhn_upper_2004; @surovell_toward_2009] to the establishment and dimension of social networks [@whallon_social_2006] as well as exchanges between groups or individuals [@wurzHowiesonsPoortBacked1999; @gamblePalaeolithicSocietiesEurope1999].
+This has frequently been done through the systematic characterization of geological and archaeological raw materials, the identification of corresponding sources for a clear differentiation between local and long-distance raw materials, and the study of their proportions within a site, often throughout different occupations and chronologies.
+These approaches and concepts have also been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula (++).
+Frequently coupled with technological analyses, the study of raw materials, especially chert, through a wide range of methods such as macroscopic, petrographic, and geochemical, have recurrently aided in answering key questions to Iberian archaeology.
+One such question is the disappearance of Neanderthals and consequent replacement by Modern Humans [@aubry_far_2022].
+Raw material studies have also contributed to understanding how different groups organized their technology, managed time, and risk (through the embedded/direct procurement approach) and chose resources during different chronological periods [@matias2016; @matias_o_2012; @pereira_raw_2021; @pereiraLithicEconomyTerritory2016].
+Furthermore, these studies frequently highlight long-distance procurement, even if in low proportions at times.
+In fact, raw material studies and the distinction between local/regional and exotic have also been successfully applied to identify long-distance contacts (\>150 km) during the Upper Paleolithic (UP) in the Côa Valley, interpreted as the result of reciprocal exchange [@aubry_raw_2004-1; @aubry_upper_2016].
+Although commonly known and identified during late prehistory [@nocete_circulation_2005; @rodriguezPrehistoricFlintExploitations2011; @costa_lithic_2022], the identification of these contacts has expanded our understanding of often large-scale social networks and contacts during the UP in central Portugal, painting a picture of a wide interconnected territory.
 However, unlike other regions of the Iberian Peninsula such as central Portugal, and despite the potential of raw material studies to provide key data on the lifeways and organization of Paleolithic hunter-gatherers through time, as well as the mobility range and existence of social networks, the behaviors of management and procurement of raw materials are still in a preliminary phase in southern Portugal, especially in the Algarve region.
-Although much is known about the technological organization of Paleolithic hunter-gatherer groups in southwestern Portugal, especially due to the archaeological site of Vale Boi (see ++), raw material studies have been scarcer.
+Although much is known about the technological organization of Paleolithic hunter-gatherer groups in southwestern Portugal, especially due to the archaeological site of Vale Boi [see @cascalheira_vale_2013; @bicho_ecodynamics_2013; @marreiros2015], raw material studies have been scarce.
 Previous studies applied macroscopic methods to characterize the cherts from Vale Boi.
-These preliminary results have shown the presence of a large component of local cherts from the Jurassic limestone formations in conjunction with a less abundant and variable allochthonous chert component throughout the several Upper Paleolithic occupations of the site [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004].
+Preliminary results showed the presence of a large component of local cherts from the Jurassic limestone formations, along with a less abundant and variable allochthonous chert component throughout the several UP occupations of the site [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004].
 @pereira_territory_2016-1 suggest that the exogenous group may have been the result of embedded procurement within a 16 km radius, while the allochthonous group may have been the result of transportation during trips to other areas of the Algarve, Alentejo, or even south Spain.
 The presence of chert from the Cretaceous formations of central Portugal has also been suggested [@bicho_ecodynamics_2013].
-Furthermore, previous studies aiming to characterize the local and regional chert sources in the Algarve region have established a comparative macroscopic and petrographic database [@belmiro2023; @pereira_presenting_2016-1], a key element to raw material studies, especially to distinguish between local and non-local raw materials and understanding changes in the archaeological record [@pop_simulating_2015; @surovell_toward_2009].
-These preliminary results and the existence of regional reference collections show the potential for a systematic raw material study at the site, using different methods and approaches, applied to the lithic assemblages throughout the several Upper Paleolithic occupations, to understand how these hunter-gatherer groups behaved through time: how they explored their landscape and lithic resources, where they moved across a territory nestled between water and land, and with whom did they establish social networks.
-To answer these questions, we aim to characterize the raw material procurement and mobility patterns of hunter-gatherer groups throughout the Upper Paleolithic of southern Portugal.
-By studying the chalcedony and chert artifacts of the UP levels of Vale Boi we aim to characterize these raw materials macroscopically and petrographically, to identify different chert groups and their respective sources.
+Furthermore, previous studies aiming to characterize the local and regional chert sources in the Algarve region have established a comparative macroscopic and petrographic database [@belmiro2023; @pereira_presenting_2016-1], a key element in raw material studies, especially to distinguish between local and non-local raw materials and understanding changes in the archaeological record [@pop_simulating_2015; @surovell_toward_2009].
+These preliminary results, coupled with the existence of regional reference collections, show the potential for a systematic raw material study at the site, using different methods and approaches and applied to the lithic assemblages throughout the several Upper Paleolithic occupations, to understand how these hunter-gatherer groups behaved through time: how they explored their landscape and lithic resources, where they moved across a territory nestled between water and land, and with whom did they establish social networks.
+To address these questions, we aim to characterize the raw material procurement and mobility patterns of hunter-gatherers throughout the Upper Paleolithic of southern Portugal.
+By studying the chalcedony and chert artifacts of the UP levels of Vale Boi we aim to characterize these raw materials macroscopically and petrographically, to identify different chert groups and their souces.
 
@@
 
-Vale Boi is located on the western coast of the Algarve region (south Portugal), in a small valley that connects south to the Atlantic current coast at around 2 km and is bordered by a limestone hill.
-This hill is characterized by limestone exposures that form rocks shelters facing west and southwest [@cascalheira_tecnologia_2010].
-The site extends for more than 10 000 sq m through the slope of the valley, through which three main occupation areas have been identified: Slope, Shelter and Terrace.
+Vale Boi is located on the western coast of the Algarve region in southern Portugal, within small valley that connects south to the Atlantic current coast at around 2 km.
+It is bordered by a limestone hill, characterized by limestone exposures that form rocks shelters facing west and southwest [@cascalheira_tecnologia_2010].
+The site extends for more than 10,000 sq m along the slope of the valley, through which three main occupation areas have been identified: Slope, Shelter and Terrace.
 The excavation methodology included the recording of all artifacts with dimensions superior to 2 cm or complete small artifacts (i.e., bladelets) with a total station, and was applied to all main areas.
 Several occupations have been identified in both the Terrace and the Shelter area.
-The Terrace includes occupations from the Holocene to the Upper Paleolithic, resulting in the identification of eight main lithostratigraphic units.
-The UP occupations start with the Gravettian (levels 8-6), followed by Proto-solutrean (bottom of 5 and 5/4E, sensu Belmiro et al., 2021), and Solutrean occupations (4D-4 and 3B).
+The Terrace includes occupations from the Holocene to the Upper Paleolithic, resulting in the identification of eight main lithostratigraphic units where several UP occupations were identified, from the Gravettian, Proto-Solutrean and Solutrean.
 The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations.
@@
 
+TABLE: Table with short layer description, attributed technocomplex, and radiocarbon dates. Will need complete field layer description and the dates we currently have for VB from JC.
+
+MAP: With geological context, location of VB, Iberian Peninsula small map.
+
 ## Geological context
 
-The Algarve region is a complex territory, characterized by several geomorphic sub-regions and geological units, which provide different lithological resources used throughout prehistory.
-There are two main geological units, the South Portuguese Zone (SPZ) in the north sector of Algarve and the Algarve basin to the south.
-The SPZ is characterized by schists, greywacke and quartzites, while the Algarve basin is characterized by limestones, dolomites, and marls.
+The Algarve region is a complex territory, marked by several geomorphic sub-regions and geological units that provide different lithological resources used throughout prehistory.
+Two main geological units characterize the region: the South Portuguese Zone (SPZ) in the north sector of Algarve and the Algarve basin to the south.
+The SPZ is characterized by schists, greywacke and quartzites, while the Algarve basin features limestones, dolomites, and marls.
 It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules and beds can be found.
-Chalcedony is located in a 15 km radius from Vale Boi, in the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013].
-The cherts can be found in different Jurassic formations throughout the Algarve, but are most frequently known for the Lower Jurassic beds and small nodules in the limestone cliffs at ++ km distance from the site, which have been previously studied and characterized [@ribeiro_evolucao_2005; @belmiro2023].
+Chalcedony is located approximately 10 km from Vale Boi, in the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013].
+The cherts can be found in several Jurassic formations throughout the Algarve, but are frequently known for the Lower Jurassic beds and small nodules in the limestone cliffs at \~20 km from the site, which have been previously studied and characterized [@ribeiro_evolucao_2005; @belmiro2023].
 Chert is also available in the less known Upper Jurassic limestone formation in western Algarve (++ km from the site), and in the center and eastern portion of the Algarve, in Middle and Upper Jurassic formations with chert nodules [@belmiro2023].
+
+QUESTION: Should I include a short description of the cherts in the text, or do this through an image or table?
 
@@
 To better characterize the different chert groups identified in the macroscopic analysis, 22 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory.
-The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels (@tbl-thinsections).
+The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels (@tbl-TS-list).
 For groups that showed macroscopic variability, several thin sections were analyzed to ensure the collection of petrographic data that reflected this variability.
@@
 
-| Type               | Sample ID | Level | TS \# |
-|--------------------|-----------|-------|-------|
-| Chalcedony         | I19-1615  | 4D    | 8     |
-| Type 2             | I19-2835  | 5     | 17    |
-| Type 2 (rmu_17)    | I19-2226  | 4E    | 18    |
-| Type 2 (altered)   | H18-1938  | 4D    | 20    |
-| Type 2 (reddening) | H19-2426  | 4D    | 3     |
-| Type 2B            | H21-3095  | 4C    | 6     |
-| Type 2B            | H21-4234  | 5     | 7     |
-| Type 15            | I20-3160  | 5     | 10    |
-| Type 7             | H20-2441  | 4E    | 19    |
-| Type 12            | I21-3252  | 4E    | 14    |
-| Type 12B           | H18-2708  | 5     | 2     |
-| CPT                | I19-3350  | 6     | 9     |
-| CPT                | J18-778   | 5     | 13    |
-| CPT                | I20-3689  | 6     | 16    |
-| Sub-type LLB       | I20-3951  | 6     | 11    |
-| Sub-type 11E       | H19-4074  | 6     | 4     |
-| Type 11F           | H19-2924  | 4E    | 1     |
-| Type 13 (01b; 01a) | L19-64    | 5     | 12    |
-| Type 13 (04)       | J18-1264  | 8     | 22    |
-| Type 13B           | I21-2966  | 4C    | 15    |
-| Oolitic            | H19-4216  | 6     | 5     |
-| White patinated    | H20-4166  | 6     | 21    |
+```{r}
+#| label: tbl-TS-list
+#| tbl-cap: A table with a thin section list.
 
-: A table with thin sections {#tbl-thinsections}
+TS_list <- read_csv(here::here('analysis/data/raw_data/TS_short_list.csv'))
+
+kable(TS_list)
+
+```
 
@@
   filter(Level !="4B") %>%  # Filtering out levels with barely any artifacts
-  mutate(Level = case_when(Level == "7C" ~ "7", # Collapsing samples from Level 7C into 7
+  mutate(Level = case_when(Level == "7C" ~ "7",
+                           Level == 8 ~ "7", # Collapsing samples from Level 7C into 7 and two samples from Level 8 into 7, since 8 disappears in field_data
                            TRUE ~ Level)) %>% 
-  mutate(Level = case_when(Level == 8 ~ "7", # Collapsing two samples from Level 8 into 7, since 8 disappears in field_data
-                           TRUE ~ Level))
+  mutate(RMTYPE = case_when(RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
+                            RMTYPE == "TYPE 15" ~ "TYPE 3",
+                            RMTYPE == "TYPE 7" ~ "TYPE 4",
+                            RMTYPE == "TYPE 12" ~ "TYPE 5",
+                            RMTYPE == "TYPE 11" ~ "TYPE 6",
+                            RMTYPE == "TYPE 13" ~ "TYPE 7",
+                            RMTYPE == "TYPE 11F" ~ "TYPE 8",
+                            RMTYPE == "TYPE 13B" ~ "TYPE 9",
+                            RMTYPE == "Oolitic" ~ "TYPE 10",
+                            RMTYPE == "GB" ~ "TYPE 11",
+                            TRUE ~ RMTYPE))
 
@@
     RMTYPE == "Type 2" ~ "TYPE 2",
+    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
+    RMTYPE == "TYPE 15" ~ "TYPE 3",
+    RMTYPE == "TYPE 7" ~ "TYPE 4",
+    RMTYPE == "TYPE 12" ~ "TYPE 5",
+    RMTYPE == "TYPE 11" ~ "TYPE 6",
+    RMTYPE == "TYPE 13" ~ "TYPE 7",
+    RMTYPE == "TYPE 11F" ~ "TYPE 8",
+    RMTYPE == "TYPE 13B" ~ "TYPE 9",
+    RMTYPE == "Oolitic" ~ "TYPE 10",
+    RMTYPE == "TYPE GB" ~ "TYPE 11",
     .default = RMTYPE))
@@
   filter(RMTYPE !="INDET") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F", "Oolitic", "GB", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   tabyl(RMTYPE, Level) %>%
@@
 @fig-rm-comparison shows that for the selected levels, chert represents only \~20-15% of the lithic assemblage.
-Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke \[++\].
-Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site \[++\].
+Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke.
+Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site [@pereira_territory_2016-1].
 Frequently, the quartz and greywacke assemblages include large amounts of shattered and/or unknapped chunks or slabs.
-Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021].
+Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiro_proto-solutrean_2020].
 Based on this, despite the low amounts of individually plotted chert artifacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.
@@
 
-context_short <- context %>% 
-  filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
-dataset_short <- T_dataset %>% 
+context_short <- field_data %>% 
+  filter(Level !="4B") %>%  # Filtering out levels with barely any artifacts
+  mutate(Level = case_when(Level == "7C" ~ "7",
+                           Level == 8 ~ "7", # Collapsing samples from Level 7C into 7 and two samples from Level 8 into 7, since 8 disappears in field_data
+                           TRUE ~ Level)) %>% 
+  mutate(Level = case_when(
+    Level == 5 & Z > 24.1 ~ "4E/5",
+    Level == 5 & Z < 24.1 ~ "5",
+    TRUE ~ Level)) %>% 
+  filter(!Level %in% c("3B", "5B", "5C"))
+
+dataset_short <- phase_dataset %>% 
   select(ID, RMTYPE)
@@
     RMTYPE == "Other" ~ "Other",
-    RMTYPE == "Chalcedony" ~ "Chalcedony",
-    RMTYPE != "Other" | RMTYPE != "Chalcedony" ~ "Chert"))
+    RMTYPE == "TYPE 1" ~ "Chalcedony",
+    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert"))
 
@@
 
-Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony groups, present throughout the stratigraphy in the Terrace and Shelter area.
+### Cherts congruent with local sources
+
+Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony groups, present throughout the stratigraphy in the Terrace and Shelter area (TABLE with groups).
 A single different group was identified in the Shelter area, with limited presence in the Terrace (n=2), and 14 varieties of chert and 2 chalcedony composed of one to three samples were identified in several levels of the Terrace and Shelter.
@@
 
-The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts.
-The chalcedony is a massive, fibrous chalcedony without allochems.
+TABLE WITH SHORT GROUP DESCRIPTION.
+
+The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts (local_grid).
+The chalcedony (Type 1; A1) is a massive, fibrous chalcedony without allochems.
 It is characterized by a colorless and translucent appearance, often partially or completely white patinated.
@@
 It is macroscopically and petrographically congruent with the hydrothermal chalcedony found in lenses in Triassic formations, also frequently characterized by irregularities and fractures.
+NEED chalcedony data from previous works from X.T.
 
@@
 In general, these are opaque to sub-translucent cherts, with colors varying brown, yellow, pale red and grey, with the frequent presence of spicules and iron oxides.
-The differences are in their colors, nodule morphology, frequency, and variety of bioclasts, something already identified in the geological samples [@belmiro2023].
+The differences are in their colors, nodule morphology, frequency, and variety of bioclasts (Group 2 including triaxon sponge spicules and echinoderm spines), something already identified in the geological samples [@belmiro2023].
 Petrographically, however, these cherts are homogeneous, except for a small number of artifacts with a banded variety within group 5.
@@
 These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite.
-Group 12 shows higher percentages of micrite/sparite.
+Type 5 shows higher percentages of micrite/sparite.
 In all samples, fossils range from common to very frequent, albeit frequently unidentifiable.
-When identifiable in thin-section, they are sponge spicules, with the rare occurrence of possible foraminifera and ostracod in a sample from type 12.
+When identifiable in thin-section, they are sponge spicules (B2, C2, H2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (H3) in a sample from type 5.
 Porosity varies between 10% and 1%, of vug type.
-Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 15, 7 and 12 to be local, as they are congruent with the described characteristics.
-Groups 2 and 15 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils.
-The macroscopic similarities of Groups 7 and 12 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the Ferrel area.
+Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 3, 4 and 5 to be local, as they are congruent with the described characteristics.
+Groups 2 and 3 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils.
+The macroscopic similarities of Groups 4 and 5 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the Ferrel area.
 As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape.
@@
 
-Five chert types, not congruent with local sources, were identified (Groups 6 to 10), including the TL group.
+FIGURE: local_grid, variability of local chert macroscopic and petrographic characteristics
 
-Group 11 corresponds to a massive cryptocrystalline quartz with a mudstone texture.
-The lack of local similar cherts, as well as identifiable fossils does not allow us to attribute a depositional environment or epoch/stage.
-However, rare spicules were identified; given the exclusivity of the sponge’s (Porifera) living environment to be marine (Flügel, 2010), group 11 may be considered a chert formed in a marine environment.
+### Cherts not congruent with local sources
+
+We identified several chert types with macroscopic and petrographic characteristics that are not congruent with the chert available in known outcrops in western Algarve (nonlocal_grid).
+
+Group 6 (A1, B1, A2, B2) corresponds to a massive cryptocrystalline quartz with a mudstone texture, from a marine environment.
 This chert is translucent and shows a variety of colors, ranging from brown, red, and gray, which seem to vary in function of the artifact’s thickness.
-Patterns are rare, however, whenever present they are laminated, with flecks, and splotches.
-In unaltered samples, bioclasts are rare.
+In unaltered samples, visible bioclasts are rare.
 The cortex is rounded, and rarely pebble-like, ranging from thin to medium thickness.
-Two sub-groups were individualized based on macroscopic characteristics such as color, translucency variation, and patterns, although they may simply reflect the use of specific nodules from within Type 11’s variability.
+Two sub-groups were individualized based on macroscopic characteristics such as color, translucency variation, and patterns, although they may simply reflect the use of specific nodules from within Type 6’s variability.
 Five thin sections were produced to better characterize the macroscopic variability of this group and sub-groups.
@@
 It is mainly composed of cryptocrystalline quartz (between 95-97% in the samples), with accessory fibrous chalcedony found replacing fossils or filling fractures, and in some samples macrocrystalline quartz, microcrystalline quartz, and micrite/sparite.
-Allochems are opaques (u), iron oxides (u), and bioclasts (r/u).
-Bioclasts are unidentifiable and poorly preserved, replaced by fibrous chalcedony and rarely by microcrystalline quartz.
-Sponge spicules were identified in a single sample.
+Allochems are opaques, iron oxides, and bioclasts.
+Bioclasts are unidentifiable and poorly preserved, with rare identifiable fossils such as sponge spicules, replaced by fibrous chalcedony and rarely by microcrystalline quartz.
 Porosity is present between \<1-1%, of vug-type when identifiable.
+
+FIGURE: nonlocal_grid, variability of non-local chert macroscopic and petrographic characteristics.
 
@@
 These Cretaceous cherts, found in secondary deposition settings, are described as frequently translucent, with geodes, with colors ranging from yellow to red or grey, and mineralogically homogeneous between them.
-Petrographic studies highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matias2016].
+Petrographic studies from the Estremadura area highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matias2016].
 When identified, sponge spicules and possible rare foraminifera have been identified [@matias2012].
+Studies of Upper Cretaceous cherts from the Torres Vedras, Sintra and Lisbon area describe cherts with similar macroscopic characteristics, with micro-cryptocrystalline quartz mudstones with bioclasts such as sponge spicules or bivalve fragments, or wackestone/packstone textures with abundant bioclasts, including ostracods and different types of foraminifers [@jordaoFlintSourcesMobility2022].
 These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from central Portugal hosted at LusoLit.
-Alike the archaeological thin sections from Type 11 and congruent with other studies [@matias2016], the thin sections produced from the Cretaceous cherts show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
+Alike the archaeological thin sections from Type 6 and congruent with other studies [@matias2016; @jordaoFlintSourcesMobility2022], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
 From these cherts, only the dark grey nodules (RT231) showed identifiable fossils which were rare ostracods.
-Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Group 11 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts from the Estremadura area in central Portugal.
+Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Group 6 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts from the Estremadura and Lisbon region in central Portugal.
 
-Group 11F corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment.
-Identified and well-preserved fossils are foraminifera from the Pfenderenidae family, which are larger benthic foraminifera from marine environments.
-The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous.
-Macroscopically, this group is characterized by gray colors and a white area, frequently at the center of the sample, which seems to be a weathering rind and where the presence of fossils is highlighted.
-The translucency ranges from sub-translucent to opaque and visible patterns are splotches, broad mottling, spots, and speckling.
-Whenever present, the cortex is rounded and thin.
-Petrographically, this group is a packstone (mudstone texture close to the edges) with a massive microstructure.
-It is composed of 97% micro-cryptocrystalline quartz, macrocrystalline quartz, and fibrous chalcedony, found mostly replacing fossils.
-Other minerals include uncommon iron oxides.
-The porosity is of vug type (3%).
-Fossil content is very frequent, albeit mostly unidentifiable and poorly preserved.
-Identified fossils are rare sponge spicules , rare ostracods with some degree of preservation, and common small, unidentified foraminifera (100 μm) of differing preservation degrees.
-Identifiable foraminifera are rare and from the Pfenderenidae family (500 μm).
-Despite the macroscopic similarities between type 11F and cherts from the Andalusian chert formations, the samples inspected with a stereomicroscope showed no visible foraminifera, especially from the ++ family.
-However, the archaeological thin section revealed that despite no foraminifera being visible macroscopically, they are visible under the microscope in thin section.
-The fact that no foraminifera are visible in the the small number of analyzed samples from the Cadiz lithotheque may simply be a result of their uncommon visibility.
-Despite this, given the lack of this specific fossil in the reference samples, the probable source of Group 11F remains unknown.
-
-Group 13 is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
-It is characterized by two colors, white (opaque) and grey (sub-translucent), with splotches, speckles and horizontally dotted laminations.
+Group 7 (C1, C2, D1, D2) is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
+It is characterized by white and grey colors.
 The cortex is from an outcrop, rounded, and varying between medium thickness to thick.
@@
 Unlike macroscopical observations, bioclasts under the thin section are all unidentifiable, although rare fossils may be bivalve shells.
-Porosity is vug (10%).
-The consultation of the lithotheque hosted at UCA allowed us to identify a sub-set of cherts with similarities to those from Type 13.
+Porosity is variable (\<1-10%).
+The consultation of the lithotheque hosted at UCA allowed us to identify a sub-set of cherts with similarities to those from Type 7 from the Middle Subbetic region (south Spain).
 These cherts are massive, sub-translucent, peloidal/oolitic packstones.
@@
 The presence of bioclasts varies between samples, ranging between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon) and echinoderm spines (rare).
-Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allows us to interpret Type 13 as coming from the Andalusian region of southern Spain.
+Similar cherts have been identified and described in previous works, describing grey cherts with peloidal and oolitic facies and occasional foraminifera [@rodriguezPrehistoricFlintExploitations2011].
+Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allows us to interpret Type 7 as coming from the Upper Jurassic formations of the Betic Systems, extending from Cadiz to Alicante.
 
-Group 13B corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone.
+Group 8 (E, F1, F2, F3) corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment.
+Identified and well-preserved fossils are foraminifera from the Pfenderenidae family, which are larger benthic foraminifera from marine environments.
+The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous.
+Macroscopically, this group is characterized by gray and white colors.
+The translucency ranges from sub-translucent to opaque.
+Whenever present, the cortex is rounded and thin.
+Petrographically, this group is a packstone (mudstone texture close to the edges) with a massive microstructure.
+It is composed of 97% micro-cryptocrystalline quartz, macrocrystalline quartz, and fibrous chalcedony, found mostly replacing fossils.
+Other minerals include uncommon iron oxides and the porosity is of vug type (3%).
+Fossil content is very frequent, albeit mostly unidentifiable and poorly preserved.
+Identified fossils are rare sponge spicules , rare ostracods with some degree of preservation, common small, unidentified foraminifera (100 μm) of differing preservation degrees, and as aformentioned foraminifera from the Pfenderenidae family (500 μm).
+Despite the macroscopic similarities between type 8 and cherts from the Middle Subbetic region chert formations, the geological reference samples showed no macroscopically visible foraminifera.
+However, our results indicate that despite no foraminifera being visible macroscopically, they are present in thin section.
+The fact that no foraminifera were identified in the small number of analyzed samples from the Cadiz lithotheque may simply be a result of their uncommon visibility.
+Another possible source may be the Upper Cretaceous wackestone/packstone cherts with foraminifers from the Lisbon region [@jordaoFlintSourcesMobility2022].
+However, these studies do not specify the existence of Pfenderenidae foraminifers.
+Given the lack of this specific fossil in the reference samples from Andaluzian chert and the central Portugal cherts, the probable source of Type 8 remains unknown.
+
+Type 9 (G1, G2) corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone.
 It is opaque and has a heterogeneous structure and is characterized by a variable color distribution: it shows a horizontal laminated and finely laminated pattern, with gray dark gray bands intercalated with light gray bands.
@@
 Although its origin is unknown, when visiting the Cadiz lithotheque, we identified a small group of banded black cherts without noticeable fossils or inclusions.
-Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Group 13B belongs to Andalusian chert formations in southern Spain.
+Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Type 9 belongs to Andalusian chert formations in southern Spain.
 
-Group Oolitic is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
+Type 10 (H1, H2) is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
 Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: brown (opaque) and black (sub-translucent) fabric.
-The quality is acceptable since it has fine granularity and a smooth feel, but shows natural fractures and occasional inclusions, such as macrocrystalline quartz or chalcedony veins that may impact the knapping.
-The cortex is difficult to identify since it is similar to the brown fabric, highly altered, and with a gradual transition.
-Petrographically this group is a packstone with a massive microstructure.
+Petrographically this chert is a packstone with a massive microstructure.
 The chosen sample for the thin section is composed of 35% micro-cryptocrystalline quartz, 53% micrite, 9% macrocrystalline quartz, and accessory fibrous chalcedony (filling fractures) and mica; the percentages of quartz and micrite might be related to the fabrics present in the artifacts.
@@
 This chert shows no similarities to any consulted reference material.
-Despite the existence of oolitic limestones in the Algarve region, no chert nodules are mentioned (++) and the visit of outcrops mentioned in the geological cartography did not allow their identification; the trade of oolitic cherts in recent prehistory may indicate however that these types of cherts are available in the landscape in long-distance (++), and their presence in Vale Boi may simply reflect similar mobility patterns.
+Despite the existence of oolitic limestones in the Algarve region, no chert nodules are mentioned (++) and the visit of outcrops mentioned in the geological cartography did not allow their identification; however, the trade of oolitic cherts in recent prehistory may indicate that these types of cherts are available in the landscape in long-distance [@nocete_circulation_2005], and their presence in Vale Boi may simply reflect similar mobility patterns.
+
+Just realized I need to add Type 11 here. –
 
@@
 
-Apart from the groups common throughout the stratigraphy of the Terrace and Shelter, group GB and samples from the TL group also show no identifiable source.
+Apart from the groups common throughout the stratigraphy of the Terrace and Shelter, Type 11 and samples from the TL group also show no identifiable source.
 The exceptions are TL01, TL10-11 and TL15-16.
-Jasper is known in the SPZ area, from the Alentejo region to Spain, above the Algarve suggesting that the source of this jasper may be the SPZ formations.
-Similarly, despite different from the frequent Cretaceous nodules, the macroscopic similarities between TL10 and TL11 and specific nodules from Central Portugal in the Lusolit lithotheque may indicate these cherts are from the Estremadura area.
+Jasper is known in the SPZ area (Missing a ref here), from the Alentejo region to Spain, above the Algarve suggesting that the source of TL01 may be from the Évora/Beja region.
+Similarly, despite different from the frequent Cretaceous nodules, the macroscopic similarities between TL10 and TL11 and specific nodules from Central Portugal in the Lusolit lithotheque may indicate these cherts are from the Estremadura/Lisbon area.
 This difference in appearance may be due to the already mentioned variability of the nodule's macroscopic characteristics.
-Finally, TL15 and TL16 show macroscopic similarities to radiolarite reference samples located at the Cadiz lithotheque, which may indicate the source of these cherts is from Andalusian geological formations.
+Finally, TL15 and TL16 show macroscopic similarities to radiolarite reference samples located at the Cadiz lithotheque, which may indicate the source of these cherts is from Middle Subbetic region [@dominguez-bella_estudio_2006; @dominguezbellaMineralesRocasSociedades2010].
+
+DOUBT: Should I include here the figure with the smaller groups, or simply leave it in the SOM?
 
@@
 
-Chert groups are present across the stratigraphy (except for GB which appears only in level 4 and in a small number of artifacts; n=2).
+Chert groups are present across the stratigraphy (except for Type 11 which appears only in level 4 and in a small number of artifacts; n=2).
 Different chert types were consistently used in the Terrace area throughout time, with different intensities of use.
@@
 In the Solutrean and Proto-solutrean levels, local cherts make up more than \~70% of the chert (n) and 80% when considered the weight.
-Despite this homogeneous pattern, the use of group 2 and group 7 marks significant differences between the Solutrean occupations (levels 4, 4C, and 4D) and the Proto-Solutrean occupations (4E and 4E/5); in the first group the yellow/brown cherts (type 2) represent \~30% of local chert use, while in the second group, the use of type 2 diminishes drastically (\~10-15%).
-In contrast, Type 7 becomes the most used chert with percentages above 35% in level 4E and \~20% in the top of level 5.
+Despite this homogeneous pattern, the use of Type 2 and Type 4 marks significant differences between the Solutrean occupations (levels 4, 4C, and 4D) and the Proto-Solutrean occupations (4E and 4E/5); in the first group the yellow/brown cherts (Type 2) represent \~30% of local chert use, while in the second group, the use of Type 2 diminishes drastically (\~10-15%).
+In contrast, Type 4 becomes the most used chert with percentages above 35% in level 4E and \~20% in the top of level 5.
 For both occupations, chert non-congruent with the known local sources represent percentages lower than 30%.
-This pattern seems to shift in the bottom of 5 and throughout the Gravettian occupation; the use of chert local cherts drops to 45-60%, while the percentages of specific non-local chert types increase significantly, such as Type 11 and Type 13.
-However, when considering the sum of the weight, the percentages of Type 11 diminish, instead representing around 20% of the chert use, and bringing for level 6 the use of local chert to 50% in comparison to chert types of non-local source.
-Other non-local cherts such as the Oolitic cherts and TL categories also show smaller increases in the Gravettian occupation.
-Chalcedony seems to be present but always in small frequencies.
+This pattern seems to shift in the bottom of 5 and throughout the Gravettian occupation; the use of chert local cherts drops to 45-60%, while the percentages of specific non-local chert types increase significantly, such as Type 6 and Type 7.
+However, when considering the sum of the weight, the percentages of Type 6 diminish, instead representing around 20% of the chert use, and bringing for level 6 the use of local chert to 50% in comparison to chert types of non-local source.
+Other non-local cherts such as Type 10 and TL categories also show smaller increases in the Gravettian occupation.
+Chalcedony (Type 1) seems to be present but always in small frequencies.
 Despite this, there is a small increase in the use of chalcedony in level 5 (both top and bottom), from lower than 5% in level 4E to 13.5% in top of 5, with a tendency to decrease in the older levels.
@@
 #| echo: false
+#| fig-width: 10
 
@@
   filter(RMTYPE != "INDET") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   dplyr::group_by(Level, RMTYPE) %>% 
@@
   filter(RMTYPE != "INDET" & WEIGHT != "0") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   dplyr::group_by(Level, RMTYPE) %>% 
@@
   filter(!Z >= "25") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   filter(RMTYPE != "INDET") %>% 
@@
   ungroup() %>% 
-  arrange(factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F", "Oolitic")))
+  arrange(factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10")))
 
@@
 #| fig-cap: Boxplot of weights.
+#| #| fig-width: 10
 
@@
   filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   ggplot(aes(RMTYPE, WEIGHT, color = RMTYPE)) +
@@
 When plotting the weight, it is apparent that most of the samples are located under the 0.3 kg mark, with median and bottom quartiles often falling under the 0.5 kg line.
-Local types, such as Type 2, Type 7 and Type 12, although mostly represented by artifacts with a small mass (\~0.3 kg), show the heaviest and highest number of heavy artifacts.
+Local types, such as Type 2, Type 4 and Type 5, although mostly represented by artifacts with a small mass (\~0.3 kg), show the heaviest and highest number of heavy artifacts.
 In comparison, other types, although showing some variation in the weight of artifacts, show lighter numbers of artifacts, and their means and medians are frequently closer to the bottom quartile.
@@
 The same raw materials found in the Terrace area are also found in the Solutrean occupation of the Shelter, albeit with some differences.
-Although with similar percentages to the local cherts, the use of local chert in the Shelter area is heavily dominated by the brown/yellow cherts that frequently characterize the Lower Jurassic outcrops of the Algarve; although the other macroscopic types are present but not significant.
-The presence of cherts not congruent with the local sources is, alike level 4, close to 25%, although dominated mostly by Type 11.
-Type GB is present in higher numbers than in level 4, however still representing less than 5% of the assemblage.
+Although with similar percentages to the local cherts, the use of local chert in the Shelter area is heavily dominated by the brown/yellow cherts (Type 2) that frequently characterize the Lower Jurassic outcrops of the Algarve; although other macroscopic types are present, they are not significant.
+The presence of cherts not congruent with the local sources is, alike level 4, close to 25%, although dominated mostly by Type 6.
+Type 11 is present in higher numbers than in level 4, however still representing less than 5% of the assemblage.
 Similarly to the Solutrean occupations of the Terrace, although the artifacts are generally small (lower weights), the local types show a higher variability in mass, with several outliers and heavy samples when compared to the non-local types.
@@
 #| echo: false
+#| #| fig-width: 10
 
@@
 shelter_count <- s_data_plots %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","TYPE GB", "Oolitic", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   dplyr::group_by(Layer, RMTYPE) %>% 
@@
 shelter_weight <- s_data_plots %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","TYPE GB", "Oolitic", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   dplyr::group_by(Layer, RMTYPE) %>% 
@@
 
-ggarrange(shelter_count, shelter_weight, labels = c("A", "B"), ncol=2, common.legend = TRUE, legend="bottom")
-
-```
-
-```{r}
-#| label: weight-shelter
-#| fig-cap: Boxplot of weights.
-
-s_data_plots %>% 
+shelter_box <- s_data_plots %>% 
   filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","TL") & Layer != "ND") %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","TYPE GB", "Oolitic"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   ggplot(aes(RMTYPE, WEIGHT, color = RMTYPE)) +
@@
         axis.ticks.x=element_blank())
+
+ggarrange(shelter_count, shelter_weight, shelter_box, labels = c("A", "B", "C"), ncol=2, nrow = 2, common.legend = TRUE, legend="bottom")
 
@@
 The macroscopic and petrographic analysis of chert and chalcedony lithic materials from the UP levels of the Terrace and Shelter areas of Vale Boi have provided crucial insights into the raw material procurement and mobility patterns of hunter-gatherers.
-These raw materials are useful proxies to identify mobility patterns since they show patterns of local and non-local procurement.
+These raw materials are useful proxies to identify mobility changes since they show patterns of local and non-local procurement.
 The prevalence of local raw materials in the Solutrean and Proto-Solutrean assemblages show a preferential use of local resources, although with discrete differences in the used nodules/outcrops.
-The presence of non-local raw materials throughout the stratigraphy suggests a continuous pattern of long-distance sourcing, which was intense during the Gravettian and slowed down with time.
+The presence of non-local raw materials throughout the stratigraphy suggests a continuous pattern of long-distance sourcing, which was intense during the Gravettian and slowed down through time.
 
 Despite differences in proportion, there seems to be continuity in the raw materials used throughout the UP, including the allochthonous sources.
-This may be explained by the tradition of shared knowledge and culture regarding the location of raw materials and other resources in the landscapes, or the maintenance of social networks through time.
-In this sense, raw materials and the connection they suggest seem to present themselves as resilient components of UP occupations and organization, and less prone to cultural or technological change.
+This may be explained by the tradition of shared knowledge and culture regarding the location of raw materials and other resources in the landscapes (Is there a good ref for this?), or the maintenance of social networks through time.
+In this sense, raw materials are resilient elements of UP occupations and social/technological organization, and less prone to cultural or technological change.
 This persistence however, may also be explained by cached raw materials that were accumulated at the site and were intensively used and reused throughout several re-occupations of the site, eventually disappearing and leading to a more intense use of the local sources, either through large nodules or large blanks that remained unused [@surovell_toward_2009; @kuhn_mousterian_1995; @hofmanSHIFTINGSANDSFOLSOMMIDLAND2024; @binford_organization_1979].
@@
 The other chert sources (Lower Jurassic and Upper Jurassic) of western Algarve are located by the beach and cliff areas as previously mentioned, between 18-15 km of Vale Boi.
-Although our current data does not allow us to identify which specific outcrops were visited based on macroscopic and petrographic observations, the presence of marine resources, such as shells, at the site shows these groups visited the coast, which means they could easily obtain raw materials in the vicinity without necessary detours [@binford_organization_1979; @surovell_toward_2009].
+Although our current data does not allow us to identify which specific outcrops were visited based on macroscopic and petrographic observations, the presence of marine resources such as shells at the site shows these groups visited the coast, which means they could easily obtain raw materials in the vicinity without necessary detours [@binford_organization_1979; @surovell_toward_2009].
 We may see these patterns of close-distance procurement and possibly caching in the presence of heavy (large mass) nodules and chunks of chert from local types, as well as the considerable amount of outliers (representing larger blocks and chunks) in the weight boxplots throughout the stratigraphy (@fig-weight).
 
+MAP/Figure with several elements: map with the several sources cited in the text (local and not local) and cited geological formations, marks exemplifying whether the sources were use or not as for example the eastern outcrops of the Algarve, color codes for type of mobility (close-distance, long-distance), side images for the geological cherts and the archaeological cherts, possibly also using the Quíros Castillo map as a secondary map to exemplify the use of plateaus in the mobility.
+What do you think?
+
 However, the discussion involving local cherts seems further complicated by the existence of similar outcrops in eastern Algarve.
-Previous studies showed that macroscopic criteria is not enough to discern whether the used sources are Lower Jurassic by the site or Middle Jurassic since they are similar macroscopically and petrographically, and with similar quality [@belmiro2023].
-In this sense, and similar to other archaeological studies [@surovell_toward_2009], we cannot discredit that some portion of the local cherts we interpret as being procured within a 20 km radius of Vale Boi, may be the result of long-distance procurement or the result of regional mobility, between \~80-100 km where the Middle Jurassic cherts are located.
-In fact, previous studies have shown that the outcrops in this region were used, due to the presence of UP archaeological sites by the outcrops \[++\].
-
+Previous studies showed that macroscopic criteria is not enough to discern whether the used sources are Lower Jurassic (west Algarve within a \~20 km radius of the site) or Middle Jurassic (east Algarve, over 80 km away) since they are similar macroscopically and petrographically, and with similar quality [@belmiro2023].
+In this sense, and following other archaeological studies [@surovell_toward_2009], we cannot discredit that some portion of the local cherts we interpret as being procured within a 20 km radius of Vale Boi, may be the result of long-distance procurement or the result of regional mobility, between \~80-100 km where the Middle Jurassic cherts are located.
 Given the similar quality of the raw materials and lack of geochemical studies, and following raw material formal and informal models for raw material procurement, we interpret that most of these raw materials are probably from the local (under 20 km sources) while a small quantity may be the result of long distance displacements.
-Similarly, the Upper Jurassic cherts in proximity to the latter sources, around 90 km, despite having macroscopic characteristics not congruent with the local sources, have not been identified in the assemblages.
+Similarly, the Upper Jurassic cherts in proximity to the latter sources, around 90 km from Vale Boi, have not been identified in the assemblages despite having macroscopic characteristics not congruent with the local sources.
 This may be explained by the medium quality and generally small size of the nodules (frequently between 3-5 cm), or the location of the outcrops, since the known outcrops are located inland, and it may be possible that they were not accessible or visible.
 
-The non-local cherts seem to show two modalities: knapped at the site, and present in larger quantities, and individual samples of variable chert and chalcedony types, but present in a limited number of artifacts.
-The allochthonous cherts with possibly identified sources seem to come from two directions: 1) north, where the Cretaceous cherts from Central Portugal were obtained, along with the small quantity of jasper; 2) west, where the peloidal/oolitic chert from south Spain can be found.
+While the regional Algarve cherts were not identified in the assemblages, the allochthonous cherts with possibly identified sources seem to come from further distances, and from two directions: 1) north, where the Cretaceous cherts from Central Portugal are located, along with the small quantity of jasper (TL01); 2) west, where the peloidal/oolitic chert from south Spain can be found.
 Despite both these sources being \~300 km away, these cherts may have been obtained at closer distances, through riverbeds with smaller transport distances, without pebble-like cortex.
@@
 Despite these high distance travel paths, studies have suggested the existence of low-cost paths to travel through the Iberian Peninsula.
-In fact, previous formal models designed to explain hunter-gatherer mobility have suggested that variables such as how topography influences mobility may be important for the distances traveled by groups or individuals [@brantinghamMeasuringForagerMobility2006].
-For example, Castillo [-@Quirós Castillo, 2020] describes how Copper age groups reduced subsistence risks by creating exchange networks and practicing regional mobility seen in the exchange of autochthonous raw materials.
+In fact, previous formal models designed to explain hunter-gatherer mobility have suggested that variables such as topography are important for the distances traveled by groups or individuals [@brantinghamMeasuringForagerMobility2006].
+For example, Quíros Castillo [-@quiroscastilloArchaeologyHistoryPeasantries2020a] describes how Copper age groups reduced subsistence risks by creating exchange networks and practicing regional mobility seen in the exchange of autochthonous raw materials.
 These exchanges would have come with excessive energy costs and risks.
-However, the author points out that there are important features about the Iberian Peninsula regarding mobility such as the low cost of moving through the plateaus and the connectivity of the Guadalquivir River valley with the Guadiana and Tagus rivers (Quirós Castillo, 2020).
-In this sense, albeit distant, the existence of the north/east directions of exchange at Vale Boi may simply reflect the use of these natural highways for hunter-gatherer mobility and explain the high percentages of allochthonous chert use.
+However, the author points out that there are important features about the Iberian Peninsula regarding mobility such as the low cost of moving through the plateaus and the connectivity of the Guadalquivir River valley with the Guadiana and Tagus rivers.
+In this sense, albeit distant, the existence of the north/east directions of long-distance mobility by hunter-gatherers at Vale Boi may simply reflect the use of these natural highways for hunter-gatherer mobility and explain the high percentages of allochthonous chert use.
 The identification of the smaller subset of unknown source allochthonous cherts and chalcedonies (TL) may reflect different types of long-distance exchanges of finished products and blanks or with different social networks, possibly even reflecting individual networks and mobility [sensu @gamblePalaeolithicSocietiesEurope1999].
 
-A higher percentage of these allochthonous cherts, especially as we see in the Gravettian occupation, may be interpreted then as a high degree of communication and mobility between the groups occupying Vale Boi and an extended social network elsewhere rather than secondary catchment.
+A higher percentage of these allochthonous cherts, especially as we see in the Gravettian occupation, may be interpreted then as a high degree of communication and mobility between the groups occupying Vale Boi and an extended social network.
 This is a significantly different interpretation from that obtained from the Gravettian technology from south Portugal.
@@
 These have been interpreted as a different exploitation of territories as well as the reflection of different ecological and ethnographic boundaries between hunter-gatherers [@marreiros2015].
-Based on these combined data, it may be possible that the technology data indeed reflects different coexistence boundaries, as the regional band is often no more than ++ km \[++\], as such not being expected to represent the full range of hunter-gatherer mobility \[++\].
-The existence of a high frequency of allochtonous raw materials may be the reflection of these contacts and exchanges between different groups, focusing on good quality raw materials in detriment of those found locally, while still maintaining boundaries between groups, expressed at the technological level.
-A similar scenario is expressed by Brantingham [@brantinghamMeasuringForagerMobility2006] when discussing social exchange; that exchange may occur between two bands occupying non-overlapping territories but with established regular meetings between, when they would exchange their raw materials, thus explaining how very long-distance sources might be identified in archaeological sites.
+Based on these combined data, it may be possible that the technological aspect of lithics indeed reflects different coexistence boundaries limited to the regional band.
+The existence of a high frequency of allochtonous raw materials may be the reflection of these contacts and exchanges between different bands, focusing on good quality raw materials in detriment of those found locally.
+A similar scenario is expressed by Brantingham [-@brantinghamMeasuringForagerMobility2006] when discussing social exchange; that exchange may occur between two bands occupying non-overlapping territories but with established regular meetings between, when they would exchange their raw materials, thus explaining how very long-distance sources might be identified in archaeological sites.
 
 A different interpretation for the high frequencies of non-local cherts during the Gravettian may be used following the "Mean per Capita Occupation Span" model by Surovell [-@surovell_toward_2009].
-According to this model, short occupations should display high proportions of transported artifacts (in this case, understood by allochthonous raw materials and artifacts produced with these raw materials), and as the duration of the occupation increases, local artifacts should become dominant.
+According to this model, short occupations should display high proportions of transported artifacts (in this case, understood by allochthonous raw materials and artifacts produced with these resources), and as the duration of the occupation increases, local artifacts should become dominant.
 Similarly, the author also suggests long-term occupations may be a product of the environment, as patchy environments and colder temperatures present risks to increased mobility, and the settlement of camp in a locally abundant environment in a scarce regional setting could lead to long-term occupations which would present a less risky option [@surovell_toward_2009][Kelly 1995].
 In this theoretical paradigm, the Gravettian occupations could be the result of high mobility and short-term occupations of Vale Boi, with possible reoccupations [sensu @surovell_toward_2009], thus showing high amounts of used non-local raw materials.
-As the climatic conditions begin to change, with the onset of Heinrich Event 2 during the Proto-Solutrean and the consequent glacial maximum that characterized the Solutrean [@sanchez_goni_millennial-scale_2010, @schmidt_rapid_2012], which caused a dryer, harsher climate with patchy vegetation \[++\], the resource abundant characteristics of an ecological niche region such as southwestern Portugal \[++\], could have decreased the mobility of hunter-gatherers.
+As the climatic conditions begin to change, with the onset of Heinrich Event 2 during the Proto-Solutrean in south Portugal [@belmiro_gravettian-solutrean_2021; @cascalheiraHunterGathererEcodynamics2013] and the consequent glacial maximum that characterized the Solutrean [@sanchez_goni_millennial-scale_2010; @schmidt_rapid_2012], which caused a dryer, harsher climate with patchy vegetation [@cascalheiraHunterGathererEcodynamics2013; @gonzalez-samperiz_steppes_2010-1], the resource abundant characteristics of an ecological niche region such as southwestern Portugal [@cascalheira_2017; @schmidt_rapid_2012], could have decreased the mobility of hunter-gatherers.
 With longer-term occupations to reduce risk, the amount of non-local raw materials from previous expeditions (obtained both through catchment at the source or through trade), would eventually reduce as local raw materials became more dominant.
 
-In general, our results build upon previous preliminary studies, where the majority of the chert used seemed to be the local Lower Jurassic cherts within a 20 km range of the archaeological site, while non-local materials composed a small percentage of used chert [@pereira_territory_2016-1].
-Previously suggested long-distance sourcing, such as the Cretaceous cherts from central Portugal are confirmed through visual and petrographic data [@pereira_territory_2016-1, ++], as well as the north and east mobility paths that connected the hunter-gatherer groups from the southwest of Portugal to the center of Portugal and the south of Spain [@pereira_territory_2016-1].
+In general, our results build upon previous preliminary studies, where the majority of the chert used seems to be the local Lower Jurassic cherts within a 20 km range of the archaeological site, while non-local materials composed a small percentage of used chert [@pereira_territory_2016-1].
+Previously suggested long-distance sourcing, such as the Cretaceous cherts from central Portugal are confirmed through visual and petrographic data [@pereira_territory_2016-1; ++], as well as the north and east mobility paths that connected the hunter-gatherer groups from the southwest of Portugal to the center of Portugal and the south of Spain [@pereira_territory_2016-1].
 These cherts and chalcedonies with different sources present a picture of hunter-gatherers with complex systems and modes of exploring the landscape, managing resources, and possibly managing risk.
-Allochthonous raw materials, even if a small percentage in the Proto-Solutrean and Solutrean occupations, the Vale Boi case study shows that hunter-gatherers occupying a marginal territory in the Iberian Peninsula and Europe, bordered south and west by sea, did not live isolated and instead demonstrate an extensive social network.
-Alike the results identified in central Portugal (++), we also identify a pattern of interconnection between the Upper Paleolithic hunter-gatherers of the Iberian Peninsula.
+Through the identification of allochthonous raw materials, even if in a small percentage in the Proto-Solutrean and Solutrean occupations, the Vale Boi case study shows that hunter-gatherers occupying a peripheric territory in the Iberian Peninsula and Europe, bordered south and west by sea, did not live isolated and instead demonstrate an extensive social network.
+Alike the results identified in central Portugal (++), in south Portugal we also identify a pattern of interconnection between the Upper Paleolithic hunter-gatherers of the Iberian Peninsula.
 
-This study showed the potential of macroscopic and petrographic analysis to identify different chert and chalcedony types and identify their possible sources.
-However, the macroscopic and petrographic observations have frequently been associated with geochemical analyses, through methods such as XRF or LA-ICP-MS to further strengthen the attribution of source.
-For future research, we aim to use of a geochemical approach in the studied collection and geological reference collection to further expand upon the discussion of mobility through the UP of Vale Boi.
+The obtained results show potential for a continuous study of these assemblages using different methods and approaches.
+For future research, we aim to use of a geochemical approach in the studied assemblages and geological reference collection to further expand upon the discussion of mobility through the UP of Vale Boi.
 Testing with fire and the creation of an altered reference collection may also aid in strengthening the raw material types and group of currently unidentifiable samples as mentioned in previous studies [@pereira_territory_2016-1].
-Finally, for future research, we aim to correlate technological data from existing and in-progress datasets to further understand how raw materials (local and non-local) were being used and managed through time and include these in the broader spectrum of the changing environmental conditions occurring throughout the Upper Paleolithic.
+Finally, we aim to correlate technological data from existing and in-progress datasets to further understand how raw materials (local and non-local) were being used and managed through time and include these in the broader spectrum of the changing environmental conditions occurring throughout the Upper Paleolithic.
 

### 2024-01-08 analysis/supplementary-materials/SOM_lithotypes.qmd
First draft

@@
 
-### TL01
+**TL01 - Jasper.** This chert is a jasper, with a massive microstructure, mudstone texture and red coloration, with frequent fractures filled with colorless, translucent fibrous chalcedony.
+The chert is opaque while the chalcedony is translucent.
+There are no visible allochems and the samples show no cortex.
+
+**TL02 - Orange chalcedony.** Fibrous chalcedony, without visible allochems. Good quality, differing from the group of white/translucent chalcedony due to its color. Translucent and without cortex.
+
+**TL03 - Massive chert with uncommon spots.** This chert is a massive mudstone, with uncommon allochems limited to uncommon dark spots which may be iron oxides. It is opaque and without cortex.
+
+**TL04 - Massive chert with inclusions and fractures.** Yellow, massive wakestone chert, with common allochems which are probably large, unidentifiable fossils and uncommon iron oxides.
+It has frequent fractures filled with fibrous chalcedony, is sub-translucent and with no cortex.
+
+**TL05 - Banded/massive chert.** This chert shows two samples, with massive and laminated microstructure variations.
+The laminations seem to be of ++ origin.
+They are grey mudstones, with uncommon allochems which are opaques creating speckling patterns.
+It is sub-translucent and without cortex.
+
+**TL06 - Sorted oolitic chert.** Massive, oolitic packstone.
+Allochems are ooids which are distributed homogeneously, highly abundant and concentrated.
+The ooids are well-sorted, normal, small (\~200-220 μm) and round, with differentiated walls and nucleus.
+The nucleus corresponds to half the width of the ooid [@flugelMicrofaciesCarbonateRocks2010].
+The chert is sub-translucent and without cortex.
+It is distinguished from the other oolitic cherts by the small, well-sorted ooids.
+
+**TL07 - Laminated (dendritic) chert.** Massive/banded chert with uncommon allochems, which are large white splotches, possibly unidentifiable fossils.
+The color variations and banding may be due to iron oxide concentrations.
+It is sub-translucent and without cortex.
+
+**TL08 - Colorless, mudstone chert.** Massive mudstone without bioclasts. It is translucent and without cortex.
+
+**TL09 - Chalcedony with mossy inclusions.** Fibrous chalcedony, with massive microstructure and packstone texture.
+Allochems include unidentified orange and red "moss"-like inclusions.
+It is translucent and without cortex.
+
+**TL10 - Massive chert with oxides.** Massive mudstone, with common allochems which are orange iron oxides in speckling or laminated patterns.
+It is sub-translucent and with cortex.
+Shows similarities to a nodule from the Rio Maior area (central Portugal) from the LusoLit lithotheque.
+
+**TL11 - Banded, translucent chert.** Banded mudstone without inclusions. It is translucent, and shows a cortex or alteration rind. Shows similarities to nodules from the Rio Maior area.
+
+**TL12 - Massive, oolitic packstone chert.** This chert is completely white and translucent, with ooids which are distributed homogeneously, highly abundant, but poorly preserved and poorly sorted.
+The ooids are not visible under the naked eye or hand lens, and can only be identified under the stereomicroscope.
+No cortex.
+
+**TL13 - Banded, translucent chert.** Banded mudstone without inclusions.
+It is translucent, without cortex but with parent rock attached.
+The top laminations may be due to the concentration of iron oxides.
+
+**TL14 - Massive chert with benthic foraminifera.** Massive mudstone, with uncommon allochems/bioclasts which are larger, benthic foraminifera (1-2 mm).
+It has two fabrics, both with foraminifera: fine and sub-translucent; coarse and opaque.
+No cortex.
+
+**TL15 - Massive brown chert with inclusions.** Massive wackestone with common, unidentifiable round inclusions, speckling and iron oxides.
+It is opaque, glossy but with dull splotches, and without cortex.
+
+**TL16 - Radiolarite with inclusions.** Massive chert with uncommon allochems which are unpreserved, unidentifiable fossils. Thin, rounded cortex.
 

### 2024-04-26 analysis/paper/paper.qmd
Pre submission commit_1

--- 2023-12-27 analysis/paper/paper.qmd
@@
 ---
-title: "Title Goes Here"
+title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
 author:
-  - Jane Doe:
+  - Joana Belmiro:
+      orcid: https://orcid.org/0000-0003-4951-2907
       correspondence: "yes"
-      email: janedoe@fosg.org
-      orcid: 0000-0003-1689-0557
+      email: jfbelmiro@ualg.pt
       institute:
-        - fosg
-        - fop
-  - John Q. Doe:
-      institute: fosg
-      orcid: 0000-0003-1689-0558
-  - Peder Ås:
-      institute: fosg
-      orcid: 0000-0003-1689-0559
-  - Juan Pérez:
-      orcid: 0000-0003-1689-0551
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
+  - Xavier Terradas:
+      orcid: https://orcid.org/0000-0002-8000-5607
       institute: 
-        - name: Acme Corporation
-  - Max Mustermann:
-      orcid: 0000-0003-1689-0552
+        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
+  - Salvador Dominguez-Bella:
+      orcid: https://orcid.org/0000-0003-3892-763X
       institute: 
-  - fosg:
-      name: Formatting Open Science Group
-      address: 23 Science Street, Eureka, Mississippi, USA
-  - fop: Federation of Planets
+        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
+  - João Cascalheira:
+      orcid: https://orcid.org/0000-0003-0321-8892
+      institute:
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
 title-block-published: "Last updated" 
-date: now
-date-format: long
 format: 
@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/springer-socpsych-author-date.csl" # Insert path for the bib-style
 abstract: |
-  Text of abstract
+  Analyses of raw materials and the distinction between local/regional and long-distance sources have proven invaluable for understanding the extensive movements, interactions, and social networks during the Upper Palaeolithic (UP) in the Iberian Peninsula.
+However, unlike other parts of Iberia, research on the management and acquisition of raw materials in the south and west of the Peninsula remains relatively underdeveloped.
+Despite significant knowledge about the technological practices of Palaeolithic hunter-gatherers from southern Portugal, particularly from studies conducted at the Vale Boi site, there's a noticeable lack of focus on raw materials.
+This paper presents the first comprehensive characterization of chert raw materials from the Gravettian, Proto-Solutrean, and Solutrean sequences at Vale Boi, using both macroscopic and petrographic techniques.
+Our study reveals that the majority of chert found at Vale Boi originates locally, within a 20 km radius, while a smaller portion comes from non-local sources, potentially indicating long-distance procurement from central Portugal.
+Additionally, the identification of cherts resembling those from southern Spain suggests a significant west-east mobility pattern throughout the UP period.
 keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+  Upper Paleolithic; Iberian Peninsula; Lithic raw materials; Petrography
 ---
@@
 
-```{r packages, echo=FALSE}
+```{r packages}
+#| label: callpackages
+
 # Call packages
@@
 library(janitor)
+library(ggpubr)
+library(knitr)
+library(gtsummary)
 
@@
 
-# Introduction
-
-Here is a citation [@Marwick2017]
-
-# Background
-
-# Methods
-
-The samples chosen for this study consist of the chert and chalcedony lithic artifacts with individual IDs (excluding buckets) from the Upper Paleolithic levels of the Terrace and Shelter areas of Vale Boi.
-The sample from the Terrace area (VB-T) consists of 3653 from levels 8 to 4, while the sample from the Shelter area (VBS) consists of 1652 from levels C to A.
-These two areas were chosen since both have detailed spatial information (as mentioned in section *IN*), several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal and geoarchaeological data.
-
-## Geo-archaeological surveys
-
-The geological data and comparative collection from local and regional cherts used in this study were obtained through previous prospections and petrographic works (\*\*), located at the Lusolit lithotheque (hosted at ICArEHB, University of Algarve).
-These samples also included cherts collected from previous prospections from other areas of Portugal (such as central Portugal).
-Hand samples were used throughout the macroscopic characterization for comparison with the archaeological materials, and the archaeological thin sections were compared with the previously studied geological chert thin sections.
-The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz was also visited, to understand the macroscopic variability from the Cadiz local and regional cherts, to discern between local and regional chert types and possible long-distance cherts.
-
-## Macroscopic characterization
-
-We applied a macroscopic analysis to both Terrace and Shelter samples, which was divided into two steps.
-The first step included a preliminary characterization using a hand lens of 10x magnification.
-This step allowed us to create groups and subgroups based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these include characteristics such as color, translucency, and feel; a data dictionary for the recorded variables can be found in the Supplementary Materials (\*\*).
-
-In the second step, we used a stereomicroscope () to observe each sample in further detail, which allowed us to better characterize the artifacts (especially regarding inclusions and alterations), and the previously established groups.
-The data dictionary used for the individual samples can also be found in Supplementary Materials (\*\*).
-Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis is simplified and focused on collecting data related to weight, cortex, and alterations.
-
-## Petrographic characterization
-
-To better characterize the different chert groups identified in the macroscopic analysis, we analyzed ++ thin sections.
-The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels.
-For groups which showed macroscopic variability, several thin sections were analysed to better characterize these cherts.
-@tbl-TS-table Petrographic thin sections were analyzed by polarized light microscopy () to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations.
-A total of 20 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory.
-The archaeological samples chosen for the petrographic study are from the Terrace area, from several levels, and focused on characterizing the macroscopic variability, reflected on the several groups and subgroups (table \*\*).
-
-```{r}
-#| label: tbl-TS-table
-#| tbl-cap: Thin section list.
-#| echo: false
-
-TS_short_list <- read_csv(here::here('analysis/data/raw_data/TS_short_list.csv'))
-
-knitr::kable(TS_short_list)
-
-```
-
-# Results
-
-```{r}
+```{r dbsetup}
 #| label: db-setup
-#| echo: false
 
@@
 T_RM_data <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_TERRACE.csv')) # Terrace RM analysis
-context <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_context.csv')) # Terrace Field DB - lithic artifact info
+context <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_context.csv')) # Terrace Field DB - lithic artefact info
 xyz <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_xyz.csv')) # Terrace Field DB - just coordinates
@@
 context <- context %>%
-  # dplyr::rename("Unit" = "j") %>% # Renaming first column
+  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column
   unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
@@
 xyz <- xyz %>%
-  #dplyr::rename(Unit = UNIT) %>% # Rename to homogenize
   unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
-  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artifact
+  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
   select(ID, X, Y, Z) %>% # Selecting important variables
@@
 T_dataset <- T_dataset %>% 
-  #filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
-  filter(Level !="4B" & Level !="7C" & Level!="8") # Filtering out levels with barely any artifacts
-
-cols_to_concat_spit <- c("Level", "Spit")
-RM_by_spit <- T_dataset %>%
-  filter(!Spit %in% c("limp_corte", "profile", " ", NA)) %>%
-  select(ID, Z, Level, Spit, RMTYPE) %>%
-  unite(col='depth', cols_to_concat_spit, sep=".", remove=FALSE) %>% 
-  dplyr::group_by(depth) %>%
-  dplyr::count(RMTYPE) %>%
-  mutate(freq = n / sum(n)*100)
+  filter(!Level %in% c("2", "3")) %>%
+  drop_na(Level) %>% 
+  mutate(Level = case_when(Level == "7C" ~ "7",
+                           Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7, since 8 disappears in field_data
+                           Level == "4B" ~ "4", # collapsing samples from 4B to 4 due to small number
+                           TRUE ~ Level)) %>% 
+  mutate(RMTYPE = case_when(RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
+                            RMTYPE == "TYPE 15" ~ "TYPE 3",
+                            RMTYPE == "TYPE 7" ~ "TYPE 4",
+                            RMTYPE == "TYPE 12" ~ "TYPE 5",
+                            RMTYPE == "TYPE 11" ~ "TYPE 6",
+                            RMTYPE == "TYPE 13" ~ "TYPE 7",
+                            RMTYPE == "TYPE 11F" ~ "TYPE 8",
+                            RMTYPE == "TYPE 13B" ~ "TYPE 9",
+                            RMTYPE == "Oolitic" ~ "TYPE 10",
+                            RMTYPE == "GB" ~ "TYPE 11",
+                            TRUE ~ RMTYPE))
 
@@
     Level == 5 & Z < 24.1 ~ "5",
-    TRUE ~ Level))
+    TRUE ~ Level)) %>% 
+  mutate(Technocomplex = case_when(
+    Level == 4 | Level == "4C" | Level == "4D" ~ "Solutrean",
+    Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
+    Level == 6 | Level == 7 ~ "Gravettian"))
 
@@
 S_RM_data <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
-S_field_data <- read_csv(here::here('analysis/data/raw_data/VB_Shelter_xyz.csv')) # Shelter coordinates
+S_field_data <- read_csv(here::here('analysis/data/raw_data/VB_Shelter_xyz_codes.csv')) # Shelter coordinates
+S_layerdata <- read_csv(here::here('analysis/data/raw_data/shelter_layers_JC.csv')) # Shelter layers
 
@@
 
+S_layerdata <- S_layerdata %>% 
+  mutate_at("Date", str_replace, "20", "") %>%
+  dplyr::rename("YEAR" = "Date") %>% 
+  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% 
+  select(ID, Layer)
+
 S_field_data <- S_field_data %>%
+  mutate_at("YEAR", str_replace, "20", "") %>%
   unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% # Unite YEAR with Id s«using cols_to_concat
-  select(ID, X, Y, Z) %>% # Selecting important variables
+  select(ID, X, Y, Z, CODE) %>% # Selecting important variables
   distinct(ID, .keep_all = TRUE)
-S_dataset <- left_join(S_RM_data, S_field_data, by = "ID")
+
+shelter_field_data <- full_join(S_field_data, S_layerdata, by = "ID")
+S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID", relationship = "many-to-many")
 
 ```
+
+# Introduction
+
+Knappable raw materials play a crucial role in understanding the mobility and lifeways of past hunter-gatherers, given the ubiquitous presence of lithic artefacts throughout prehistory.
+More than mere rocks with which stone tools were produced, the management of these resources is intimately connected to a group’s technological, social, and cultural organization, potentially influencing the group’s overall survivability [@oestmoFormalModelingApproach2017; @binfordOrganizationFormationProcesses1979; @bleedOptimalDesignHunting1986; @bousmanHuntergathererAdaptationsEconomic1993; @torrenceTimeBudgetingHuntergatherer1983; @gouldLithicProcurementCentral1985].
+Several key topics in the study of hunter-gatherer behaviour and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccall_behavioral_2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallon_social_2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999].
+This is often achieved by systematically characterizing geological and archaeological raw materials and establishing correlations between samples from both origins.
+The ultimate aim is to pinpoint corresponding sources, enabling a precise differentiation between local, regional, and long-distance raw materials, as well as to examine their distribution within a site over different occupations and periods.
+These approaches and concepts have also been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula [@aubry_upper_2016; @aubryEconomyLithicRaw2009; @garcia-rojasGreatStepForward2021; @gomezdesolerPanadellaChertMontmaneu2020; @herrero-alonsoManagementLithicRaw2020; @matias_raw_2016; @ortegaMobilitatDesplacamentsDels2003; @pereiraLithicEconomyTerritory2016; @pereira_raw_2021; @pereira_flint_2022; @ramacciottiMovingLandFirst2022; @sanchezdelatorreNewDataChert2023; @soto_procurement_2016], and contributed to understanding how different groups organized their technology, managed time and risk, and explored the available resources during different chronological periods.
+
+Specifically, analyses of raw materials and the distinction between local/regional and long-distance sources have been instrumental in tracing extensive movements, contacts, and social networks throughout the Upper Palaeolithic (UP).
+Noteworthy examples include studies in northwestern Iberia (Galicia) during the Solutrean period [@hermidaSilexNWPeninsula2016], northern Iberia (Asturias) during the Magdalenian occupations at Las Caldas [@corchonrodriguezMobiliteTerritoiresRelations2016], northeastern Iberia (Catalunya) during the Aurignacian and Gravettian periods at Arbreda cave [@marreirosExploringLithicVariability2016; @ortegaMobilitatDesplacamentsDels2003], northern Portugal during the Gravettian/Solutrean periods in the Côa Valley [@aubry_upper_2016; @aubryRawMaterialProcurement2004], and inland central Iberia (Guadalajara) during the Proto-Solutrean and Solutrean periods at Peña Capón [@sanchezdelatorreNewDataChert2023].
+Although such contacts are commonly recognized and identified in later prehistoric periods [@costa_lithic_2022; @noceteRelacionesCentroPeriferia2005; @rodriguezPrehistoricFlintExploitations2011], the recognition of these connections has broadened our comprehension of the often extensive social networks and contacts during the UP in Iberia, illustrating a vast interconnected territory.
+
+However, in contrast to other areas of the Iberia, and despite the potential of raw material studies to shed light on the lifeways and organizational structures of Palaeolithic hunter-gatherers, research on the management and procurement of raw materials during the UP in southern and western Iberia remains in its initial stages.
+Specifically, in southern Portugal, even though there is considerable knowledge about the technological organization of Palaeolithic hunter-gatherer groups—largely due to research conducted at the Vale Boi archaeological site [see @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @marreirosLithicTechnologyGravettian2015]—studies focusing on raw materials are notably infrequent.
+This situation is attributed to the absence of a detailed and comprehensive characterization of regional lithic resources in the Algarve region (southern Portugal) and the subsequent creation of a reference collection, which is a critical component for conducting raw material studies.
+Such efforts are particularly crucial for distinguishing between local and non-local raw materials and for understanding variations within the archaeological record [@pop_simulating_2015; @surovellBehavioralEcologyLithic2009].
+Recently, we have made significant strides by introducing a reference collection and establishing a framework for raw material studies in the region, through a detailed macroscopic and petrographic database of chert resources in the Algarve region [@belmiroCreatingFramesReference2023].
+This initiative paves the way for systematic investigations into raw material procurement strategies during the Upper Palaeolithic, especially concerning the Vale Boi site.
+
+Vale Boi holds a pivotal role in the study of UP adaptations on the Iberian Peninsula, providing invaluable insights into the subsistence and technological strategies of hunter-gatherer populations [@belmiro_gravettian-solutrean_2021; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraTerritorialityOrganizationTechnology2019; @horta_role_2019; @manneIntensiveSubsistencePractices2012; @marreirosExploringLithicVariability2016; @pereira_territory_2016-1].
+The rich archaeological record from the site, indicative of prolonged human occupation, is essential for understanding raw material management practices in UP contexts.
+
+Preliminary studies at Vale Boi, employing macroscopic analysis, identified a predominant use of local cherts from Jurassic limestone formations, complemented by a lesser extent of non-local cherts throughout various UP occupations [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004].
+According to Pereira et al.
+[-@pereira_territory_2016-1], the local cherts likely originated within a 16 km radius, reflecting embedded procurement strategies.
+In contrast, the presence of non-local cherts points to extended sourcing networks, possibly involving travel to other regions of the Algarve, Alentejo, or southern Spain.
+The presence of chert from the Cretaceous formations of central Portugal has also been suggested [@bicho_ecodynamics_2013].
+
+These results highlight the potential for systematic raw material studies at the site, employing various methods and approaches to analyze the lithic assemblages from multiple UP occupations.
+Such research aims to comprehend the behaviours of these hunter-gatherer groups over time: how they navigated and utilized their landscape and lithic resources, their movements across a territory bordered by the ocean to the south and west, a large river to the east and mountain ranges to the north, and the nature of the social networks they established.
+Vale Boi stands as a key archaeological site for this study since it is currently the only site in southern Portugal with a long-term occupation spanning most of the UP and allowing the exploration of raw material use and mobility trends through time [@cascalheira_2017].
+Furthermore, previous studies have highlighted the role of Vale Boi in understanding different patterns of territory exploitation and contact with other regions of the Peninsula throughout the UP.
+During the Solutrean, Vale Boi has been interpreted as an interface between the several sites of Central Portugal and Southern Spain, making it a key site to identify extended social networks that promoted the exchange of information and culture [@cascalheiraValeBoiAlgarve2013; @cascalheira_2017; @cascalheiraSolutrenseEmPortugal2013].
+
+On the contrary, techno-typological studies from the Gravettian occupations of Vale Boi instead show marked differences between regions and suggest a more limited circulation of people and information in southern Iberia [@marreirosLithicTechnologyVariability2013].
+
+To investigate the applicability of these patterns to the management and circulation of lithic raw materials, and to examine the prevailing hypotheses concerning UP mobility and social network exchanges, this paper presents the first comprehensive characterization of chert raw materials derived from the Gravettian, Proto-Solutrean, and Solutrean (c.
+32-19 ka cal BP) sequence at Vale Boi.
+
+## Site description
+
+Vale Boi is located on the western coast of the Algarve region in southern Portugal, within a small valley that extends southward to the Atlantic coast, approximately 2 km away (@fig-map b).
+It is bordered by limestone outcrops that form rock shelters facing west and southwest [@bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011].
+The site extends for more than 10,000 sq meters along the slope of the valley, through which three main areas have been excavated between 2000 and 2019: Slope, Shelter, and Terrace.
+
+The Terrace encompasses occupations from the UP to the Early Neolithic, resulting in the identification of eight main lithostratigraphic units.
+The UP sequence includes several occupations attributed to the Gravettian (levels 8 to 6) between 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between 26 and 24 ka cal BP, and Solutrean (level 4D, 4C, 4C, 4 and 3) between 24 and 20 ka cal BP [@belmiro_gravettian-solutrean_2021; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017].
+The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations.
+From these, the Solutrean levels show the most intensive occupation, with three, well-preserved, archaeological horizons (layers C to A) and dated from 24 to 22 ka cal BP [@cascalheiraValeBoiAlgarve2013; @cascalheiraTecnologiaLiticaSolutrense2010; @cascalheira_2017].
+
+```{r}
+#| label: tbl-c14
+#| tbl-cap: Radiocarbon dates for Vale Boi Upper Paleolithic sequence of the Terrace area and Solutrean of the Shelter area, partially adapted from Cascalheira et al.
+[-@cascalheira_2017] and Belmiro et al.
+[-@belmiro_gravettian-solutrean_2021], and including new dates.
+All dates were calibrated or recalibrated using the IntCal20 and Marine20 curves [@reimerIntCal20NorthernHemisphere2020; @heatonMarine20MarineRadiocarbon2020] in OxCal online v.4.4.4 [@ramseyBayesianAnalysisRadiocarbon2009].
+For shells a DR value of 123 ± 58 was used, obtained through the CHRONO Marine Reservoir Database at http://calib.org/marine/.
+
+c14_tbl <- read_csv(here::here('analysis/data/raw_data/C14_VB_calBP.csv'))
+c14_20 <- read_csv(here::here('analysis/data/raw_data/IntCal20_Marine20_Oxcal.csv'))
+dates_cal <- left_join(c14_tbl, c14_20, by = "Lab-ref")
+dates_cal <- dates_cal %>% 
+  dplyr::rename("Date Cal BP" = "Concatenate") %>% 
+  dplyr::select(Area, Level, Culture, `Lab-ref`, Date, SD, `Date Cal BP`, Material)
+
+kable(dates_cal, "simple")
+
+```
+
+```{r}
+#| label: fig-map
+#| fig-cap: Iberian Peninsula map with locations and geological formations mentioned throughout the paper.
+a) Detail of the central Portugal region where the Cenomanian formations with chert nodules are located.
+b) Detail of southwestern Portugal, with chert and chalcedony bearing formations, outcrops, and location of Vale Boi.
+c) Detail of eastern Algarve with chert bearing formations.
+The figure was produced using ArcGIS Pro 3.2.
+
+knitr::include_graphics(here::here("analysis/figures/map_layout.png"))
+```
+
+## Regional geological background
+
+The Algarve region is a complex territory, marked by several geomorphic sub-regions and geological units that provide different lithological resources used throughout prehistory.
+Two of the main geological units which characterize the Algarve are: 1) the Baixo Alentejo Flyschoid Group, part of the South Portuguese Zone [@fernandes_new_2012; @oliveira_carta_1984-1; @onezimeNewGeodynamicInterpretation2003], located in the north sector of the region, and characterized by schists, greywackes and quartzites; and 2) the Algarve basin in the south, which is characterized by limestones, dolomites and marls [@terrinha_bacia_2006].
+It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules and beds can be found [@manuppellaNoticiaExplicativaFolha1987; @manuppellaNoticiaExplicativaFolha2007; @ribeiro_evolucao_2005; @rocha_noticia_1979; @rocha_noticia_1983; @terrinha_bacia_2006].
+Following Luedtke [-@luedtke_archaeologists_1992], we use the term chert to refer to sedimentary rocks composed primarily of microcrystalline quartz, which includes other possible varieties such as jasper, agate, or chalcedony.
+The term chalcedony is used in its petrographic sense and refers to cherts with a fibrous quartz structure.
+The availability of regional resources containing cherts and the characterization of their lithologies has already been published by our team [@belmiroCreatingFramesReference2023], and we herein present the summarized main findings.
+The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b).
+These cherts are frequently opaque with yellow, grey, purple, or red colours, occasionally with bands or laminations, and with frequently visible sponge spicules.
+Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being around 20 cm, and frequently characterized by high levels of dolomitization or the presence of fractures.
+Chert sources were also identified in two formations from the eastern area of the Algarve (@fig-map c).
+The Middle Jurassic Malhão formation chert outcrops were identified at around 80-100 km east of the site and characterized by chert nodules with similar macroscopic features to those found in western Algarve.
+The Upper Jurassic Jordana formation chert outcrops were identified at around 90 km east of Vale Boi and characterized by chert nodules with significantly different macroscopic features to those of other formations, also with frequently small sizes (\~5 cm), difficult removal from the parent rock and nonexistence of secondary deposits.
+Chalcedony outcrops approximately 10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes (Bicho et al., 2013).
+The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.
+
+# Materials and methods
+
+To characterize the siliceous raw materials used during the UP occupations at Vale Boi and to identify their geological sources, we focused on chalcedony and chert archaeological materials from the Terrace and Shelter areas.
+These two areas were chosen since both have detailed spatial information, several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal, and geoarchaeological data.
+Chert was chosen for this study as it is one of the three main used raw materials at the archaeological site of Vale Boi alongside quartz and greywacke.
+Other raw materials such as chalcedony, dolerite, and schists have also been identified [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @pereira_territory_2016-1].
+Using the field database with all individually coordinated lithic artefacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials.
+@fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents between \~50-70% of the lithic assemblage from the Solutrean layers A, B, and C.
+Based on previous lithic studies in the terrace and shelter areas, the remaining 80% of other raw materials are mostly composed of quartz and greywacke [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009].
+Although systematic raw material research has not been conducted on these materials, according to the regional geological context, they have been interpreted as local and readily accessible in the vicinity of the site [@pereira_territory_2016-1].
+Frequently, the quartz and greywacke assemblages include large amounts of shattered and/or unknapped chunks or slabs.
+Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020].
+Based on this, despite the low amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.
+The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].
+
+```{r}
+#| label: fig-rm-comparison
+#| fig-cap: "Comparison of chert, chalcedony, and other raw materials piece plotted in the Terrace (a) and Shelter (b) areas of Vale Boi."
+
+# Terrace rm comparison
+## Data preparation
+context_comp <- field_data %>%
+  mutate(Level = case_when(Level == "4B" ~ "4", # Collapsing samples from level 4B into 4
+                           Level == "7C" ~ "7",
+                           Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7
+                           TRUE ~ Level)) %>% 
+  mutate(Level = case_when( # Dividing 5 as previous
+    Level == 5 & Z > 24.1 ~ "4E/5",
+    Level == 5 & Z < 24.1 ~ "5",
+    TRUE ~ Level)) %>% 
+  filter(!Level %in% c("3B", "5B", "5C")) %>% 
+  dplyr::filter(Code %in% c("LITHIC", "POINT"))
+
+dataset_comp <- phase_dataset %>% 
+  select(ID, RMTYPE)
+T_comp <- full_join(context_comp, dataset_comp, by = "ID")
+T_comp["RMTYPE"][is.na(T_comp["RMTYPE"])] <- "Other" # All NA's (other RMs) turned into "Other"
+
+T_comp <- T_comp %>% 
+ mutate(Rawmaterial = case_when(
+    RMTYPE == "Other" ~ "Other",
+    RMTYPE == "TYPE 1" ~ "Chalcedony",
+    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category
+  dplyr::filter(Level %in% c("4", "4C", "4D", "4E", "4E/5", "6", "7")) # Choosing only UP levels (excluding 3)
+
+## Barplot
+T_comp_bar <- T_comp %>%
+  dplyr::group_by(Level, Rawmaterial) %>% 
+  dplyr::summarise(n = n()) %>%
+  mutate(percent = n/sum(n)) %>% 
+  ggplot(aes(x=Level, y = percent, fill = Rawmaterial)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  theme_minimal() + #type of theme
+  labs(x = "Terrace levels", y = "Count %") +# change labels +# change labels
+  guides(fill=guide_legend(title=NULL)) + # remove legend title
+  geom_text(aes(label=ifelse(percent >= 0.07, paste0(sprintf("%.1f", percent*100),"%"),"")),
+                position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_text(size = 10, vjust = -1))+
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+
+# Shelter rm comparison
+## Data preparation
+S_field_comp <- S_field_data %>% 
+  filter(CODE == "L") # Filtering only lithics
+S_data_comp <- S_dataset %>% 
+  dplyr::select(ID, RMTYPE)
+S_comp <- full_join(S_field_comp, S_data_comp, by = "ID")
+S_layerdata <- S_layerdata %>% 
+  distinct() # Keeping only individual IDs to avoid repetitions
+S_comp <- left_join(S_comp, S_layerdata, by = "ID")
+S_comp["RMTYPE"][is.na(S_comp["RMTYPE"])] <- "Other" # All NA's (other RMs) turned into "Other"
+S_comp["Layer"][is.na(S_comp["Layer"])] <- "No info" # Unknown Shelter levels turned into "No info" to avoid NA
+
+S_comp <- S_comp %>% 
+ mutate(Rawmaterial = case_when(
+    RMTYPE == "Other" ~ "Other",
+    RMTYPE == "TYPE 1" ~ "Chalcedony",
+    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category
+  dplyr::mutate(Layer = case_when(
+    Layer == "c" ~ "C",
+    TRUE ~ Layer)) %>% 
+  filter(!Layer %in% c("D", "Z")) # Choosing only Solutrean occupation layers (A, B and C)
+
+## Barplot
+S_comp_bar <- S_comp %>%
+  dplyr::group_by(Layer, Rawmaterial) %>% 
+  dplyr::summarise(n = n()) %>%
+  mutate(percent = n/sum(n)) %>% 
+  ggplot(aes(x=Layer, y = percent, fill = Rawmaterial)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  theme_minimal() + #type of theme
+  labs(x = "Shelter levels", y = "Count %") +# change labels
+  guides(fill=guide_legend(title=NULL)) + # remove legend title
+  geom_text(aes(label=ifelse(percent >= 0.07, paste0(sprintf("%.1f", percent*100),"%"),"")),
+                position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_text(size = 10, vjust = -1)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+
+## Unused barplot: here to understand how the chert ratio changes when A, B and C are collapsed since "No info" has most of the other rms.
+#S_comp %>% 
+#  dplyr::group_by(Rawmaterial) %>% 
+#  dplyr::summarise(n = n()) %>%
+#  mutate(percent = n/sum(n)) %>% 
+#  ggplot(aes(x=Rawmaterial, y = percent, fill = Rawmaterial)) +
+#  geom_bar(stat = "identity", width = 0.5) +
+#  scale_y_continuous(labels = scales::percent_format()) +
+#  theme_minimal() + #type of theme
+#  labs(x = "Shelter levels", y = "Count %") +# change labels
+#  guides(fill=guide_legend(title=NULL)) + # remove legend title
+#  geom_text(aes(label=ifelse(percent >= 0.07, paste0(sprintf("%.1f", percent*100),"%"),"")),
+#                position=position_stack(vjust=0.5), colour="white", size=3) +
+#  theme(axis.title.x = element_text(size = 10, vjust = -1)) +
+#  scale_fill_manual(values = c("#C74B50", "#D49B54", "#845460", "#EAD3CB", "#BDC7C9",
+#                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+
+## Terrace and shelter barplots
+ggarrange(T_comp_bar, S_comp_bar, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
+
+```
+
+We employed both macroscopic and petrographic methods to achieve a more comprehensive characterization of the materials, aiming to offset the inherent limitations of each method [@luedtke_archaeologists_1992].
+
+Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustillo_is_2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=5318).
+The distinct features of various local cherts facilitate their differentiation from cherts of neighbouring origins.
+Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustillo_is_2009; @tarrino_materias_2013].
+Therefore, we implemented an initial phase of macroscopic analysis, subsequently augmented by petrographic study to diminish the subjectivity of the analysis and enhance the material characterization with petrographic data.
+The synergistic application of these two methods has been previously validated in the characterization of Algarve region cherts, following recent prospection efforts [@belmiroCreatingFramesReference2023].
+Hence, an extensive macroscopic and petrographic collection is available for comparative analysis with archaeological materials, employing consistent methodologies.
+
+This reference collection is the Lusolit lithotheque, hosted physically at ICArEHB (University of Algarve, Faro) and digitally at www.lusolit.icarehb.com, and aided in the characterization and source attribution of cherts.
+Hand samples were used throughout the macroscopic characterization for comparison with the archaeological materials, and the archaeological thin sections were compared with the previously studied geological chert thin sections.
+To identify the source of other lithologies which are not congruent with the regional cherts and identify possible long-distance chert procurement, we used hand samples and thin sections of cherts from central Portugal also available in the Lusolit lithotheque.
+The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz (UCA) was also visited to understand the macroscopic variability of the local and regional cherts of the lower Guadalquivir basin, within the Atlantic strip of south-western Spain (present-day provinces of Huelva and Cádiz).
+Following previous studies, we classify the distance of chert sources from the site as local (1-30 km from the site), regional (30-120 km), and long-distance (\>120 km) [@herrero-alonsoManagementLithicRaw2020; @tarrinoFlintRawMaterial2015].
+
+The macroscopic analysis of the archaeological materials was divided into two steps.
+The first step included a preliminary characterization using a hand lens of 10x magnification.
+This step allowed us to create types and subtypes based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these include characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992], and a data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1).
+In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types.
+The data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1).
+Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis is simplified and focused on collecting data related to weight, cortex, and alterations.
+This first step was applied to a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 8 to 4) and 2805 chert artefacts from the Solutrean sequence of the Shelter area (layers C to A).
+
+While the advantages of employing a macroscopic methodology for analysing the cherts of Vale Boi are notable, a limitation of this approach is its inability to unequivocally characterize and categorize artefacts that exhibit surface alterations typical of an open-air site, including those related to fire, since weathered and altered surface may destroy visible structures and even alter the chemical composition [@delluniversita_development_2019].
+Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversita_development_2019; @gomezdesolerPanadellaChertMontmaneu2020; @soto_procurement_2016].
+For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), they were analysed using the individual sample dataset, but grouped in a category of indeterminate cherts (INDET).
+
+As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artefacts, corresponding to 40% of the total sample (n=1027).
+This is due to the frequent alterations of the sample, especially related to patinas and alteration rinds, pits, and fire/heat alterations.
+For all other levels in the Terrace, indeterminate represents on average \~28% of the chert artefacts, while in the Shelter the frequency of INDET is slightly lower (\~20-25%).
+
+```{r}
+#| label: fig-terrace-alterations
+#| fig-cap: Samples with extensive alterations which prevented the identification of chert type.
+a) artefact with cracks, white patina and reddening covering 100% of the surface; b) artefact with surface alterations, displaying a portion with patina (brown) and a portion without patina (grey), cleaned with Hydrochloric acid.
+
+knitr::include_graphics(here::here("analysis/figures/terrace_alterations.png"))
+
+```
+
+```{r}
+#| label: tbl-indet-freq
+#| tbl-cap: N and percentage of grouped and indeterminate (INDET) artefacts by level in the Terrace (levels 4 to 7) and Shelter (A to C) areas.
+
+terrace_indet_table <- phase_dataset %>% 
+  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>%
+  tabyl(Group,Level) %>% 
+  adorn_totals("row") %>%
+  adorn_percentages("col") %>%
+  adorn_pct_formatting(2) %>%
+  adorn_ns("front")
+
+shelter_indet_table <- S_dataset %>% 
+  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
+  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
+  filter(!Layer %in% c("D", "ND")) %>% 
+  tabyl(Group,Layer) %>% 
+  adorn_totals("row") %>%
+  adorn_percentages("col") %>%
+  adorn_pct_formatting(2) %>%
+  adorn_ns("front")
+
+indet_table <- full_join(terrace_indet_table, shelter_indet_table, by = "Group")
+
+kable(head(indet_table))
+
+```
+
+To better characterize the different chert types identified in the macroscopic analysis and aid in minimizing the inherent caveats of macroscopic methods, twenty-two thin sections were prepared at the Thin Section Lab (Toul, France).
+The samples chosen for this analysis include at least one sample from each type present throughout the stratigraphy of the Terrace area, from different archaeological levels (@tbl-TS-list).
+For types that showed high macroscopic variability (e.g.
+translucency and colour variation or differentiated bioclast facies), thin sections from sub-types were analysed to collect petrographic data that reflected this variability.
+Petrographic thin sections were analysed by polarized light microscopy (Nikon LV100ND) to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations.
+
+```{r}
+#| label: tbl-TS-list
+#| tbl-cap: List of archaeological thin sections, with the summarised description of observed macroscopic variability and sub-types.
+
+TS_list <- read_csv(here::here('analysis/data/raw_data/TS_short_list.csv'))
+
+kable(TS_list)
+
+```
+
+The complete petrographic analysis for all thin sections studied in this paper (including macroscopic and microscopic figures) can be found as Supplementary Information Materials at our online repository (++).
+A detailed macroscopic and petrographic description and figures for each type and sub-type of chert and chalcedony (Online Resource 2) and the data dictionary used for the petrographic analysis (Online Resource 1) can be found in Supplementary Information.
+The complete set of images of the geological thin sections can also be consulted in the LusoLit online database.
+
+The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online research compendium (https://doi.org/10.17605/OSF.IO/DBFT2).
+We used the rrtools package by Marwick et al.
+[-@marwickPackagingDataAnalytical2018] to create a research compendium and write a reproducible journal article.
+The provided files include the complete set of raw data used in the analysis, along with a custom R project (Wickham, 2015) containing the code required to generate all tables and figures.
+To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY [additional details can be found in @marwickComputationalReproducibilityArchaeological2017].
+
+# Results
 
@@
 
-The three main used raw materials at the archaeological site of Vale Boi are quartz, greywacke, and chert.
-Using the field database with all individually coordinated lithic artifacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials.
-Since some lithic assemblages of the Terrace area are currently under study, the data regarding the percentage of raw materials in the “Other” category was not available for all archaeological levels.
-@fig-rm-comparison shows that for the selected levels, and using single coordinated artifacts (excluding buckets), chert represents only \~20-15% of the lithic assemblage.
-Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke.
-Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site.
-Frequently, the quartz and greywacke assemblage includes large amounts of shattered and/or unknapped chunks or slabs.
-Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz.
-Based on this, despite the low amounts of individually plotted chert artifacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.
+## Characterization and source attribution
+
+Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony types, present throughout the stratigraphy in the Terrace and Shelter areas (a summarised characterization table can be found in the Supplementary Information, Online Resource 3).
+A distinct type was identified exclusively in the Shelter area, with only a limited presence in the Terrace (n=2), and 16 varieties, each consisting of one to three samples, were identified across various levels of both the Terrace and Shelter.
+The latter were clustered into a category named Trace Lithotypes (TL).
+Through the comparison with the reference collections, we identified cherts congruent with the local sources and cherts that are likely not local due to significant differences in appearance, fossil content, and petrography when compared to the cherts available in the geological setting located within a 30 km radius of Vale Boi.
+
+### Cherts congruent with local sources
+
+The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts (@fig-local-types).
+The chalcedony (Type 1; @fig-local-types a) is identified as a massive, fibrous variant devoid of allochems (@fig-local-types a2).
+It represents 6.6% of the overall Terrace cherts (excluding INDET) and 2.6% in the Shelter.
+It presents a colourless and translucent appearance, frequently exhibiting partial or complete white patination (@fig-local-types a1).
+This type is marked by extensive crystallizations, irregularities, and fractures, coupled with post-depositional alterations that challenge the identification and preservation of the cortex.
+Both macroscopic and petrographic analyses confirm its congruence with hydrothermal chalcedony located within lenses in Triassic volcanic formations, which are similarly marked by irregularities and fractures.
 
 ```{r}
-#| label: tbl-indet-freq
-#| tbl-cap: N and percentage of grouped and indeterminate artifacts by level.
+#| label: tbl-cortex_tab
+#| tbl-cap: Frequency of cortex type and thickness by raw material from the Terrace area.
 
-indet_table <- phase_dataset %>% 
-  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
-  tabyl(Group,Level) %>% 
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
-  adorn_ns("front")
+# Cortex table
+## Preparing cortex db Terrace
+cortex_tab <- phase_dataset %>% 
+  select(RMTYPE, CORTEX, CORTEXAPPEARANCE, CORTEXTHICKNESS) %>%
+  filter(CORTEX == "YES") %>% # filtering only samples with cortex
+  filter(RMTYPE != "INDET") %>% 
+  filter(RMTYPE != "TYPE 11") %>% # removing Type 11 since it barely has samples
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL")))
+
+cortex_tab <- cortex_tab %>% # dropping the caps and organizing value order
+  mutate(CORTEXAPPEARANCE = case_when(
+    CORTEXAPPEARANCE == "ROUNDED" ~ "Rounded",
+    CORTEXAPPEARANCE == "PEBBLE" ~ "Pebble",
+    CORTEXAPPEARANCE == "OUTCROP" | CORTEXAPPEARANCE == "Outcrop" ~ "Irregular",
+    CORTEXAPPEARANCE == "OTHER" | CORTEXAPPEARANCE == "Other" | CORTEXAPPEARANCE == "INDETERMINATE" ~ "Indeterminate",
+    TRUE ~ CORTEXAPPEARANCE)) %>%
+  dplyr::mutate(CORTEXTHICKNESS = case_when(
+    CORTEXTHICKNESS == "MEDIUM" ~ "Medium",
+    CORTEXTHICKNESS == "THICK" ~ "Thick",
+    CORTEXTHICKNESS == "THIN" ~ "Thin",
+    TRUE ~ CORTEXTHICKNESS)) %>% 
+  mutate(CORTEXAPPEARANCE = factor(CORTEXAPPEARANCE, levels = c("Irregular", "Rounded", "Pebble", "Indeterminate"))) %>%
+  mutate(CORTEXTHICKNESS = factor(CORTEXTHICKNESS, levels = c("Thin", "Medium", "Thick"))) %>% 
+  drop_na() # removing missing values that may be cortex=NO instead passed as YES
+
+## gt summary table
+cortex_tab %>%
+  select(RMTYPE, CORTEXAPPEARANCE, CORTEXTHICKNESS) %>% 
+  tbl_summary(by=RMTYPE, label = list(CORTEXAPPEARANCE ~ "Cortex type", CORTEXTHICKNESS ~ "Thickness")) %>% 
+  modify_header(label ~ "**Variable**") %>% 
+  bold_labels()
 
@@
 
-As previously mentioned, whenever too altered to confidently attribute a group, the artifacts were instead classified as indeterminable (INDET).
-As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artifacts, corresponding to 40% of the total sample (n=1027).
-This is due to the frequent alterations of the sample, especially related with patinas and alteration rinds, pits and fire/heat alterations.
-For all other levels, indeterminate represents in mean \~28% of the chert artifacts.
+The cherts are massive micro-cryptocrystalline quartz from a marine environment (@fig-local-types b-h).
+They represent \~60% of the overall Terrace cherts and \~80% of the chert in the Shelter.
+Macroscopically, these types show variability which resulted in their separation in four different macroscopic types (Type 2 to 5).
+In general, these are opaque to sub-translucent cherts, with colours varying from brown, yellow, pale red, and grey, with the frequent presence of spicules and iron oxides.
+The differences are in their colours, nodule morphology, frequency, and variety of bioclasts (Type 2 including triaxon sponge spicules and echinoderm spines), something already identified in the geological samples [@belmiroCreatingFramesReference2023].
+Whenever present, the cortex seems to show a mixture of irregular cortex and rounded cortex, both from an outcrop source (@tbl-cortex_tab).
+Petrographically, however, these cherts are homogeneous, except for a small number of artefacts with a banded variety within type 5.
+The macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.
+These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite.
+Type 5 shows higher percentages of micrite/sparite.
+In all samples, fossils range from common to very frequent, albeit frequently unidentifiable.
+When identifiable in thin section, they are sponge spicules (@fig-local-types b2, c2, h2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (@fig-local-types h3) in a sample from type 5.
+Porosity varies between 10% and 1%, of vug type.
+Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiroCreatingFramesReference2023], we interpret Types 2, 3, 4, and 5 to be local, as they are congruent with the described characteristics representative of cherts from the Lower Jurassic (Carixian) and Middle Jurassic Malhão formations of the Algarve basin.
+Types 2 and 3 are congruent with the frequently available brown/yellow and red nodules and beds (@fig-local-types b-d), ranging from uncommon to very frequent fossils.
+The macroscopic similarities of Types 4 and 5 are more limited (@fig-local-types e-h), as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the inland Lower Jurassic outcrops (@fig-map b).
+As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape.
+This is especially relevant since the inland outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].
 
 ```{r}
-#| label: fig-rm-comparison
-#| fig-cap: "Comparison of chert, chalcedony and other raw materials coordinate in the terrace area of the archaeological site of Vale Boi."
-#| echo: false
+#| label: fig-local-types
+#| fig-cap: Variability of local chert types and main petrographic characteristics.
+a) Type 1 macroscopic appearance (a1) and thin-section (a2).
+b) Type 3 macroscopic appearance (b1) and thin-section (b2).
+c-d) Type 2 macroscopic appearance (c1-d1) and thin-section (c2-d2).
+e-f) Type 4 macroscopic appearance (e-f1) and thin-section (f2-f3).
+g-h) Type 5 macroscopic appearance (g-h1) and thin section (h2-h3).
 
-context_short <- context %>% 
-  filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
-dataset_short <- T_dataset %>% 
-  select(ID, RMTYPE)
-T_short <- full_join(context_short, dataset_short, by = "ID")
-T_short["RMTYPE"][is.na(T_short["RMTYPE"])] <- "Other"
-
-T_short <- T_short %>% 
- mutate(Rawmaterial = case_when(
-    RMTYPE == "Other" ~ "Other",
-    RMTYPE == "Chalcedony" ~ "Chalcedony",
-    RMTYPE != "Other" | RMTYPE != "Chalcedony" ~ "Chert"))
-
-#T_short_table <- T_short %>% 
-#   tabyl(Rawmaterial, Level) %>%
-# adorn_totals(c("row", "col")) %>%
-#  adorn_percentages("col") %>%
-#  adorn_pct_formatting(2) %>%
-#  adorn_ns("front")
-
-T_short %>%
-  dplyr::group_by(Level, Rawmaterial) %>% 
-  dplyr::summarise(n = n()) %>%
-  mutate(percent = n/sum(n)) %>% 
-  ggplot(aes(x=Level, y = percent, fill = Rawmaterial)) +
-  geom_bar(stat = "identity", width = 0.5) +
-  scale_y_continuous(labels = scales::percent_format()) +
-  theme_minimal() + #type of theme
-  labs(x = "Level", y = NULL) +# change labels
-  guides(fill=guide_legend(title=NULL)) + # remove legend title
-  geom_text(aes(label=ifelse(percent >= 0.07, paste0(sprintf("%.1f", percent*100),"%"),"")),
-                position=position_stack(vjust=0.5), colour="white", size=3) +
-  theme(axis.title.x = element_text(size = 10, vjust = -1))
+knitr::include_graphics(here::here("analysis/figures/local_grid.png"))
 
@@
 
-Through the macroscopic and petrographic analysis we identified 24 types of chert and three types of chalcedony.
-Of these, one type of chalcedony and 9 types of chert are composed of several samples (\>10) and present across both areas (Terrace and Shelter) and throughout the Upper Paleolithic occupations of the Terrace area.
-Thin sections were produced for the lithics of the latter groups.
-One type had several samples (\>10; Type GB) but was only identified in the Shelter area, while 14 varieties of chert and 2 chalcedony are composed of one to three samples and restricted to specific levels.
-The latter varieties were collapsed into a category named TL (Trace lithotypes).
+### Cherts not congruent with local sources
 
-We present here a short description of each identified macroscopic and petrographic type.
-A detailed macroscopic and petrographic description and figures for each type and sub-type of chert and chalcedony can be found in the SOM (++).
-The petrographic analysis PDF files for all thin sections studied in this paper (including macroscopic and microscopic figures) can also be found in our GitHub page (++) as well as the OSF repository (++).
+We identified six chert types with macroscopic and petrographic characteristics that are not congruent with the chert available in known outcrops in western Algarve (@fig-distance-types).
+Despite their scarce representation in the archaeological record, these raw materials of presumably allochthonous origin are a valuable source of information.
+Often these lithological types can be considered as clear lithological markers of mobility, which can contribute to support hypotheses about mobility and contacts with neighbouring populations and territories.
 
-## Type description
+```{r}
+#| label: fig-distance-types
+#| fig-cap: Variability of non-chert types and main petrographic characteristics.
+a-b) Type 6 macroscopic appearance (a1-b1) and thin-section (a2-b2); b1-b2 correspond to a sample covered 100% in white patina.
+c-d) Type 7 macroscopic appearance (c1-d1) and thin-section (c2-d2) of peloidal facies (c1-c2) and oolitic facies (d1-d2).
+e-f) Type 8 macroscopic appearance (e-f1) and thin-section (f2-f3); white and black arrows correspond to foraminifera.
+g) Type 9 macroscopic appearance (g1) and thin-section (g2).
+Type 10 macroscopic appearance (h1) and thin-section (h2).
 
-The chalcedony is a massive, fibrous chalcedony mudstone.
-It is characterized by a colorless and translucent appearance, often partially or completely white patinated.
-It has good to medium quality, due to frequent large crystallizations, irregularities or fractures which impact the knapping.
-In thin section, the chalcedony is composed by fibrous chalcedony, without allochems or bioclasts, and no porosity (\<1%).
-Radiating spherulitic textures are very frequent.
+knitr::include_graphics(here::here("analysis/figures/nonlocal_grid.png"))
 
-The first chert groups identified at Vale Boi are massive micro-cryptocrystalline quartzes from a marine environment.
-Petrographically, these cherts are homogeneous, except for a small amount of artifacts with a banded variety within type 12.
-These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite.
-Group 12 shows higher percentages of micrite/sparite.
-In all samples, fossils range from common to very frequent, albeit frequently unidentifiable.
-When identifiable in thin-section, they are sponge spicules, with the rare occurrence of possible foraminifera and ostracod in a sample from type 12.
-Porosity varies between 10% and 1%, of vug type.
-Macroscopically however, these groups show variability, which resulted in their separation in 4 different macroscopic groups; this separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.
+```
 
-Group 2 is characterized by single-color and multiple-color samples, ranging between yellow, grayish brown, pale red, and white, and with variable patterns such as splotched, broad mottling, speckling, and laminations.
-They are opaque, with sub-translucent areas mainly focused in the white areas of the sample.
-Whenever present, the cortex can be rounded or irregular, varying between thin to thick.
-Inclusions include black speckles and red speckles, which are probably organics and iron oxides.
-Bioclasts are frequently unidentifiable, but also include monaxon sponge spicules, bivalve shells replaced by quartz or chalcedony, and rarely triaxon sponge spicules.
-In the shelter area, the variety of identified fossils also includes possible uncommon ostracods, a possible small foraminifer, and echinoderm spines.
-
-Group 15 is characterized by light gray and reddish gray colors and varies between opaque to sub-translucent.
-It is characterized by the presence of splotches (often coarser and sub-translucent), speckling, and fine lamination.
-Whenever present, the cortex is white, rounded, and of medium thickness.
-Allochems such as iron oxides or opaques are also visible macroscopically as black speckling or rare red speckling.
-The bioclasts are frequently unidentifiable, due to poor preservation and are replaced by microcrystalline quartz or fibrous chalcedony, except for uncommon sponge spicules.
-
-Type 7 shows a heterogeneous structure with two types of fabrics.
-The interior of the samples and the nodules is gray, and sub-translucent with a single mica mineral, flecks, and speckles.
-The outer edge and within the grey fabric as splotches or broad mottling are white, opaque, and with yellow speckles.
-The cortex is white, with a sharp transition and with a medium to thick width.
-The morphology of the cortex in fully cortical samples is originally available as irregular (slightly oval) nodules within the parent rock.
-Whenever identifiable, bioclasts are limited to sponge spicules.
-
-Type 12 chert has a conchoidal fracture (albeit uneven at times) and acceptable to low quality due to the lack of homogeneity and frequent fractures, as well as a coarser fabric in certain areas.
-It has gray colors and ranges from opaque to sub-translucent.
-The group displays frequent patterns: splotched, banded, broad mottling, and speckling.
-The cortex is from an outcrop, frequently rounded, but also irregular.
-Based on reference samples, the chert seems to be enclosed within the parent rock as irregular or tabular nodules).
-The cortex is thin, with a sharp transition, and some artifacts show chunks of the parent rock attached to the cortex.
-Bioclasts are uncommon, and when identifiable are sponge spicules.
-
-Group 11 corresponds to a massive cryptocrystalline quartz with a mudstone texture.
-The lack of local similar cherts, as well as identifiable fossils does not allow us to attribute a depositional environment or epoch/stage.
-However, rare spicules were identified; given the exclusivity of the sponge’s (Porifera) living environment to be marine (Flügel, 2010), group 11 may be considered a chert formed in a marine environment.
-This chert is translucent and shows a variety of colors, ranging from brown, red, and gray, which seem to vary in function of the artifact’s thickness.
-Patterns are rare, however, whenever present they are laminated, with flecks, and splotches.
-In unaltered samples, bioclasts are rare.
-The cortex is rounded, and rarely pebble-like, ranging from thin to medium thickness.
-Two sub-groups were individualized based on macroscopic characteristics such as color, translucency variation, and patterns, although they may simply reflect the use of specific nodules from within Type 11’s variability.
-Five thin sections were produced to better characterize the macroscopic variability of this group and sub-groups.
-The petrographic analysis shows the samples are characterized by a mudstone texture and massive microstructure.
-It is mainly composed of cryptocrystalline quartz (between 95-97% in the samples), with accessory fibrous chalcedony found replacing fossils or filling fractures, and in some samples macrocrystalline quartz, microcrystalline quartz, and micrite/sparite.
-Allochems are opaques (u), iron oxides (u), and bioclasts (r/u).
-Bioclasts are unidentifiable and poorly preserved, replaced by fibrous chalcedony and rarely by microcrystalline quartz.
-Sponge spicules were identified in a single sample.
+Type 6 (@fig-distance-types a-b) corresponds to a massive cryptocrystalline quartz with a mudstone texture, from a marine environment.
+It represents \~18% of the overall Terrace cherts and 12% of the chert in the Shelter.
+This chert is translucent and shows a variety of colours, ranging from brown, red, and grey, which seem to vary in function of the artefact’s thickness.
+In unaltered samples, visible bioclasts are rare.
+The cortex is mostly rounded, and rarely pebble-like (FIG), ranging from thin to medium thickness.
+Two sub-types were individualized based on macroscopic characteristics such as colour, translucency variation, and patterns, although they may simply reflect the use of specific nodules from within Type 6’s variability.
+Five thin sections were produced to better characterize the macroscopic variability of this type and sub-types.
+The petrographic analysis shows the samples are characterized by a massive microstructure, mainly composed of cryptocrystalline quartz (between 95-97% in the samples), with accessory fibrous chalcedony found replacing fossils or filling fractures (@fig-distance-types a2-b2), and in some samples macrocrystalline quartz, microcrystalline quartz, and micrite/sparite.
+Allochems are opaques, iron oxides, and bioclasts.
+Bioclasts are unidentifiable and poorly preserved, with rare identifiable fossils such as sponge spicules, replaced by fibrous chalcedony and rarely by microcrystalline quartz.
 Porosity is present between \<1-1%, of vug-type when identifiable.
 
-Group 11F corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment.
-Identified and well-preserved fossils are foraminifera from the Pfenderenidae family, which are larger benthic foraminifera from marine environments.
+The similarities between chert artefacts at the UP occupations of Vale Boi and the chert from the Rio Maior region (Estremadura Portuguesa; @fig-map a) have been previously suggested although without systematic studies to ascertain the attribution [@bicho_ecodynamics_2013; @pereira_territory_2016-1].
+These Cretaceous cherts, mostly found in secondary deposition settings, are described as frequently translucent, with geodes, colours ranging from yellow to red or grey, and mineralogically homogeneous between them.
+Petrographic studies from the Estremadura area highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matias_raw_2016].
+When identified, sponge spicules and possible rare foraminifera have been identified [@matiasAprovisionamentoMateriasprimasLiticas2012].
+Studies of Upper Cretaceous cherts located in the Lisbon area describe cherts with similar macroscopic characteristics (@fig-map a), with micro-cryptocrystalline quartz mudstones with bioclasts such as sponge spicules or bivalve fragments, or wackestone/packstone textures with abundant bioclasts, including ostracods and different types of foraminifers [@jordaoFlintSourcesMobility2022].
+These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from central Portugal hosted at LusoLit.
+Alike the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matias_raw_2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
+From these cherts, only the dark grey nodules (RT231; SOM) showed identifiable fossils which were rare ostracods.
+Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Type 6 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts (Upper Cenomanian formations) from central Portugal, between \~215-250 km north of Vale Boi (@fig-map a).
+
+Type 7 (@fig-distance-types c-d) is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
+In the terrace, it represents 10% of the overall Terrace cherts and less than 1% of the chert in the Shelter.
+It is characterized by white and grey colours.
+The cortex is from an outcrop, rounded and in lesser frequencies (\<30%) irregular, and varying between medium thickness to thick.
+This type shows several facies, with different types and concentrations of allochems and bioclasts, such as peloids (@fig-distance-types c1), ooids (@fig-distance-types d1), and sponge spicules.
+A detailed scheme of the facies can be found in the Supplementary Information (Online Resource 2).
+The petrography results show this chert has a massive microstructure and packstone texture (@fig-distance-types c2-d2).
+It is composed mainly of micro-cryptocrystalline quartz (80%), fibrous chalcedony (10%), and micrite (10%).
+Allochems are opaques, iron oxides, ooids, and peloids.
+The ooids are poorly preserved and only uncommonly show concentric lamellae in plane-polarized light.
+Unlike macroscopical observations, bioclasts under the thin section are all unidentifiable, although rare fossils may be bivalve shells.
+Porosity is variable (\<1-10%).
+The consultation of the lithotheque hosted at UCA allowed us to identify a subset of cherts with similarities to those from Type 7 from the Middle Subbetic region (Betic Systems, south Spain, @fig-map).
+These cherts are massive, sub-translucent, peloidal/oolitic packstones.
+Allochems include very frequent peloids and ooids.
+The peloids are densely arranged in the samples, more or less visible depending on the alterations to the surface or the thickness.
+The ooids are poorly sorted, with oval or round shapes, and replaced by quartz.
+Uncommonly the ooids show a preserved nucleus.
+The presence of bioclasts varies between samples, ranging between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon), and echinoderm spines (rare).
+Similar cherts have been identified and described in previous works, describing grey cherts with peloidal and oolitic facies and occasional foraminifera [@rodriguezPrehistoricFlintExploitations2011].
+Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allow us to interpret Type 7 as coming from the Upper Jurassic formations of the Betic Systems, located in the south and southeastern Iberian Peninsula, at least 250 km east from the site (@fig-map).
+
+Type 8 (@fig-distance-types, e-f) corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment.
+It represents \~2% of the overall Terrace cherts and \~5% of the chert in the Shelter.
+Identified and well-preserved fossils are foraminifera from the Pfenderenidae family (@fig-distance-types e, f2), which are larger benthic foraminifera from marine environments [@boudagher-fadelChapterMesozoicLarger2008].
 The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous.
-Macroscopically, this group is characterized by gray colors and a white area, frequently at the center of the sample, which seems to be a weathering rind and where the presence of fossils is highlighted.
-The translucency ranges from sub-translucent to opaque and visible patterns are splotches, broad mottling, spots, and speckling.
+Macroscopically, this type is characterized by grey and white colours (@fig-distance-types e-f1).
+The translucency ranges from sub-translucent to opaque.
 Whenever present, the cortex is rounded and thin.
-Petrographically, this group is a packstone (mudstone texture close to the edges) with a massive microstructure.
+Petrographically, this type is a packstone (mudstone texture close to the edges) with a massive microstructure.
 It is composed of 97% micro-cryptocrystalline quartz, macrocrystalline quartz, and fibrous chalcedony, found mostly replacing fossils.
-Other minerals include uncommon iron oxides.
-The porosity is of vug type (3%).
-Fossil content is very frequent, albeit mostly unidentifiable and poorly preserved.
-Identified fossils are rare sponge spicules , rare ostracods with some degree of preservation, and common small, unidentified foraminifera (100 μm) of differing preservation degrees.
-Identifiable foraminifera are rare and from the Pfenderenidae family (500 μm).
+Other minerals include uncommon iron oxides and the porosity is of vug type (3%).
+Fossil content is very frequent (@fig-distance-types f2-f3), albeit mostly unidentifiable and poorly preserved.
+Identified fossils are rare sponge spicules, rare ostracods with some degree of preservation, common small, unidentified foraminifera (100 μm) of differing preservation degrees, and as aforementioned, foraminifera from the Pfenderenidae family (500 μm).
+Despite the macroscopic similarities between type 8 and cherts from the Middle Subbetic region chert formations (@fig-map), the geological reference samples showed no macroscopically visible foraminifera.
+However, our results indicate that despite no foraminifera being visible macroscopically, they are present in thin section.
+The fact that no foraminifera were identified in the small number of analysed samples from the Cadiz lithotheque may simply reflect their uncommon visibility, possibly due to patination.
+Another possible source may be the Upper Cretaceous wackestone/packstone cherts with foraminifers from the Lisbon region [@jordaoFlintSourcesMobility2022].
+However, these studies do not specify the existence of Pfenderenidae foraminifers.
+Given the lack of this specific fossil in the reference samples from the Andalusian chert \~250 km east and southeast, and the central Portugal cherts 215-250 km north, the probable source of Type 8 remains unknown.
 
-Group 13 is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
-It is characterized by two colors, white (opaque) and grey (sub-translucent), with splotches, speckles and horizontally dotted laminations.
-The cortex is from an outcrop, rounded, and varying between medium thickness to thick.
-This group shows several facies, with different types and concentrations of allochems and bioclasts, of which identifiable are peloids, ooids, and sponge spicules.
-A detailed scheme of the facies can be found in the SOM (++).
-The petrography results show this chert has a massive microstructure and packstone texture.
-It is composed mainly of micro-cryptocrystalline quartz (80%), fibrous chalcedony (10%), and micrite (10%).
-Allochems are opaques, iron oxides, ooids and peloids.
-The ooids are poorly preserved and only uncommonly show concentric lamellae in plane polarized light.
-Unlike macroscopical observations, bioclasts under the thin section are all unidentifiable, although rare fossils may be bivalve shells.
-Porosity is vug (10%).
-
-Group 13B corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone.
-It is opaque and has a heterogeneous structure and is characterized by a variable color distribution: it shows a horizontal laminated and finely laminated pattern, with gray dark gray bands intercalated with light gray bands.
-Whenever present, the cortex is rounded (from an outcrop source) and is generally thin.
+Type 9 (@fig-distance-types g) corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone.
+It corresponds to less than \~2% of the overall cherts in the Terrace and Shelter areas.
+It is opaque, has a heterogeneous structure and is characterized by a variable colour distribution: it shows a horizontal laminated and finely laminated pattern, with grey dark bands intercalated with light grey bands (@fig-distance-types g1).
+Whenever present, the cortex is rounded (FIG) and is generally thin.
 In specific samples, the cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers.
-Petrographically this chert is a banded packstone (banding of sedimentary origin), composed mostly of micro-cryptocrystalline quartz (48%), micrite (40%), dolomite (10%), and accessory fibrous chalcedony and mica (muscovite).
+Petrographically this chert is a banded packstone (banding of sedimentary origin; @fig-distance-types g2), composed mostly of micro-cryptocrystalline quartz (48%), micrite (40%), dolomite (10%), and accessory fibrous chalcedony and mica (muscovite).
 The allochems are peloids, unidentifiable bioclasts, iron oxides, and opaque minerals.
 Porosity is low (1%).
+Although its origin is unknown, when visiting the Cadiz lithotheque, we identified a small type of banded black cherts without noticeable fossils or inclusions.
+Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Type 9 belongs to Andalusian chert formations in southern Spain, 250 km east from the site (@fig-map).
 
-Group Oolitic is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
-Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: brown (opaque) and black (sub-translucent) fabric.
-The quality is acceptable since it has fine granularity and a smooth feel, but shows natural fractures and occasional inclusions, such as macrocrystalline quartz or chalcedony veins that may impact the knapping.
-The cortex is difficult to identify since it is similar to the brown fabric, highly altered, and with a gradual transition.
-Petrographically this group is a packstone with a massive microstructure.
-The chosen sample for the thin section is composed of 35% micro-cryptocrystalline quartz, 53% micrite, 9% macrocrystalline quartz, and accessory fibrous chalcedony (filling fractures) and mica; the percentages of quartz and micrite might be related to the fabrics present in the artifacts.
+Type 10 (@fig-distance-types h) is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
+It represents \~1% of the overall Terrace cherts and less than 0.1% of the chert in the Shelter.
+Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: brown (opaque) and black (sub-translucent) fabric (@fig-distance-types h1).
+The sample chosen for the thin section is composed of 35% micro-cryptocrystalline quartz, 53% micrite, 9% macrocrystalline quartz, and accessory fibrous chalcedony (filling fractures) and mica; the percentages of quartz and micrite might be related to the fabrics present in the artefacts.
 Allochems are common iron oxides and very frequent ooids.
@@
 The porosity is intraparticle and vug (10%).
-The ooids are distributed homogeneously across the sample (although clearly macroscopically visible in the brown fabric) and are highly abundant and concentrated.
-They are poorly sorted, varying between 500 and 20 μm.
-They show round-to-elliptical shapes, and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010], pp.
-144\].
+The ooids are distributed homogeneously across the sample (although macroscopically visible in the brown fabric) and are highly abundant and concentrated.
+They are poorly sorted, varying between 500 and 20 μm (@fig-distance-types h2).
+They show round-to-elliptical shapes, and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010].
+This chert shows no similarities to any consulted reference material.
+Despite the existence of oolitic limestones in the Algarve region correspondent to Middle and Upper Jurassic formations and often described as whiteish (Rocha et al., 1979), the presence of oolitic chert in the region is not recognized in geological and archaeological literature [@cardoso_sobre_2018; @rocha_noticia_1979; @verissimo_jazidas_2004] and the visit of outcrops mentioned in the geological cartography did not allow their identification.
 
-Finally, group TL is characterized by several samples with differing characteristics from the established larger groups.
-Based on their different characteristics they have been interpreted as exotic or long-distance raw materials.
-A detailed description of these types can also be found in the SOM (++).
-TL01 is a jasper, with a massive microstructure and red coloration, with frequent fractures filled with colorless, translucent fibrous chalcedony.
-TL02 is an orange chalcedony of good quality, differing from the group of white/translucent chalcedony due to its color.
-TL03 is a massive chert with uncommon spots.
-TL04 is a yellow massive chert, with inclusions and fractures filled with fibrous chalcedony.
-TL05 is a grey, banded/massive chert, where the banding seems to be of ++ origin.
-TL06 is a sub-translucent, oolitic packstone chert.
-Unlike the oolitic cherts in the previous groups, this chert shows well-sorted and small (200-220 μm), round ooids with differentiated walls and cores.
-TL07 is characterized by a laminated dendritic pattern, with uncommon inclusions.
-TL08 is a translucent and colorless, massive chert, without inclusions.
-TL09 is a translucent chalcedony with a massive microstructure and packstone texture.
-It is characterized by unidentified, moss-like orange/red inclusions, and no other allochems or bioclasts.
-TL10 is a massive chert with frequent iron oxide speckling or banding.
-TL11 is a banded, translucent chert with no inclusions.
-TL12 is a massive oolitic packstone chert, white and translucent.
-The ooids are not visible under the naked eyes, and can only be identified under the stereomicroscope.
-TL13 is a translucent, banded chert without inclusions.
-TL14 is a massive chert with large benthic foraminifera (1-2 mm).
-TL15 is a massive brown chert with common iron oxides and unidentifiable fossils.
-And finally, TL 16 is a radiolarite with unidentifiable fossils.
+Type 11 (@fig-small-types) is a banded/laminated mudstone, characterized by the presence of a conchoidal fracture (within the nodule) and a non-conchoidal, laminar fracture (at the edges of the nodule).
+This chert type is grey, opaque, and without identifiable allochems.
+This chert shows a limited presence at the archaeological site as previously mentioned, constricted to the Solutrean occupations, and corresponding to less than 0.1% of the Terrace cherts and \~2% of Shelter cherts.
 
-## Source attribution
+```{r}
+#| label: fig-small-types
+#| fig-cap: Type 11 and TL cherts macroscopic variability.
 
-Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 15, 7 and 12 to be local, as they are congruent with the described characteristics.
-Groups 2 and 15 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils.
-The macroscopic similarities of Groups 7 and 12 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, as mostly located in the Ferrel area.
-As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape.
-This is especially relevant, since the Ferrel outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiro2023].
+knitr::include_graphics(here::here("analysis/figures/small_grid.png"))
 
-The similarities between chert artifacts at the UP occupations of Vale Boi and the chert from the Rio Maior region (Estremadura Portuguesa) have been previously suggested although without systematic studies to ascertain the attribution.
+```
 
-These Cretaceous cherts, found in secondary deposition settings, are described as frequently translucent, with geodes, with colors ranging from yellow to red or grey, and mineralogically homogeneous between them.
-Petrographic studies highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matias2016] .
-When identified, sponge spicules and possible rare foraminifera have been identified [@matias2012].
-These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from central Portugal hosted at LusoLit.
+Finally, the TL types (@fig-small-types) are characterized by several samples with differing characteristics from the previously established types, and without correspondence to the local cherts and chalcedony from the Algarve region previously described [@belmiroCreatingFramesReference2023].
+These are composed mostly of blanks or retouched tools, rarely with cortex, and rarely the same TL type is found in different archaeological layers.
+A detailed description of these types can be found in the SOM (SOM).
+In both areas, the TL cherts compose less than 1% of the cherts.
 
-To aid on the comparison, we compared the lithics attributed to Type 11 to hand-samples from the LusoLit lithotheque, previously collected from the Rio Maior area.
-Similarly, three thin sections were produced of geological samples with different macroscopic characteristics from the Central Portugal area to compare with the archaeological thin sections.
-Alike the archaeological thin sections from Type 11 and congruent with other studies [@matias2016], these thin sections show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
-From these cherts, only the dark grey nodules (RT231) showed identifiable fossils which were rare ostracods.
-
-Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Group 11 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts from the Estremaduran area in central Portugal.
-
-The peloidal/oolitic packstone cherts are not congruent with the cherts from the Algarve area, making them not local or regional.
-The consultation of the lithotheque hosted at UCA allowed us to identify a sub-set of cherts with similarities to those from Type 13.
-These cherts are massive, sub-translucent, peloidal/oolitic packstones.
-Allochems include very frequent peloids and ooids.
-The peloids are densely arranged in the samples, more or less visible depending on the alterations to the surface or the thicknes.
-The ooids are poorly sorted, with oval or round shapes and replaced by quartz.
-Uncommonly the ooids show a preserved nucleus.
-The presence of bioclasts varies between samples, randing between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon) and echinoderm spines (rare).
-@fig-type13-source exemplifies several of the facies identified in Type 13 cherts from the Terrace and Shelter areas (left) and matching samples from the UCA lithotheque (right).
-Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allows us to interpret Type 13 as coming from the Andalusian region of southern Spain.
-
-Group 13B is a small group but present throughout the whole stratigraphy of the Terrace and in the Solutrean occupation of the Shelter, and with significant different characteristics from the local cherts of the Algarve region.
-Although its origin is unknown, when visiting the Cadiz lithotheque, we identified a small group of banded black cherts without noticeable fossils or inclusions.
-Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Group 13B belongs to Andalusian chert formations in southern Spain.
-
-From the identified groups, only group 11F and Oolitic remain unknown.
-Despite the macroscopic similarities between type 11F and cherts from the Andalusian chert formations, the samples inspected with a stereomicroscope showed no visible foraminifera, especially from the ++ family.
-However, the archaeological thin section revealed that despite no foraminifera being visible macroscopically, they are visible under the microscope in thin section.
-The fact that no foraminifera are visible in the the small number of analyzed samples from the Cadiz lithotheque may simply be a result of their uncommon visibility.
-Despite this, given the lack of this specific fossil in the reference samples, the probable source of Group 11F remains unknown.
-Similarly, the black oolitic chert shows no similarities to any consulted reference material.
-Despite the existence of oolitic limestones in the Algarve region, no chert nodules are mentioned (++) and the visit of outcrops mentioned in the geological cartography did not allow their identification; the trade of oolitic cherts in recent prehistory may indicate however that these types of cherts are available in the landscape in long-distance (++), and their presence in Vale Boi may simply reflect similar mobility patterns.
-
-Apart from the groups common throughout the stratigraphy of the Terrace and Shelter, group GB and samples from the TL group also show no identifiable source.
+Apart from the types common throughout the stratigraphy of the Terrace and Shelter, Type 11 and samples from the TL type also show no identifiable source.
 The exceptions are TL01, TL10-11 and TL15-16.
-Jasper is known in the SPZ area, from the Alentejo region to Spain, above the Algarve.
-Although no petrographic data is available, the macroscopic similarities between TL01 and a hand-sample hosted at Luso-lit seem to indicate that the source of this jasper is the SPZ formations.
-Similarly, despite different from the frequent Cretaceous nodules, the macroscopic similarities between TL10 and TL11 and specific nodules from Central Portugal in the Luso-lit lithotheque may indicate these cherts are from the Estremadura area.
-This difference in appearance may be due to the already mentioned variability of the nodule's macroscopic characteristics.
-Finally, TL15 and TL16 show macroscopic similarities to radiolarite reference samples located at the Cadiz lithotheque, which may indicate the source of these cherts is from Andalusian geological formations.
+Jasper is known in the South Portuguese Zone area [@oliveira_carta_1984-1], from south/central Portugal to Spain, above the Algarve (from Aljustrel to Beja; @fig-map), suggesting that the source of TL01 may be from south/central Portugal at a minimum distance of 100 km north and northeast from Vale Boi.
+Similarly, despite being different from the most frequent Cretaceous nodules, the macroscopic similarities between TL10 and TL11 and specific nodules from Central Portugal in the Lusolit lithotheque may indicate these cherts are from the Estremadura/Lisbon area (@fig-map a).
+This difference in appearance may be due to the already-mentioned variability of the nodule’s macroscopic characteristics.
+Finally, TL15 and TL16 show macroscopic similarities to radiolarite reference samples located at the Cadiz lithotheque, which may indicate the source of these cherts is from the Middle Subbetic region [@dominguezbellaMineralesRocasSociedades2010; @dominguez-bella_estudio_2006].
 
@@
 
-Chert groups are present across the stratigraphy (except for GB which appears only in level 4 and in a small number of artifacts; n=2).
-Different chert types were consistently used in the Terrace area throughout time, with different intensities of use.
-@fig-type-per-level shows the percentage of chalcedony and chert types used throughout the stratigraphy in the Terrace area of Vale Boi; this data excludes the unidentifiable and altered samples, categorized in a group named INDET.
-In the Solutrean and Proto-solutrean levels, local cherts make up more than \~70% of the chert.
-Despite this homogeneous pattern, the use of group 2 and group 7 marks significant differences between the Solutrean occupations (levels 4, 4C, and 4D) and the Proto-solutrean occupations (4E and 4E/5); in the first group the yellow/brown cherts (type 2) represent \~30% of local chert use, while in the second group, the use of type 2 diminishes drastically (\~10-15%), while Type 7 becomes the most used chert with \~38% in level 4E and \~23% in the top of level 5.
-For both occupations, chert non-congruent with the known local sources represent percentages lower than 30%.
-This pattern seems to shift in the bottom of 5 and throughout the Gravettian occupation; the use of chert local cherts drops to 45-60%, while the percentages of specific non-local chert types increase significantly, such as Type 11 and Type 13.
-Other non-local cherts such as the Oolitic cherts and TL categories also show smaller increases in the Gravettian occupation.
-As seen in previous studies, chalcedony seems to be present but always in small frequencies.
-Despite this, there is a small increase in the use of chalcedony in level 5 (both top and bottom), from lower than 5% in level 4E to 13.5% in top of 5, with a tendency to decrease in the older levels.
+### Terrace area
+
+Most chert types are consistently present across the Terrace stratigraphy (except for Type 11 which appears only in level 4 and in a small number of artefacts; n=2), although there are statistically significant differences in chert types used throughout the different levels (p-value \<2.22e-16, FIG.).
+In the Solutrean (levels 4, 4C, and 4D) and Proto-Solutrean (levels 4E and 4E/5) occupations, local cherts make up more than \~70% of the chert and 80% when considering the weight.
+Despite this homogeneous pattern, local types show statistically significant differences between the two technocomplexes.
+In the Solutrean levels, the yellow/brown cherts (Type 2) are present in the highest percentage, at \~30%.
+In contrast, the grey/white cherts (Type 4) become the most used chert with percentages above 35% in level 4E and \~20% in the top of level 5.
+For both technocomplexes, chert non-congruent with the known local sources represent percentages lower than 30%.
+This pattern seems to shift at the start of the Proto-Solutrean horizon (level 5) and the Gravettian occupations (levels 6 and 7); in these levels, the percentages of local cherts drop to 45-60%, while the percentages of specific non-local chert types increase significantly, such as Type 6 (increasing from \~10% during the Solutrean to 30%).
+However, when considering the sum of the weight, the percentages of Type 6 diminish, instead representing around 20% of the chert use, and bringing in level 6 the use of local chert to 50% in comparison to chert types of non-local sources.
 
@@
 #| label: fig-type-per-level
-#| fig-cap: "Chalcedony and chert types per level, without unidentifiable samples. Only percentages superior to the 5% threshold are presented in the figure."
-#| echo: false
+#| fig-cap: Terrace area chalcedony and chert type frequency (a) and weight sum (b) per level, without unidentifiable samples. Only percentages superior to the 5% threshold are shown in the figure.
+#| fig-width: 10
 
-phase_dataset %>%
+count_table <- phase_dataset %>%
   filter(RMTYPE != "INDET") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   dplyr::group_by(Level, RMTYPE) %>% 
@@
   mutate(percent = n/sum(n)) %>% 
-  ggplot(aes(x=Level, y = percent, fill = RMTYPE)) +
+  mutate(Technocomplex = case_when(
+    Level == 4 | Level == "4C" | Level == "4D" ~ "Solutrean",
+    Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
+    Level == 6 | Level == 7 ~ "Gravettian")) %>% 
+  mutate(Technocomplex = factor(Technocomplex, levels = c("Solutrean", "Proto-Solutrean", "Gravettian")))
+
+count_plot <- ggplot(count_table, aes(x=Level,y=percent, fill = RMTYPE)) +
   geom_bar(stat = "identity", width = 0.5) +
+  facet_grid(.~Technocomplex, scales = "free_x") +
   scale_y_continuous(labels = scales::percent_format()) +
-  theme_minimal() + #type of theme
-  labs(x = "Level", y = NULL) +# change labels
+  #theme_minimal() + #type of theme
+  labs(y = "Count %") +# change labels
   guides(fill=guide_legend(title=NULL)) + # remove legend title
@@
                 position=position_stack(vjust=0.5), colour="white", size=3) +
-  theme(axis.title.x = element_text(size = 10, vjust = -1)) %>% 
+  theme(axis.title.x = element_blank()) +
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
@@
 
-T_dataset %>%
+weight_table <- phase_dataset %>%
+  filter(RMTYPE != "INDET" & WEIGHT != "0") %>% 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  dplyr::group_by(Level, RMTYPE) %>% 
+  dplyr::summarise(wsum = sum(WEIGHT)) %>% 
+  mutate(percent = wsum/sum(wsum)) %>% 
+  mutate(Technocomplex = case_when(
+    Level == 4 | Level == "4C" | Level == "4D" ~ "Solutrean",
+    Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
+    Level == 6 | Level == 7 ~ "Gravettian")) %>% 
+  mutate(Technocomplex = factor(Technocomplex, levels = c("Solutrean", "Proto-Solutrean", "Gravettian")))
+
+  weight_plot <- ggplot(weight_table, aes(x = Level, y = percent, fill = RMTYPE)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  facet_grid(.~Technocomplex, scales = "free_x") +
+  scale_y_continuous(labels = scales::percent_format()) +
+  #theme_minimal() + #type of theme
+  labs(x = "Level", y = "Weight %") +# change labels
+  guides(fill=guide_legend(title=NULL)) + # remove legend title
+  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
+                position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_blank()) + 
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+
+ggarrange(count_plot, weight_plot, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
+
+```
+
+```{r}
+#| label: chisq
+#| include: false
+
+library(corrplot)
+
+chisq <- phase_dataset %>% 
+  filter(RMTYPE != "INDET") %>% 
+  filter(RMTYPE != "TYPE 11")
+
+chisq_tab <- table(chisq$RMTYPE, chisq$Level)
+chisq_matrix <- as.table(as.matrix(chisq_tab))
+chisq_rm <- chisq.test(chisq_tab)
+round(chisq_rm$residuals, 3)
+
+corrplot(chisq_rm$residuals, method = "color", is.corr = FALSE)
+
+mosaicplot(chisq_tab, shade = TRUE, las = 2, main = "RMTYPE")
+
+library(vcd)
+vcd_labels <- list(RMTYPE = c("TL", "1", "10", "2", "3", "4",
+                                   "5", "6", "7", "8", "9"))
+mosaic(~ Level + RMTYPE,
+      gp = shading_max,
+      data = chisq, 
+      set_labels = vcd_labels,
+      rot_labels = c(0, 0))
+
+```
+
+```{r}
+#| label: rm-tables
+#| include:  false
+
+# Tables f Terrace and Shelter with n(%) of chert types both overall by area and by level. Perhaps going to SOM (?)
+terrace_table <- phase_dataset %>%
+  filter(RMTYPE != "INDET") %>% 
+  select(RMTYPE, Level) %>% 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1", "TYPE 2", "TYPE 3", "TYPE 4", "TYPE 5", "TYPE 6",
+                                            "TYPE 7", "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11","TL"))) %>%
+  tbl_summary(include = c(RMTYPE), by = Level) %>% 
+  add_overall()
+
+S_dataset["Layer"][is.na(S_dataset["Layer"])] <- "No info"
+shelter_table <- S_dataset %>% 
+  filter(RMTYPE != "INDET") %>%
+  filter(Layer != "D") %>% 
+  select(RMTYPE, Layer) %>% 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1", "TYPE 2", "TYPE 3", "TYPE 4", "TYPE 5", "TYPE 6",
+                                            "TYPE 7", "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11","TL"))) %>%
+  tbl_summary(include = c(RMTYPE), by = Layer) %>% 
+  add_overall()
+
+```
+
+```{r}
+#| label: extra-graph
+#| include: false
+
+# Violin plot for RMtype per depth
+phase_dataset %>%
   filter(Z != "0") %>% 
   filter(!Z >= "25") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   filter(RMTYPE != "INDET") %>% 
-  ggplot(aes(RMTYPE, Z, color = RMTYPE)) +
-  geom_violin(trim = FALSE, scale="width", bw=0.1) + 
-  labs(title="Violin plot", 
-       subtitle="City Mileage vs Class of vehicle",
+  ggplot(aes(RMTYPE, Z, fill = RMTYPE)) +
+  geom_violin(trim = FALSE, scale="width", adjust = .5) + 
+  geom_point(position = position_jitter(seed = 1, width = 0.2), alpha = 0.1, color = "darkorange") +
+  geom_hline(yintercept = c(24.5,24.3,24.1,23.8,23.45), colour="#BB0000", linetype="dashed") +
+  labs(title="Groups per depth", 
+       subtitle="Chert groups in the Terrace area by depth",
        caption="Source: mpg",
-       x="Class of Vehicle",
-       y="City Mileage") +
-  stat_summary(fun.data=mean_sdl, geom="errorbar", color="black", width=0.2)
-  ggsave("violin_plot.png")
-
-T_dataset %>% 
-  filter(Z != "0") %>% 
-  filter(!Z >= "25") %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","GB", "Oolitic", "TL"))) %>%
-  dplyr::group_by(Z, RMTYPE) %>% 
-  dplyr::summarise(n = n()) %>%
-  ggplot(aes(x = n, y = Z)) +
-  geom_point(alpha = 0.1, color = "coral") +
-  facet_grid(cols = vars(RMTYPE)) +
-  labs(x = "Count", y = "Depth (cm)") +
+       x="",
+       y="Depth") +
+  stat_summary(fun.data=mean_sdl, geom="errorbar", color="black", width=0.2) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
   theme(axis.text.x=element_blank(),
@@
 
-The same raw materials found in the Terrace area are also found in the Solutrean occupation of the Shelter, albeit with some differences.
-Although with similar percentages to the local cherts, the use of local chert in the Shelter area is heavily dominated by the brown/yellow cherts that frequently characterize the Lower Jurassic outcrops of the Algarve; although the other macroscopic types are present but not significant.
-The presence of cherts not congruent with the local sources is, alike level 4, close to 25%, although dominated mostly by Type 11.
-Type GB is present in higher numbers than in level 4, however still representing less than 5% of the assemblage.
+When plotting the weight, it is apparent that most of the samples are located under the 0.3 kg mark, with median and bottom quartiles often falling under the 0.5 kg line (@fig-weight).
+Local types, such as Type 2, Type 4, and Type 5, although mostly represented by artefacts with a small mass (\~0.3 kg), show the heaviest and highest number of heavy artefacts.
+In comparison, other types, although showing some variation in the weight of artefacts, show smaller artefacts and their means and medians are frequently closer to the bottom quartile.
+As seen in Figure 3 B, non-local types show larger masses from level 5 to level 7.
 
 ```{r}
-#| label: fig-type-per-level_shelter
-#| fig-cap: "Chalcedony and chert types per level, without unidentifiable samples. Only percentages superior to the 5% threshold are presented in the figure."
-#| echo: false
+#| label: fig-weight
+#| fig-cap: "Terrace boxplots of chert type weight per level."
+#| #| fig-width: 10
 
-S_dataset %>%
-  filter(RMTYPE != "INDET") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
-                                        "TYPE 12", "TYPE 13","TYPE 13B", 
-                                        "TYPE 11", "TYPE 11F","TYPE GB", "Oolitic", "TL"))) %>%
-  dplyr::group_by(AREA, RMTYPE) %>% 
+# Weight boxplots terrace
+phase_dataset %>% 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(Technocomplex = factor(Technocomplex, levels = c("Solutrean", "Proto-Solutrean", "Gravettian"))) %>% 
+  ggplot(aes(RMTYPE, WEIGHT, color = RMTYPE)) +
+  geom_boxplot() +
+  facet_wrap(~Technocomplex + Level) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Weight") +
+  coord_cartesian(ylim =  c(0, 40)) +
+  scale_color_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+
+```
+
+### Shelter area
+
+The same raw materials found in the Terrace area are also found in the Solutrean occupation of the Shelter, albeit with some differences (@fig-analysis_shelter).
+The local cherts make up between \~70-80% of the sample and are mainly composed of the brown/yellow cherts (Type 2) that frequently characterize the Lower Jurassic outcrops of the Algarve (@fig-analysis_shelter a-b).
+The presence of cherts not congruent with the local sources is, like level 4 in the Terrace, close to 25%, although dominated mostly by Type 6 (\~15%).
+Similarly to the Solutrean occupations of the Terrace, although the artefacts are generally small (lower weights), the local types show a higher variability in mass, with several outliers and heavy samples when compared to the non-local types (@fig-analysis_shelter c).
+
+```{r}
+#| label: fig-analysis_shelter
+#| fig-cap: "a-b) Shelter area chalcedony and chert type frequency (a) and weight sum (b) per level, without unidentifiable samples.
+Only percentages superior to the 5% threshold are shown in the figure.
+c): Shelter boxplots of chert type weight per level."
+#| #| fig-width: 10
+
+# Type per level plots fr Shelter
+## Preparing db for the barplots
+s_data_plots <- S_dataset %>% # sorting missing values and filtering non-solutrean layers
+  filter(RMTYPE != "INDET" & WEIGHT != "0") %>% 
+  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
+  filter(CODE != "B" & Layer != "D") %>% #removing buckets (B)
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL")))
+
+## Count barplot
+shelter_count <- s_data_plots %>%
+  dplyr::group_by(Layer, RMTYPE) %>% 
   dplyr::summarise(n = n()) %>%
   mutate(percent = n/sum(n)) %>% 
-  ggplot(aes(x = AREA, y = percent, fill = RMTYPE)) +
+  ggplot(aes(x = Layer, y = percent, fill = RMTYPE)) +
   geom_bar(stat = "identity", width = 0.5) +
   scale_y_continuous(labels = scales::percent_format()) +
-  theme_minimal() + #type of theme
-  labs(x = "Level", y = NULL) +# change labels
+  #theme_minimal() + #type of theme
+  labs(y = "Count %") +# change labels
   guides(fill=guide_legend(title=NULL)) + # remove legend title
@@
                 position=position_stack(vjust=0.5), colour="white", size=3) +
-  theme(axis.title.x = element_text(size = 10, vjust = -1)) %>% 
+  theme(axis.title.x = element_blank()) + 
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
@@
                               "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+## Weight barplot
+shelter_weight <- s_data_plots %>%
+  dplyr::group_by(Layer, RMTYPE) %>% 
+  dplyr::summarise(wsum = sum(WEIGHT)) %>% 
+  mutate(percent = wsum/sum(wsum)) %>% 
+  ggplot(aes(x = Layer, y = percent, fill = RMTYPE)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  #theme_minimal() + #type of theme
+  labs(y = "Weight %") +# change labels
+  guides(fill=guide_legend(title=NULL)) + # remove legend title
+  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
+                position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_blank()) + 
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+## Weight boxplot
+shelter_box <- s_data_plots %>% 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","TL") & Layer != "ND") %>%
+  ggplot(aes(RMTYPE, WEIGHT, color = RMTYPE)) +
+  geom_boxplot() +
+  facet_wrap(.~Layer, ncol = 1) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y = "Weight") +
+  #geom_jitter(shape=16, position=position_jitter(0.2)) +
+  coord_cartesian(ylim =  c(0, 40)) +
+  scale_color_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), legend.position = "none", axis.title.x = element_blank())
 
-```
-
-```{r}
-#| label: fig-type13-source
-#| fig-cap: Comparison between chert samples of different facies from Type 13 of the Terrace area (left) and samples from the Universidad de Cádiz lithotheque.
-
-knitr::include_graphics(here::here("analysis/figures/Terrace_UCA_comparison.png"))
-
-```
-
-```{r}
-#| label: tbl-TL
-#| tbl-cap: A table of TL
-
-TL_table_T <- phase_dataset %>% 
-  filter(RMTYPE=="TL") %>% 
-  select("Level", "SUBTYPE")
-TL_table_S <- S_dataset %>% 
-  filter(RMTYPE=="TL") %>% 
-  mutate_at("LAYER", ~replace_na(.,"Shelter")) %>% 
-  select("LAYER", "SUBTYPE") %>% 
-  rename(Level = LAYER)
-
-TL_table <- rbind(TL_table_T, TL_table_S)
-TL_table <- TL_table %>% 
-  tabyl(SUBTYPE, Level) %>% 
-  adorn_totals(c("row","col"))
+## Arrange plots in a single figure
+gg1 <- ggarrange(shelter_count, shelter_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend="bottom")
+ggarrange(gg1, shelter_box, labels = c(" ", "c", ncol = 2, common.legend = TRUE, legend = "bottom"))
 
@@
 
-# Conclusion
+The macroscopic and petrographic analysis of chert and chalcedony lithic materials from the UP levels of the Terrace and Shelter areas at Vale Boi yield essential insights into the procurement of raw materials and the mobility patterns of hunter-gatherers in southwestern Iberian Peninsula.
+These materials serve as valuable indicators for identifying changes in mobility, as they reveal patterns of both local and non-local procurement.
+The dominance of local raw materials in all assemblages indicates a heavy reliance on nearby resources.
+However, the observed variation in the frequencies of local types may reflect differences in mobility strategies or preferences for certain nodules.
+The detection of non-local raw materials throughout the stratigraphy points to a persistent use of long-distance chert sources.
+This continuity could be attributed to a tradition of shared knowledge and culture regarding the locations of raw materials and other resources within the landscape, or the preservation of social networks over time.
+@cascalheira_2017 acknowledge the adoption of specific behavioural strategies by Vale Boi hunter-gatherers in response to the regional ecological conditions, which were sustained throughout the UP at the site.
+Practices such as grease rendering or the functional specialization of raw materials are seen as deliberate actions with significant cultural implications for future generations, influencing the adaptive choices made over time.
+Accordingly, chert raw materials appear to be integral to these behavioural strategies, established from the site's earliest occupations and perpetuated throughout the UP.
+
+The local chalcedony and chert indicate that the groups were sourcing a significant portion of their raw materials within approximately a 20 km radius of the site.
+The direct procurement of these materials is plausible, given that known chalcedony sources are situated to the west, about 6-7 km from Vale Boi, and chert sources are found to the east, within the inland Lower Jurassic outcrops, roughly 8 km away (@fig-map b).
+This hypothesis is supported by the potential existence of UP occupations near these outcrops, which might reflect quarrying and chert catchment activities during prehistory [@zambujo_jazidas_1998].
+Evidence of short-distance procurement includes the presence of heavy nodules and large chunks of chert from local varieties with parent rock still attached, as well as a significant number of outliers (representing larger blocks and chunks) in the weight boxplots across the stratigraphy (Figure 4).
+These may have been discarded due to natural fractures.
+It's also possible that local cherts were procured from farther distances, perhaps as part of embedded activities.
+The other chert sources from the Lower and Upper Jurassic periods in the western Algarve, located along the beach and cliff areas as previously mentioned, are between 15-18 km southwest of Vale Boi (@fig-map b).
+While our current data do not pinpoint which specific outcrops were exploited based on macroscopic and petrographic analysis, the discovery of marine resources such as shells at the site indicates that these groups visited the coast [@belmiro_gravettian-solutrean_2021; @mannePryingNewMeaning2011].
+This suggests that raw materials could have been conveniently acquired in coastal vicinities without significant detours [@binfordOrganizationFormationProcesses1979; @surovellBehavioralEcologyLithic2009]
+
+However, the discussion around local cherts becomes more complex due to the presence of similar outcrops in the eastern Algarve.
+Previous research has indicated that macroscopic criteria alone are insufficient to determine whether the cherts used originated from Lower Jurassic sources in the western Algarve (within approximately a 20 km radius of the site) or from Middle Jurassic sources approximately 80 km to the east (@fig-map c).
+This is because the cherts are similar both macroscopically and petrographically [@belmiroCreatingFramesReference2023].
+A similar issue is noted by Surovell [-@surovellBehavioralEcologyLithic2009], who described a scenario at the late Pleistocene archaeological site of Krmpotich (USA), where a specific type of chert found at the site could be sourced from both local and long-distance outcrops.
+Surovell suggested that some of these cherts might have been imported from distant sources or previous campsites.
+Consequently, we cannot rule out the possibility that some of the cherts we have identified as being sourced within a 20 km radius of Vale Boi could actually be the result of long-distance procurement and regional mobility, potentially from areas approximately 80-100 km away where Middle Jurassic cherts are located.
+Given the similarity in the quality of the raw materials and the absence of geochemical studies, and in light of the study and model mentioned earlier, we surmise that most of these raw materials likely originate from local sources (within 20 km), although a small portion may be attributable to regional movements.
+
+Similarly, Upper Jurassic cherts located about 90 km east of Vale Boi (@fig-map c) have not been detected in the assemblages.
+This absence could be attributed to the limited use and transport of these nodules from their sources to Vale Boi, potentially due to the small size of the nodules (often between 3-5 cm), the difficulty of detaching them from the parent rock, or challenges related to their accessibility and visibility, given that the known outcrops are situated inland, approximately 10 km from the present coastline.
+
+While the regional Upper Jurassic cherts situated less than 100 km east of the site were not identified within the assemblages, the allochthonous cherts with potentially identified sources seem to originate from greater distances and in two main directions: 1) from the north, where Cretaceous cherts from Central Portugal are found (@fig-map a), along with a small amount of jasper (TL01); and 2) from the east, where peloidal/oolitic chert from western Andalusia is present (@fig-map), and possibly jasper, transported by the Guadiana river at the current border between Portugal and Spain.
+The scarcity of pebble cortex and the predominance of rounded cortex suggest that these cherts were likely not collected from beaches or riverbeds far from the outcrops but were possibly gathered closer to the source areas.
+These identified axes of long-distance movement to the north and east could be influenced by the topography, affecting mobility patterns.
+Indeed, formal models that seek to explain hunter-gatherer mobility have highlighted the significance of factors such as topography on the distances travelled by groups or individuals [@brantinghamMeasuringForagerMobility2006].
+For instance, Quíros Castillo [-@quiroscastilloArchaeologyHistoryPeasantries2020] discusses how Copper Age groups mitigated subsistence risks by establishing exchange networks and engaging in regional mobility, as evidenced by the trade of allochthonous raw materials, despite the high energy costs and risks associated with these exchanges.
+Nonetheless, the author notes certain characteristics of the Iberian Peninsula that facilitate mobility, such as the low cost of traversing the plateaus and the interconnectivity of the Guadalquivir River valley with the Guadiana and Tagus rivers.
+Thus, the observed long-distance mobility patterns in the north/east directions by hunter-gatherers at Vale Boi might reflect the exploitation of these natural corridors for movement, explaining the substantial use of allochthonous chert.
+The presence of a smaller subset of allochthonous cherts of unknown origin (TL) could represent different forms of long-distance trade of finished goods and blanks, or interactions with various social networks, potentially even indicating individual networks and mobility strategies [sensu @gamblePalaeolithicSocietiesEurope1999].
+
+The significant presence of allochthonous cherts during the Gravettian occupation at Vale Boi suggests extensive communication and mobility among the groups residing there, indicating an expansive social network.
+This finding diverges from the anticipated patterns.
+In the context of the Solutrean occupations, one would expect a high frequency of non-local raw materials due to the broadening of social networks.
+In contrast, the Gravettian period, especially in light of the technological findings from southern Portugal, would typically be associated with a reduced reliance on non-local raw materials.
+
+Technological analyses of lithic assemblages have unveiled notable differences between Gravettian assemblages from central and southern Portugal, pointing to the existence of two distinct regional Gravettian traditions in southern Iberia: the Vicentine Gravettian at Vale Boi, and the Mediterranean Gravettian identified in UP sites in southern Spain [@marreirosLithicTechnologyGravettian2015; @marreirosLithicTechnologyVariability2013].
+These traditions reflect adaptations to the unique local and regional ecological settings [@bradtmoller_lithic_2016] and evolving environmental conditions [@marreirosLithicTechnologyVariability2013], potentially fostering distinct ethnographic divisions among hunter-gatherer communities [@marreirosLithicTechnologyGravettian2015].
+
+This analysis suggests that lithic technological traits may delineate different coexistence boundaries, confined to specific regional bands and aligned with the ecological dynamics of southwestern Iberia.
+Consequently, one might expect territorial or cultural demarcations to influence material usage patterns, with a shift towards a greater reliance on local resources and a diminished use of non-local raw materials during the Gravettian.
+However, the observed prevalence of allochthonous materials could reflect not merely long-distance movements within a single cultural framework but rather the result of interactions and exchanges between different groups.
+Such exchanges, focused on securing raw materials not locally available, underscore the complexity of prehistoric social networks.
+
+This perspective aligns with Brantingham's [-@brantinghamMeasuringForagerMobility2006] discussion on social exchange, which posits that exchanges likely occurred between groups occupying distinct territories, who would meet regularly to trade raw materials.
+This model helps explain the presence of materials from distant sources at archaeological sites, illustrating the nuanced and interconnected nature of prehistoric social landscapes.
+
+An alternative interpretation of the varying frequencies of non-local cherts during different periods at Vale Boi can be informed by the "Mean per Capita Occupation Span" model proposed by Surovell [-@surovellBehavioralEcologyLithic2009] and the "Distance/Frequency Index (DFI)" outlined by Grove et al.
+[@groveMovingFarMoving2023].
+These models hypothesize that shorter-term occupations are likely to show a higher proportion of non-local artefacts, whereas with longer-term settlements, the prevalence of local artefacts increases.
+This pattern may be attributed to environmental influences: challenging conditions such as fragmented habitats and cooler temperatures limit mobility, making long-term settlement in areas rich in local resources a preferable, lower-risk option [@surovellBehavioralEcologyLithic2009].
+This pattern is mirrored in the archaeological record of Vale Boi, where, aside from the earlier Gravettian and Magdalenian phases, subsequent occupations seem to adopt a logistical-settlement strategy, positioning Vale Boi as a long-term basecamp [@cascalheira_2017].
+
+In this context, the Gravettian phase at Vale Boi could reflect a period of enhanced mobility and brief occupations, possibly interspersed with reoccupations, indicative of a significant use of non-local raw materials.
+This scenario parallels observations from the Late Gravettian in the French Massif Central, where hunter-gatherer groups are known to have transported entire or pre-formed blocks of raw material over considerable distances for short-term, hunting-related site occupations [@delvigneGeoresourcesTechnoculturalExpressions2019].
+However, the shift toward colder and drier climatic conditions marked by Heinrich Event 2 during the Proto-Solutrean, followed by the Last Glacial Maximum in the Solutrean period [@cascalheiraHunterGathererEcodynamics2013; @gonzalez-samperiz_steppes_2010-1; @sanchez_goni_millennial-scale_2010; @schmidt_rapid_2012; @belmiro_gravettian-solutrean_2021], resulted in a more challenging environment with sparser vegetation [@cascalheiraHunterGathererEcodynamics2013; @gonzalez-samperiz_steppes_2010-1].
+These changes likely constrained the mobility of hunter-gatherers, fostering longer-term settlements within the resource-rich niche of southwestern Portugal [@cascalheira_2017; @schmidt_rapid_2012].
+Consequently, as settlements extended in duration, the reliance on non-local raw materials, whether sourced directly or acquired through trade, diminished in favour of increasingly dominant local raw materials.
+
+# Conclusions
+
+Our findings build upon earlier studies indicating that the majority of chert utilized at Vale Boi originates from local Lower Jurassic sources within a 20 km radius of the site, with non-local materials constituting a minor portion of the chert used [@pereira_territory_2016-1].
+The hypothesis of long-distance procurement of Cretaceous cherts from central Portugal, situated north of Vale Boi, is corroborated by visual and petrographic evidence [@pereira_territory_2016-1; @bicho_ecodynamics_2013], and this pattern persists throughout the UP.
+Furthermore, the identification of cherts bearing macroscopic resemblance to the peloidal/oolitic Upper Jurassic cherts from the Betic System in southern Spain confirms a west-east mobility axis, previously inferred from techno-typological studies, spanning the entirety of the UP period.
+
+These findings reveal a landscape of hunter-gatherers engaged in complex practices of landscape exploration, resource management, and possibly risk management.
+The presence of allochthonous raw materials, albeit in small quantities during the Proto-Solutrean and Solutrean phases, underscores that the inhabitants of Vale Boi—situated on the periphery of the Iberian Peninsula and Europe, bordered by the sea to the south and west—were not isolated.
+Instead, they were part of an extensive social network, a pattern of interconnectivity that resonates with findings across western Europe.
+In southern Portugal, this interconnection among UP hunter-gatherers from diverse and distant regions of the Iberian Peninsula is similarly observed.
+
+The outcomes of this research pave the way for continued exploration of these assemblages through diverse methodologies.
+Future directions include employing a geochemical approach to both the studied assemblages and the geological reference collection, aiming to enrich discussions on mobility throughout the UP at Vale Boi.
+This approach is anticipated to provide diagnostic elements to distinguish between cherts from Lower and Middle Jurassic units, thereby offering a comprehensive view of regional resource utilization and mobility patterns.
+
+Additionally, experimental replication and the creation of a reference collection modified through heat treatment could enhance the identification of raw material types and currently indistinguishable samples, as noted in prior research [@pereira_territory_2016-1].
+Lastly, we intend to integrate technological data from current and ongoing datasets to delve deeper into the usage and management of raw materials (both local and non-local) over time, their association with different site uses, and to contextualize these practices within the broader framework of human adaptations to changing environmental conditions throughout the UP.
 
 # Acknowledgements
+
+We would like to thank the Fundacão para a Ciência e Tecnologia for the funding essential for our research, and ICArEHB for funding and providing access to the archaeological materials.
+We would also like to thank Dr.
+David Ortega for the provided insight and expertise that greatly assisted the research and Dr.
+Célia Gonçalves for the invaluable assistance with the maps and imagery created for this paper.
+
+# Declarations
+
+## Funding
+
+This study was supported by the Fundacão para a Ciência e Tecnologia (https://www.fct.pt/), attributed to Joana Belmiro with the contract reference 2020.08722.BD (DOI: 10.54499/2020.08722.BD), and João Cascalheira with the contract reference DL57/2016/CP1361/CT0026.
+This study was also supported by the Seed Fund’21 attributed to Joana Belmiro, by ICArEHB – The interdisciplinary Center for Archaeology and Evolution of Human Behaviour, funded by the Fundacão para a Ciência e Tecnologia (https://www.fct.pt/) under program UIDP/04211/2020.
+Open Access was funded by Fundacão para a Ciência e Tecnologia (https://www.fct.pt/) and Universidade do Algarve.
+
+## Competing Interests
+
+The authors have no competing interests to declare that are relevant to the content of this article.
+
+## Data availability
+
+The authors confirm that all data generated and analysed during this study are available to the public in the Open Science Framework repository at: ++.
+
+## Author Contributions
+
+Study conception and design was created by Joana Belmiro, João Cascalheira and Xavier Terradas.
+Material preparation, data collection and analysis were performed by Joana Belmiro.
+The first draft of the manuscript was written by Joana Belmiro and all authors commented on previous versions of the manuscript.
+All authors read and approved the final manuscript.
+
+# Supplementary Information
+
+**Online Resource 1**: Data dictionaries for macroscopic (group and individual) and petrographic analyses.
+
+**Online Resource 2**: Terrace and shelter lithotypes complete macroscopic and petrographic description.
+
+**Online resource 3**: Summarised type macroscopic and petrographic features and epoch/stage identification.
 
@@
 # what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+#if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
 ```

### 2024-04-26 analysis/supplementary-materials/SOM_datadictionaries.qmd
Pre submission commit_1

@@
 ---
-title: "Data dictionaries"
+title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia, Journal of Paleolithic Archaeology"
+subtitle: "Online Resource 1: Data dictionaries for macroscopic and petrographic analyses"
 author:
-  - Jane Doe:
+  - Joana Belmiro:
+      orcid: https://orcid.org/0000-0003-4951-2907
       correspondence: "yes"
-      email: janedoe@fosg.org
-      orcid: 0000-0003-1689-0557
+      email: jfbelmiro@ualg.pt
       institute:
-        - fosg
-        - fop
-  - John Q. Doe:
-      institute: fosg
-      orcid: 0000-0003-1689-0558
-  - Peder Ås:
-      institute: fosg
-      orcid: 0000-0003-1689-0559
-  - Juan Pérez:
-      orcid: 0000-0003-1689-0551
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
+  - Xavier Terradas:
+      orcid: https://orcid.org/0000-0002-8000-5607
       institute: 
-        - name: Acme Corporation
-  - Max Mustermann:
-      orcid: 0000-0003-1689-0552
+        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
+  - Salvador Dominguez-Bella:
+      orcid: https://orcid.org/0000-0003-3892-763X
+      institute: 
+        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
+  - João Cascalheira:
+      orcid: https://orcid.org/0000-0003-0321-8892
+      institute:
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
 title-block-published: "Last updated"  
 format: 
-  pdf:
+  docx:
     classoption: landscape
-    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
+    reference-doc: "../templates/template_landscape.docx" # Insert path for the DOCX file
 execute:
@@
 bibliography: ../paper/references.bib
+toc: true
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
@@
 
-# 1. Data dictionaries
+# Macroscopic type characterization
 
-## 1.1. Group characterization
+```{r}
+#| label: tbl-macrogroup
+#| tbl-cap: Data dictionary for macroscopic characterization of the archaeological chert types and sub-types.
 
-## 1.2. Single characterization
+tibble::tribble(
+  ~Variable, ~"Allowed values", ~Description, ~Reference,
+  "Type", "-", "Chert type ID.", "-",
+  "Sub-type", "-", "Chert subtype ID.", "-",
+  "Color", "-", "Color according to the Munsell Soil Color Chart.", "-",
+  "Fabric", "Homogeneous, Heterogeneous", "Distribution of color, luster, translucency, feel and pattern in the sample.
+**Heterogeneous**: presence of different characteristics within one single sample.", "@luedtke_archaeologists_1992",
+  "Luster", "Shiny, Medium, Dull", "The character of light reflected by minerals.
+**Shiny**: very intense luster; **Medium**: some luster but it is not very intense; **Dull**: no luster.", "@crandell_macroscopic_2005-1 and @luedtke_archaeologists_1992",
+  "Translucency - qualitative", "Highly translucent, Translucent, Sub-translucent, Opaque", "Degree to which the light can penetrate a material when held 5cm from a 40w lamp.
+**Highly translucent**: transparent or near transparent; **Translucent**: silhouettes can be seen through the sample’s thin part and/or light passes through the thick parts; **Sub-translucent**: light only passes through the thin parts; **Opaque**: no or almost no light passes through.", "@bressyCaracterisationGestionSilex2002 and @crandell_macroscopic_2005-1",
+  "Translucency - quantitative", "-", "Measurement in mm of the thickness where light begins to pass through in translucent to sub-translucent samples.", "-",
+  "Feel", "Smooth, Semi-smooth, Rough", "A fingernail is dragged across the sample’s surface.
+**Smooth**: nothing can be felt; **Semi-smooth**: slight rough feel; **Rough**: if the sample is distinctly rough.", "@bressyCaracterisationGestionSilex2002 and @crandell_macroscopic_2005-1",
+  "Pattern", "Shaded, Spotted, Splotched, Broad mottling, Marbled mottling, Speckling, Flecks, Banded, Streaked, Laminated, Finely laminated", "Type of pattern present on the sample.
+**Shaded**: gradient color or texture change; **Spotted**: circle patterns less than 30% of the surface area; **Splotched**: irregular shapes less than 30% of the surface area; **Broad mottling**: large, irregular blotching, covering more than 30% of the surface; **Marbled mottling**: large relatively round shapes, covering more than 30% of the surface; **Speckling**: small dots, well distributed over the surface; **Flecks**: small dots, grouped together; **Banded**: regular lines thicker than 1 cm; **Streaked**: less regular and wider banding; **Laminated**: lines with thickness inferior to 1 cm; **Finely laminated**: series of lines with thickness inferior to 1 mm.", "@crandell_macroscopic_2005-1",
+  "Cortex type", "Outcrop, Rounded, Pebble, Indeterminate", "**Outcrop**: uneven and low level of alteration to the original cortical surface; **Rounded**: homogeneous and rounded cortex, with possible slight alteration to the original cortical surface; **Pebble**: Smooth and highly weathered surface, with high degree of alteration to the cortical surface.", "-",
+  "Cortex maximum thickness", "-", "Measurement in mm of the thickest cortex area.", "-",
+  "Cortex minimum thickness", "-", "Measurement in mm of the thinnest cortex area.", "-",
+  "Cortex transition", "Sharp, Gradual", "**Sharp**: transition between cortex and chert is sharp, with an obvious change in colors and textures; **Gradual**: cortex and chert show a gradual transition with gradient colors and textures", "-",
+  "Allochems", "-", "Macroscopically visible allochems (excluding bioclasts).", "-",
+  "Bioclasts", "-", "Macroscopically visible bioclasts.", "-",
+  "Fracture", "Conchoidal, Uneven", "Presence or absence of a conchoidal fracture.", "-",
+  "Quality", "Good, Acceptable, Low", "Combination of: 1) homogeneity, understood as the lack of fractures, flaws and cracks; 2) fine granularity and smooth feel; 3) lack of inclusions that due to their nature or distribution impact the knapping.", "@luedtke_chert_1978 and @brandl_genesis_2013"
+) |>
+  knitr::kable()
 
-## 1.3. Petrographic characterization
+```
 
-# 2. References
+# Macroscopic individual analysis
+
+```{r}
+#| label: tbl-macroind
+#| tbl-cap: Data dictionary for macroscopic analysis of archaeological chert artifacts.
+
+tibble::tribble(
+  ~Variable, ~"Allowed values", ~Description, ~Reference,
+  "Area", "Terrace, Shelter", "Archaeological artifact provenance area from Vale Boi.", "-",
+  "Level", "-", "Individual artefact archaeological level.", "-",
+  "ID", "-", "Individual artefact ID.", "-",
+  "Type", "-", "Type of chert characterized with macroscopic and microscopic methods.", "-",
+  "Sub-type", "-", "Subtype of chert characterized with macroscopic and microscopic methods.", "-",
+  "Weight", "-", "Weight of individual artefacts measured in gr.", "-",
+  "Alterations", "Cracks, Potlids, Reddening, Gloss, White patina, Patina, Other", "**Cracks**: surface shrinkage cracks or crazing, described as the result of contact to high temperatures; **Potlids**: small, circular and concave negative where a fragment appears to have popped off; **Reddening**: changes to the surface of the chert from the original color to a reddish/brown color; **Gloss**: glossy luster which differs from the original chert's luster, present in all the artifacts or in specific removals; **White patina**: film, rind or layer produced on the surface of the chert visible as a color change from the original color to a white color; **Patina**: film, rind or layer produced on the surface of the chert visible as a color change from the original color (excluding white patina); **Other**: other types of alterations to the original appearance of chert, such as rolled edges or pits.", "@hurst_patination_1961; @glauberman_flint_2012; @luedtke_chert_1978; @abdolahzadehHeatExposedLithics2023; @pattersonThermalDamageChert1995; @purdyFracturesArchaeologist1975",
+  "Cortex", "Yes, No", "Presence or absence of cortex.", "-",
+  "Cortex thickness", "Thin, Medium, Thick", "Thickness of cortex measured in cm, following defined intervals.
+**Thin**: cortex thickness equal or inferior to 0.5 cm; **Medium**: 0.5-2 cm; **Thick**: >2 cm.", "-",
+  "Cortex appearance", "Outcrop, Rounded, Pebble, Indeterminate", "**Outcrop**: uneven and low level of alteration to the original cortical surface; **Rounded**: homogeneous and rounded cortex, with possible slight alteration to the original cortical surface; **Pebble**: Smooth and highly weathered surface, with high degree of alteration to the cortical surface", "-",
+  "Inclusions", "-", "Macroscopically visible inclusions (allochems and bioclasts).", "-"
+) |>
+  knitr::kable()
+```
+
+# Petrographic analysis
+
+```{r}
+#| label: tbl-petrography
+#| tbl-cap: Data dictionary for petrographic characterization of thin sections.
+
+library(tibble)
+
+tibble::tribble(
+  ~Variable, ~"Allowed values", ~Description, ~Reference,
+  "ID", "-", "Individual thinsection ID.", "-",
+  "Outcrop/level", "-", "Outcrop name or archaeological level.", "-",
+  "Lithology", "-", "Type of rock.", "-",
+  "Texture", "Mudstone, Wackestone, Packstone, Grainstone, Boundstone, Other", "**Mudstone**: muddy carbonate rock containing less then 10 % grains; **Wackestone**: mud-supported carbonate rock containing more than 10 % grains; **Packstone**: grain-supported muddy carbonate rock; **Grainstone**: mud-free carbonate rocks, which are grain supported; **Boundstone**: carbonate rocks showing signs of being bound during deposition.", "@dunhamClassificationCarbonateRocks1962",
+  "Microstructure", "Homogeneous, Banded, Laminar, Nodular, Brechoid, Other", "Distribution of crystals and clasts within the rock at a microscopic scale.
+**Homogeneous**: equally spread in the rock; **Banded**: distributed in bands; **Nodular**: distributed in clumps; **Brechoid**: fracturing of the rock irregularly.", "@castro_dorado_petrografibasica_1989",
+  "Orthochem", "-", "Materials formed in two ways: 1) deposited directly from supersaturated aqueous solutions due to chemical reactions or evaporation; 2) formed by the replacement of existing sedimentary materials.", "@vernonMicrostructuresSedimentaryRocks2018",
+  "Orthochem type", "Essential (ES), Accessory (AC), Secondary (SE)", "**Essential**: minerals that form more than 5% of the volume of the rock; **Accessory**: Minerals with proportion of less than 5% of the volume of the rock; **Secondary**: products of the alteration (hydrothermal or physical), independent of the proportion within the rock.", "@castro_dorado_petrografibasica_1989",
+  "Orthochem description", "-", "General description of the orthochem and where it is identified.", "-",
+  "Orthochem (%)", "-", "Approximate percentage of the orthochem’s presence in the total thin-section area.", "-",
+  "Allochem", "-", "Material formed by the movement and reorganization into new shapes by chemical, physical or biological processes within the depositional basin (ex.
+ooliths, fecal pellets, iron oxide minerals).", "@vernonMicrostructuresSedimentaryRocks2018",
+  "Allochem (freq)", "Rare, Uncommon, Common, Very frequent", "**Rare**: present one or two elements; **Uncommon**: present three to 10 elements; **Common**: present 11 to 20 elements; Very frequent:> 20 elements.", "-",
+  "Bioclast", "-", "Also known as skeletal particles, are the remains (complete or fragmented) of the hard parts of carbonate-secreeting organisms.", "@adamsAtlasSedimentaryRocks1988",
+  "Bioclast (freq)", "Rare, Uncommon, Common, Very frequent", "**Rare**: present one or two elements; **Uncommon**: present three to 10 elements; Common: present 11 to 20 elements; **Very frequent**:> 20 elements.", "-",
+  "Porosity (%)", "-", "Approximate frequency of effective porosity.", "-",
+  "Porosity type", "Interparticle, Moldic, Fenestral, Fracture, Vuggy, Shelter, Other", "**Interparticle**: porosity between particles; **Moldic**: Porosity formed by selective removal of an individual constituent of the rock; **Fenestral**: Pores larger than grain-supported interstices (interparticle); **Fracture**: Porosity formed by fracturing; **Vug**: Pores larger then 1/16 mm in diameter and somewhat equant in shape; **Shelter**: Porosity created by the sheltering effect of large sedimentary particles; **Other**: Other types of porosity.", "@choquetteGeologicNomenclatureClassification1970",
+  "Sedimentary structures", "Parallel lamination, Convoluted lamination, Bands/zonations, Burrow, Other", "**Parallel lamination**: sedimentary strata less than 10 mm thick, recognizable due to variation in structure or composition and more or less parallel bounding surfaces; **Convoluted lamination**: symmetrical about a vertical plane or leaning and asymmetrical, and usually exhibit narrow vertical upturned laminae, often truncated at the top, separated by a broader synclinal downfolds; **Bands/zonations**: limited areas with different characteristics related to changes in the sedimentation or cementation process; **Burrow**: bioturbation structures caused by activity of an organism that disrupts the stratification features; **Other**: other types of sedimentary structures.", "@middletonEncyclopediaSedimentsSedimentary2005"
+  ) |>
+knitr::kable()
+```
+
+# References

### 2024-04-26 analysis/supplementary-materials/SOM_lithotypes.qmd
Pre submission commit_1

@@
 ---
-title: "Terrace and shelter lithotype complete description"
+title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia, Journal of Paleolithic Archaeology"
+subtitle: "Online Resource 2: Terrace and shelter lithotypes complete description"
 author:
-  - Jane Doe:
+  - Joana Belmiro:
+      orcid: https://orcid.org/0000-0003-4951-2907
       correspondence: "yes"
-      email: janedoe@fosg.org
-      orcid: 0000-0003-1689-0557
+      email: jfbelmiro@ualg.pt
       institute:
-        - fosg
-        - fop
-  - John Q. Doe:
-      institute: fosg
-      orcid: 0000-0003-1689-0558
-  - Peder Ås:
-      institute: fosg
-      orcid: 0000-0003-1689-0559
-  - Juan Pérez:
-      orcid: 0000-0003-1689-0551
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
+  - Xavier Terradas:
+      orcid: https://orcid.org/0000-0002-8000-5607
       institute: 
-        - name: Acme Corporation
-  - Max Mustermann:
-      orcid: 0000-0003-1689-0552
+        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
+  - Salvador Dominguez-Bella:
+      orcid: https://orcid.org/0000-0003-3892-763X
+      institute: 
+        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
+  - João Cascalheira:
+      orcid: https://orcid.org/0000-0003-0321-8892
+      institute:
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
 title-block-published: "Last updated"  
@@
   comment: "#>"
-  fig-path: "../figures/"
-  fig-dpi: 600
 filters:
@@
   comment = "#>",
-  fig.path = "../figures/",
   dpi = 300,
-  out.width = "100%",
+  out.width = "70%",
   fig.align = "center"
@@
 
-# Chert group description
+# Chert type description
 
-## Group 2
+## Type 1
 
-Group 2 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment.
-The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Pliensbachian.
+Type 1 corresponds to a hydrothermal, fibrous chalcedony.
+The macroscopic and petrographic association of this type with the local Triassic chalcedonies from the west Algarve basin locate its epoch in the Triassic.
 
-It is a fine-grained chert, with a conchoidal fracture and good quality (see the Data dictionary for macroscopic analysis document for details on the attribution of quality to raw materials used in this analysis).
-It is macroscopically heterogeneous in color, patterns, and textural characteristics; due to this, initially, two sub-groups were created: single-color @fig-type2-variability and multiple-color () groups.
+It is a fine-grained chert, with a conchoidal fracture and acceptable quality, due to the frequent fractures and porosity within the lenses and nodules (see the Data dictionary for macroscopic analysis document for details on the attribution of quality to raw materials used in this analysis).
+It is homogeneous, without color but frequently white patinated.
+The artifacts are dull (or with a waxy appearance), translucent (whenever not patinated) and smooth to semi-smooth.
 
-The single-color group's main colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), grayish brown (10YR 5/2) and white (2.5Y 8/1).
+Cortex is difficult to identify due to the frequent patina, fractures and porosity, although whenever present, it is thin.
+Although found in lenses in the outcrops, the artifacts do not display any tabular characteristics (cortex in two opposed faces).
+
+Results from thin-section show this chert is composed by 100% fibrous chalcedony, without allochems or bioclasts.
+
+```{r}
+#| label: fig-type1-variability
+#| fig-cap: Chert macroscopic variability within type 1 (chalcedony).
+
+knitr::include_graphics("../figures/SOM/type1_macromeso.png")
+
+```
+
+## Type 2
+
+Type 2 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment (@fig-type2-variability and @fig-type2Bvariability).
+The macroscopic and petrographic association of this type with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Carixian.
+
+It is a fine-grained chert, with a conchoidal fracture and good quality.
+It is macroscopically heterogeneous in color, patterns, and textural characteristics; due to this, initially, two sub-types were created: single-color (@fig-type2-variability) and multiple-color (@fig-type2Bvariability) types.
+
+The single-color type's main colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), grayish brown (10YR 5/2) and white (2.5Y 8/1).
 The color variability and presence of allochems and bioclasts produce macroscopic patterns such as splotched, speckling, and laminations.
@@
 Translucency is opaque and the feel is smooth to semi-smooth.
-The multiple-color group shows a variable color distribution: colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), pale red (2.5YR 6/2), and white (2.5Y 8/1).
-The color variability creates macroscopic patterns such as splotched, broad mottling, speckling, and laminations (Fig.
-\@ref(fig:type2Bvariability)).
-The structure is similar to group single-color, although slightly heterogeneous as translucency varies from opaque to sub-translucent, with light passing through at around 3.5 mm, limited to the white areas.
+The multiple-color type shows a variable color distribution: colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), pale red (2.5YR 6/2), and white (2.5Y 8/1).
+The color variability creates macroscopic patterns such as splotched, broad mottling, speckling, and laminations (@fig-type2Bvariability).
+The structure is similar to type single-color, although slightly heterogeneous as translucency varies from opaque to sub-translucent, with light passing through at around 3.5 mm, limited to the white areas.
 
-For both sub-groups, whenever present, the cortex can be rounded or irregular and from an outcrop, varying between ++ to ++ mm in thickness, frequently with a sharp transition.
+For both sub-types, whenever present, the cortex can be rounded or irregular and from an outcrop, varying between 10 to 0.5 mm in thickness, frequently with a sharp transition.
 
@@
 Bioclasts include white, round speckles, spicule-shaped skeletal remains, monaxon sponge spicules, bivalve shells (rare, hereon r) replaced by quartz or chalcedony, and rarely triaxon sponge spicules.
-In the single-color group fossil remains range from uncommon to frequent, while in the multiple-color group fossils range from uncommon to very frequent; in all cases, the bioclasts are frequently unidentifiable.
+In the single-color type fossil remains range from uncommon to frequent, while in the multiple-color type fossils range from uncommon to very frequent; in all cases, the bioclasts are frequently unidentifiable.
 
-In the shelter area, the variety of identified fossils also includes possible ostracods (uncommon, hereon u), a possible small foraminifer (Fig.
-\@ref(fig:shelter2fossils) d), and echinoderm spines (Fig.
-\@ref(fig:shelter2fossils) a-c).
+In the shelter area, the variety of identified fossils also includes possible ostracods (uncommon, hereon u), a possible small foraminifer (@fig-shelter2fossils d), and echinoderm spines (@fig-shelter2fossils a-c).
 
-Five thin sections were produced to account for the macroscopic variability of this group.
+Five thin sections were produced to account for the macroscopic variability of this type.
 All thin sections show a wackestone texture and massive microstructure.
@@
 All samples have very frequent unidentifiable fossils.
-In samples from multiple-color group, although in lesser frequencies, monaxon sponge spicules (H21-3095) and a possible triaxone sponge spicule (H21-4234) are identifiable.
+In samples from multiple-color type, although in lesser frequencies, monaxon sponge spicules (H21-3095) and a possible triaxone sponge spicule (H21-4234) are identifiable.
 All samples show vug-type porosity between 1-5%.
@@
 #| label: fig-type2-variability
-#| fig-cap: Chert macroscopic variability within group 2 (single color).
-#| fig-width: 8
+#| fig-cap: Chert macroscopic variability within type 2 (single color).
 
-knitr::include_graphics("../figures/SOM/Group2_macromeso_b.png")
+knitr::include_graphics("../figures/SOM/type2_macromeso.png")
 
@@
 
-```{r type2Bvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 2 (multiple color).", out.width = '80%'}
+```{r}
+#| label: fig-type2Bvariability
+#| fig-cap: Chert macroscopic variability within type 2 (multiple color).
 
-knitr::include_graphics("../figures/SOM/Group2B_macromeso.png")
+knitr::include_graphics("../figures/SOM/type2B_macromeso.png")
 
@@
 
-```{r shelter2fossils, echo=FALSE, fig.cap="Identified fossil variability from group 2 from the Shelter area.
-Fossils present in the figures are: a-b, crinoid stem section with 4 series; c, cross-section of crinoid columnal with pentagonal stem plate; d, possible foraminifer (black arrow).", out.width = '80%'}
+```{r}
+#| label: fig-shelter2fossils
+#| fig-cap: "Identified fossil variability from type 2 from the Shelter area.
+Fossils present in the figures are: a-b, crinoid stem section with 4 series; c, cross-section of crinoid columnal with pentagonal stem plate; d, possible foraminifer (black arrow)."
 
@@
 
-A group of samples was individualized from group 2 due to color and textural differences (named 2R during the analysis process).
-These are interpreted as alterations to group 2 cherts, possibly due to heat, although further experimentation is needed.
+A type of samples was individualized from type 2 due to color and textural differences (named 2R during the analysis process; @fig-type2Rvariabilitymeso).
+These are interpreted as alterations to type 2 cherts, possibly due to heat, although further experimentation is needed.
 The color distribution is variable and it has a heterogeneous structure, possibly due to the alterations.
@@
 
-```{r type2Rvariabilitymeso, echo=FALSE, fig.cap="Archaeological samples attributed to Type 2R. ", out.width = '80%'}
+```{r}
+#| label: fig-type2Rvariabilitymeso
+#| fig-cap: Archaeological samples attributed to Type 2 with reddening alterations.
 
-knitr::include_graphics("../figures/SOM/Group2R_macromeso-01.png")
+knitr::include_graphics("../figures/SOM/type2R_macromeso-01.png")
 
@@
 A thin section was made to confirm whether the color and textural differences are solely macroscopic, and do not represent a different petrographic type.
-The sample (H19-2426 from Level 4D) shows no outstanding differences from the samples from group 2.
+The sample (H19-2426 from Level 4D) shows no outstanding differences from the samples from type 2.
 
-## Group 15
+## Type 3
 
-Group 15 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment.
-The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Pliensbachian.
+Type 3 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment (@fig-type3variability).
+The macroscopic and petrographic association of this type with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Carixian.
 
-Group 15 is a fine to medium chert, with a conchoidal fracture and good quality.
+Type 3 is a fine to medium chert, with a conchoidal fracture and good quality.
 The color distribution is uniform, despite subtle changes to gray.
@@
 The splotches and lamination adjacent to the cortex seem to be sub-translucent as well, showing more translucency than the chert itself.
-The feel is semi-smooth and the patterns are splotches, speckling, concentric lamination (limited to a small number of samples), and fine lamination (found only between the cortex and the chert).
+The feel is semi-smooth and the patterns are splotches (@fig-type3variability d), speckling, concentric lamination (limited to a small number of samples), and fine lamination (found only between the cortex and the chert).
 
 Whenever present, the cortex seems to be from an outcrop and rounded.
-Despite the brown coloration possibly due to post-depositional alterations, the original color is white.
+Despite the brown coloration possibly due to post-depositional alterations, the original color is white (@fig-type3variability a-c).
 Cortex thickness ranges from 3.8 mm to 1.2 mm, and has a sharp transition.
 One single sample from Level 6 shows a gradual cortex transition.
-Although the original morphology of this chert is unknown, the presence of cortex on two opposed extremities in a sample (H19-3196, \@ref(fig:type15variability) a) indicates the chert can be found in irregularly rounded nodules within the parent rock.
+Although the original morphology of this chert is unknown, the presence of cortex on two opposed extremities in a sample (H19-3196, @fig-type3variability a) indicates the chert can be found in irregularly rounded nodules within the parent rock.
 
@@
 
-```{r type15variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 15.", out.width = '80%'}
+```{r}
+#| label: fig-type3variability
+#| fig-cap: Chert macroscopic variability within type 3.
 
-knitr::include_graphics("../figures/SOM/Group15_macromeso.png")
+knitr::include_graphics("../figures/SOM/type3_macromeso.png")
 ```
 
-## Group 7
+## Type 4
 
-Group 7 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment.
-The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin locate their epoch and stage in the Lower Jurassic, Pliensbachian.
+Type 4 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment (@fig-type4variability).
+The macroscopic and petrographic association of this type with the local cherts from the west Algarve basin locate their epoch and stage in the Lower Jurassic, Carixian.
 
@@
 
-Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: grey (G) and white (W) fabric.
+Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: grey and white fabric.
 The grey fabric corresponds to the interior of the nodule.
 It is light gray (5Y 7/1, 7/2), light olive gray (5Y 6/2), and gray (5Y 6/1, 5/1), sub-translucent (light passing through in the thin areas, around 3.2 mm thickness) and has medium luster.
-The visible inclusions are limited to a single mica mineral (\~50 μm), white flecks (concentrated near the outer edges of white fabric), and speckles, which may be circular, unidentifiable fossil remains, and rare spicule-like remains.
+The visible inclusions are limited to a single mica mineral (\~50 μm), white flecks (concentrated near the outer edges of white fabric), and speckles, which may be circular (@fig-type4variability b-d), unidentifiable fossil remains, and rare spicule-like remains.
 In layer 6 spicule-shaped remains are uncommon.
@@
 This fabric is white (5Y 8/1, 10YR 8/1) and very pale brown (10YR 8/2, 8/4), opaque and dull.
-In the pale brown splotches, inclusions are yellow speckles (possibly iron oxides, Fig.
-\@ref(fig:type7variability) a).
+In the pale brown splotches, inclusions are yellow speckles (possibly iron oxides; @fig-type4variability a).
 The white fabric generally has barely visible fossils and uncommon unidentifiable fossils replaced by chalcedony or quartz.
@@
 The cortex is white or light brown with a sharp transition between 1.4 mm to 10 mm thick.
-The morphology of the cortex in fully cortical samples (i.e., I18-2099; \@ref(fig:type7variability) d) may indicate that this chert is originally available as irregular (slightly oval) nodules within the parent rock.
+The morphology of the cortex in fully cortical samples (i.e., I18-2099; @fig-type4variability d) may indicate that this chert is originally available as irregular (slightly oval) nodules within the parent rock.
 
-Petrography results show this group is a massive wackestone composed mostly of micro-cryptocrystalline quartz (98%), with accessory macrocrystalline quartz and fibrous chalcedony.
+Petrography results show this type is a massive wackestone composed mostly of micro-cryptocrystalline quartz (98%), with accessory macrocrystalline quartz and fibrous chalcedony.
 Allochems are iron oxides (c), opaques (vf), and bioclasts (c).
@@
 
-```{r type7variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 7.", out.width = '80%'}
+```{r}
+#| label: fig-type4variability
+#| fig-cap: Chert macroscopic variability within type 4.
 
-knitr::include_graphics("../figures/SOM/Group7_macromeso.png")
+knitr::include_graphics("../figures/SOM/type4_macromeso.png")
 
@@
 
-## Group 12
+## Type 5
 
-Group 12 corresponds to a massive and banded micro-cryptocrystalline quartz from a marine environment and possibly a shelf depositional environment.
-The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin, as well as the large size of the nodules and presence of parent rock, allow us to interpret this chert as local, possibly the Lower Jurassic although the exact outcrop is not identifiable.
+Type 5 corresponds to a massive and banded micro-cryptocrystalline quartz from a marine environment and possibly a shelf depositional environment (@fig-type5variability).
+The macroscopic and petrographic association of this type with the local cherts from the west Algarve basin, as well as the large size of the nodules and presence of parent rock, allow us to interpret this chert as local, possibly the Lower Jurassic although the exact outcrop is not identifiable.
 
 This chert is fine to medium-grained, with a conchoidal fracture (albeit uneven at times).
-The quality is acceptable to low due to the lack of homogeneity and frequent fractures, as well as a coarser fabric in certain areas.
+The quality is acceptable to low due to the lack of homogeneity and frequent fractures (@fig-type5variability c), as well as a coarser fabric in certain areas.
 It has a heterogeneous structure, frequently showing different characteristics within one single artifact.
@@
 4 mm of thickness.
-The group displays frequent patterns, mostly related to this heterogeneity, but also due to patination; these are splotched, banded, broad mottling, and speckling.
+The type displays frequent patterns, mostly related to this heterogeneity, but also due to patination; these are splotched, banded, broad mottling, and speckling.
 
 The cortex is from an outcrop, frequently rounded, but also irregular.
-Based on reference samples, the chert seems to be enclosed within the parent rock as irregular or tabular nodules (Fig.
-\@ref(fig:type12variability) b).
-The cortex is thin (\~0.5 mm), with a sharp transition, and some artifacts show chunks of the parent rock attached to the cortex (Fig.
-\@ref(fig:type12variability) a).
+Based on reference samples, the chert seems to be enclosed within the parent rock as irregular or tabular nodules (@fig-type5variability b).
+The cortex is thin (\~0.5 mm), with a sharp transition, and some artifacts show chunks of the parent rock attached to the cortex (@fig-type5variability a).
 
-Based on the presence of bands, two sub-groups were created: 12 (Fig.
-\@ref(fig:type12variability) a-c) and 12B (Fig.
-\@ref(fig:type12variability) d), the latter characterized by banding.
+Based on the presence of bands, two sub-types were created: 5 (@fig-type5variability a-c) and 5B (@fig-type5variability d), the latter characterized by banding.
 Two thin sections were produced to characterize the macroscopic variability.
-The petrographic results show that group 12 is a wackestone, with a massive (group 12) or banded (group 12B) microstructure.
+The petrographic results show that type 5 is a wackestone, with a massive (type 5) or banded (type 5B) microstructure.
 They are mainly composed of micro-cryptocrystalline quartz (75% and 72%, respectively) and micrite (20% in both), and in smaller percentages macrocrystalline quartz and fibrous chalcedony.
 Allochems are unidentified opaques (vf), iron oxides (u), and bioclasts (c).
-The sub-groups show larger differences in the porosity, with 12 showing 1% vug porosity while 12B shows 10% vug and moldic porosity, concentrated in the banded areas with more opaques.
+The sub-types show larger differences in the porosity, with 5 showing 1% vug porosity while 5B shows 10% vug and moldic porosity, concentrated in the banded areas with more opaques.
 
 Bioclasts are uncommon both macroscopically and in thin section: sponge spicules and round, unidentifiable fossils can be seen by the naked eye or under the microscope.
-In thin section however, several other unidentified fossils are present, some of which may be an ostracod (both in 12 and 12B) and a foraminifer (12).
+In thin section however, several other unidentified fossils are present, some of which may be an ostracod (both in 5 and 5B) and a foraminifer (5).
 
-```{r type12variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 12.", out.width = '80%'}
+```{r}
+#| label: fig-type5variability
+#| fig-cap: Chert macroscopic variability within type 5.
 
-knitr::include_graphics("../figures/SOM/Group12_macromeso.png")
+knitr::include_graphics("../figures/SOM/type5_macromeso.png")
 
@@
 
-## Group 11
+## Type 6
 
-Group 11 corresponds to a massive cryptocrystalline quartz with mudstone texture.
+Type 6 corresponds to a massive cryptocrystalline quartz with mudstone texture (@fig-type6variability).
 The lack of local similar cherts, as well as identifiable fossils does not allow us to attribute a depositional environment or epoch/stage.
-However, rare spicules were identified; given the exclusivity of sponge's (Porifera) living environment to be marine [@flugelMicrofaciesCarbonateRocks2010], group 11 may be considered a chert formed in a marine environment.
+However, rare spicules were identified; given the exclusivity of sponge's (Porifera) living environment to be marine [@flugelMicrofaciesCarbonateRocks2010], type 6 may be considered a chert formed in a marine environment.
+The similarities with cherts from Central Portugal indicates these cherts were formed in Cretaceous formations from the Lisbon and Estremadura area.
 
@@
 The structure is heterogeneous, with shiny to medium luster and semi-smooth to smooth feel.
-This group of translucent cherts shows a variety of colors: grayish brown (2.5Y 5/2), light brownish gray (10YR 6/2), brown (7.5 YR 4/2), reddish brown (2.5YR 4/3) and dark gray (10YR 4/1).
+This type of translucent cherts shows a variety of colors: grayish brown (2.5Y 5/2), light brownish gray (10YR 6/2), brown (7.5 YR 4/2), reddish brown (2.5YR 4/3) and dark gray (10YR 4/1).
 The color distribution is variable; in some cases, the color seems to be related to the thickness of the sample; thicker parts show darker colors, while thinner parts show lighter colors.
@@
 
-The cortex is rounded, from an outcrop source, and rarely rounded and smooth like a pebble. It is frequently around 0.5 to ++ mm thick and the transition is gradual.
+The cortex is rounded, from an outcrop source, and rarely rounded and smooth like a pebble. It is frequently around 0.5 to 1 mm thick and the transition is gradual.
 
-```{r typeCPTvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 11.", out.width = '80%'}
+```{r}
+#| label: fig-type6variability
+#| fig-cap: Chert macroscopic variability within type 6.
 
-knitr::include_graphics("../figures/SOM/GroupCPT_macromeso.png")
+knitr::include_graphics("../figures/SOM/type6_macromeso.png")
 
@@
 
-### Sub-group LLB
+### Sub-type 6LLB
 
-A sub-group named LLB (corresponding to the light brownish gray colors) was created during the macroscopic analysis process due to the presence of a group of artifacts with homogeneous characteristics (and refits) within the previous group.
+A sub-type named 6LLB (corresponding to the light brownish gray colors) was created during the macroscopic analysis process due to the presence of a type of artifacts with homogeneous characteristics (and refits) within the previous type (@fig-subtype6variability a-b).
 
-This sub-group is characterized by white (10YR 8/1), light brownish gray (10YR 6/2), and brown (7.5YR 5/2) colors.
+This sub-type is characterized by white (10YR 8/1), light brownish gray (10YR 6/2), and brown (7.5YR 5/2) colors.
 It has a heterogeneous structure, with two fabrics: white and brown.
@@
 
-```{r typeLLB11Evariability, echo=FALSE, fig.cap="Chert macroscopic variability within the sub-group LLB (a-b) and 11E (c-d).", out.width = '100%'}
+```{r}
+#| label: fig-subtype6variability
+#| fig-cap: Chert macroscopic variability within the sub-type LLB (a-b) and 6E (c-d).
 
-knitr::include_graphics("../figures/SOM/groupLLB_11E_variability.png")
+knitr::include_graphics("../figures/SOM/type6LLB_6E_variability.png")
 
@@
 
-### Sub-group 11E
+### Sub-type 6E
 
-Similarly to the previous sub-group, a sub-group named 11E was created during the macroscopic analysis process due to the presence of a group of artifacts with homogeneous characteristics (and refits).
+Similarly to the previous sub-type, a sub-type named 6E was created during the macroscopic analysis process due to the presence of a type of artifacts with homogeneous characteristics (and refits; @fig-subtype6variability c-d).
 
-This sub-group is characterized by a heterogeneous structure.
+This sub-type is characterized by a heterogeneous structure.
 The colors are brown (10YR 5/3), gray (2.5Y 5/1), light gray (10YR 7/2) and white (10YR 8/1) The translucency is variable, some areas are translucent (light passing through at 8 mm of thickness), while the splotches are sub-translucent.
@@
 
-Five thin sections were produced to better characterize the macroscopic variability of this group and sub-groups.The results show no outstanding differences between the samples and sub-groups.
+Five thin sections were produced to better characterize the macroscopic variability of this type and sub-types.
+The results show no outstanding differences between the samples and sub-types.
 The petrographic analysis shows the samples are characterized by a mudstone texture and massive microstructure.
@@
 
-## Group 11F
+## Type 7
 
-Group 11F corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment.
-Identified and well-preserved fossils are foraminifera from the Pfenderenidae family, which are larger benthic foraminifera from marine environments.
+Type 7 is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment (@fig-type7variability), due to the type, sorting, and preservation conditions of the ooids.
+
+It is a fine-grained chert, with a conchoidal fracture.
+The quality is good.
+It has a heterogeneous structure, with two fabrics: 1) white fabric; 2) grey fabric.
+The white fabric is characterized by dull luster, opaque translucency, and semi-smooth feel.
+The gray fabric has gray (10YR 7/1 to 5/1) and light brownish gray (10YR 6/2) colors, and varies between medium to dull luster and it is opaque to sub-translucent, with light passing through at around 1.9 mm of thickness in some areas.
+The feel is semi-smooth and the patterns are splotched, speckled, and laminated dotted horizontally.
+
+The cortex is from an outcrop, rounded, and varying between a minimum of 3.4 mm of thickness and a maximum of 12 mm of thickness.
+The cortex transition is frequently sharp, although some samples show a gradual transition.
+
+This type shows several facies, with different types and concentrations of allochems and bioclasts (@fig-type7facies).
+Facies 01a shows a wackestone to packstone texture, with visible small fossils (very frequently below 50 μm).
+Fossils include round white fossils, with grey or non-colored walls; sponge spicules, monaxon pointed at one end or tetraxon.
+A possible bivalve shell replaced by quartz or chalcedony was identified on a single sample.
+Other fossils are unidentifiable, but very frequently they are round and light, scattered within the chert.
+Facies 01b is similar to 01a, but composed mainly of peloids or unpreserved ooids, with uncommon presence of spicules.
+Facies 01c shows a decrease in the visible fossils (wackestone texture), and identifiable fossils are spicules and small round fossils with walls.
+Facies 02 is characterized by a packstone texture, composed mostly of larger (\~200-500 μm, uncommonly 1 cm) round or oval fossils, white or grey, completely or partially replaced by quartz or chalcedony.
+The smaller round fossils with walls present in Facies 01 are uncommon.
+Due to poor preservation, these fossils are unidentifiable.
+Facies 03 is mostly white, with faint round shapes; this facies is interpreted as a white patinated area of both Facies 01 and 02.
+Facies 04 is characterized by a packstone texture, and composed of ooids (\~50-200 μm, uncommonly 500 μm), which are completely or partially replaced by quartz and chalcedony.
+In this facies, the chert is light (white or light grey) and it seems to be sub-translucent.
+Fossils seen in other facies are not visible.
+
+The petrography results for a sample with Facies 01a and 01b (L19-64) and Facies 04 (J18-1264) show this chert has a massive microstructure and packstone texture.
+It is composed mainly of micro-cryptocrystalline quartz (80-85%), fibrous chalcedony (10-5%), and micrite (10%).
+Allochems are opaques (u), iron oxides (vf/u), ooids (vf/u) and peloids (vf).
+The ooids are poorly preserved and only uncommonly show concentric lamellae in plane polarized light.
+Unlike macroscopical observations, bioclasts under the thin section are all unidentifiable, although rare fossils may be bivalve shells.
+Porosity is vug (10%) or <1% in Facies 04.
+
+```{r}
+#| label: fig-type7variability
+#| fig-cap: Chert macroscopic variability within type 7.
+
+knitr::include_graphics("../figures/SOM/type7_macromeso.png") 
+```
+
+```{r}
+#| label: fig-type7facies
+#| fig-cap: Several fossil facies within Type 7.
+
+knitr::include_graphics("../figures/SOM/type7_facies.png")
+```
+
+## Type 8
+
+Type 8 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment (@fig-type8variability).
+Identified and well-preserved fossils are foraminifera from the Pfenderenidae family, which are larger benthic foraminifera from marine environments (@fig-type8fossils a, c-d).
 The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous.
 
-Macroscopically, this group is characterized by gray colors (Gley 1 6/ and 5/) and a white (10YR 8/1) area, frequently at the center of the sample, which seems to be a weathering rind.
-However, the white broad mottling highlights the presence of very frequent bioclasts (\@ref(fig:type11Fvariability)).
-The translucency is variable: the grey areas are sub-translucent (light passing through at 2.5/3 mm of thickness), while the broad mottling is opaque; due to the presence of this sort of alteration in all artifacts from this group, a correct measurement of translucency was not possible.
+Macroscopically, this type is characterized by gray colors (Gley 1 6/ and 5/) and a white (10YR 8/1) area, frequently at the center of the sample, which seems to be a weathering rind.
+However, the white broad mottling highlights the presence of very frequent bioclasts (@fig-type8fossils).
+The translucency is variable: the grey areas are sub-translucent (light passing through at 2.5/3 mm of thickness), while the broad mottling is opaque; due to the presence of this sort of alteration in all artifacts from this type, a correct measurement of translucency was not possible.
 The luster ranges from shiny to medium, the feel is smooth and the grain is fine.
@@
 
-Petrographically, this group is a packstone (mudstone texture close to the edges) with a massive microstructure.
+Petrographically, this type is a packstone (mudstone texture close to the edges) with a massive microstructure.
 It is composed of 97% micro-cryptocrystalline quartz.
@@
 Larger unidentifiable fossils (\~200 μm; r) are frequently replaced by several generations of microcrystalline quartz, macrocrystalline quartz, and fibrous chalcedony.
-Identified fossils are sponge spicules (r; visible only as ghosts and poorly preserved), ostracods (r) with some degree of preservation, and common small foraminifera (100 μm) of differing preservation degrees (Fig.
-\@ref(fig:11Ffossils)).
-Identifiable foraminifera are rare and from the Pfenderenidae family (Fig.
-\@ref(fig:11Ffossils), black arrows); these have been identified both through thin section (Fig.
-\@ref(fig:11Ffossils) d) and macroscopically (Fig.
-\@ref(fig:11Ffossils) a, c).
+Identified fossils are sponge spicules (r; visible only as ghosts and poorly preserved), ostracods (r) with some degree of preservation, and common small foraminifera (100 μm) of differing preservation degrees (@fig-type8fossils).
+Identifiable foraminifera are rare and from the Pfenderenidae family (@fig-type8fossils, black arrows); these have been identified both through thin section (@fig-type8fossils d) and macroscopically (@fig-type8fossils a, c).
 They are well-preserved, replaced by microcrystalline quartz, and measure around 500 μm.
 
-```{r type11Fvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 11F.", out.width = '80%'}
+```{r}
+#| label: fig-type8variability
+#| fig-cap: Chert macroscopic variability within type 8.
 
-knitr::include_graphics("../figures/SOM/Group11F_macromeso.png")
+knitr::include_graphics("../figures/SOM/type8_macromeso.png")
 
@@
 
-```{r 11Ffossils, echo=FALSE, fig.cap="Foraminifera from Group 11F from the Terrace and Shelter area.
+```{r}
+#| label: fig-type8fossils
+#| fig-cap: "Foraminifera from Type 8 from the Terrace and Shelter area.
 Figures a-c were obtained through a stereomicroscope; figure d is a thin section in plane polarized light.
-Black arrows: foraminifera from the Pfenderenidae family; White arrow: other unidentified foraminifera.", out.width = '80%'}
+Black arrows: foraminifera from the Pfenderenidae family; White arrow: other unidentified foraminifera."
 
-knitr::include_graphics("../figures/SOM/group11F_FOSSILS.png")
+knitr::include_graphics("../figures/SOM/type8_FOSSILS.png")
 
@@
 
-## Group 13
+## Type 9
 
-Group 13 is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
-
-It is a fine-grained chert, with a conchoidal fracture.
-The quality is good.
-It has a heterogeneous structure, with two fabrics: 1) white fabric; 2) grey fabric.
-The colors are .
-The white fabric is characterized by dull luster, opaque translucency, and semi-smooth feel.
-The grey fabric varies between medium to dull luster and it is opaque to sub-translucent, with light passing through at around 1.9 mm of thickness in some areas.
-The feel is semi-smooth and the patterns are splotched, speckled, and laminated dotted horizontally.
-
-The cortex is from an outcrop, rounded, and varying between a minimum of 3.4 mm of thickness and a maximum of 12 mm of thickness.
-The cortex transition is frequently sharp, although some samples show a gradual transition.
-
-This group shows several facies, with different types and concentrations of allochems and bioclasts (Fig.
-\@ref(fig:type13facies)).
-Facies 01a shows a wackestone to packstone texture, with visible small fossils (very frequently below 50 μm).
-Fossils include round white fossils, with grey or non-colored walls; sponge spicules, monaxon pointed at one end or tetraxon.
-A possible bivalve shell replaced by quartz or chalcedony was identified on a single sample.
-Other fossils are unidentifiable, but very frequently they are round and light, scattered within the chert.
-Facies 01b is similar to 01a, but composed mainly of peloids or unpreserved ooids, with uncommon presence of spicules.
-Facies 01c shows a decrease in the visible fossils (wackestone texture), and identifiable fossils are spicules and small round fossils with walls.
-Facies 02 is characterized by a packstone texture, composed mostly of larger (\~200-500 μm, uncommonly 1 cm) round or oval fossils, white or grey), completely or partially replaced by quartz or chalcedony.
-The smaller round fossils with walls present in Facies 01 are uncommon.
-Due to poor preservation, these fossils are unidentifiable.
-Facies 03 is mostly white, with faint round shapes; this facies is interpreted as a white patinated area of both Facies 01 and 02.
-Facies 04 is characterized by a packstone texture, and composed of ooids (\~50-200 μm, uncommonly 500 μm), which are completely or partially replaced by quartz and chalcedony.
-In this facies, the chert is light (white or light grey) and it seems to be sub-translucent.
-Fossils seen in other facies are not visible.
-
-The petrography results for a sample with Facies 01a and 01b (L19-64) show this chert has a massive microstructure and packstone texture.
-It is composed mainly of micro-cryptocrystalline quartz (80%), fibrous chalcedony (10%), and micrite (10%).
-Allochems are opaques (u), iron oxides (vf), ooids (vf) and peloids (vf).
-The ooids are poorly preserved and only uncommonly show concentric lamellae in plane polarized light.
-Unlike macroscopical observations, bioclasts under the thin section are all unidentifiable, although rare fossils may be bivalve shells.
-Porosity is vug (10%).
-
-```{r type13variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 11F.", out.width = '80%'}
-
-knitr::include_graphics("../figures/SOM/Group13_macromeso.png")
-
-```
-
-```{r type13facies, echo=FALSE, fig.cap="Several fossil facies within Type 13.", out.width = '80%'}
-
-knitr::include_graphics("../figures/SOM/GROUP13_facies-01.png")
-```
-
-## Type 13B
-
-Group 13B corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone.
+Type 9 corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone (@fig-type9variability).
 
@@
 It is characterized by a sharp transition.
-In specific samples (i.e., I21-3546, \@ref(fig:type13variability) a) cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers.
+In specific samples (i.e., I21-3546; @fig-type9variability a) cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers.
 In this case, the sample indicates a bed thickness of at least \~19 mm in that particular area of the outcrop.
@@
 
-```{r type13Bvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 13B.", out.width = '80%'}
+```{r}
+#| label: fig-type9variability
+#| fig-cap: Chert macroscopic variability within type 9.
 
-knitr::include_graphics("../figures/SOM/Group13B_macromeso.png")
+knitr::include_graphics("../figures/SOM/type9_macromeso.png")
 
@@
 
-## Oolitic chert
+## Type 10
 
-This group is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids.
+This type is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids (@fig-type10variability).
 
@@
 The black fabric is black (7.5YR 2.5/1), shiny, and sub-translucent (light passing through in the thin areas, around 1 to 2 mm thickness).
-It corresponds to the fabric which creates horizontal laminated patterns within the brown fabric (Fig.
-\@ref(fig:ooiliticvariability)c).
+It corresponds to the fabric which creates horizontal laminated patterns within the brown fabric (@fig-type10variability c).
 This chert is fine-grained with a conchoidal fracture.
@@
 
-Petrographically this group is a packstone with a massive microstructure.
+Petrographically this type is a packstone with a massive microstructure.
 The chosen sample for thin section is composed of 35% micro-cryptocrystalline quartz, 53% micrite, 9% macrocrystalline quartz, and accessory fibrous chalcedony (filling fractures) and mica; the percentages of quartz and micrite might be related to the amount of fabrics present in the artifacts.
@@
 They are poorly sorted, varying between 500 and 20 μm.
-They show round to elliptical shapes, and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010], pp.
-144\].
+They show round to elliptical shapes, and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010].
 When preserved, the nucleus represents half the width of the ooid and the cortex is normal.
 
-```{r ooiliticvariability, echo=FALSE, fig.cap="Chert macroscopic variability within the Oolitic chert group.", out.width = '80%'}
+```{r}
+#| label: fig-type10variability
+#| fig-cap: Chert macroscopic variability within type 10.
 
-knitr::include_graphics("../figures/SOM/oolitic_macromeso.png")
+knitr::include_graphics("../figures/SOM/type10_macromeso.png")
 
@@
 
+## Type 11
+
+Type 11 corresponds to a banded mudstone of unknown formation environment (@fig-TL).
+It is fine-grained, with a conchoidal fracture at the center of the nodule, but laminar fractures in some areas, possibly resulting in small, slab-like products, which resemble a schist.
+As such, the quality is acceptable and knapping limited to the inner part of the nodule.
+It has a heterogeneous structure and is characterized by variable color distributions: it shows a dark gray (5Y 4/1) interior, intercalated with bands and laminations of gray (5Y 6/1) and light olive gray (5Y 6/2) colors.
+The chert is dull and opaque, with a semi-smooth feel.
+No inclusions or bioclasts are visible.
+
+No cortex is identified.
+Instead, the top and bottom of the nodule transition from a chert with conchoidal fracture to laminar fracturing with some degree of alteration.
+The presence of two parallel planes in the samples indicates the nodule was recovered from a bedded outcrop.
+
 ## Trace lithotypes
+
+The TL types correspond to cherts which are not congruent with the known local sources, composed of 1-3 samples, limited to certain archaeological layers in the Terrace and Shelter area.
+16 different TL types have been identified (@fig-TL):
 
@@
 
-**TL02 - Orange chalcedony.** Fibrous chalcedony, without visible allochems. Good quality, differing from the group of white/translucent chalcedony due to its color. Translucent and without cortex.
+**TL02 - Orange chalcedony.** Fibrous chalcedony, without visible allochems. Good quality, differing from the type of white/translucent chalcedony due to its color. Translucent and without cortex.
 
@@
 
+```{r}
+#| label: fig-TL
+#| fig-cap: Type 11 and trace lithotypes present in the Terrace and Shelter areas.
+
+knitr::include_graphics("../figures/small_grid.png")
+
+```
+
 # References

### 2024-04-26 README.Rmd
Update README.Rmd

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
 ```
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Belmiro, J., Terradas, X., Dominguez-Bella, S., Cascalheira, J., (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Belmiro, J., Terradas, X., Dominguez-Bella, S., Cascalheira, J., (`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Title`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/10.17605/OSF.IO/DBFT2>
 
@@
 
-You can download the compendium as a zip from from this URL:
+You can download the compendium as a zip from this URL:
 [master.zip](/archive/master.zip). After unzipping:

### 2024-06-28 analysis/paper/paper.qmd
- Chi-square SOM - Final JC revisions

@@
   Analyses of raw materials and the distinction between local/regional and long-distance sources have proven invaluable for understanding the extensive movements, interactions, and social networks during the Upper Palaeolithic (UP) in the Iberian Peninsula.
-However, unlike other parts of Iberia, research on the management and acquisition of raw materials in the south and west of the Peninsula remains relatively underdeveloped.
-Despite significant knowledge about the technological practices of Palaeolithic hunter-gatherers from southern Portugal, particularly from studies conducted at the Vale Boi site, there's a noticeable lack of focus on raw materials.
-This paper presents the first comprehensive characterization of chert raw materials from the Gravettian, Proto-Solutrean, and Solutrean sequences at Vale Boi, using both macroscopic and petrographic techniques.
-Our study reveals that the majority of chert found at Vale Boi originates locally, within a 20 km radius, while a smaller portion comes from non-local sources, potentially indicating long-distance procurement from central Portugal.
-Additionally, the identification of cherts resembling those from southern Spain suggests a significant west-east mobility pattern throughout the UP period.
+However, unlike other parts of Iberia, research on the management and acquisition of raw materials in the south and west of Iberia remains relatively underdeveloped.
+Despite significant knowledge about the technological practices of Palaeolithic hunter-gatherers from Southern Portugal, particularly from studies conducted at the site of Vale Boi, there is a noticeable lack of focus on raw materials management.
+This paper presents the first comprehensive characterization of chert raw materials from the Gravettian, Proto-Solutrean, and Solutrean occupations at Vale Boi, using both macroscopic and petrographic techniques.
+Our study reveals that the majority of chert found at Vale Boi originates locally, within a 20 km radius.
+However, a non-neglectable portion of the chert comes from non-local sources, indicating >200 km raw material circulation from Central Portugal and Southern Spain.
 keywords: |
@@
 # Call packages
+# This includes all necessary packages to run the code
 library(readr)
@@
 context <- context %>%
-  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column
-  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
-  select(ID, Level, Spit, Code) %>% # Selecting important variables
+  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column from Unit to UNIT to use cols_to_concat
+  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
+  select(ID, Level, Spit, Code) %>% # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
   distinct(ID, .keep_all = TRUE)
@@
 xyz <- xyz %>%
-  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
+  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
   filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
@@
 T_dataset <- T_dataset %>% 
-  filter(!Level %in% c("2", "3")) %>%
-  drop_na(Level) %>% 
+  filter(!Level %in% c("2", "3")) %>% # Filtering out Upper levels
+  drop_na(Level) %>% # Dropping artifacts without level
   mutate(Level = case_when(Level == "7C" ~ "7",
@@
 
-Knappable raw materials play a crucial role in understanding the mobility and lifeways of past hunter-gatherers, given the ubiquitous presence of lithic artefacts throughout prehistory.
-More than mere rocks with which stone tools were produced, the management of these resources is intimately connected to a group’s technological, social, and cultural organization, potentially influencing the group’s overall survivability [@oestmoFormalModelingApproach2017; @binfordOrganizationFormationProcesses1979; @bleedOptimalDesignHunting1986; @bousmanHuntergathererAdaptationsEconomic1993; @torrenceTimeBudgetingHuntergatherer1983; @gouldLithicProcurementCentral1985].
-Several key topics in the study of hunter-gatherer behaviour and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccall_behavioral_2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallon_social_2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999].
-This is often achieved by systematically characterizing geological and archaeological raw materials and establishing correlations between samples from both origins.
-The ultimate aim is to pinpoint corresponding sources, enabling a precise differentiation between local, regional, and long-distance raw materials, as well as to examine their distribution within a site over different occupations and periods.
-These approaches and concepts have also been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula [@aubry_upper_2016; @aubryEconomyLithicRaw2009; @garcia-rojasGreatStepForward2021; @gomezdesolerPanadellaChertMontmaneu2020; @herrero-alonsoManagementLithicRaw2020; @matias_raw_2016; @ortegaMobilitatDesplacamentsDels2003; @pereiraLithicEconomyTerritory2016; @pereira_raw_2021; @pereira_flint_2022; @ramacciottiMovingLandFirst2022; @sanchezdelatorreNewDataChert2023; @soto_procurement_2016], and contributed to understanding how different groups organized their technology, managed time and risk, and explored the available resources during different chronological periods.
-
-Specifically, analyses of raw materials and the distinction between local/regional and long-distance sources have been instrumental in tracing extensive movements, contacts, and social networks throughout the Upper Palaeolithic (UP).
-Noteworthy examples include studies in northwestern Iberia (Galicia) during the Solutrean period [@hermidaSilexNWPeninsula2016], northern Iberia (Asturias) during the Magdalenian occupations at Las Caldas [@corchonrodriguezMobiliteTerritoiresRelations2016], northeastern Iberia (Catalunya) during the Aurignacian and Gravettian periods at Arbreda cave [@marreirosExploringLithicVariability2016; @ortegaMobilitatDesplacamentsDels2003], northern Portugal during the Gravettian/Solutrean periods in the Côa Valley [@aubry_upper_2016; @aubryRawMaterialProcurement2004], and inland central Iberia (Guadalajara) during the Proto-Solutrean and Solutrean periods at Peña Capón [@sanchezdelatorreNewDataChert2023].
-Although such contacts are commonly recognized and identified in later prehistoric periods [@costa_lithic_2022; @noceteRelacionesCentroPeriferia2005; @rodriguezPrehistoricFlintExploitations2011], the recognition of these connections has broadened our comprehension of the often extensive social networks and contacts during the UP in Iberia, illustrating a vast interconnected territory.
-
-However, in contrast to other areas of the Iberia, and despite the potential of raw material studies to shed light on the lifeways and organizational structures of Palaeolithic hunter-gatherers, research on the management and procurement of raw materials during the UP in southern and western Iberia remains in its initial stages.
-Specifically, in southern Portugal, even though there is considerable knowledge about the technological organization of Palaeolithic hunter-gatherer groups—largely due to research conducted at the Vale Boi archaeological site [see @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @marreirosLithicTechnologyGravettian2015]—studies focusing on raw materials are notably infrequent.
-This situation is attributed to the absence of a detailed and comprehensive characterization of regional lithic resources in the Algarve region (southern Portugal) and the subsequent creation of a reference collection, which is a critical component for conducting raw material studies.
+Knappable raw materials play a crucial role in understanding the mobility and lifeways of past hunter-gatherers, given the ubiquitous presence of lithic artefacts throughout Prehistory.
+More than mere rocks with which stone tools were produced, the management of these resources is intimately connected to a group’s technological, social, and cultural organisation, potentially influencing the group’s overall survivability [@oestmoFormalModelingApproach2017; @binfordOrganizationFormationProcesses1979; @bleedOptimalDesignHunting1986; @bousmanHuntergathererAdaptationsEconomic1993; @torrenceTimeBudgetingHuntergatherer1983; @gouldLithicProcurementCentral1985].
+
+Several key topics in the study of hunter-gatherer behaviour and organization have been explored in different geographic regions and chronological periods through raw material analysis, such as modalities of procurement and mobility strategies [@ambroseSocialEcologicalModels1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccallBehavioralEcologicalModels2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallonSocialNetworksInformation2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999].
+These analyses are often achieved by systematically characterizing geological and archaeological raw materials and establishing correlations between samples from both origins.
+The ultimate aim is to identify the corresponding sources, enabling, for example, a precise differentiation between local and non-local raw materials, and examine their distribution within a site over different occupations and periods.
+
+These approaches and concepts have also been extensively and successfully applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula [e.g., @aubryUpperPalaeolithicLithic2016; @aubryEconomyLithicRaw2009; @garcia-rojasGreatStepForward2021; @gomezdesolerPanadellaChertMontmaneu2020; @herrero-alonsoManagementLithicRaw2020; @matiasRawMaterialSourcing2016; @ortegaMobilitatDesplacamentsDels2003; @pereiraLithicEconomyTerritory2016; @pereiraRawMaterialProcurement2021; @pereira_flint_2022; @ramacciottiMovingLandFirst2022; @sanchezdelatorreNewDataChert2023; @sotoProcurementMobilityLate2016; @rodriguezPrehistoricFlintExploitations2011; @costa_lithic_2022; @noceteRelacionesCentroPeriferia2005], and contributed to understanding how different groups explored and managed the available lithic resources.
+
+In this context, raw material analyses and the distinction between local, regional and long-distance sources have been instrumental in tracing extensive movements, contacts, and social networks throughout the Upper Palaeolithic (UP) throughout Iberia.
+Noteworthy examples include studies in Northwestern Iberia (Galicia) during the Solutrean period [@hermidaSilexNWPeninsula2016], Northern Iberia (Asturias) during the Magdalenian occupations at Las Caldas [@corchonrodriguezMobiliteTerritoiresRelations2016], Northeastern Iberia (Catalunya) during the Aurignacian and Gravettian periods at Arbreda cave [@marreirosExploringLithicVariability2016; @ortegaMobilitatDesplacamentsDels2003], Northern Portugal during the Gravettian/Solutrean periods in the Côa Valley [@aubryUpperPalaeolithicLithic2016; @aubryRawMaterialProcurement2004], and inland Central Iberia (Guadalajara) during the Proto-Solutrean and Solutrean periods at Peña Capón [@sanchezdelatorreNewDataChert2023].
+Ultimately, the recognition of these connections has broadened our comprehension of mobility and social organization during the UP in Iberia, illustrating a vast interconnected territory.
+
+However, in contrast to other areas of Iberia, and despite the potential of raw material studies to shed light on the lifeways and organisational structures of Palaeolithic hunter-gatherers, research on the procurement and management of raw materials during the UP in Southern and Western Iberia remains in its initial stages.
+Specifically, in Southern Portugal, even though there is considerable knowledge about the technological organisation of Palaeolithic hunter-gatherers—largely due to research conducted at the Vale Boi archaeological site [see e.g., @belmiro_gravettian-solutrean_2021; @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @marreirosLithicTechnologyGravettian2015; @cascalheiraTecnologiaLiticaSolutrense2010; @horta_role_2019]—raw material studies are notably infrequent and use mainly macroscopic methodologies.
+
+This situation is attributed to the absence of a detailed and comprehensive characterization of regional lithic resources in the Algarve region (Southern Portugal) and the subsequent lack of a complete reference collection produced by complementing characterization methodologies, which is a critical component for conducting raw material studies.
 Such efforts are particularly crucial for distinguishing between local and non-local raw materials and for understanding variations within the archaeological record [@pop_simulating_2015; @surovellBehavioralEcologyLithic2009].
-Recently, we have made significant strides by introducing a reference collection and establishing a framework for raw material studies in the region, through a detailed macroscopic and petrographic database of chert resources in the Algarve region [@belmiroCreatingFramesReference2023].
-This initiative paves the way for systematic investigations into raw material procurement strategies during the Upper Palaeolithic, especially concerning the Vale Boi site.
-
-Vale Boi holds a pivotal role in the study of UP adaptations on the Iberian Peninsula, providing invaluable insights into the subsistence and technological strategies of hunter-gatherer populations [@belmiro_gravettian-solutrean_2021; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraTerritorialityOrganizationTechnology2019; @horta_role_2019; @manneIntensiveSubsistencePractices2012; @marreirosExploringLithicVariability2016; @pereira_territory_2016-1].
-The rich archaeological record from the site, indicative of prolonged human occupation, is essential for understanding raw material management practices in UP contexts.
-
-Preliminary studies at Vale Boi, employing macroscopic analysis, identified a predominant use of local cherts from Jurassic limestone formations, complemented by a lesser extent of non-local cherts throughout various UP occupations [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004].
-According to Pereira et al.
-[-@pereira_territory_2016-1], the local cherts likely originated within a 16 km radius, reflecting embedded procurement strategies.
-In contrast, the presence of non-local cherts points to extended sourcing networks, possibly involving travel to other regions of the Algarve, Alentejo, or southern Spain.
-The presence of chert from the Cretaceous formations of central Portugal has also been suggested [@bicho_ecodynamics_2013].
-
-These results highlight the potential for systematic raw material studies at the site, employing various methods and approaches to analyze the lithic assemblages from multiple UP occupations.
-Such research aims to comprehend the behaviours of these hunter-gatherer groups over time: how they navigated and utilized their landscape and lithic resources, their movements across a territory bordered by the ocean to the south and west, a large river to the east and mountain ranges to the north, and the nature of the social networks they established.
-Vale Boi stands as a key archaeological site for this study since it is currently the only site in southern Portugal with a long-term occupation spanning most of the UP and allowing the exploration of raw material use and mobility trends through time [@cascalheira_2017].
-Furthermore, previous studies have highlighted the role of Vale Boi in understanding different patterns of territory exploitation and contact with other regions of the Peninsula throughout the UP.
-During the Solutrean, Vale Boi has been interpreted as an interface between the several sites of Central Portugal and Southern Spain, making it a key site to identify extended social networks that promoted the exchange of information and culture [@cascalheiraValeBoiAlgarve2013; @cascalheira_2017; @cascalheiraSolutrenseEmPortugal2013].
-
-On the contrary, techno-typological studies from the Gravettian occupations of Vale Boi instead show marked differences between regions and suggest a more limited circulation of people and information in southern Iberia [@marreirosLithicTechnologyVariability2013].
-
-To investigate the applicability of these patterns to the management and circulation of lithic raw materials, and to examine the prevailing hypotheses concerning UP mobility and social network exchanges, this paper presents the first comprehensive characterization of chert raw materials derived from the Gravettian, Proto-Solutrean, and Solutrean (c.
-32-19 ka cal BP) sequence at Vale Boi.
+
+Recently, we have made significant strides by introducing such a reference collection and establishing a framework for raw material studies in the region, through a detailed macroscopic and petrographic database of chert resources in the Algarve region [@belmiroCreatingFramesReference2023].
+This initiative paves the way for systematic investigations into raw material procurement strategies during the Upper Palaeolithic, especially concerning the Vale Boi site, where previous raw material studies have already shown complex patterns of chert procurement and management.
+These preliminary studies at Vale Boi, employing only macroscopic methods, pointed to a predominant use of local cherts [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004], likely originating within a 16 km radius and reflecting embedded procurement strategies [@pereira_territory_2016-1].
+To a lesser extent, non-local cherts were also identified throughout the various UP occupations [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004], pointing to extended sourcing networks, possibly involving sourcing from other regions of Southern Portugal and Southern Spain.
+The presence of chert from the Cretaceous formations of Central Portugal has also been suggested [@bicho_ecodynamics_2013].
+
+These results highlight the potential for systematic raw material studies at the site to shed light on the lithic resource procurement strategies, mobility patterns, and social networks of the hunter-gatherer groups of Southwestern Iberia throughout the UP.
+
+Vale Boi stands as a key archaeological site for this study since it is currently the only site in Southern Portugal with a long-term occupation spanning most of the UP and allowing the exploration of raw material procurement and mobility trends through time [@cascalheira_2017].
+It also holds a pivotal role in the study of UP adaptations on the Iberian Peninsula, providing invaluable insights into the subsistence and technological strategies of hunter-gatherer populations [@belmiro_gravettian-solutrean_2021; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraTerritorialityOrganizationTechnology2019; @horta_role_2019; @manneIntensiveSubsistencePractices2012; @marreirosExploringLithicVariability2016; @pereira_territory_2016-1].
+
+Furthermore, previous studies have highlighted the role of Vale Boi in understanding different patterns of territory exploitation and contact with other regions of Iberia throughout the UP.
+During the Solutrean, Vale Boi has been interpreted as an important connection point between the core Solutrean areas of Central Portugal and Southern Spain, making it a key site for identifying extended social networks that promoted the exchange of information and culture [@cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @cascalheiraSolutrenseEmPortugal2013].
+
+On the contrary, techno-typological studies from the Gravettian occupations of Vale Boi instead show marked differences between regions and suggest a more limited circulation of people and information in Southern Iberia [@marreirosLithicTechnologyVariability2013].
+
+Based on these interpretations and studies, a set of hypotheses and expectations regarding chert procurement and human mobility throughout the UP at Vale Boi can be proposed.
+On one hand, it would be expected that Solutrean occupations are characterized by the predominant use of local raw materials but with a considerable percentage of cherts obtained from long-distance sources, possibly through exchange with different groups in Iberia.
+Conversely, it would be expected that Gravettian occupations would be mostly composed of local cherts, with a more limited presence of long-distance cherts.
+
+To investigate the applicability of these patterns to the procurement and circulation of lithic raw materials, and to examine the prevailing hypotheses concerning UP mobility and social network exchanges, this paper focuses on the following questions: 1) what types of chert were the hunter-gatherers of Vale Boi using and where did they come from?; 2) are there different patterns of chert procurement throughout time?; 3) and if so, how can these be related to patterns of land use and the circulation of people/ideas, as suggested through technological studies of the same lithic assemblages?
+As a result, our study presents the first comprehensive characterisation of chert raw materials, through macroscopic and microscopic analysis, of UP occupations in Southwestern Iberia, derived from the Gravettian, Proto-Solutrean, and Solutrean (c.
+32-19 ka cal BP) occupations at Vale Boi.
 
@@
 It is bordered by limestone outcrops that form rock shelters facing west and southwest [@bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011].
-The site extends for more than 10,000 sq meters along the slope of the valley, through which three main areas have been excavated between 2000 and 2019: Slope, Shelter, and Terrace.
+The site extends for more than 10,000 sq meters along the slope of the valley, through which three main areas were excavated between 2000 and 2019: Slope, Terrace, and Shelter.
+For this study, the chosen areas were the Terrace and Shelter areas (see Section 2 Materials and Methods).
 
 The Terrace encompasses occupations from the UP to the Early Neolithic, resulting in the identification of eight main lithostratigraphic units.
-The UP sequence includes several occupations attributed to the Gravettian (levels 8 to 6) between 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between 26 and 24 ka cal BP, and Solutrean (level 4D, 4C, 4C, 4 and 3) between 24 and 20 ka cal BP [@belmiro_gravettian-solutrean_2021; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017].
+The UP sequence includes several occupations attributed to the Gravettian (levels 8 to 6) between c.
+32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between c.
+26 and 24 ka cal BP, and Solutrean (level 4D, 4C, 4C, 4 and Lower 3) between c.
+24 and 20 ka cal BP [@belmiro_gravettian-solutrean_2021; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017].
+
 The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations.
-From these, the Solutrean levels show the most intensive occupation, with three, well-preserved, archaeological horizons (layers C to A) and dated from 24 to 22 ka cal BP [@cascalheiraValeBoiAlgarve2013; @cascalheiraTecnologiaLiticaSolutrense2010; @cascalheira_2017].
-
-```{r}
-#| label: tbl-c14
-#| tbl-cap: Radiocarbon dates for Vale Boi Upper Paleolithic sequence of the Terrace area and Solutrean of the Shelter area, partially adapted from Cascalheira et al.
-[-@cascalheira_2017] and Belmiro et al.
-[-@belmiro_gravettian-solutrean_2021], and including new dates.
-All dates were calibrated or recalibrated using the IntCal20 and Marine20 curves [@reimerIntCal20NorthernHemisphere2020; @heatonMarine20MarineRadiocarbon2020] in OxCal online v.4.4.4 [@ramseyBayesianAnalysisRadiocarbon2009].
-For shells a DR value of 123 ± 58 was used, obtained through the CHRONO Marine Reservoir Database at http://calib.org/marine/.
-
-c14_tbl <- read_csv(here::here('analysis/data/raw_data/C14_VB_calBP.csv'))
-c14_20 <- read_csv(here::here('analysis/data/raw_data/IntCal20_Marine20_Oxcal.csv'))
-dates_cal <- left_join(c14_tbl, c14_20, by = "Lab-ref")
-dates_cal <- dates_cal %>% 
-  dplyr::rename("Date Cal BP" = "Concatenate") %>% 
-  dplyr::select(Area, Level, Culture, `Lab-ref`, Date, SD, `Date Cal BP`, Material)
-
-kable(dates_cal, "simple")
-
-```
+From these, the Solutrean levels show the most intensive occupation, with three, well-preserved, archaeological horizons (layers C to A) and dated to c.
+24-22 ka cal BP [@cascalheiraValeBoiAlgarve2013; @cascalheiraTecnologiaLiticaSolutrense2010; @cascalheira_2017].
 
@@
 #| fig-cap: Iberian Peninsula map with locations and geological formations mentioned throughout the paper.
-a) Detail of the central Portugal region where the Cenomanian formations with chert nodules are located.
-b) Detail of southwestern Portugal, with chert and chalcedony bearing formations, outcrops, and location of Vale Boi.
-c) Detail of eastern Algarve with chert bearing formations.
+a) Detail of the Central Portugal region where the Cenomanian formations with chert nodules are located.
+b) Detail of Southwestern Portugal, with chert and chalcedony bearing formations, outcrops, and location of Vale Boi.
+c) Detail of Eastern Algarve with chert-bearing formations.
 The figure was produced using ArcGIS Pro 3.2.
@@
 
-The Algarve region is a complex territory, marked by several geomorphic sub-regions and geological units that provide different lithological resources used throughout prehistory.
-Two of the main geological units which characterize the Algarve are: 1) the Baixo Alentejo Flyschoid Group, part of the South Portuguese Zone [@fernandes_new_2012; @oliveira_carta_1984-1; @onezimeNewGeodynamicInterpretation2003], located in the north sector of the region, and characterized by schists, greywackes and quartzites; and 2) the Algarve basin in the south, which is characterized by limestones, dolomites and marls [@terrinha_bacia_2006].
-It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules and beds can be found [@manuppellaNoticiaExplicativaFolha1987; @manuppellaNoticiaExplicativaFolha2007; @ribeiro_evolucao_2005; @rocha_noticia_1979; @rocha_noticia_1983; @terrinha_bacia_2006].
-Following Luedtke [-@luedtke_archaeologists_1992], we use the term chert to refer to sedimentary rocks composed primarily of microcrystalline quartz, which includes other possible varieties such as jasper, agate, or chalcedony.
+The Algarve region is a complex territory, marked by several geomorphic sub-regions and geological units that provide different lithological resources.
+Two of the main geological units which characterize the Algarve are: 1) the Baixo Alentejo Flyschoid Group, part of the South Portuguese Zone [@fernandesNewEvidenceConcerning2012; @oliveira_carta_1984-1; @onezimeNewGeodynamicInterpretation2003], located in the north sector of the region, and characterized by schists, greywackes and quartzites; and 2) the Algarve basin in the south, which is characterized by limestones, dolomites and marls [@terrinha_bacia_2006].
+It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules/beds can be found [@manuppellaNoticiaExplicativaFolha1987; @manuppellaNoticiaExplicativaFolha2007; @ribeiro_evolucao_2005; @rocha_noticia_1979; @rocha_noticia_1983; @terrinha_bacia_2006].
+
+The availability of regional resources containing cherts and the characterization of their lithologies has already been published by our team [@belmiroCreatingFramesReference2023], and we herein present the summarized main findings.
+The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of Westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b).
+These cherts are frequently opaque with yellow, grey, purple, or red colours, occasionally with bands or laminations, and with frequently visible sponge spicules.
+Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being \~20 cm, and frequently characterized by high levels of dolomitization and fractures.
+
+Chert sources were also identified in two formations from the eastern area of the Algarve (@fig-map c).
+The Middle Jurassic Malhão formation chert outcrops were identified at around 80-100 km east of the site and characterized by chert nodules with similar macroscopic features to those found in Western Algarve.
+The Upper Jurassic Jordana formation chert outcrops were identified at around 90 km east of Vale Boi and characterized by chert nodules with significantly different macroscopic features to those of other formations, with frequently small sizes (\~5 cm), difficult removal from the parent rock and unidentified secondary deposits.
+
+Chalcedony outcrops \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes (Bicho et al., 2013).
+The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.
+
+# Materials and methods
+
+To characterize the siliceous raw materials used during the UP occupations at Vale Boi and to identify their geological sources, we focused on chalcedony and chert archaeological materials from the Terrace and Shelter areas.
+These two areas were chosen since both have detailed spatial information, well-defined chronological sequences, and previous studies providing lithic technology, faunal, and geoarchaeological data.
+
+In this study, the terms chert and chalcedony follow the definitions provided by Luedtke (1992).
+We use the term chert to refer to sedimentary rocks composed primarily of microcrystalline quartz, which includes other possible varieties such as jasper, agate, or chalcedony.
 The term chalcedony is used in its petrographic sense and refers to cherts with a fibrous quartz structure.
-The availability of regional resources containing cherts and the characterization of their lithologies has already been published by our team [@belmiroCreatingFramesReference2023], and we herein present the summarized main findings.
-The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b).
-These cherts are frequently opaque with yellow, grey, purple, or red colours, occasionally with bands or laminations, and with frequently visible sponge spicules.
-Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being around 20 cm, and frequently characterized by high levels of dolomitization or the presence of fractures.
-Chert sources were also identified in two formations from the eastern area of the Algarve (@fig-map c).
-The Middle Jurassic Malhão formation chert outcrops were identified at around 80-100 km east of the site and characterized by chert nodules with similar macroscopic features to those found in western Algarve.
-The Upper Jurassic Jordana formation chert outcrops were identified at around 90 km east of Vale Boi and characterized by chert nodules with significantly different macroscopic features to those of other formations, also with frequently small sizes (\~5 cm), difficult removal from the parent rock and nonexistence of secondary deposits.
-Chalcedony outcrops approximately 10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes (Bicho et al., 2013).
-The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.
-
-# Materials and methods
-
-To characterize the siliceous raw materials used during the UP occupations at Vale Boi and to identify their geological sources, we focused on chalcedony and chert archaeological materials from the Terrace and Shelter areas.
-These two areas were chosen since both have detailed spatial information, several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal, and geoarchaeological data.
+
 Chert was chosen for this study as it is one of the three main used raw materials at the archaeological site of Vale Boi alongside quartz and greywacke.
 Other raw materials such as chalcedony, dolerite, and schists have also been identified [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @pereira_territory_2016-1].
-Using the field database with all individually coordinated lithic artefacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials.
-@fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents between \~50-70% of the lithic assemblage from the Solutrean layers A, B, and C.
-Based on previous lithic studies in the terrace and shelter areas, the remaining 80% of other raw materials are mostly composed of quartz and greywacke [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009].
+
+Using the field database with all individually coordinated lithic artefacts (all pieces larger than \~2 cm), we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials.
+@fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents \~50-70% of the lithic assemblage from the Solutrean layers A, B, and C.
+Based on previous lithic studies in the Terrace and Shelter areas, the remaining 80% of other raw materials are mostly composed of quartz and greywacke [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009].
+
 Although systematic raw material research has not been conducted on these materials, according to the regional geological context, they have been interpreted as local and readily accessible in the vicinity of the site [@pereira_territory_2016-1].
-Frequently, the quartz and greywacke assemblages include large amounts of shattered and/or unknapped chunks or slabs.
-Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020].
-Based on this, despite the low amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.
+Frequently, the quartz and greywacke assemblages include large amounts of shattered and unknapped chunks or slabs.
+Previous studies from the Terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020].
+Based on this, despite the smaller amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.
 The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].
@@
     TRUE ~ Layer)) %>% 
-  filter(!Layer %in% c("D", "Z")) # Choosing only Solutrean occupation layers (A, B and C)
+  filter(!Layer %in% c("D", "Z", "No info")) # Choosing only Solutrean occupation layers (A, B and C)
 
@@
 
-Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustillo_is_2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=5318).
+Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustilloMacroscopicClassificationFlint2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=5318).
 The distinct features of various local cherts facilitate their differentiation from cherts of neighbouring origins.
-Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustillo_is_2009; @tarrino_materias_2013].
+Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustilloMacroscopicClassificationFlint2009; @tarrino_materias_2013].
 Therefore, we implemented an initial phase of macroscopic analysis, subsequently augmented by petrographic study to diminish the subjectivity of the analysis and enhance the material characterization with petrographic data.
-The synergistic application of these two methods has been previously validated in the characterization of Algarve region cherts, following recent prospection efforts [@belmiroCreatingFramesReference2023].
-Hence, an extensive macroscopic and petrographic collection is available for comparative analysis with archaeological materials, employing consistent methodologies.
-
+The synergistic application of these two methods has been previously validated in the characterization of Algarve region cherts, following recent geological survey efforts [@belmiroCreatingFramesReference2023].
+Hence, an extensive macroscopic and petrographic collection was already available for comparative analysis with archaeological materials, employing consistent methodologies.
 This reference collection is the Lusolit lithotheque, hosted physically at ICArEHB (University of Algarve, Faro) and digitally at www.lusolit.icarehb.com, and aided in the characterization and source attribution of cherts.
+
 Hand samples were used throughout the macroscopic characterization for comparison with the archaeological materials, and the archaeological thin sections were compared with the previously studied geological chert thin sections.
-To identify the source of other lithologies which are not congruent with the regional cherts and identify possible long-distance chert procurement, we used hand samples and thin sections of cherts from central Portugal also available in the Lusolit lithotheque.
-The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz (UCA) was also visited to understand the macroscopic variability of the local and regional cherts of the lower Guadalquivir basin, within the Atlantic strip of south-western Spain (present-day provinces of Huelva and Cádiz).
+To identify the source of other lithologies which are not congruent with the regional cherts and identify possible long-distance chert procurement, we used hand samples and thin sections of cherts from Central Portugal also available in the Lusolit lithotheque.
+The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz (UCA) was also visited to understand the macroscopic variability of the local and regional cherts of the lower Guadalquivir basin, within the Atlantic strip of Southwestern Spain (present-day provinces of Huelva and Cádiz).
+
 Following previous studies, we classify the distance of chert sources from the site as local (1-30 km from the site), regional (30-120 km), and long-distance (\>120 km) [@herrero-alonsoManagementLithicRaw2020; @tarrinoFlintRawMaterial2015].
@@
 The first step included a preliminary characterization using a hand lens of 10x magnification.
-This step allowed us to create types and subtypes based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these include characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992], and a data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1).
-In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types.
-The data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1).
-Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis is simplified and focused on collecting data related to weight, cortex, and alterations.
-This first step was applied to a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 8 to 4) and 2805 chert artefacts from the Solutrean sequence of the Shelter area (layers C to A).
-
-While the advantages of employing a macroscopic methodology for analysing the cherts of Vale Boi are notable, a limitation of this approach is its inability to unequivocally characterize and categorize artefacts that exhibit surface alterations typical of an open-air site, including those related to fire, since weathered and altered surface may destroy visible structures and even alter the chemical composition [@delluniversita_development_2019].
-Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversita_development_2019; @gomezdesolerPanadellaChertMontmaneu2020; @soto_procurement_2016].
+This step allowed us to create types and subtypes based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1).
+In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types - the data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1).
+Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis was simplified and focused on collecting data related to weight, cortex, and alterations.
+This first step was applied to a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 8 to 4) and 1665 chert artefacts from the Solutrean sequence of the Shelter area (layers C to A).
+
+While the advantages of employing a macroscopic methodology for analysing the cherts of Vale Boi are notable, a limitation of this approach is its inability to unequivocally characterize and categorize artefacts that exhibit surface alterations typical of an open-air site, including those related to fire, since weathered and altered surface may destroy visible structures and even alter the chemical composition [@delluniversitaDevelopmentMultiparametricCharacterisation2019].
+Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversitaDevelopmentMultiparametricCharacterisation2019; @gomezdesolerPanadellaChertMontmaneu2020; @sotoProcurementMobilityLate2016].
 For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), they were analysed using the individual sample dataset, but grouped in a category of indeterminate cherts (INDET).
 
-As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artefacts, corresponding to 40% of the total sample (n=1027).
-This is due to the frequent alterations of the sample, especially related to patinas and alteration rinds, pits, and fire/heat alterations.
+As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artefacts, corresponding to \~40% of the total sample (n=1027).
+This is due to the frequent alterations of the artifacts, especially related to patinas and alteration rinds, pits, and fire/heat alterations.
 For all other levels in the Terrace, indeterminate represents on average \~28% of the chert artefacts, while in the Shelter the frequency of INDET is slightly lower (\~20-25%).
@@
 
-To better characterize the different chert types identified in the macroscopic analysis and aid in minimizing the inherent caveats of macroscopic methods, twenty-two thin sections were prepared at the Thin Section Lab (Toul, France).
+As previously mentioned, the macroscopic study was followed by a petrographic approach.
+To better characterize the different chert types identified in the macroscopic analysis and aid in minimizing the inherent caveats of macroscopic methods, 22 thin sections were prepared at the Thin Section Lab (Toul, France).
 The samples chosen for this analysis include at least one sample from each type present throughout the stratigraphy of the Terrace area, from different archaeological levels (@tbl-TS-list).
 For types that showed high macroscopic variability (e.g.
-translucency and colour variation or differentiated bioclast facies), thin sections from sub-types were analysed to collect petrographic data that reflected this variability.
-Petrographic thin sections were analysed by polarized light microscopy (Nikon LV100ND) to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations.
+translucency and colour variation or differentiated bioclast facies), thin sections from sub-types were analysed to collect petrographic data that could reflect this variability.
+Petrographic thin sections were analysed by polarized light microscopy (Nikon LV100ND) to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any other alterations.
 
@@
 
-The complete petrographic analysis for all thin sections studied in this paper (including macroscopic and microscopic figures) can be found as Supplementary Information Materials at our online repository (++).
+The complete petrographic analysis for all thin sections studied in this paper (including macroscopic and microscopic figures) can be found at our online research compendium (https://doi.org/10.17605/OSF.IO/DBFT2).
 A detailed macroscopic and petrographic description and figures for each type and sub-type of chert and chalcedony (Online Resource 2) and the data dictionary used for the petrographic analysis (Online Resource 1) can be found in Supplementary Information.
@@
 
-The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online research compendium (https://doi.org/10.17605/OSF.IO/DBFT2).
+The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online research compendium.
 We used the rrtools package by Marwick et al.
 [-@marwickPackagingDataAnalytical2018] to create a research compendium and write a reproducible journal article.
-The provided files include the complete set of raw data used in the analysis, along with a custom R project (Wickham, 2015) containing the code required to generate all tables and figures.
+The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickhamPackagesOrganizeTest2015] containing the code required to generate all tables and figures.
 To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY [additional details can be found in @marwickComputationalReproducibilityArchaeological2017].
@@
 The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts (@fig-local-types).
-The chalcedony (Type 1; @fig-local-types a) is identified as a massive, fibrous variant devoid of allochems (@fig-local-types a2).
+The chalcedony (Type 1; @fig-local-types a) is identified as a massive, fibrous variety without allochems (@fig-local-types a2).
 It represents 6.6% of the overall Terrace cherts (excluding INDET) and 2.6% in the Shelter.
@@
 Whenever present, the cortex seems to show a mixture of irregular cortex and rounded cortex, both from an outcrop source (@tbl-cortex_tab).
-Petrographically, however, these cherts are homogeneous, except for a small number of artefacts with a banded variety within type 5.
-The macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.
+Petrographically, however, these cherts are homogeneous, except for a small number of artefacts with a banded variety within Type 5.
 These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite.
@@
 In all samples, fossils range from common to very frequent, albeit frequently unidentifiable.
-When identifiable in thin section, they are sponge spicules (@fig-local-types b2, c2, h2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (@fig-local-types h3) in a sample from type 5.
+When identifiable in thin section, they are sponge spicules (@fig-local-types b2, c2, h2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (@fig-local-types h3) in a sample from Type 5.
 Porosity varies between 10% and 1%, of vug type.
+Despite the petrographic homogeneity between local types, this macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.
+
 Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiroCreatingFramesReference2023], we interpret Types 2, 3, 4, and 5 to be local, as they are congruent with the described characteristics representative of cherts from the Lower Jurassic (Carixian) and Middle Jurassic Malhão formations of the Algarve basin.
@@
 
-We identified six chert types with macroscopic and petrographic characteristics that are not congruent with the chert available in known outcrops in western Algarve (@fig-distance-types).
-Despite their scarce representation in the archaeological record, these raw materials of presumably allochthonous origin are a valuable source of information.
-Often these lithological types can be considered as clear lithological markers of mobility, which can contribute to support hypotheses about mobility and contacts with neighbouring populations and territories.
+We identified six chert types with macroscopic and petrographic characteristics that are not congruent with the chert available in known outcrops in Western Algarve (@fig-distance-types).
 
@@
 #| label: fig-distance-types
-#| fig-cap: Variability of non-chert types and main petrographic characteristics.
+#| fig-cap: Variability of non-local types and main petrographic characteristics.
 a-b) Type 6 macroscopic appearance (a1-b1) and thin-section (a2-b2); b1-b2 correspond to a sample covered 100% in white patina.
@@
 Type 6 (@fig-distance-types a-b) corresponds to a massive cryptocrystalline quartz with a mudstone texture, from a marine environment.
-It represents \~18% of the overall Terrace cherts and 12% of the chert in the Shelter.
+It represents \~18% of the overall Terrace cherts and \~12% of the chert in the Shelter.
 This chert is translucent and shows a variety of colours, ranging from brown, red, and grey, which seem to vary in function of the artefact’s thickness.
 In unaltered samples, visible bioclasts are rare.
-The cortex is mostly rounded, and rarely pebble-like (FIG), ranging from thin to medium thickness.
+The cortex is mostly rounded, and rarely pebble-like, ranging from thin to medium thickness.
 Two sub-types were individualized based on macroscopic characteristics such as colour, translucency variation, and patterns, although they may simply reflect the use of specific nodules from within Type 6’s variability.
@@
 
-The similarities between chert artefacts at the UP occupations of Vale Boi and the chert from the Rio Maior region (Estremadura Portuguesa; @fig-map a) have been previously suggested although without systematic studies to ascertain the attribution [@bicho_ecodynamics_2013; @pereira_territory_2016-1].
+The similarities between chert artefacts at the UP occupations of Vale Boi and the chert from the Rio Maior region (Central Portugal; @fig-map a) have been previously suggested although without systematic studies to ascertain the attribution [@bicho_ecodynamics_2013; @pereira_territory_2016-1].
 These Cretaceous cherts, mostly found in secondary deposition settings, are described as frequently translucent, with geodes, colours ranging from yellow to red or grey, and mineralogically homogeneous between them.
-Petrographic studies from the Estremadura area highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matias_raw_2016].
+Petrographic studies from Central Portugal highlight the presence of iron oxide accumulations and rare fossil ghosts or frequently difficult to see in thin section [@matiasRawMaterialSourcing2016].
 When identified, sponge spicules and possible rare foraminifera have been identified [@matiasAprovisionamentoMateriasprimasLiticas2012].
+
 Studies of Upper Cretaceous cherts located in the Lisbon area describe cherts with similar macroscopic characteristics (@fig-map a), with micro-cryptocrystalline quartz mudstones with bioclasts such as sponge spicules or bivalve fragments, or wackestone/packstone textures with abundant bioclasts, including ostracods and different types of foraminifers [@jordaoFlintSourcesMobility2022].
-These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from central Portugal hosted at LusoLit.
-Alike the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matias_raw_2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
-From these cherts, only the dark grey nodules (RT231; SOM) showed identifiable fossils which were rare ostracods.
-Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Type 6 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts (Upper Cenomanian formations) from central Portugal, between \~215-250 km north of Vale Boi (@fig-map a).
+These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from Central Portugal hosted at LusoLit.
+Alike the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matiasRawMaterialSourcing2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
+From these cherts, only the dark grey nodules (RT231) showed identifiable fossils which were rare ostracods.
+
+Based on the macroscopic and petrographic similarities to the geological references from Central Portugal and congruence with previous studies from this area, we interpreted Type 6 (and sub-types) as belonging to the macroscopically variable Cretaceous cherts (Upper Cenomanian formations) from Central Portugal, between \~215-250 km (as the crow flies) north of Vale Boi (@fig-map a).
 
@@
 Porosity is variable (\<1-10%).
+
 The consultation of the lithotheque hosted at UCA allowed us to identify a subset of cherts with similarities to those from Type 7 from the Middle Subbetic region (Betic Systems, south Spain, @fig-map).
@@
 The presence of bioclasts varies between samples, ranging between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon), and echinoderm spines (rare).
-Similar cherts have been identified and described in previous works, describing grey cherts with peloidal and oolitic facies and occasional foraminifera [@rodriguezPrehistoricFlintExploitations2011].
-Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allow us to interpret Type 7 as coming from the Upper Jurassic formations of the Betic Systems, located in the south and southeastern Iberian Peninsula, at least 250 km east from the site (@fig-map).
+Similar cherts have also been identified and described in previous works, describing grey cherts with peloidal and oolitic facies and occasional foraminifera [@rodriguezPrehistoricFlintExploitations2011].
+Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allow us to interpret Type 7 as coming from the Upper Jurassic formations of the Betic Systems, located in the south and southeastern Iberian Peninsula, at least 250 km (as the crow flies) east from Vale Boi (@fig-map).
 
@@
 Fossil content is very frequent (@fig-distance-types f2-f3), albeit mostly unidentifiable and poorly preserved.
-Identified fossils are rare sponge spicules, rare ostracods with some degree of preservation, common small, unidentified foraminifera (100 μm) of differing preservation degrees, and as aforementioned, foraminifera from the Pfenderenidae family (500 μm).
+Identified fossils are rare sponge spicules, rare ostracods with some degree of preservation, common small, unidentified foraminifera (100 μm) of differing preservation degrees, and as mentioned above, foraminifera from the Pfenderenidae family (500 μm).
 Despite the macroscopic similarities between type 8 and cherts from the Middle Subbetic region chert formations (@fig-map), the geological reference samples showed no macroscopically visible foraminifera.
 However, our results indicate that despite no foraminifera being visible macroscopically, they are present in thin section.
-The fact that no foraminifera were identified in the small number of analysed samples from the Cadiz lithotheque may simply reflect their uncommon visibility, possibly due to patination.
-Another possible source may be the Upper Cretaceous wackestone/packstone cherts with foraminifers from the Lisbon region [@jordaoFlintSourcesMobility2022].
+The fact that no foraminifera were identified in the small number of analysed samples from the UCA lithotheque may simply reflect their uncommon visibility, possibly due to patination.
+Another possible source for Type 8 cherts may be the Upper Cretaceous wackestone/packstone cherts with foraminifers from the Lisbon region [@jordaoFlintSourcesMobility2022].
 However, these studies do not specify the existence of Pfenderenidae foraminifers.
-Given the lack of this specific fossil in the reference samples from the Andalusian chert \~250 km east and southeast, and the central Portugal cherts 215-250 km north, the probable source of Type 8 remains unknown.
+Given the lack of this specific fossil in the reference samples from the Betic chert \~250 km east and southeast, and the Central Portugal cherts 215-250 km north, the probable source of Type 8 remains unknown.
 
@@
 It is opaque, has a heterogeneous structure and is characterized by a variable colour distribution: it shows a horizontal laminated and finely laminated pattern, with grey dark bands intercalated with light grey bands (@fig-distance-types g1).
-Whenever present, the cortex is rounded (FIG) and is generally thin.
+Whenever present, the cortex is rounded and is generally thin.
 In specific samples, the cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers.
@@
 Porosity is low (1%).
-Although its origin is unknown, when visiting the Cadiz lithotheque, we identified a small type of banded black cherts without noticeable fossils or inclusions.
-Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Type 9 belongs to Andalusian chert formations in southern Spain, 250 km east from the site (@fig-map).
+Although its origin is unknown, when visiting the UCA lithotheque, we identified a small type of banded black cherts without noticeable fossils or inclusions.
+Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Type 9 belongs to Betic chert formations in Southern Spain, 250 km east (as the crow flies) from the site (@fig-map).
 
@@
 The ooids are distributed homogeneously across the sample (although macroscopically visible in the brown fabric) and are highly abundant and concentrated.
-They are poorly sorted, varying between 500 and 20 μm (@fig-distance-types h2).
-They show round-to-elliptical shapes, and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010].
+They are poorly sorted, varying between 500 and 20 μm (@fig-distance-types h2), show round-to-elliptical shapes and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010].
 This chert shows no similarities to any consulted reference material.
-Despite the existence of oolitic limestones in the Algarve region correspondent to Middle and Upper Jurassic formations and often described as whiteish (Rocha et al., 1979), the presence of oolitic chert in the region is not recognized in geological and archaeological literature [@cardoso_sobre_2018; @rocha_noticia_1979; @verissimo_jazidas_2004] and the visit of outcrops mentioned in the geological cartography did not allow their identification.
+Despite the existence of oolitic limestones in the Algarve region correspondent to Middle and Upper Jurassic formations and often described as whiteish (Rocha et al., 1979), the presence of oolitic chert in the region is not recognized in geological and archaeological literature [@cardosoSobrePresencaLaminas2018; @rocha_noticia_1979; @verissimo_jazidas_2004] and the visit of outcrops mentioned in the geological cartography did not allow their identification.
 
@@
 This chert type is grey, opaque, and without identifiable allochems.
-This chert shows a limited presence at the archaeological site as previously mentioned, constricted to the Solutrean occupations, and corresponding to less than 0.1% of the Terrace cherts and \~2% of Shelter cherts.
+This chert shows a limited presence at the archaeological site as previously mentioned, restricted to the Solutrean occupations, and corresponding to less than 0.1% of the Terrace cherts and \~2% of Shelter cherts.
 
@@
 These are composed mostly of blanks or retouched tools, rarely with cortex, and rarely the same TL type is found in different archaeological layers.
-A detailed description of these types can be found in the SOM (SOM).
+A detailed description of these types can be found in the Supplermentary Information (Online Resource 2).
 In both areas, the TL cherts compose less than 1% of the cherts.
 
-Apart from the types common throughout the stratigraphy of the Terrace and Shelter, Type 11 and samples from the TL type also show no identifiable source.
-The exceptions are TL01, TL10-11 and TL15-16.
-Jasper is known in the South Portuguese Zone area [@oliveira_carta_1984-1], from south/central Portugal to Spain, above the Algarve (from Aljustrel to Beja; @fig-map), suggesting that the source of TL01 may be from south/central Portugal at a minimum distance of 100 km north and northeast from Vale Boi.
+Apart from the types common throughout the stratigraphy of the Terrace and Shelter, Type 11 and samples from the TL type also show no identifiable source. The exceptions are TL01, TL10-11 and TL15-16.
+
+Jasper is known in the South Portuguese Zone area [@oliveira_carta_1984-1], from Southern/Central Portugal to Spain, north of the Algarve (from Aljustrel to Beja; @fig-map), suggesting that the source of TL01 may be from south/central Portugal at a minimum distance of 100 km north and northeast from Vale Boi (as the crow flies).
+
 Similarly, despite being different from the most frequent Cretaceous nodules, the macroscopic similarities between TL10 and TL11 and specific nodules from Central Portugal in the Lusolit lithotheque may indicate these cherts are from the Estremadura/Lisbon area (@fig-map a).
 This difference in appearance may be due to the already-mentioned variability of the nodule’s macroscopic characteristics.
-Finally, TL15 and TL16 show macroscopic similarities to radiolarite reference samples located at the Cadiz lithotheque, which may indicate the source of these cherts is from the Middle Subbetic region [@dominguezbellaMineralesRocasSociedades2010; @dominguez-bella_estudio_2006].
+
+Finally, TL15 and TL16 show macroscopic similarities to radiolarite reference samples located at the Cadiz lithotheque, which may indicate the source of these cherts is from the Middle Subbetic region [@dominguezbellaMineralesRocasSociedades2010; @dominguez-bellaEstudioMateriasPrimas2006].
 
@@
 
-Most chert types are consistently present across the Terrace stratigraphy (except for Type 11 which appears only in level 4 and in a small number of artefacts; n=2), although there are statistically significant differences in chert types used throughout the different levels (p-value \<2.22e-16, FIG.).
-In the Solutrean (levels 4, 4C, and 4D) and Proto-Solutrean (levels 4E and 4E/5) occupations, local cherts make up more than \~70% of the chert and 80% when considering the weight.
+Most chert types described above are consistently present across the Terrace stratigraphy (except for Type 11 which appears only in level 4 and in a small number of artefacts; n=2).
+However, being ubiquitous, we identified a difference in the percentages of chert types throughout the stratigraphy.
+As such, a chi-square test of independence was performed to determine whether the percentages of chert types are related to the archaeological layer.
+The results are significant, X-square = 817.79, p \< 2.2e-16, showing that there are statistically significant differences in the chert types present throughout the occupations.
+A full description of chi-square test and Pearson residuals can be found in the Supplementary Information (Online Resource 4).
+
+In the Gravettian occupations (levels 6 and 7) and at the start of the Proto-Solutrean (level 5) local cherts make up \~45-60%, while the combined non-local cherts make up \~65-40%.
+Level 6 shows the highest percentages of all non-local chert use, with specific types like Type 6 representing \~30%, although when considering the sum of weight, the percentages reduce, and non-local chert represents \~50% of total chert.
+The percentages of local chert increase in the following occupations.
+In the Proto-Solutrean (levels 4E and 4E/5) and Solutrean (levels 4, 4C, and 4D) local cherts make up more than \~70% of the sample and 80% when considering the sum of weight, while non-local cherts decrease significantly and represent percentages lower than 30%.
 Despite this homogeneous pattern, local types show statistically significant differences between the two technocomplexes.
-In the Solutrean levels, the yellow/brown cherts (Type 2) are present in the highest percentage, at \~30%.
-In contrast, the grey/white cherts (Type 4) become the most used chert with percentages above 35% in level 4E and \~20% in the top of level 5.
-For both technocomplexes, chert non-congruent with the known local sources represent percentages lower than 30%.
-This pattern seems to shift at the start of the Proto-Solutrean horizon (level 5) and the Gravettian occupations (levels 6 and 7); in these levels, the percentages of local cherts drop to 45-60%, while the percentages of specific non-local chert types increase significantly, such as Type 6 (increasing from \~10% during the Solutrean to 30%).
-However, when considering the sum of the weight, the percentages of Type 6 diminish, instead representing around 20% of the chert use, and bringing in level 6 the use of local chert to 50% in comparison to chert types of non-local sources.
+During the top Proto-Solutrean occupations the grey/white cherts (Type 4) are the most present chert with percentages between 35-20%, while in the Solutrean levels, the yellow/brown cherts (Type 2) are present in the highest percentage, at \~30%.
 
@@
   filter(RMTYPE != "INDET") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(Level = case_when(
+    Level == "4C" | Level == "4D" ~ "4",
+    Level == "4E/5" ~ "4E",
+    Level == "7" ~ "6",
+    .default = Level)) %>% 
+  mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" | RMTYPE == "TYPE 10" | RMTYPE == "TYPE 11" | RMTYPE == "TL" ~ "Non-local")) %>% 
+  #mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                       # "TYPE 5", "TYPE 6","TYPE 7", 
+                                        #"TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   dplyr::group_by(Level, RMTYPE) %>% 
@@
   geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
-                position=position_stack(vjust=0.5), colour="white", size=3) +
-  theme(axis.title.x = element_blank()) +
+                position=position_stack(vjust=0.5), colour="white", size=5) +
+  theme(axis.text.x = element_text(size = 16),
+        axis.text.y = element_text(size = 16),  
+        axis.title.x = element_blank(),
+        axis.title.y = element_text(size = 12, angle = 90, hjust = .5, vjust = .5),
+        strip.text = element_text(size = 16),
+        legend.text = element_text(size = 16)) +
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
@@
 In comparison, other types, although showing some variation in the weight of artefacts, show smaller artefacts and their means and medians are frequently closer to the bottom quartile.
-As seen in Figure 3 B, non-local types show larger masses from level 5 to level 7.
+As seen in Figure 3 B, non-local types show larger weights from level 5 to level 7.
 
@@
 #| label: fig-weight
-#| fig-cap: "Terrace boxplots of chert type weight per level."
+#| fig-cap: "Terrace boxplots of chert type weight per level. Diamond shape within the boxplots represents weight mean."
 #| #| fig-width: 10
@@
   mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
-  filter(CODE != "B" & Layer != "D") %>% #removing buckets (B)
+  filter(CODE != "B" & Layer != "D" & Layer != "ND") %>% #removing buckets (B)
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
@@
   #geom_jitter(shape=16, position=position_jitter(0.2)) +
-  coord_cartesian(ylim =  c(0, 40)) +
+  coord_cartesian(ylim =  c(0, 35)) +
   scale_color_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
@@
 
-The macroscopic and petrographic analysis of chert and chalcedony lithic materials from the UP levels of the Terrace and Shelter areas at Vale Boi yield essential insights into the procurement of raw materials and the mobility patterns of hunter-gatherers in southwestern Iberian Peninsula.
-These materials serve as valuable indicators for identifying changes in mobility, as they reveal patterns of both local and non-local procurement.
-The dominance of local raw materials in all assemblages indicates a heavy reliance on nearby resources.
-However, the observed variation in the frequencies of local types may reflect differences in mobility strategies or preferences for certain nodules.
-The detection of non-local raw materials throughout the stratigraphy points to a persistent use of long-distance chert sources.
-This continuity could be attributed to a tradition of shared knowledge and culture regarding the locations of raw materials and other resources within the landscape, or the preservation of social networks over time.
-@cascalheira_2017 acknowledge the adoption of specific behavioural strategies by Vale Boi hunter-gatherers in response to the regional ecological conditions, which were sustained throughout the UP at the site.
-Practices such as grease rendering or the functional specialization of raw materials are seen as deliberate actions with significant cultural implications for future generations, influencing the adaptive choices made over time.
-Accordingly, chert raw materials appear to be integral to these behavioural strategies, established from the site's earliest occupations and perpetuated throughout the UP.
-
+The macroscopic and petrographic analysis of chert and chalcedony lithic materials from the UP levels of the Terrace and Shelter areas at Vale Boi yielded essential insights into the procurement of raw materials and the mobility patterns of hunter-gatherers in Southwestern Iberia.
+These materials serve as valuable indicators for identifying changes in mobility, as they reveal patterns of both local and non-local procurement and have allowed us to answer questions about landscape exploitation and territory movement through the presence (or absence) of local, regional, and long-distance cherts.
+
+The patterns of high percentages of local raw materials in all assemblages indicate a heavy reliance on nearby resources.
 The local chalcedony and chert indicate that the groups were sourcing a significant portion of their raw materials within approximately a 20 km radius of the site.
@@
 This hypothesis is supported by the potential existence of UP occupations near these outcrops, which might reflect quarrying and chert catchment activities during prehistory [@zambujo_jazidas_1998].
-Evidence of short-distance procurement includes the presence of heavy nodules and large chunks of chert from local varieties with parent rock still attached, as well as a significant number of outliers (representing larger blocks and chunks) in the weight boxplots across the stratigraphy (Figure 4).
-These may have been discarded due to natural fractures.
-It's also possible that local cherts were procured from farther distances, perhaps as part of embedded activities.
-The other chert sources from the Lower and Upper Jurassic periods in the western Algarve, located along the beach and cliff areas as previously mentioned, are between 15-18 km southwest of Vale Boi (@fig-map b).
-While our current data do not pinpoint which specific outcrops were exploited based on macroscopic and petrographic analysis, the discovery of marine resources such as shells at the site indicates that these groups visited the coast [@belmiro_gravettian-solutrean_2021; @mannePryingNewMeaning2011].
-This suggests that raw materials could have been conveniently acquired in coastal vicinities without significant detours [@binfordOrganizationFormationProcesses1979; @surovellBehavioralEcologyLithic2009]
-
-However, the discussion around local cherts becomes more complex due to the presence of similar outcrops in the eastern Algarve.
-Previous research has indicated that macroscopic criteria alone are insufficient to determine whether the cherts used originated from Lower Jurassic sources in the western Algarve (within approximately a 20 km radius of the site) or from Middle Jurassic sources approximately 80 km to the east (@fig-map c).
-This is because the cherts are similar both macroscopically and petrographically [@belmiroCreatingFramesReference2023].
-A similar issue is noted by Surovell [-@surovellBehavioralEcologyLithic2009], who described a scenario at the late Pleistocene archaeological site of Krmpotich (USA), where a specific type of chert found at the site could be sourced from both local and long-distance outcrops.
-Surovell suggested that some of these cherts might have been imported from distant sources or previous campsites.
-Consequently, we cannot rule out the possibility that some of the cherts we have identified as being sourced within a 20 km radius of Vale Boi could actually be the result of long-distance procurement and regional mobility, potentially from areas approximately 80-100 km away where Middle Jurassic cherts are located.
-Given the similarity in the quality of the raw materials and the absence of geochemical studies, and in light of the study and model mentioned earlier, we surmise that most of these raw materials likely originate from local sources (within 20 km), although a small portion may be attributable to regional movements.
-
-Similarly, Upper Jurassic cherts located about 90 km east of Vale Boi (@fig-map c) have not been detected in the assemblages.
+Evidence of short-distance procurement at Vale Boi includes the presence of heavy nodules and large chunks of chert from local varieties with parent rock still attached, as well as a significant number of outliers (representing larger blocks and chunks) in the weight boxplots across the stratigraphy (Figure 4), that may have been discarded due to natural fractures.
+
+It is also possible that local cherts were procured from farther distances, perhaps as part of embedded activities [@pereira_territory_2016-1].
+The chert sources from the Lower and Upper Jurassic periods in Western Algarve, located along the beach and cliff areas, can be found between 15-18 km southwest of Vale Boi (@fig-map b).
+While our current data is not able to pinpoint which specific outcrops were exploited based on macroscopic and petrographic analysis, the identification of persistent coastal adaptions at Vale Boi [@bichoPleistoceneHuntergathererCoastal2022] and exploitation of marine resources such as shells at the site indicates that these groups visited the coast and used the available coastal resources throughout all of the UP [@belmiro_gravettian-solutrean_2021; @mannePryingNewMeaning2011].
+This suggests that raw materials could have been conveniently acquired in coastal vicinities without significant detours [@binfordOrganizationFormationProcesses1979; @surovellBehavioralEcologyLithic2009].
+
+The limitations of macroscopic and petrographic methods in discerning between outcrops due to the chert nodule similarities have also impaired the specific distinction between Lower Jurassic sources in Western Algarve (within approximately a 20 km radius of the site) and Middle Jurassic cherts located in Eastern Algarve (approximately 80 km to the east; @fig-map c).
+As such, the presence of these regional resources at Vale Boi is uncertain.
+
+A similar issue was noted by @surovellBehavioralEcologyLithic2009, who described a scenario at the Late Pleistocene archaeological site of Krmpotich (USA), where a specific type of chert found at the site could be sourced from both local and long-distance outcrops.
+The author suggested that some of these cherts might have been imported from distant sources or previous campsites.
+
+Consequently, we cannot rule out the possibility that some of the cherts we have identified as being sourced within a 20 km radius of Vale Boi may actually be the result of regional procurement and mobility, potentially from areas approximately 80-100 km away, where the Middle Jurassic cherts are located.
+Given the similarity of the raw materials and the absence of geochemical studies, and considering the study and model mentioned above, most of these cherts likely originate from local sources, while a minor portion may be attributable to regional movements.
+
+In contrast to the Middle Jurassic cherts, the regional Upper Jurassic cherts located \~90 km east of Vale Boi (@fig-map c) have easily discernible characteristics.
+However, they have not been detected in the assemblages.
 This absence could be attributed to the limited use and transport of these nodules from their sources to Vale Boi, potentially due to the small size of the nodules (often between 3-5 cm), the difficulty of detaching them from the parent rock, or challenges related to their accessibility and visibility, given that the known outcrops are situated inland, approximately 10 km from the present coastline.
-
-While the regional Upper Jurassic cherts situated less than 100 km east of the site were not identified within the assemblages, the allochthonous cherts with potentially identified sources seem to originate from greater distances and in two main directions: 1) from the north, where Cretaceous cherts from Central Portugal are found (@fig-map a), along with a small amount of jasper (TL01); and 2) from the east, where peloidal/oolitic chert from western Andalusia is present (@fig-map), and possibly jasper, transported by the Guadiana river at the current border between Portugal and Spain.
-The scarcity of pebble cortex and the predominance of rounded cortex suggest that these cherts were likely not collected from beaches or riverbeds far from the outcrops but were possibly gathered closer to the source areas.
-These identified axes of long-distance movement to the north and east could be influenced by the topography, affecting mobility patterns.
+This may corroborate the previously observed interest of Pleistocene hunter-gatherers in coastal environments in Iberia [@schmidtRapidClimateChange2012], which would have been characterized by rich marine productivity due to upwelling especially in the Atlantic coast [@bichoPleistoceneHuntergathererCoastal2022; @bicho_at_2008], as well as the aforementioned adaptations and use of coastal resources of hunter-gatherers from Vale Boi [@bichoPleistoceneHuntergathererCoastal2022; @mannePryingNewMeaning2011].
+
+So far, we have discussed the presence of cherts whose outcrops are in relative proximity to Vale Boi and show the full usage of local chert variability and the selective usage of regional cherts.
+As previously mentioned, however, it is the combined identification of local and long-distance cherts that allows us to have a comprehensive idea of the total exploited territory, as well as the possible existence of exchanges and social networks between groups occupying different territories.
+
+Although a wide variety of chert types incongruent with the regional reference collection were identified throughout the stratigraphy and classified as non-local, we were only able to ascertain the possible sources of three types.
+The long-distance cherts with identified sources come from distances of over \~200 km and in two main directions: 1) from the north, where Cretaceous cherts from Central Portugal are found (Figure 1 a), along with a small amount of jasper (TL01); and 2) from the east, where peloidal/oolitic chert from Western Andalusia is present (@fig-map), and possibly jasper, transported by the Guadiana river at the current border between Portugal and Spain.
+
+In fact, Vale Boi’s proximity to a coastal setting, with access to beaches, as well as rivers such as the Guadiana River, where raw material nodules could be transported from far-away distances, could explain the presence of long-distance cherts.
+In this case, the presence of cherts with sources from over 200 km would be explained by their procurement in secondary settings and a possible local movement range.
+However, the scarcity of pebble cortex and the predominance of rounded cortex suggest that these cherts were likely not collected from beaches or riverbeds far from the outcrops but were possibly gathered closer to the source areas [@kuhnUpperPaleolithicRaw2004].
+As such, the presence of these long-distance cherts is better explained by the extended movement of human groups in the south-north and west-east axes.
+
+However, it is relevant to ask why human groups occupying Vale Boi were obtaining a considerable amount of their chert from such long distances and such specific directions, either directly through procurement or through exchanges with other groups.
+Although the source of the other non-local cherts is still unknown, making the directions and distances to obtain them also unknown, the identified axes of long-distance movement to the north and east may be the result of the influence of the topography on mobility patterns.
 Indeed, formal models that seek to explain hunter-gatherer mobility have highlighted the significance of factors such as topography on the distances travelled by groups or individuals [@brantinghamMeasuringForagerMobility2006].
-For instance, Quíros Castillo [-@quiroscastilloArchaeologyHistoryPeasantries2020] discusses how Copper Age groups mitigated subsistence risks by establishing exchange networks and engaging in regional mobility, as evidenced by the trade of allochthonous raw materials, despite the high energy costs and risks associated with these exchanges.
-Nonetheless, the author notes certain characteristics of the Iberian Peninsula that facilitate mobility, such as the low cost of traversing the plateaus and the interconnectivity of the Guadalquivir River valley with the Guadiana and Tagus rivers.
-Thus, the observed long-distance mobility patterns in the north/east directions by hunter-gatherers at Vale Boi might reflect the exploitation of these natural corridors for movement, explaining the substantial use of allochthonous chert.
-The presence of a smaller subset of allochthonous cherts of unknown origin (TL) could represent different forms of long-distance trade of finished goods and blanks, or interactions with various social networks, potentially even indicating individual networks and mobility strategies [sensu @gamblePalaeolithicSocietiesEurope1999].
-
-The significant presence of allochthonous cherts during the Gravettian occupation at Vale Boi suggests extensive communication and mobility among the groups residing there, indicating an expansive social network.
-This finding diverges from the anticipated patterns.
-In the context of the Solutrean occupations, one would expect a high frequency of non-local raw materials due to the broadening of social networks.
-In contrast, the Gravettian period, especially in light of the technological findings from southern Portugal, would typically be associated with a reduced reliance on non-local raw materials.
-
-Technological analyses of lithic assemblages have unveiled notable differences between Gravettian assemblages from central and southern Portugal, pointing to the existence of two distinct regional Gravettian traditions in southern Iberia: the Vicentine Gravettian at Vale Boi, and the Mediterranean Gravettian identified in UP sites in southern Spain [@marreirosLithicTechnologyGravettian2015; @marreirosLithicTechnologyVariability2013].
-These traditions reflect adaptations to the unique local and regional ecological settings [@bradtmoller_lithic_2016] and evolving environmental conditions [@marreirosLithicTechnologyVariability2013], potentially fostering distinct ethnographic divisions among hunter-gatherer communities [@marreirosLithicTechnologyGravettian2015].
-
-This analysis suggests that lithic technological traits may delineate different coexistence boundaries, confined to specific regional bands and aligned with the ecological dynamics of southwestern Iberia.
-Consequently, one might expect territorial or cultural demarcations to influence material usage patterns, with a shift towards a greater reliance on local resources and a diminished use of non-local raw materials during the Gravettian.
-However, the observed prevalence of allochthonous materials could reflect not merely long-distance movements within a single cultural framework but rather the result of interactions and exchanges between different groups.
-Such exchanges, focused on securing raw materials not locally available, underscore the complexity of prehistoric social networks.
+
+Similarly, friction of terrain cartography for Iberia has shown that terrain can produce different costs of travelling on foot, which differ from those directly observed on current-day maps and direct distances [@diazdelrioWhatIberianCopper2020].
+In this context, distance is less related to measurements and more related to the time it costs to go from point A to point B [@zilhaoLatePersistenceMiddle2021].
+For example, @diazdelrioWhatIberianCopper2020 highlights the existence of low-cost pathways through plateaus and the connectivity between the river valleys of Central and South Iberia (Guadalquivir, Guadiana and Tagus).
+
+Using the same friction of terrain cartography, @zilhaoLatePersistenceMiddle2021 also identifies the possible role of terrain in hunter-gatherer mobility in the Upper Paleolithic.
+For example, the author highlights that the Parpalló bifacial point is found across Central/South Spain and Portugal during the Upper Solutrean, and thus is considered the index fossil for this period in the region.
+From Westernmost Central Portugal to Easternmost Central Spain (west-east axis across Iberia), there is a distance of \~800-900 km as the crow flies.
+However, the Parpalló point is not found in Northern Spain (Catalunya), only 200 km north of the easternmost site where the index fossil was identified.
+The author suggests that this separation may be explained by the friction of terrain, which divides Iberia into two halves, creating a quicker mobility pathway between Central/South Spain and Portugal in opposition to the higher-cost time necessary to travel to Northern Spain.
+
+As such, the observed long-distance mobility patterns in the north/east directions by hunter-gatherers at Vale Boi might reflect this Central and Southern Iberia territory dynamic and the exploitation of the natural corridors for movement, explaining the substantial use of long-distance chert.
+In this case, the presence of a smaller subset of non-local cherts of unknown origin (TL) but unknapped at the site could represent different forms of long-distance trade of finished goods and blanks, or interactions with various social networks, potentially even indicating individual networks and mobility strategies [sensu @gamblePalaeolithicSocietiesEurope1999].
+
+Another key idea about these mobility patterns is not only their distance and direction but also their continuity throughout the UP.
+The detection of the same types of non-local raw materials throughout the stratigraphy at Vale Boi points to a persistent use of long-distance chert sources.
+This continuity cannot only be explained by the natural topographic connectivity between Central and South Iberia but can also be attributed to a tradition of shared knowledge and culture regarding the locations of raw materials and other resources within the landscape, or the preservation of social networks over time.
+
+For example, @cascalheira_2017 acknowledge the adoption of specific behavioural strategies by Vale Boi hunter-gatherers in response to the regional ecological conditions, which were sustained throughout the UP at the site.
+Practices such as grease rendering or the functional specialization of raw materials are seen as deliberate actions with significant cultural implications for future generations, influencing the adaptive choices made over time.
+Accordingly, chert raw material acquisition and management appear to be integral to these behavioural strategies, established from the site’s earliest occupations and perpetuated throughout the UP.
+
+However, despite the consistent presence of local and long-distance chert throughout the stratigraphy, we identified trends of changing frequencies in the different occupations, possibly also providing different interpretations about mobility and social behaviour through time.
+For example, the significant presence of non-local cherts during the Gravettian occupations at Vale Boi suggests extensive communication and mobility among the groups occupying the site.
+As previously established, however, while it would be expected for Solutrean groups to have high mobility and extended social networks, the opposite would be expected for the Gravettian occupations.
+As such, our results, especially those from the Gravettian occupations, diverge from the anticipated patterns.
+
+Specifically, technological analyses of lithic assemblages have unveiled notable differences between Gravettian assemblages from the Atlantic and Mediterranean coastal regions, pointing to the existence of two distinct regional Gravettian traditions in Southern Iberia: the Vicentine Gravettian at Vale Boi, and the Mediterranean Gravettian identified in UP sites in Southern Spain [@marreirosLithicTechnologyGravettian2015; @marreirosLithicTechnologyVariability2013].
+These traditions are thought to reflect adaptations to the unique local and regional ecological settings [@bradtmoller_lithic_2016] and evolving environmental conditions [@marreirosLithicTechnologyVariability2013], potentially fostering distinct ethnographic divisions among hunter-gatherer communities [@marreirosLithicTechnologyGravettian2015].
+Consequently, it would be expected for territorial or cultural demarcations to influence material usage patterns, with a shift towards a greater reliance on local resources and a diminished use of non-local raw materials during the Gravettian at Vale Boi.
+
+However, the observed prevalence of non-local chert materials could reflect not merely long-distance movements within a single cultural framework but rather the result of interactions and exchanges between different cultural groups, thus explaining the existence of distinct regional technological characteristics.
 
 This perspective aligns with Brantingham's [-@brantinghamMeasuringForagerMobility2006] discussion on social exchange, which posits that exchanges likely occurred between groups occupying distinct territories, who would meet regularly to trade raw materials.
-This model helps explain the presence of materials from distant sources at archaeological sites, illustrating the nuanced and interconnected nature of prehistoric social landscapes.
-
-An alternative interpretation of the varying frequencies of non-local cherts during different periods at Vale Boi can be informed by the "Mean per Capita Occupation Span" model proposed by Surovell [-@surovellBehavioralEcologyLithic2009] and the "Distance/Frequency Index (DFI)" outlined by Grove et al.
-[@groveMovingFarMoving2023].
+This model helps explain the presence of materials from distant sources at archaeological sites, illustrating the nuanced and interconnected nature of prehistoric social landscapes [@whallonSocialNetworksInformation2006].
+
+In this sense, interpreting the ubiquitous presence of long-distance chert at Vale Boi as a result of exchange between groups and social networks informs us that even during chronological periods where ethnographic boundaries have been suggested, there is communication and exchange between groups.
+This further cements the idea of interconnectedness across the Iberian territory during the UP, expanding the extensive line of contacts and networks identified across North and Central Iberia [@aubryRawMaterialProcurement2004; @aubryUpperPalaeolithicLithic2016; @corchonrodriguezMobiliteTerritoiresRelations2016; @hermidaSilexNWPeninsula2016; @marreirosExploringLithicVariability2016; @ortegaMobilitatDesplacamentsDels2003; @sanchezdelatorreNewDataChert2023] to Southwesternmost Iberia.
+
+However, although social networks may explain the presence of long-distance cherts at Vale Boi even during periods of demarked regional boundaries, the high/low presence of long-distance raw materials at the site may not be the direct result of a higher reliance on these networks, or their expansion or shrinkage.
+
+An alternative interpretation of the varying frequencies of non-local cherts during different periods at Vale Boi can be explained by the “Mean per Capita Occupation Span” model proposed by @surovellBehavioralEcologyLithic2009, and the “Distance/Frequency Index (DFI)” outlined by @groveMovingFarMoving2023.
 These models hypothesize that shorter-term occupations are likely to show a higher proportion of non-local artefacts, whereas with longer-term settlements, the prevalence of local artefacts increases.
 This pattern may be attributed to environmental influences: challenging conditions such as fragmented habitats and cooler temperatures limit mobility, making long-term settlement in areas rich in local resources a preferable, lower-risk option [@surovellBehavioralEcologyLithic2009].
-This pattern is mirrored in the archaeological record of Vale Boi, where, aside from the earlier Gravettian and Magdalenian phases, subsequent occupations seem to adopt a logistical-settlement strategy, positioning Vale Boi as a long-term basecamp [@cascalheira_2017].
-
-In this context, the Gravettian phase at Vale Boi could reflect a period of enhanced mobility and brief occupations, possibly interspersed with reoccupations, indicative of a significant use of non-local raw materials.
-This scenario parallels observations from the Late Gravettian in the French Massif Central, where hunter-gatherer groups are known to have transported entire or pre-formed blocks of raw material over considerable distances for short-term, hunting-related site occupations [@delvigneGeoresourcesTechnoculturalExpressions2019].
-However, the shift toward colder and drier climatic conditions marked by Heinrich Event 2 during the Proto-Solutrean, followed by the Last Glacial Maximum in the Solutrean period [@cascalheiraHunterGathererEcodynamics2013; @gonzalez-samperiz_steppes_2010-1; @sanchez_goni_millennial-scale_2010; @schmidt_rapid_2012; @belmiro_gravettian-solutrean_2021], resulted in a more challenging environment with sparser vegetation [@cascalheiraHunterGathererEcodynamics2013; @gonzalez-samperiz_steppes_2010-1].
-These changes likely constrained the mobility of hunter-gatherers, fostering longer-term settlements within the resource-rich niche of southwestern Portugal [@cascalheira_2017; @schmidt_rapid_2012].
+
+This pattern is mirrored in the archaeological record of Vale Boi, where, aside from the earlier Gravettian of levels 6 and 7, subsequent occupations seem to adopt a logistical-settlement strategy, positioning Vale Boi as a long-term basecamp [@cascalheira_2017].
+In this context, the Gravettian phase at Vale Boi could reflect a period of enhanced mobility and brief occupations, possibly interspersed with reoccupations, indicative of a significant import of non-local raw materials.
+This scenario parallels observations from the Late Gravettian in the French Massif Central, where hunter-gatherer groups are known to have transported entire or pre-formed blocks of raw material over considerable distances (\> 200 km) for short-term, hunting-related site occupations [@delvigneGeoresourcesTechnoculturalExpressions2019].
+The authors also note that in the following occupations, during the Badegoulian and Magdalenian, occupation duration increases along with a significant increase of local and semi-local raw materials.
+
+Similarly, the lower percentages of non-local cherts characterizing the phases following the Gravettian at Vale Boi could reflect a period of increased site occupation.
+In this context, and following the model by @surovellBehavioralEcologyLithic2009, the shift toward colder and drier climatic conditions marked by Heinrich Event 2 (HE2) during the Proto-Solutrean, followed by the Last Glacial Maximum in the Solutrean period [@belmiro_gravettian-solutrean_2021; @cascalheiraHunterGathererEcodynamics2013; @gonzalez-samperiz_steppes_2010-1; @sanchez_goni_millennial-scale_2010; @schmidtRapidClimateChange2012], could have resulted in a more challenging environment with sparser vegetation [@cascalheiraHunterGathererEcodynamics2013; @gonzalez-samperiz_steppes_2010-1].
+These changes likely constrained the mobility of hunter-gatherers, fostering longer-term settlements within the resource-rich niche of Southwestern Portugal [@cascalheira_2017; @schmidtRapidClimateChange2012].
 Consequently, as settlements extended in duration, the reliance on non-local raw materials, whether sourced directly or acquired through trade, diminished in favour of increasingly dominant local raw materials.
@@
 
-Our findings build upon earlier studies indicating that the majority of chert utilized at Vale Boi originates from local Lower Jurassic sources within a 20 km radius of the site, with non-local materials constituting a minor portion of the chert used [@pereira_territory_2016-1].
-The hypothesis of long-distance procurement of Cretaceous cherts from central Portugal, situated north of Vale Boi, is corroborated by visual and petrographic evidence [@pereira_territory_2016-1; @bicho_ecodynamics_2013], and this pattern persists throughout the UP.
-Furthermore, the identification of cherts bearing macroscopic resemblance to the peloidal/oolitic Upper Jurassic cherts from the Betic System in southern Spain confirms a west-east mobility axis, previously inferred from techno-typological studies, spanning the entirety of the UP period.
+Our findings build upon earlier studies indicating that the majority of chert utilized at Vale Boi originates from local Lower Jurassic sources within a 20 km radius of the site [@pereira_territory_2016-1], with non-local materials constituting a relevant portion of the chert used.
+The hypothesis of long-distance procurement of Cretaceous cherts from Central Portugal, situated north of Vale Boi, is corroborated by visual and petrographic evidence [@pereira_territory_2016-1; @bicho_ecodynamics_2013], and this pattern persists throughout the UP.
+Furthermore, the identification of cherts bearing macroscopic resemblance to the peloidal/oolitic Upper Jurassic cherts from the Betic System in Southern Spain confirms a west-east mobility axis, previously inferred from techno-typological studies, spanning also the entirety of the UP period.
 
 These findings reveal a landscape of hunter-gatherers engaged in complex practices of landscape exploration, resource management, and possibly risk management.
-The presence of allochthonous raw materials, albeit in small quantities during the Proto-Solutrean and Solutrean phases, underscores that the inhabitants of Vale Boi—situated on the periphery of the Iberian Peninsula and Europe, bordered by the sea to the south and west—were not isolated.
-Instead, they were part of an extensive social network, a pattern of interconnectivity that resonates with findings across western Europe.
-In southern Portugal, this interconnection among UP hunter-gatherers from diverse and distant regions of the Iberian Peninsula is similarly observed.
+The presence of allochthonous raw materials, albeit in small quantities during the Proto-Solutrean and Solutrean phases, underscores that the inhabitants of Vale Boi—situated on the periphery of Iberia and Europe, bordered by the sea to the south and west, and confined to the patchy refugia created by the worsening climatic conditions of the HE2 and following Glacial Maximum [@schmidtRapidClimateChange2012]—were certainly not isolated.
+Instead, they were part of an extensive social network, a pattern of interconnectivity that resonates with findings across Western Europe.
+
+In Southern Portugal, this interconnection among UP hunter-gatherers from diverse and distant regions of Iberia is similarly observed.
+This was already expected for the Proto-Solutrean and Solutrean occupations, but less so for the Gravettian where social boundaries were expected based on previously identified technological facies between regions [@marreirosLithicTechnologyGravettian2015].
+Instead, our results show a high presence of long-distance raw materials during the Gravettian occupations, which may indicate that social networks were established from the first UP occupations in South Portugal and maintained through time.
 

### 2024-06-28 analysis/supplementary-materials/SOM_chisquare.qmd
- Chi-square SOM - Final JC revisions

---
title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia, Journal of Paleolithic Archaeology"
subtitle: "Online Resource 4: Statistical analysis of Terrace chert types"
author:
  - Joana Belmiro:
      orcid: https://orcid.org/0000-0003-4951-2907
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute: 
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - Salvador Dominguez-Bella:
      orcid: https://orcid.org/0000-0003-3892-763X
      institute: 
        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated"  
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: ../paper/references.bib
toc: false
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

```{r}
#| label: callpackages

# Call packages
# This includes all necessary packages to run the code
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(knitr)
library(corrplot)
library(vcd)
```

```{r}
#| label: db-setup

# Terrace DB preparation (repeated from main document)
# Read DBs
T_RM_data <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_TERRACE.csv')) # Terrace RM analysis
context <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_context.csv')) # Terrace Field DB - lithic artefact info
xyz <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_xyz.csv')) # Terrace Field DB - just coordinates

# Df cleaning and uniformizing
cols_to_concat <- c("UNIT", "ID") # Vector to concatenate Unit and ID in one single variable (e.g. A6-100) - to use with both Field databases
context$ID <- str_squish(context$ID) # Remove empty spaces from ID field
xyz$ID <- str_squish(xyz$ID) # Remove empty spaces from UNIT field

# Concatenate UNIT and ID variables in one single variable (e.g. A6-100)
context <- context %>%
  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column
  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
  select(ID, Level, Spit, Code) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

xyz <- xyz %>%
  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
  select(ID, X, Y, Z) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

field_data <- full_join(xyz, context, by = "ID") # Join field tables
T_dataset <- left_join(T_RM_data, field_data, by = "ID") # Join field and RM type tables

T_dataset <- T_dataset %>% 
  filter(!Level %in% c("2", "3")) %>%
  drop_na(Level) %>% 
  mutate(Level = case_when(Level == "7C" ~ "7",
                           Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7, since 8 disappears in field_data
                           Level == "4B" ~ "4", # collapsing samples from 4B to 4 due to small number
                           TRUE ~ Level)) %>% 
  mutate(RMTYPE = case_when(RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
                            RMTYPE == "TYPE 15" ~ "TYPE 3",
                            RMTYPE == "TYPE 7" ~ "TYPE 4",
                            RMTYPE == "TYPE 12" ~ "TYPE 5",
                            RMTYPE == "TYPE 11" ~ "TYPE 6",
                            RMTYPE == "TYPE 13" ~ "TYPE 7",
                            RMTYPE == "TYPE 11F" ~ "TYPE 8",
                            RMTYPE == "TYPE 13B" ~ "TYPE 9",
                            RMTYPE == "Oolitic" ~ "TYPE 10",
                            RMTYPE == "GB" ~ "TYPE 11",
                            TRUE ~ RMTYPE))

# Mutate level 5 to add phases: top of 5 (4E/5) and bottom of 5 (5), following Belmiro et al. (2020).
phase_dataset <- T_dataset %>% 
  mutate(Level = case_when(
    Level == 5 & Z > 24.1 ~ "4E/5",
    Level == 5 & Z < 24.1 ~ "5",
    TRUE ~ Level)) %>% 
  mutate(Technocomplex = case_when(
    Level == 4 | Level == "4C" | Level == "4D" ~ "Solutrean",
    Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
    Level == 6 | Level == 7 ~ "Gravettian"))
```

```{r}
#| label: chisq
#| include: false

chisq <- phase_dataset %>% 
  filter(RMTYPE != "INDET") %>% 
  filter(RMTYPE != "TYPE 11")

chisq_tab <- table(chisq$RMTYPE, chisq$Level)
chisq_matrix <- as.table(as.matrix(chisq_tab))
chisq_rm <- chisq.test(chisq_tab)
round(chisq_rm$residuals, 3)

```

The chert raw material analysis of the Terrace and Shelter areas of the archaeological site of Vale Boi revealed the presence of varied types of local and non-local chert and chalcedony throughout the stratigraphy. However, we identified differences in the percentages of these types throughout the several Upper Paleolithic occupations in the Terrace area.

To explore the statistical significance of these patterns in the Terrace area, a chi-square test of independence was performed to determine whether the chert types are related to the archaeological layer. The results are significant, X-square = 817.79, *p* \< 2.2e-16, showing statistically significant differences in the chert types present throughout the occupations. @fig-mosaicplot shows a mosaic plot visualising the residuals of the test and @fig-corrplots shows the visual correlations between the variables in level and chert type. These show different associations and patterns throughout the different levels and occupations.

In the Solutrean occupations, Type 2 and Type 3 cherts are relatively more frequent in Level 4, and although to a lesser degree, Type 5 is relatively more frequent in Level 4C. For the Solutrean occupations, all positive correlations relate to local cherts.

For the Proto-Solutrean occupations, Type 4 cherts are relatively more frequent in Levels 4E and 4E/5, and Type 1 chalcedony is relatively more frequent in Level 4E/5. Similar to the Solutrean levels, only local cherts and chalcedony show positive correlations. In comparison, Type 6 and 7 are relatively more frequent in Level 5, while Type 4 is relatively less frequent than expected. This shows a significant shift from positive correlations of local cherts to non-local.

The Gravettian occupations, especially Level 6, show markedly different patterns. TL cherts, Type 10, Type 6 and Type 9 cherts (all non-local) are relatively more frequent in Level 6. The same patterns can also be seen in @fig-corrplots, although there are expanded (although weaker) correlations between non-local chert types and Level 6 and 7.

```{r}
#| label: fig-mosaicplot
#| fig-cap: Mosaic plot showing standardized Peason's residuals from the chi-square test showing the dependencies of the different archaeological levels and chert types.

vcd_labels <- list(RMTYPE = c("TL", "1", "10", "2", "3", "4",
                                   "5", "6", "7", "8", "9"))
mosaic(~ Level + RMTYPE,
      gp = shading_max,
      data = chisq, 
      set_labels = vcd_labels,
      rot_labels = c(0, 0))
```

```{r}
#| label: fig-corrplots
#| fig-cap: Correlation plot using Pearson's residuals from the chi-square test, showing positive and negative correlations between archaeological levels and chert types.

corrplot(chisq_rm$residuals, method = "color", is.corr = FALSE)

```

### 2024-07-15 analysis/paper/paper.qmd
- Correction of shelter graphs and text

@@
   unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% 
-  select(ID, Layer)
+  select(ID, Layer) %>% 
+  distinct(ID, .keep_all = TRUE)
 
@@
   unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% # Unite YEAR with Id s«using cols_to_concat
-  select(ID, X, Y, Z, CODE) %>% # Selecting important variables
+  select(ID, X, Y, Z, CODE) %>%
+  filter(CODE == "L") %>% # Selecting important variables
   distinct(ID, .keep_all = TRUE)
 
-shelter_field_data <- full_join(S_field_data, S_layerdata, by = "ID")
-S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID", relationship = "many-to-many")
+shelter_field_data <- left_join(S_field_data, S_layerdata, by = "ID")
+S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID")
 
@@
 Using the field database with all individually coordinated lithic artefacts (all pieces larger than \~2 cm), we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials.
-@fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents \~50-70% of the lithic assemblage from the Solutrean layers A, B, and C.
+@fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents \~35-50% of the lithic assemblage from the Solutrean layers A, B, and C.
 Based on previous lithic studies in the Terrace and Shelter areas, the remaining 80% of other raw materials are mostly composed of quartz and greywacke [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009].
@@
 ## Data preparation
-context_comp <- field_data %>%
+context_comp <- field_data %>% # Using the field data to obtain lithic info beyond the analysed chert/chalcedony
   mutate(Level = case_when(Level == "4B" ~ "4", # Collapsing samples from level 4B into 4
@@
   filter(!Level %in% c("3B", "5B", "5C")) %>% 
-  dplyr::filter(Code %in% c("LITHIC", "POINT"))
-
-dataset_comp <- phase_dataset %>% 
-  select(ID, RMTYPE)
-T_comp <- full_join(context_comp, dataset_comp, by = "ID")
+  dplyr::filter(Code %in% c("LITHIC", "POINT")) # Field data includes ALL artifacts; filtering only lithics
+
+dataset_comp <- phase_dataset %>% # Using the analysed DB to get chert/chalcedony data
+  select(ID, RMTYPE) # Selecting only ID (to join dbs and RMTYPE to know whether chert or chalcedont)
+T_comp <- full_join(context_comp, dataset_comp, by = "ID") # Joining by ID.
+the RMTYPE column has the RM type from the analysis and NA for those which were not analysed, and correspond to non-chert artifacts
 T_comp["RMTYPE"][is.na(T_comp["RMTYPE"])] <- "Other" # All NA's (other RMs) turned into "Other"
@@
  mutate(Rawmaterial = case_when(
-    RMTYPE == "Other" ~ "Other",
-    RMTYPE == "TYPE 1" ~ "Chalcedony",
-    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category
+    RMTYPE == "Other" ~ "Other", # Keeping Other as Other
+    RMTYPE == "TYPE 1" ~ "Chalcedony", # Naming Chalcedony
+    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category (All but Other and TYPE1)
   dplyr::filter(Level %in% c("4", "4C", "4D", "4E", "4E/5", "6", "7")) # Choosing only UP levels (excluding 3)
@@
 ## Data preparation
-S_field_comp <- S_field_data %>% 
-  filter(CODE == "L") # Filtering only lithics
+S_field_comp <- S_field_data
 S_data_comp <- S_dataset %>% 
@@
   filter(RMTYPE != "INDET") %>% 
-  mutate(Level = case_when(
-    Level == "4C" | Level == "4D" ~ "4",
-    Level == "4E/5" ~ "4E",
-    Level == "7" ~ "6",
-    .default = Level)) %>% 
-  mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" | RMTYPE == "TYPE 10" | RMTYPE == "TYPE 11" | RMTYPE == "TL" ~ "Non-local")) %>% 
-  #mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                       # "TYPE 5", "TYPE 6","TYPE 7", 
-                                        #"TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
   dplyr::group_by(Level, RMTYPE) %>% 
@@
   geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
-                position=position_stack(vjust=0.5), colour="white", size=5) +
-  theme(axis.text.x = element_text(size = 16),
-        axis.text.y = element_text(size = 16),  
-        axis.title.x = element_blank(),
-        axis.title.y = element_text(size = 12, angle = 90, hjust = .5, vjust = .5),
-        strip.text = element_text(size = 16),
-        legend.text = element_text(size = 16)) +
+                position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_blank()) +
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",

### 2024-07-16 analysis/paper/paper.qmd
- Title update on paper and SOM docs

@@
 ---
-title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
+title: "Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
 author:

### 2024-07-16 analysis/supplementary-materials/SOM_chisquare.qmd
- Title update on paper and SOM docs

@@
 ---
-title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia, Journal of Paleolithic Archaeology"
+title: "Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
 subtitle: "Online Resource 4: Statistical analysis of Terrace chert types"

### 2024-07-16 analysis/supplementary-materials/SOM_datadictionaries.qmd
- Title update on paper and SOM docs

@@
 ---
-title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia, Journal of Paleolithic Archaeology"
+title: "Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
 subtitle: "Online Resource 1: Data dictionaries for macroscopic and petrographic analyses"

### 2024-07-16 analysis/supplementary-materials/SOM_lithotypes.qmd
- Title update on paper and SOM docs

@@
 ---
-title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia, Journal of Paleolithic Archaeology"
+title: "Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
 subtitle: "Online Resource 2: Terrace and shelter lithotypes complete description"

### 2024-07-16 analysis/paper/paper.qmd
- Bibliography corrections - Replace thin-section for thin section in main document

@@
 abstract: |
-  Analyses of raw materials and the distinction between local/regional and long-distance sources have proven invaluable for understanding the extensive movements, interactions, and social networks during the Upper Palaeolithic (UP) in the Iberian Peninsula.
+  Analyses of raw materials and the distinction between local/regional and long-distance sources have proven invaluable for understanding the extensive movements, interactions, and social networks during the Upper Palaeolithic in the Iberian Peninsula.
 However, unlike other parts of Iberia, research on the management and acquisition of raw materials in the south and west of Iberia remains relatively underdeveloped.
@@
 
-The authors confirm that all data generated and analysed during this study are available to the public in the Open Science Framework repository at: ++.
+The authors confirm that all data generated and analysed during this study are available to the public in the Open Science Framework repository at: https://doi.org/10.17605/OSF.IO/DBFT2.
 
@@
 
+**Online resource 4**: Statistical analysis of Terrace chert types.
+
 <!-- The following line inserts a page break  -->

### 2024-07-16 analysis/paper/paper.qmd
- Corrected filter in the shelter graph in stratigraphy.

@@
 #| fig-cap: Variability of local chert types and main petrographic characteristics.
-a) Type 1 macroscopic appearance (a1) and thin-section (a2).
-b) Type 3 macroscopic appearance (b1) and thin-section (b2).
-c-d) Type 2 macroscopic appearance (c1-d1) and thin-section (c2-d2).
-e-f) Type 4 macroscopic appearance (e-f1) and thin-section (f2-f3).
+a) Type 1 macroscopic appearance (a1) and thin section (a2).
+b) Type 3 macroscopic appearance (b1) and thin section (b2).
+c-d) Type 2 macroscopic appearance (c1-d1) and thin section (c2-d2).
+e-f) Type 4 macroscopic appearance (e-f1) and thin section (f2-f3).
 g-h) Type 5 macroscopic appearance (g-h1) and thin section (h2-h3).
@@
 #| fig-cap: Variability of non-local types and main petrographic characteristics.
-a-b) Type 6 macroscopic appearance (a1-b1) and thin-section (a2-b2); b1-b2 correspond to a sample covered 100% in white patina.
-c-d) Type 7 macroscopic appearance (c1-d1) and thin-section (c2-d2) of peloidal facies (c1-c2) and oolitic facies (d1-d2).
-e-f) Type 8 macroscopic appearance (e-f1) and thin-section (f2-f3); white and black arrows correspond to foraminifera.
-g) Type 9 macroscopic appearance (g1) and thin-section (g2).
-Type 10 macroscopic appearance (h1) and thin-section (h2).
+a-b) Type 6 macroscopic appearance (a1-b1) and thin section (a2-b2); b1-b2 correspond to a sample covered 100% in white patina.
+c-d) Type 7 macroscopic appearance (c1-d1) and thin section (c2-d2) of peloidal facies (c1-c2) and oolitic facies (d1-d2).
+e-f) Type 8 macroscopic appearance (e-f1) and thin section (f2-f3); white and black arrows correspond to foraminifera.
+g) Type 9 macroscopic appearance (g1) and thin section (g2).
+Type 10 macroscopic appearance (h1) and thin section (h2).
 
@@
   mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
-  filter(CODE != "B" & Layer != "D" & Layer != "ND") %>% #removing buckets (B)
+  filter(CODE != "B" & Layer != "D" & Layer != "ND" & Layer != "No info") %>% #removing buckets (B)
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",

### 2024-10-21 analysis/paper/paper.qmd
Fig2 corrected to include level 5 (previously missing from code)

@@
     RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category (All but Other and TYPE1)
-  dplyr::filter(Level %in% c("4", "4C", "4D", "4E", "4E/5", "6", "7")) # Choosing only UP levels (excluding 3)
+  dplyr::filter(Level %in% c("4", "4C", "4D", "4E", "4E/5","5", "6", "7")) # Choosing only UP levels (excluding 3)
 

### 2024-10-21 analysis/paper/paper.qmd
Corrected legend from the indeterminate frequency table. N to number

@@
 #| label: tbl-indet-freq
-#| tbl-cap: N and percentage of grouped and indeterminate (INDET) artefacts by level in the Terrace (levels 4 to 7) and Shelter (A to C) areas.
+#| tbl-cap: Number and percentage of grouped and indeterminate (INDET) artefacts by level in the Terrace (levels 4 to 7) and Shelter (A to C) areas.
 

### 2024-10-22 analysis/paper/paper.qmd
- Reviewer alterations

@@
 During the Solutrean, Vale Boi has been interpreted as an important connection point between the core Solutrean areas of Central Portugal and Southern Spain, making it a key site for identifying extended social networks that promoted the exchange of information and culture [@cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @cascalheiraSolutrenseEmPortugal2013].
-
 On the contrary, techno-typological studies from the Gravettian occupations of Vale Boi instead show marked differences between regions and suggest a more limited circulation of people and information in Southern Iberia [@marreirosLithicTechnologyVariability2013].
@@
 
-To investigate the applicability of these patterns to the procurement and circulation of lithic raw materials, and to examine the prevailing hypotheses concerning UP mobility and social network exchanges, this paper focuses on the following questions: 1) what types of chert were the hunter-gatherers of Vale Boi using and where did they come from?; 2) are there different patterns of chert procurement throughout time?; 3) and if so, how can these be related to patterns of land use and the circulation of people/ideas, as suggested through technological studies of the same lithic assemblages?
+To investigate the applicability of these patterns to the procurement and circulation of lithic raw materials, and to examine the prevailing hypotheses concerning UP mobility and social network exchanges, this paper focuses on the following questions: 1) what types of chert were the hunter-gatherers of Vale Boi using and where did they come from?; 2) are there different patterns of chert use throughout time?; 3) and if so, how can these be related to patterns of land use and the circulation of people/ideas, as suggested through technological studies of the same lithic assemblages?
 As a result, our study presents the first comprehensive characterisation of chert raw materials, through macroscopic and microscopic analysis, of UP occupations in Southwestern Iberia, derived from the Gravettian, Proto-Solutrean, and Solutrean (c.

### 2024-10-23 analysis/supplementary-materials/SOM_datadictionaries.qmd
Added variables to better correspond to the other SI documents

@@
   "Cortex transition", "Sharp, Gradual", "**Sharp**: transition between cortex and chert is sharp, with an obvious change in colors and textures; **Gradual**: cortex and chert show a gradual transition with gradient colors and textures", "-",
+  "Morphology of the nodules", "Irregular or rounded nodules, tabular, bedded, lenses", "Inferred morphology of the chert based on cortical samples, unknapped chunks or preserved geological features such as encasing rock attached.", "@trewinSEDIMENTARYROCKSChert2005",
   "Allochems", "-", "Macroscopically visible allochems (excluding bioclasts).", "-",

### 2024-11-04 analysis/paper/paper.qmd
Reviews from reviewer #1

@@
   select(ID, Layer) %>% 
-  distinct(ID, .keep_all = TRUE)
+  distinct(ID, .keep_all = TRUE) %>%
+  filter(!Layer %in% c("D", "Z"))
 
@@
 S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID")
+S_dataset <- S_dataset %>% 
+  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
+  filter(!Layer == "ND")
 
@@
 24-22 ka cal BP [@cascalheiraValeBoiAlgarve2013; @cascalheiraTecnologiaLiticaSolutrense2010; @cascalheira_2017].
+These occupation levels were identified under blocks of limestone, which collapsed from the rockshelter ceiling [@cascalheiraTecnologiaLiticaSolutrense2010].
+
+Both these areas have been previously interpreted as seasonal residential camps, repeatedly used for extended stays, due to the abundance of lithic debitage, stone tools, heat-cracked rocks related to grease rendering activities, large quantities of faunal remains (both marine and terrestrial) and the presence of ornaments and portable art [@manneIntensiveSubsistencePractices2012].
+The analysis of lithic assemblages and retouched frequency of lithic assemblages from the Terrace and Shelter areas also corroborate this interpretation [@cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009; @cascalheira_2017].
+Cascalheira et al.
+[-@cascalheira_2017] show that, with the exception of the Early Gravettian occupations of the Terrace area and Magdalenian occupation of the Shelter, all other UP occupations are composed of high-density assemblages with low degree of retouch, correspondent to an residential base-camp occupation, for extended periods of time.
 
@@
 The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of Westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b).
+The Lower Jurassic nodules and beds can be found in abundant outcrops, but also frequently in sub-primary or secondary deposition settings in proximity to the outcrops possibly due to natural erosion.
+However, in the case of the inland outcrop of Ferrel, the outcrops is partially destroyed due to recent human action, and chert can be found scattered as chunks or larger nodules.
+The Upper Jurassic nodules can be found sub-primary deposition and of easy access, or in secondary deposition.
 These cherts are frequently opaque with yellow, grey, purple, or red colours, occasionally with bands or laminations, and with frequently visible sponge spicules.
@@
 
-Chalcedony outcrops \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes (Bicho et al., 2013).
+Chalcedony outcrops \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013], although previous studies also note the presence of chalcedony (with frequent presence of cortex) in the immediate vicinity of the site [@pereira_territory_2016-1; @gibajabaoProvenienceTechnologyMorphology2013].
 The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.
@@
 These two areas were chosen since both have detailed spatial information, well-defined chronological sequences, and previous studies providing lithic technology, faunal, and geoarchaeological data.
+For this, we focused on layers that spanned the UP occupations, characterized by good preservation, a significant number of lithic finds and with radiocarbon dating (as mentioned in section Site description).
 
@@
 
-Using the field database with all individually coordinated lithic artefacts (all pieces larger than \~2 cm), we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials.
+Using the field database with all individually coordinated lithic artefacts (all pieces larger than \~2 cm; [@belmiroProtoSolutreanLithicTechnology2020]), we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials.
 @fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents \~35-50% of the lithic assemblage from the Solutrean layers A, B, and C.
@@
 Although systematic raw material research has not been conducted on these materials, according to the regional geological context, they have been interpreted as local and readily accessible in the vicinity of the site [@pereira_territory_2016-1].
-Frequently, the quartz and greywacke assemblages include large amounts of shattered and unknapped chunks or slabs.
+Frequently, the quartz and greywacke assemblages include large amounts of shattered and unknapped chunks or slabs, and have been interpreted in great part as functionally specialised in activities related to the fragmentation of bones in the case of quartz and greywack [@cascalheira_2017], or in grease rendering in the case of quartz [@manneIntensiveSubsistencePractices2012].
 Previous studies from the Terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020].
 Based on this, despite the smaller amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.
-The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].
+The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site, such as a Vale Comprido point, the fossil-director for the Proto-Solutrean in Western Portugal [@belmiro_gravettian-solutrean_2021].
 
@@
     TRUE ~ Layer)) %>% 
-  filter(!Layer %in% c("D", "Z", "No info")) # Choosing only Solutrean occupation layers (A, B and C)
+  filter(Layer %in% c("A", "B", "C")) # Choosing only Solutrean occupation layers (A, B and C)
 
@@
 
-Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustilloMacroscopicClassificationFlint2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=5318).
+Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustilloMacroscopicClassificationFlint2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=4458).
 The distinct features of various local cherts facilitate their differentiation from cherts of neighbouring origins.
@@
 The first step included a preliminary characterization using a hand lens of 10x magnification.
-This step allowed us to create types and subtypes based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1).
+This step allowed us to create types and subtypes based on similar macroscopic characteristics frequently acknowledged to be useful to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1).
 In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types - the data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1).
 Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis was simplified and focused on collecting data related to weight, cortex, and alterations.
-This first step was applied to a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 8 to 4) and 1665 chert artefacts from the Solutrean sequence of the Shelter area (layers C to A).
+This first step was applied to a total of 4458 artefacts: a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 7 to 4) and 831 chert artefacts from the Solutrean sequence of the Shelter area (layers A, B, C; all samples without an attributed layer on the database were excluded from the analysis).
 
@@
 Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversitaDevelopmentMultiparametricCharacterisation2019; @gomezdesolerPanadellaChertMontmaneu2020; @sotoProcurementMobilityLate2016].
-For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), they were analysed using the individual sample dataset, but grouped in a category of indeterminate cherts (INDET).
-
-As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artefacts, corresponding to \~40% of the total sample (n=1027).
+For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), and given the unexistence of a reference collection for local altered cherts, they were analysed using the individual sample dataset (Online Resource 1) but grouped in a category of indeterminate cherts (INDET).
+This category included key alterations such as heat/fire related alterations which clearly impacted the macroscopic characteristics of the samples, fully changing its color, luster and texture, and covering them in crazings (e.g., @fig-terrace-alterations a).
+Similarly, it also included alteration rinds that extended through most of the sample, impeding the identification of its original color, texture, translucency, inclusions and fossils (e.g., @fig-terrace-alterations b) and the presence of extended pitting.
+However, whenever surface alterations were present in only a limited area of the sample leaving recognizable macroscopic features and fossils, or when key elements were distinguishable to make a characterization (e.g., reddening in the Type 2 cherts, Online Resource 2, figure s5), the samples were not collapsed into the INDET category.
+
+As seen in @tbl-indet-freq, level 4 shows the highest percentages of indeterminate artefacts, corresponding to \~40% of the total sample (n=1037).
 This is due to the frequent alterations of the artifacts, especially related to patinas and alteration rinds, pits, and fire/heat alterations.
@@
 Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony types, present throughout the stratigraphy in the Terrace and Shelter areas (a summarised characterization table can be found in the Supplementary Information, Online Resource 3).
-A distinct type was identified exclusively in the Shelter area, with only a limited presence in the Terrace (n=2), and 16 varieties, each consisting of one to three samples, were identified across various levels of both the Terrace and Shelter.
+A distinct type (Type 11) was identified exclusively in the Shelter area, with only a limited presence in the Terrace (n=2), and 16 varieties, each consisting of one to three samples, were identified across various levels of both the Terrace and Shelter.
 The latter were clustered into a category named Trace Lithotypes (TL).
 Through the comparison with the reference collections, we identified cherts congruent with the local sources and cherts that are likely not local due to significant differences in appearance, fossil content, and petrography when compared to the cherts available in the geological setting located within a 30 km radius of Vale Boi.
-
-### Cherts congruent with local sources
+@tbl-type-source shows the different types of identified cherts and possible sources whenever attribution was possible.
+
+```{r}
+
+#| label: tbl-type-source
+#| tbl-cap: Chert types identified during the analysis, and potential sources.
+The distance column refers to the distance of the potential source from Vale Boi, as the crow flies.
+Only the TL samples with possible identified sources were included in the table.
+
+tibble::tribble(
+  ~"Chert type", ~"Potential source", ~"Distance from source",
+  "Type 1", "Triassic volcanic formations ", " ",
+  "Type 2-5", "Lower Carixian formations, Southwestern Portugal", "<20 km",
+  "Type 6", "Upper Cretaceous formations, Central Portugal", "~215-250 km",
+  "Type 7", "Betic Systems, Southern Spain", "~250 km",
+  "Type 8", "-", "-",
+  "Type 9", "Betic Systems, Southern Spain", "~250 km",
+  "Type 10", "-", "-",
+  "Type 11", "-", "-",
+  "TL01", "South Portuguese Zone, Southern/Central Portugal", ">100 km",
+  "TL10-11", "Upper Cretaceous formations, Central Portugal", "~215-250 km",
+  "TL15-16", "Betic Systems, Southern Spain", "~250 km"
+) |>
+  knitr::kable()
+
+```
+
+### Cherts congruent with local sources (Types 1-5)
 
@@
 This type is marked by extensive crystallizations, irregularities, and fractures, coupled with post-depositional alterations that challenge the identification and preservation of the cortex.
-Both macroscopic and petrographic analyses confirm its congruence with hydrothermal chalcedony located within lenses in Triassic volcanic formations, which are similarly marked by irregularities and fractures.
+Both macroscopic and petrographic analyses confirm its congruence with hydrothermal chalcedony located within lenses in Triassic volcanic formations, in the known outcrops of Monte do Tolo, Taboal and Cerro Velho (\<10 km west of Vale Boi, as the crow flies; @fig-map b), which are similarly marked by irregularities and fractures.
+
+The cherts are massive micro-cryptocrystalline quartz from a marine environment (@fig-local-types b-h).
+They represent \~60% of the overall Terrace cherts and \~80% of the chert in the Shelter.
+Macroscopically, these types show variability which resulted in their separation in four different macroscopic types (Type 2 to 5).
+In general, these are opaque to sub-translucent cherts, with colours varying from brown, yellow, pale red, and grey, with the frequent presence of spicules and iron oxides.
+The differences are in their colours, nodule morphology, frequency, and variety of bioclasts (Type 2 including triaxon sponge spicules and echinoderm spines), something already identified in the geological samples [@belmiroCreatingFramesReference2023].
+As seen in table @tbl-cortex_tab, whenever present, the cortex seems to show a mixture of irregular cortex and rounded cortex, both from an outcrop source.
+Petrographically, however, these cherts are homogeneous, except for a small number of artefacts with a banded variety within Type 5.
+These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite.
+Type 5 shows higher percentages of micrite/sparite.
+In all samples, fossils range from common to very frequent, albeit frequently unidentifiable.
+When identifiable in thin section, they are sponge spicules (@fig-local-types b2, c2, h2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (@fig-local-types h3) in a sample from Type 5.
+Porosity varies between 10% and 1%, of vug type.
+Despite the petrographic homogeneity between local types, this macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.
 
@@
 #| label: tbl-cortex_tab
-#| tbl-cap: Frequency of cortex type and thickness by raw material from the Terrace area.
+#| tbl-cap: Frequency of cortex type and thickness by raw material from the Terrace area, characterized during the individual analysis of the lithic assemblage.
 
@@
 
-The cherts are massive micro-cryptocrystalline quartz from a marine environment (@fig-local-types b-h).
-They represent \~60% of the overall Terrace cherts and \~80% of the chert in the Shelter.
-Macroscopically, these types show variability which resulted in their separation in four different macroscopic types (Type 2 to 5).
-In general, these are opaque to sub-translucent cherts, with colours varying from brown, yellow, pale red, and grey, with the frequent presence of spicules and iron oxides.
-The differences are in their colours, nodule morphology, frequency, and variety of bioclasts (Type 2 including triaxon sponge spicules and echinoderm spines), something already identified in the geological samples [@belmiroCreatingFramesReference2023].
-Whenever present, the cortex seems to show a mixture of irregular cortex and rounded cortex, both from an outcrop source (@tbl-cortex_tab).
-Petrographically, however, these cherts are homogeneous, except for a small number of artefacts with a banded variety within Type 5.
-These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite.
-Type 5 shows higher percentages of micrite/sparite.
-In all samples, fossils range from common to very frequent, albeit frequently unidentifiable.
-When identifiable in thin section, they are sponge spicules (@fig-local-types b2, c2, h2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (@fig-local-types h3) in a sample from Type 5.
-Porosity varies between 10% and 1%, of vug type.
-Despite the petrographic homogeneity between local types, this macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.
-
-Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiroCreatingFramesReference2023], we interpret Types 2, 3, 4, and 5 to be local, as they are congruent with the described characteristics representative of cherts from the Lower Jurassic (Carixian) and Middle Jurassic Malhão formations of the Algarve basin.
+Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiroCreatingFramesReference2023], we interpret Types 2, 3, 4, and 5 to be local, as they are congruent with the described characteristics representative of cherts from the Lower Jurassic found in the known outcrops Cabo S.
+Vicente, Foz dos Fornos, Ponta dos Altos, Praia do Belixe and Ferrel (Carixian; \<20 km from Vale Boi, as the crow flies), possibly Upper Jurassic Kimmeridgian outcrops of Mareta (\~15 km from Vale Boi, as the crow flies) and Middle Jurassic Malhão formations (80-100 km East of Vale Boi, as the crow flies) of the Algarve basin.
 Types 2 and 3 are congruent with the frequently available brown/yellow and red nodules and beds (@fig-local-types b-d), ranging from uncommon to very frequent fossils.
-The macroscopic similarities of Types 4 and 5 are more limited (@fig-local-types e-h), as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the inland Lower Jurassic outcrops (@fig-map b).
-As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape.
-This is especially relevant since the inland outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].
+These can be found in all previously mentioned outcrops.
+The macroscopic similarities of Types 4 and 5 are more limited (@fig-local-types e-h), as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the inland Lower Jurassic outcrop of Ferrel (@fig-map b).
+As such, we interpret these cherts as local, but possibly from nodules and inland outcrops that are no longer frequently available or visible in the landscape.
+This is especially relevant since the inland outcrops of Ferrel are currently mostly destroyed and can be found in poor preservation conditions, due to recent human intevention and construction work, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].
 
@@
 #| fig-cap: Variability of local chert types and main petrographic characteristics.
-a) Type 1 macroscopic appearance (a1) and thin section (a2).
-b) Type 3 macroscopic appearance (b1) and thin section (b2).
-c-d) Type 2 macroscopic appearance (c1-d1) and thin section (c2-d2).
-e-f) Type 4 macroscopic appearance (e-f1) and thin section (f2-f3).
-g-h) Type 5 macroscopic appearance (g-h1) and thin section (h2-h3).
+a) Type 1 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope of white patina and fractures; a2) microscopic view of thin section (XPL), with visible fibrous chalcedony composition and without allochems.
+b) Type 3 macroscopic and microscopic views: b1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils; b2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides.
+c-d) Type 2 macroscopic and microscopic views of multicoloured (c) and single-colour (d) varieties.
+c1) macroscopic view with a stereomicroscope of the multicoloured bands with iron oxides and unidentifiable fossils; d1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils; c2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; d2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides.
+e-f) Type 4 macroscopic and microscopic views: e) macroscopic view with a stereomicroscope of unidentifiable fossils and white splotches; f1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils within white and yellow splotches; f2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides; f3) microscopic view of thin section (PPL) with a concentration of fossils (possibly sponge spicules) and a fracture.
+g-h) Type 5 macroscopic and microscopic views: g) macroscopic view with a stereomicroscope; h1) macroscopic view with a stereomicroscope with visible unidentifiable fossils; h2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; h3) microscopic view of thin section (XPL) with small unidentifiable fossils, iron oxides and a possible ostracod replaced by fibrous chalcedony and macrocrystalline quartz.
 
@@
 
-### Cherts not congruent with local sources
+### Cherts not congruent with local sources (Types 6-11 and TL)
 
@@
 #| fig-cap: Variability of non-local types and main petrographic characteristics.
-a-b) Type 6 macroscopic appearance (a1-b1) and thin section (a2-b2); b1-b2 correspond to a sample covered 100% in white patina.
-c-d) Type 7 macroscopic appearance (c1-d1) and thin section (c2-d2) of peloidal facies (c1-c2) and oolitic facies (d1-d2).
-e-f) Type 8 macroscopic appearance (e-f1) and thin section (f2-f3); white and black arrows correspond to foraminifera.
-g) Type 9 macroscopic appearance (g1) and thin section (g2).
-Type 10 macroscopic appearance (h1) and thin section (h2).
+a-b) Type 6 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope with a small degree of white patina and no visible fossils; b1) macroscopic view with a stereomicroscope of a thick alteration rind but translucent interior where some unidentifiable fossils are visible; a2) microscopic view of thin section (PPL) with a mudstone texture; b2) microscopic of thin section (XPL) with small unidentifiable fossils and a laminar microstructure.
+c-d) Type 7 macroscopic and microscopic views of the peloidal facies (c) and oolitic facies (d): c1) macroscopic view with a stereomicroscope of peloids and sponge spicules; d1) macroscopic view with a stereomicroscope of poorly preserved ooids; c2) microscopic view of thin section (PPL) with packstone texture composed mainly of peloids; d2) microscopic view of thin section (PPL) with packstone texture composed mainly of ooids.
+e-f) Type 8 macroscopic and microscopic views, with foraminifera represented by the white and black arrows: e) macroscopic view with a stereomicroscope with abundant unidentifiable fossils and foraminifera; f1) macroscopic view with a stereomicroscope with abundant unidentifiable fossils but no visible foraminifera; f2) microscopic view of thin section (PPL) with packstone texture composed of unidentifiable fossils and foraminifera which are not visible macroscopically; f3) microscopic view of thin section (PPL) with packstone texture composed of unidentifiable fossils and foraminifera which not are visible macroscopically.
+g) Type 9 macroscopic and microscopic views: g1) macroscopic view with a stereomicroscope of unidentifiable fossils and two different fabrics; g2) microscopic view of thin section (XPL) with packstone texture composed of concentrations of unidentifiable fossils and opaques which create a banded microstructure.
+h) Type 10 macroscopic and microscopic views: h1) macroscopic view with a stereomicroscope of ooids and fractures filled with chalcedony; h2) microscopic view of thin section (PPL) with packstone texture composed of ooids with different levels of preservation.
 
@@
 In unaltered samples, visible bioclasts are rare.
-The cortex is mostly rounded, and rarely pebble-like, ranging from thin to medium thickness.
+The cortex is mostly rounded, and rarely pebble-like, ranging from thin to medium thickness (@tbl-cortex_tab).
 Two sub-types were individualized based on macroscopic characteristics such as colour, translucency variation, and patterns, although they may simply reflect the use of specific nodules from within Type 6’s variability.
@@
 Alike the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matiasRawMaterialSourcing2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
-From these cherts, only the dark grey nodules (RT231) showed identifiable fossils which were rare ostracods.
+From these Cretaceous cherts from Central Portugal, only the dark grey nodules (sample RT231, with petrographic descriptions available in our online research compendium) showed identifiable fossils which were rare ostracods.
 
@@
 It is characterized by white and grey colours.
-The cortex is from an outcrop, rounded and in lesser frequencies (\<30%) irregular, and varying between medium thickness to thick.
+The cortex is from an outcrop, rounded and in lesser frequencies (\<30%) irregular, and varying between medium thickness to thick (@tbl-cortex_tab).
 This type shows several facies, with different types and concentrations of allochems and bioclasts, such as peloids (@fig-distance-types c1), ooids (@fig-distance-types d1), and sponge spicules.
@@
 The translucency ranges from sub-translucent to opaque.
-Whenever present, the cortex is rounded and thin.
+Whenever present, the cortex is rounded and thin (@tbl-cortex_tab).
 Petrographically, this type is a packstone (mudstone texture close to the edges) with a massive microstructure.
@@
 However, our results indicate that despite no foraminifera being visible macroscopically, they are present in thin section.
-The fact that no foraminifera were identified in the small number of analysed samples from the UCA lithotheque may simply reflect their uncommon visibility, possibly due to patination.
+The fact that no foraminifera were identified in the small number of analysed samples from the UCA lithotheque with similarities to Type 8 (n=3) may simply reflect their uncommon visibility, possibly due to patination.
 Another possible source for Type 8 cherts may be the Upper Cretaceous wackestone/packstone cherts with foraminifers from the Lisbon region [@jordaoFlintSourcesMobility2022].
@@
 It is opaque, has a heterogeneous structure and is characterized by a variable colour distribution: it shows a horizontal laminated and finely laminated pattern, with grey dark bands intercalated with light grey bands (@fig-distance-types g1).
-Whenever present, the cortex is rounded and is generally thin.
+Whenever present, the cortex is rounded and is generally thin (@tbl-cortex_tab).
 In specific samples, the cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers.
@@
 #| label: fig-small-types
-#| fig-cap: Type 11 and TL cherts macroscopic variability.
+#| fig-cap: Macroscopic view with a stereomicroscope of non-local Type 11 and TL cherts.
+Type 11 – detail of banded patterns composed of parallel different fabrics; TL01 – detail of the jasper and fractures filled with chalcedony; TL02 – detail of the chalcedony with unidentified orange inclusions; TL03 – detail of two round inclusions, possibly iron oxides; TL04 - detail of massive chert with fractures filled with chalcedony; TL05 – detail of the banded pattern within the grey mudstone texture; TL06 - detail of the small, finely sorted ooids; TL07 – detail of the dendritic laminations, possibly composed of iron oxides; TL08 – detail of the mudstone structure, without the presence of allochems; TL09a/b – detail of the mossy, orange and red inclusions within the chalcedony; TL10 – detail of the speckling and laminated patterns within the mudstone structure; TL11 – detail of the banding within the mudstone structure; TL12 – detail of the poorly preserved ooids, only visible under the stereomicroscope; TL13 – detail of the bands and laminations within the mudstone texture fabric; TL14 – detail of an unidentified benthic foraminifera; TL15 – detail of the interface between the massive brown fabric and splotches; TL16 - detail of the of the massive structure with unidentifiable inclusions.
 
@@
 
-Finally, the TL types (@fig-small-types) are characterized by several samples with differing characteristics from the previously established types, and without correspondence to the local cherts and chalcedony from the Algarve region previously described [@belmiroCreatingFramesReference2023].
-These are composed mostly of blanks or retouched tools, rarely with cortex, and rarely the same TL type is found in different archaeological layers.
-A detailed description of these types can be found in the Supplermentary Information (Online Resource 2).
+Finally, the TL types (@fig-small-types) are characterized by several samples without macroscopic correspondence to the previously identified types or to the local cherts and chalcedony from the Algarve region previously described by Belmiro et al.
+[-@belmiroCreatingFramesReference2023].
+These are composed mostly of blanks or retouched tools, rarely with cortex (@tbl-cortex_tab), and rarely the same TL type is found in different archaeological layers.
+A detailed description of these types can be found in the Supplementary Information (Online Resource 2).
 In both areas, the TL cherts compose less than 1% of the cherts.
@@
 Most chert types described above are consistently present across the Terrace stratigraphy (except for Type 11 which appears only in level 4 and in a small number of artefacts; n=2).
-However, being ubiquitous, we identified a difference in the percentages of chert types throughout the stratigraphy.
-As such, a chi-square test of independence was performed to determine whether the percentages of chert types are related to the archaeological layer.
-The results are significant, X-square = 817.79, p \< 2.2e-16, showing that there are statistically significant differences in the chert types present throughout the occupations.
-A full description of chi-square test and Pearson residuals can be found in the Supplementary Information (Online Resource 4).
+However, being ubiquitous, we identified a difference in the percentages of chert types throughout the stratigraphy (@fig-type-per-level).
 
@@
 In the Proto-Solutrean (levels 4E and 4E/5) and Solutrean (levels 4, 4C, and 4D) local cherts make up more than \~70% of the sample and 80% when considering the sum of weight, while non-local cherts decrease significantly and represent percentages lower than 30%.
-Despite this homogeneous pattern, local types show statistically significant differences between the two technocomplexes.
+Despite the increase of local cherts during the Proto-Solutrean and Solutrean, there are still statistically significant differences between the local types present in the two technocomplexes.
 During the top Proto-Solutrean occupations the grey/white cherts (Type 4) are the most present chert with percentages between 35-20%, while in the Solutrean levels, the yellow/brown cherts (Type 2) are present in the highest percentage, at \~30%.
@@
 #| label: fig-type-per-level
-#| fig-cap: Terrace area chalcedony and chert type frequency (a) and weight sum (b) per level, without unidentifiable samples. Only percentages superior to the 5% threshold are shown in the figure.
+#| fig-cap: Terrace area chalcedony and chert type frequency (a) and weight sum (b) per level, without unidentifiable samples.
+Only percentages superior to the 5% threshold are shown in the figure.
+Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts.
+Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors).
+Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors).
 #| fig-width: 10
@@
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
-                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))
 
@@
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
-                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
+                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
   theme(axis.text.x=element_blank(),
@@
 
-When plotting the weight, it is apparent that most of the samples are located under the 0.3 kg mark, with median and bottom quartiles often falling under the 0.5 kg line (@fig-weight).
-Local types, such as Type 2, Type 4, and Type 5, although mostly represented by artefacts with a small mass (\~0.3 kg), show the heaviest and highest number of heavy artefacts.
+When plotting the weight, it is apparent that most of the samples are located under the 10 gr mark, with median and bottom quartiles often falling under the 5 gr line (@fig-weight).
+Local types, such as Type 2, Type 4, and Type 5, although mostly represented by artefacts with a small mass (\~3 gr), show the heaviest and highest number of heavy artefacts.
 In comparison, other types, although showing some variation in the weight of artefacts, show smaller artefacts and their means and medians are frequently closer to the bottom quartile.
@@
 #| label: fig-weight
-#| fig-cap: "Terrace boxplots of chert type weight per level. Diamond shape within the boxplots represents weight mean."
+#| fig-cap: "Terrace boxplots of chert type weight (gr) per level.
+Diamond shape within the boxplots represents weight mean.
+Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts.
+Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors).
+Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors)."
 #| #| fig-width: 10
@@
   mutate(Technocomplex = factor(Technocomplex, levels = c("Solutrean", "Proto-Solutrean", "Gravettian"))) %>%
-  ggplot(aes(RMTYPE, WEIGHT, color = RMTYPE)) +
-  geom_boxplot() +
+  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
   facet_wrap(~Technocomplex + Level) +
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight") +
+  labs(y="Weight (gr)") +
   coord_cartesian(ylim =  c(0, 40)) +
-  scale_color_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
-                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
   theme(axis.text.x=element_blank(),
@@
 The same raw materials found in the Terrace area are also found in the Solutrean occupation of the Shelter, albeit with some differences (@fig-analysis_shelter).
-The local cherts make up between \~70-80% of the sample and are mainly composed of the brown/yellow cherts (Type 2) that frequently characterize the Lower Jurassic outcrops of the Algarve (@fig-analysis_shelter a-b).
+The local cherts make up between \~65-75% (considering both counts and weight) of the sample and are mainly composed of the brown/yellow cherts (Type 2) that frequently characterize the Lower Jurassic outcrops of the Algarve (@fig-analysis_shelter a-b).
 The presence of cherts not congruent with the local sources is, like level 4 in the Terrace, close to 25%, although dominated mostly by Type 6 (\~15%).
@@
 Only percentages superior to the 5% threshold are shown in the figure.
-c): Shelter boxplots of chert type weight per level."
+c): Shelter boxplots of chert type weight (gr) per level.
+Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts.
+Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors).
+Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors)."
 #| #| fig-width: 10
@@
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
-                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))
 ## Weight barplot
@@
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
-                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))
+
+#ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
+  #geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
 ## Weight boxplot
@@
   filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","TL") & Layer != "ND") %>%
-  ggplot(aes(RMTYPE, WEIGHT, color = RMTYPE)) +
-  geom_boxplot() +
+  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
   facet_wrap(.~Layer, ncol = 1) +
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y = "Weight") +
+  labs(y = "Weight (gr)") +
   #geom_jitter(shape=16, position=position_jitter(0.2)) +
   coord_cartesian(ylim =  c(0, 35)) +
-  scale_color_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
-                              "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8")) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
   theme(axis.text.x=element_blank(),
@@
 The local chalcedony and chert indicate that the groups were sourcing a significant portion of their raw materials within approximately a 20 km radius of the site.
-The direct procurement of these materials is plausible, given that known chalcedony sources are situated to the west, about 6-7 km from Vale Boi, and chert sources are found to the east, within the inland Lower Jurassic outcrops, roughly 8 km away (@fig-map b).
+The direct procurement of these materials is plausible, given that known chalcedony primary sources are situated to the west, about 10 km from Vale Boi (or possibly in the vicinity of the site), and chert sources are found to the east, within the inland Lower Jurassic outcrops, roughly 8 km away (@fig-map b).
 This hypothesis is supported by the potential existence of UP occupations near these outcrops, which might reflect quarrying and chert catchment activities during prehistory [@zambujo_jazidas_1998].
-Evidence of short-distance procurement at Vale Boi includes the presence of heavy nodules and large chunks of chert from local varieties with parent rock still attached, as well as a significant number of outliers (representing larger blocks and chunks) in the weight boxplots across the stratigraphy (Figure 4), that may have been discarded due to natural fractures.
+Evidence of short-distance procurement at Vale Boi includes the presence of heavy nodules (@fig-weight) with little indication of knapping and large chunks of chert from local varieties, often with parent rock still attached.
 
@@
 
-The limitations of macroscopic and petrographic methods in discerning between outcrops due to the chert nodule similarities have also impaired the specific distinction between Lower Jurassic sources in Western Algarve (within approximately a 20 km radius of the site) and Middle Jurassic cherts located in Eastern Algarve (approximately 80 km to the east; @fig-map c).
+The limitations of macroscopic and petrographic methods in discerning between outcrops due to the chert nodule similarities have also impaired the specific distinction between Lower Jurassic and Upper Jurassic sources in Western Algarve (within approximately a 20 km radius of the site) and Middle Jurassic cherts located in Eastern Algarve (approximately 80 km to the east; @fig-map c).
 As such, the presence of these regional resources at Vale Boi is uncertain.
@@
 
-So far, we have discussed the presence of cherts whose outcrops are in relative proximity to Vale Boi and show the full usage of local chert variability and the selective usage of regional cherts.
-As previously mentioned, however, it is the combined identification of local and long-distance cherts that allows us to have a comprehensive idea of the total exploited territory, as well as the possible existence of exchanges and social networks between groups occupying different territories.
-
-Although a wide variety of chert types incongruent with the regional reference collection were identified throughout the stratigraphy and classified as non-local, we were only able to ascertain the possible sources of three types.
-The long-distance cherts with identified sources come from distances of over \~200 km and in two main directions: 1) from the north, where Cretaceous cherts from Central Portugal are found (Figure 1 a), along with a small amount of jasper (TL01); and 2) from the east, where peloidal/oolitic chert from Western Andalusia is present (@fig-map), and possibly jasper, transported by the Guadiana river at the current border between Portugal and Spain.
+A similar case is observed for the jaspers located in the SPZ [@oliveira_carta_1984-1], located at a minimum of 100 km distance from Vale Boi, as the crow flies.
+Despite being a source located at closer distances when compared to the cherts of Central Portugal or from Southwestern Spain, the jasper occurs in trace amounts (n=3) and only in the Gravettian layers (layer 6).
+Occurring North and Northeast, the limited amount of jasper in the assemblage may either represent a limited use of these resources in detriment of the cherts from the Lisbon and Rio Maior area, or suggest that they were used and knapped elsewhere, but not transported, either by the same groups or through exchanges by different groups.
+
+So far, we have discussed the presence of cherts whose outcrops are in relative proximity to Vale Boi and show the full usage of local chert variability (albeit with uncertainty regarding the Upper Jurassic cherts of Western Algarve due to their macroscopic similarities) and the selective usage of regional cherts within the Algarve.
+As previously mentioned, however, it is the combined identification of local and long-distance cherts that allows us to have a better idea of the total exploited territory, as well as the possible existence of exchanges and social networks between groups occupying different territories.
+
+Although a wide variety of chert types incongruent with the regional reference collection were identified throughout the stratigraphy and classified as non-local, we were only able to ascertain the possible sources of three types: Type 6, Type 7 and Type 9.
+The long-distance cherts with identified sources come from distances of over \~200 km and in two main directions: 1) from the north, where Cretaceous cherts from Central Portugal are found (Figure 1 a); and 2) from the east, where peloidal/oolitic chert from Western Andalusia is present (@fig-map), and possibly jasper, transported by the Guadiana river at the current border between Portugal and Spain.
 
@@
 
-As such, the observed long-distance mobility patterns in the north/east directions by hunter-gatherers at Vale Boi might reflect this Central and Southern Iberia territory dynamic and the exploitation of the natural corridors for movement, explaining the substantial use of long-distance chert.
-In this case, the presence of a smaller subset of non-local cherts of unknown origin (TL) but unknapped at the site could represent different forms of long-distance trade of finished goods and blanks, or interactions with various social networks, potentially even indicating individual networks and mobility strategies [sensu @gamblePalaeolithicSocietiesEurope1999].
+As such, the observed long-distance transfer of cherts in the north/east directions by hunter-gatherers at Vale Boi might reflect this Central and Southern Iberia territory dynamic and the exploitation of the natural corridors for movement, explaining the substantial use of long-distance chert.
+In this case, the presence of a smaller subset of non-local cherts of unknown origin (TL) but not knapped at the site could represent different forms of long-distance trade of finished goods and blanks, or interactions with various social networks, potentially even indicating individual networks and mobility strategies [sensu @gamblePalaeolithicSocietiesEurope1999].
 
@@
 For example, @cascalheira_2017 acknowledge the adoption of specific behavioural strategies by Vale Boi hunter-gatherers in response to the regional ecological conditions, which were sustained throughout the UP at the site.
-Practices such as grease rendering or the functional specialization of raw materials are seen as deliberate actions with significant cultural implications for future generations, influencing the adaptive choices made over time.
+Practices such as grease rendering or the functional specialization of raw materials (e.g., the use of quartz and greywacke for bone breakage activities or quartz for grease rendering) are seen as deliberate actions with significant cultural implications for future generations, influencing the adaptive choices made over time.
 Accordingly, chert raw material acquisition and management appear to be integral to these behavioural strategies, established from the site’s earliest occupations and perpetuated throughout the UP.
@@
 
-These findings reveal a landscape of hunter-gatherers engaged in complex practices of landscape exploration, resource management, and possibly risk management.
+These findings reveal a landscape of hunter-gatherers engaged in complex practices of resource management, and possibly risk management.
 The presence of allochthonous raw materials, albeit in small quantities during the Proto-Solutrean and Solutrean phases, underscores that the inhabitants of Vale Boi—situated on the periphery of Iberia and Europe, bordered by the sea to the south and west, and confined to the patchy refugia created by the worsening climatic conditions of the HE2 and following Glacial Maximum [@schmidtRapidClimateChange2012]—were certainly not isolated.
-Instead, they were part of an extensive social network, a pattern of interconnectivity that resonates with findings across Western Europe.
+Instead, they were part of an extensive social network, a pattern of interconnectivity that resonates with findings across Western Europe [@hermidaSilexNWPeninsula2016; @corchonrodriguezMobiliteTerritoiresRelations2016; @ortegaMobilitatDesplacamentsDels2003; @aubryUpperPalaeolithicLithic2016; @aubryRawMaterialProcurement2004; @sanchezdelatorreNewDataChert2023].
 
@@
 
-**Online resource 4**: Statistical analysis of Terrace chert types.
-
 <!-- The following line inserts a page break  -->

### 2024-11-04 analysis/paper/paper.qmd
- revision for reviewer 2 - correction of data csv file to british english (thank you Jack)

@@
 
-Based on these interpretations and studies, a set of hypotheses and expectations regarding chert procurement and human mobility throughout the UP at Vale Boi can be proposed.
+Based on the notions that the role of Vale Boi during the Solutrean as a connection point between different regions may have promoted the exchange of information and culture, including raw materials from long-distances, and the limited circulation of people during the Gravettian may have isolated groups occupying Vale Boi, making them mostly reliant on the available local or regional resources, a set of hypotheses and expectations regarding chert procurement and human mobility throughout the UP at Vale Boi can be proposed.
 On one hand, it would be expected that Solutrean occupations are characterized by the predominant use of local raw materials but with a considerable percentage of cherts obtained from long-distance sources, possibly through exchange with different groups in Iberia.
@@
 Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being \~20 cm, and frequently characterized by high levels of dolomitization and fractures.
+Petrographically, they are composed mainly of microcrystalline quartz, with wackstone packstone textures, small percentages of macrocrystalline quartz, fibrous chalcedony and allochems composed of oxides and poorly preserved bioclasts.
+When identifiable, the bioclasts are Echinoderms, Radiolarians, Sponge spicules and bivalve shells.
+The Upper Jurassic cherts of Mareta show similar petrographic characteristics, but with abundant Calcispheres which are not always apparent under the stereomicroscope.
 
@@
 The Middle Jurassic Malhão formation chert outcrops were identified at around 80-100 km east of the site and characterized by chert nodules with similar macroscopic features to those found in Western Algarve.
+Petrographically they are composed mainly by microcrystalline quartz and characterized by wackestone textures, while dolomite, chalcedony and macrocrystalline quartz are also present in small percentages (\<10%).
+Allochems are iron oxides, and albeit poorly preserved, several bioclasts can be identified: Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres and possibly Tentaculites.
 The Upper Jurassic Jordana formation chert outcrops were identified at around 90 km east of Vale Boi and characterized by chert nodules with significantly different macroscopic features to those of other formations, with frequently small sizes (\~5 cm), difficult removal from the parent rock and unidentified secondary deposits.
+Petrographically they are composed mostly of microcrystalline quartz, with wackstone to packstone textures, and small percentages of fibrous chalcedony and dolomite (\<10%).
+They show frequent iron oxides and very frequent bioclasts, albeit frequently poorly preserved and unidentifiable; identified fossils are fossils are Calcispheres, Bivalve shells, Sponge spicules, Ostracods, Echinoderms and Gastropods.
 
@@
 The term chalcedony is used in its petrographic sense and refers to cherts with a fibrous quartz structure.
+Since this definition of chalcedony implies a destructive method which cannot be applied to all archaeological chalcedony samples, we use the term based on the confirmed correlation between the petrographic characteristics of chalcedony and their macroscopic appearance, both through geological and archaeological thin sections.
 
@@
 Consequently, we cannot rule out the possibility that some of the cherts we have identified as being sourced within a 20 km radius of Vale Boi may actually be the result of regional procurement and mobility, potentially from areas approximately 80-100 km away, where the Middle Jurassic cherts are located.
-Given the similarity of the raw materials and the absence of geochemical studies, and considering the study and model mentioned above, most of these cherts likely originate from local sources, while a minor portion may be attributable to regional movements.
+Given the similarity of the raw materials and the absence of geochemical studies which have shown to be key to differentiate between cherts belonging to the same geological formations or cherts formed under similar conditions [@gomezdesolerMultitechniqueApproachCharacterization2023], and considering the studies and model mentioned above, most of these cherts likely originate from local sources, while a minor portion may be attributable to regional movements.
 
@@
 Although a wide variety of chert types incongruent with the regional reference collection were identified throughout the stratigraphy and classified as non-local, we were only able to ascertain the possible sources of three types: Type 6, Type 7 and Type 9.
-The long-distance cherts with identified sources come from distances of over \~200 km and in two main directions: 1) from the north, where Cretaceous cherts from Central Portugal are found (Figure 1 a); and 2) from the east, where peloidal/oolitic chert from Western Andalusia is present (@fig-map), and possibly jasper, transported by the Guadiana river at the current border between Portugal and Spain.
+The long-distance cherts with identified sources come from distances of over \~200 km and in two main directions: 1) from the north, where Cretaceous cherts from Central Portugal are found (@fig-map a); and 2) from the east, where peloidal/oolitic chert from Western Andalusia is present (@fig-map), and possibly jasper, transported by the Guadiana river at the current border between Portugal and Spain.
 

### 2024-11-05 analysis/paper/paper.qmd
- corrected refs - grammar corrections

@@
 
-Several key topics in the study of hunter-gatherer behaviour and organization have been explored in different geographic regions and chronological periods through raw material analysis, such as modalities of procurement and mobility strategies [@ambroseSocialEcologicalModels1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccallBehavioralEcologicalModels2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallonSocialNetworksInformation2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999].
+Several key topics in the study of hunter-gatherer behaviour and organisation have been explored in different geographic regions and chronological periods through raw material analysis, such as modalities of procurement and mobility strategies [@ambroseSocialEcologicalModels1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccallBehavioralEcologicalModels2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallonSocialNetworksInformation2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999].
 These analyses are often achieved by systematically characterizing geological and archaeological raw materials and establishing correlations between samples from both origins.
@@
 Cascalheira et al.
-[-@cascalheira_2017] show that, with the exception of the Early Gravettian occupations of the Terrace area and Magdalenian occupation of the Shelter, all other UP occupations are composed of high-density assemblages with low degree of retouch, correspondent to an residential base-camp occupation, for extended periods of time.
+[-@cascalheira_2017] show that, with the exception of the Early Gravettian occupations of the Terrace area and Magdalenian occupation of the Shelter, all other UP occupations are composed of high-density assemblages with low degree of retouch, correspondent to a residential base-camp occupation, for extended periods of time.
 
@@
 The Algarve region is a complex territory, marked by several geomorphic sub-regions and geological units that provide different lithological resources.
-Two of the main geological units which characterize the Algarve are: 1) the Baixo Alentejo Flyschoid Group, part of the South Portuguese Zone [@fernandesNewEvidenceConcerning2012; @oliveira_carta_1984-1; @onezimeNewGeodynamicInterpretation2003], located in the north sector of the region, and characterized by schists, greywackes and quartzites; and 2) the Algarve basin in the south, which is characterized by limestones, dolomites and marls [@terrinha_bacia_2006].
-It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules/beds can be found [@manuppellaNoticiaExplicativaFolha1987; @manuppellaNoticiaExplicativaFolha2007; @ribeiro_evolucao_2005; @rocha_noticia_1979; @rocha_noticia_1983; @terrinha_bacia_2006].
+Two of the main geological units which characterize the Algarve are: 1) the Baixo Alentejo Flyschoid Group, part of the South Portuguese Zone [@fernandesNewEvidenceConcerning2012; @oliveiraCartaGeologicaPortugal1984; @onezimeNewGeodynamicInterpretation2003], located in the north sector of the region, and characterized by schists, greywackes and quartzites; and 2) the Algarve basin in the south, which is characterized by limestones, dolomites and marls [@terrinhaBaciaAlgarveEstratigrafia2006].
+It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules/beds can be found [@manuppellaNoticiaExplicativaFolha1987; @manuppellaNoticiaExplicativaFolha2007; @ribeiro_evolucao_2005; @rochaNoticiaExplicativaFolha1979; @rochaNoticiaExplicativaFolha1983; @terrinhaBaciaAlgarveEstratigrafia2006].
 
@@
 
-We employed both macroscopic and petrographic methods to achieve a more comprehensive characterization of the materials, aiming to offset the inherent limitations of each method [@luedtke_archaeologists_1992].
+We employed both macroscopic and petrographic methods to achieve a more comprehensive characterization of the materials, aiming to offset the inherent limitations of each method [@luedtkeArchaeologistGuideChert1992].
 
@@
 The distinct features of various local cherts facilitate their differentiation from cherts of neighbouring origins.
-Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustilloMacroscopicClassificationFlint2009; @tarrino_materias_2013].
+Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustilloMacroscopicClassificationFlint2009; @tarrinoMateriasPrimasLiticas2013].
 Therefore, we implemented an initial phase of macroscopic analysis, subsequently augmented by petrographic study to diminish the subjectivity of the analysis and enhance the material characterization with petrographic data.
@@
 The first step included a preliminary characterization using a hand lens of 10x magnification.
-This step allowed us to create types and subtypes based on similar macroscopic characteristics frequently acknowledged to be useful to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1).
+This step allowed us to create types and subtypes based on similar macroscopic characteristics frequently acknowledged to be useful to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtkeArchaeologistGuideChert1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1).
 In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types - the data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1).
@@
 #| label: fig-local-types
-#| fig-cap: Variability of local chert types and main petrographic characteristics.
+#| fig-cap: "Variability of local chert types and main petrographic characteristics.
 a) Type 1 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope of white patina and fractures; a2) microscopic view of thin section (XPL), with visible fibrous chalcedony composition and without allochems.
@@
 e-f) Type 4 macroscopic and microscopic views: e) macroscopic view with a stereomicroscope of unidentifiable fossils and white splotches; f1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils within white and yellow splotches; f2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides; f3) microscopic view of thin section (PPL) with a concentration of fossils (possibly sponge spicules) and a fracture.
-g-h) Type 5 macroscopic and microscopic views: g) macroscopic view with a stereomicroscope; h1) macroscopic view with a stereomicroscope with visible unidentifiable fossils; h2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; h3) microscopic view of thin section (XPL) with small unidentifiable fossils, iron oxides and a possible ostracod replaced by fibrous chalcedony and macrocrystalline quartz.
+g-h) Type 5 macroscopic and microscopic views: g) macroscopic view with a stereomicroscope; h1) macroscopic view with a stereomicroscope with visible unidentifiable fossils; h2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; h3) microscopic view of thin section (XPL) with small unidentifiable fossils, iron oxides and a possible ostracod replaced by fibrous chalcedony and macrocrystalline quartz."
 
@@
 #| label: fig-distance-types
-#| fig-cap: Variability of non-local types and main petrographic characteristics.
+#| fig-cap: "Variability of non-local types and main petrographic characteristics.
 a-b) Type 6 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope with a small degree of white patina and no visible fossils; b1) macroscopic view with a stereomicroscope of a thick alteration rind but translucent interior where some unidentifiable fossils are visible; a2) microscopic view of thin section (PPL) with a mudstone texture; b2) microscopic of thin section (XPL) with small unidentifiable fossils and a laminar microstructure.
@@
 g) Type 9 macroscopic and microscopic views: g1) macroscopic view with a stereomicroscope of unidentifiable fossils and two different fabrics; g2) microscopic view of thin section (XPL) with packstone texture composed of concentrations of unidentifiable fossils and opaques which create a banded microstructure.
-h) Type 10 macroscopic and microscopic views: h1) macroscopic view with a stereomicroscope of ooids and fractures filled with chalcedony; h2) microscopic view of thin section (PPL) with packstone texture composed of ooids with different levels of preservation.
+h) Type 10 macroscopic and microscopic views: h1) macroscopic view with a stereomicroscope of ooids and fractures filled with chalcedony; h2) microscopic view of thin section (PPL) with packstone texture composed of ooids with different levels of preservation."
 
@@
 This chert shows no similarities to any consulted reference material.
-Despite the existence of oolitic limestones in the Algarve region correspondent to Middle and Upper Jurassic formations and often described as whiteish (Rocha et al., 1979), the presence of oolitic chert in the region is not recognized in geological and archaeological literature [@cardosoSobrePresencaLaminas2018; @rocha_noticia_1979; @verissimo_jazidas_2004] and the visit of outcrops mentioned in the geological cartography did not allow their identification.
+Despite the existence of oolitic limestones in the Algarve region correspondent to Middle and Upper Jurassic formations and often described as whiteish (Rocha et al., 1979), the presence of oolitic chert in the region is not recognized in geological and archaeological literature [@cardosoSobrePresencaLaminas2018; @rochaNoticiaExplicativaFolha1979; @verissimo_jazidas_2004] and the visit of outcrops mentioned in the geological cartography did not allow their identification.
 
@@
 #| label: fig-small-types
-#| fig-cap: Macroscopic view with a stereomicroscope of non-local Type 11 and TL cherts.
-Type 11 – detail of banded patterns composed of parallel different fabrics; TL01 – detail of the jasper and fractures filled with chalcedony; TL02 – detail of the chalcedony with unidentified orange inclusions; TL03 – detail of two round inclusions, possibly iron oxides; TL04 - detail of massive chert with fractures filled with chalcedony; TL05 – detail of the banded pattern within the grey mudstone texture; TL06 - detail of the small, finely sorted ooids; TL07 – detail of the dendritic laminations, possibly composed of iron oxides; TL08 – detail of the mudstone structure, without the presence of allochems; TL09a/b – detail of the mossy, orange and red inclusions within the chalcedony; TL10 – detail of the speckling and laminated patterns within the mudstone structure; TL11 – detail of the banding within the mudstone structure; TL12 – detail of the poorly preserved ooids, only visible under the stereomicroscope; TL13 – detail of the bands and laminations within the mudstone texture fabric; TL14 – detail of an unidentified benthic foraminifera; TL15 – detail of the interface between the massive brown fabric and splotches; TL16 - detail of the of the massive structure with unidentifiable inclusions.
+#| fig-cap: "Macroscopic view with a stereomicroscope of non-local Type 11 and TL cherts.
+Type 11 – detail of banded patterns composed of parallel different fabrics; TL01 – detail of the jasper and fractures filled with chalcedony; TL02 – detail of the chalcedony with unidentified orange inclusions; TL03 – detail of two round inclusions, possibly iron oxides; TL04 - detail of massive chert with fractures filled with chalcedony; TL05 – detail of the banded pattern within the grey mudstone texture; TL06 - detail of the small, finely sorted ooids; TL07 – detail of the dendritic laminations, possibly composed of iron oxides; TL08 – detail of the mudstone structure, without the presence of allochems; TL09a/b – detail of the mossy, orange and red inclusions within the chalcedony; TL10 – detail of the speckling and laminated patterns within the mudstone structure; TL11 – detail of the banding within the mudstone structure; TL12 – detail of the poorly preserved ooids, only visible under the stereomicroscope; TL13 – detail of the bands and laminations within the mudstone texture fabric; TL14 – detail of an unidentified benthic foraminifera; TL15 – detail of the interface between the massive brown fabric and splotches; TL16 - detail of the of the massive structure with unidentifiable inclusions."
 
@@
 
-Jasper is known in the South Portuguese Zone area [@oliveira_carta_1984-1], from Southern/Central Portugal to Spain, north of the Algarve (from Aljustrel to Beja; @fig-map), suggesting that the source of TL01 may be from south/central Portugal at a minimum distance of 100 km north and northeast from Vale Boi (as the crow flies).
+Jasper is known in the South Portuguese Zone area [@oliveiraCartaGeologicaPortugal1984], from Southern/Central Portugal to Spain, north of the Algarve (from Aljustrel to Beja; @fig-map), suggesting that the source of TL01 may be from south/central Portugal at a minimum distance of 100 km north and northeast from Vale Boi (as the crow flies).
 
@@
 In the Proto-Solutrean (levels 4E and 4E/5) and Solutrean (levels 4, 4C, and 4D) local cherts make up more than \~70% of the sample and 80% when considering the sum of weight, while non-local cherts decrease significantly and represent percentages lower than 30%.
-Despite the increase of local cherts during the Proto-Solutrean and Solutrean, there are still statistically significant differences between the local types present in the two technocomplexes.
+Despite the increase of local cherts during the Proto-Solutrean and Solutrean, there are still significant differences between the local types present in the two technocomplexes.
 During the top Proto-Solutrean occupations the grey/white cherts (Type 4) are the most present chert with percentages between 35-20%, while in the Solutrean levels, the yellow/brown cherts (Type 2) are present in the highest percentage, at \~30%.
@@
 Only percentages superior to the 5% threshold are shown in the figure.
-Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts.
-Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors).
-Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors).
+Type 1 to 5 (yellow and red colours) represent local chalcedony and cherts.
+Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colours).
+Type 8, 10 and 11 represent non-local cherts with non-identified provenience (blue and green colours).
 #| fig-width: 10
@@
 When plotting the weight, it is apparent that most of the samples are located under the 10 gr mark, with median and bottom quartiles often falling under the 5 gr line (@fig-weight).
-Local types, such as Type 2, Type 4, and Type 5, although mostly represented by artefacts with a small mass (\~3 gr), show the heaviest and highest number of heavy artefacts.
+Local types, such as Type 2, Type 4, and Type 5, although mostly represented by artefacts with a medium mass (\~5-10 gr), show the heaviest and highest number of heavy artefacts.
 In comparison, other types, although showing some variation in the weight of artefacts, show smaller artefacts and their means and medians are frequently closer to the bottom quartile.
-As seen in Figure 3 B, non-local types show larger weights from level 5 to level 7.
+As seen in (@fig-weight), non-local types show larger weights from level 5 to level 7.
 
@@
 Diamond shape within the boxplots represents weight mean.
-Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts.
-Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors).
-Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors)."
+Type 1 to 5 (yellow and red colours) represent local chalcedony and cherts.
+Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colours).
+Type 8, 10 and 11 represent non-local cherts with non-identified provenience (blue and green colours)."
 #| #| fig-width: 10
@@
 c): Shelter boxplots of chert type weight (gr) per level.
-Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts.
-Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors).
-Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors)."
+Type 1 to 5 (yellow and red colours) represent local chalcedony and cherts.
+Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colours).
+Type 8, 10 and 11 represent non-local cherts with non-identified provenience (blue and green colours)."
 #| #| fig-width: 10
@@
 
-A similar case is observed for the jaspers located in the SPZ [@oliveira_carta_1984-1], located at a minimum of 100 km distance from Vale Boi, as the crow flies.
+A similar case is observed for the jaspers located in the SPZ [@oliveiraCartaGeologicaPortugal1984], located at a minimum of 100 km distance from Vale Boi, as the crow flies.
 Despite being a source located at closer distances when compared to the cherts of Central Portugal or from Southwestern Spain, the jasper occurs in trace amounts (n=3) and only in the Gravettian layers (layer 6).

### 2024-11-05 analysis/paper/paper.qmd
- Small grammar improvements (thank you Jack and Cedric)

@@
 Our study reveals that the majority of chert found at Vale Boi originates locally, within a 20 km radius.
-However, a non-neglectable portion of the chert comes from non-local sources, indicating >200 km raw material circulation from Central Portugal and Southern Spain.
+However, a non-negligible portion of the chert comes from non-local sources, indicating >200 km raw material circulation from Central Portugal and Southern Spain.
 keywords: |
@@
 
-Based on the notions that the role of Vale Boi during the Solutrean as a connection point between different regions may have promoted the exchange of information and culture, including raw materials from long-distances, and the limited circulation of people during the Gravettian may have isolated groups occupying Vale Boi, making them mostly reliant on the available local or regional resources, a set of hypotheses and expectations regarding chert procurement and human mobility throughout the UP at Vale Boi can be proposed.
+Based on the notions that a) during the Solutrean Vale Boi served as a connection point between different regions which may have promoted the exchange of information and culture, including raw materials from long-distances; and b) the limited circulation of people during the Gravettian may have isolated groups occupying Vale Boi, making them mostly reliant on the available local or regional resources, we propose a set of hypotheses and expectations regarding chert procurement and human mobility throughout the UP at Vale Boi.
 On one hand, it would be expected that Solutrean occupations are characterized by the predominant use of local raw materials but with a considerable percentage of cherts obtained from long-distance sources, possibly through exchange with different groups in Iberia.
@@
 The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of Westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b).
-The Lower Jurassic nodules and beds can be found in abundant outcrops, but also frequently in sub-primary or secondary deposition settings in proximity to the outcrops possibly due to natural erosion.
-However, in the case of the inland outcrop of Ferrel, the outcrops is partially destroyed due to recent human action, and chert can be found scattered as chunks or larger nodules.
-The Upper Jurassic nodules can be found sub-primary deposition and of easy access, or in secondary deposition.
-These cherts are frequently opaque with yellow, grey, purple, or red colours, occasionally with bands or laminations, and with frequently visible sponge spicules.
+The Lower Jurassic nodules and beds can be found in abundant outcrops, but also frequently in sub-primary or secondary deposition settings in proximity to the outcrops, possibly due to natural erosion.
+However, in the case of the inland outcrop of Ferrel, the outcrop is partially destroyed due to recent human action, and chert can be found scattered as chunks or larger nodules.
+The Upper Jurassic nodules can be found in sub-primary deposition and are easily accessible, or in secondary deposition.
+These cherts are frequently opaque with yellow, grey, purple or red colours, occasionally with bands or laminations, and frequently with visible sponge spicules.
 Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being \~20 cm, and frequently characterized by high levels of dolomitization and fractures.
-Petrographically, they are composed mainly of microcrystalline quartz, with wackstone packstone textures, small percentages of macrocrystalline quartz, fibrous chalcedony and allochems composed of oxides and poorly preserved bioclasts.
+Petrographically, they are composed mainly of microcrystalline quartz, with wackestone packstone textures, small percentages of macrocrystalline quartz, fibrous chalcedony and allochems composed of oxides and poorly preserved bioclasts.
 When identifiable, the bioclasts are Echinoderms, Radiolarians, Sponge spicules and bivalve shells.
@@
 Petrographically they are composed mostly of microcrystalline quartz, with wackstone to packstone textures, and small percentages of fibrous chalcedony and dolomite (\<10%).
-They show frequent iron oxides and very frequent bioclasts, albeit frequently poorly preserved and unidentifiable; identified fossils are fossils are Calcispheres, Bivalve shells, Sponge spicules, Ostracods, Echinoderms and Gastropods.
-
-Chalcedony outcrops \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013], although previous studies also note the presence of chalcedony (with frequent presence of cortex) in the immediate vicinity of the site [@pereira_territory_2016-1; @gibajabaoProvenienceTechnologyMorphology2013].
+They show frequent iron oxides and very frequent bioclasts, albeit often poorly preserved and unidentifiable; identified fossils are Calcispheres, Bivalve shells, Sponge spicules, Ostracods, Echinoderms and Gastropods.
+
+Chalcedony can be found \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013].
+Previous studies also note the presence of chalcedony (with frequent presence of cortex) in the immediate vicinity of the site [@pereira_territory_2016-1; @gibajabaoProvenienceTechnologyMorphology2013].
 The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.
@@
 
-Chert was chosen for this study as it is one of the three main used raw materials at the archaeological site of Vale Boi alongside quartz and greywacke.
+Chert was chosen for this study as it is one of the three main raw materials used at the archaeological site of Vale Boi alongside quartz and greywacke.
 Other raw materials such as chalcedony, dolerite, and schists have also been identified [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @pereira_territory_2016-1].
@@
 Based on this, despite the smaller amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.
-The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site, such as a Vale Comprido point, the fossil-director for the Proto-Solutrean in Western Portugal [@belmiro_gravettian-solutrean_2021].
+The importance of chalcedony can also be seen in its use to produce formal tool kits during specific technocomplexes at the site, such as a Vale Comprido point, the fossil-director for the Proto-Solutrean in Western Portugal [@belmiro_gravettian-solutrean_2021] or a laurel leaf preform in the Solutrean layers [@gibajabaoProvenienceTechnologyMorphology2013].
 
@@
 Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversitaDevelopmentMultiparametricCharacterisation2019; @gomezdesolerPanadellaChertMontmaneu2020; @sotoProcurementMobilityLate2016].
-For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), and given the unexistence of a reference collection for local altered cherts, they were analysed using the individual sample dataset (Online Resource 1) but grouped in a category of indeterminate cherts (INDET).
+For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), and given the inexistence of a reference collection for local altered cherts, they were analysed using the individual sample dataset (Online Resource 1) but grouped in a category of indeterminate cherts (INDET).
 This category included key alterations such as heat/fire related alterations which clearly impacted the macroscopic characteristics of the samples, fully changing its color, luster and texture, and covering them in crazings (e.g., @fig-terrace-alterations a).
@@
 ```{r}
-
 #| label: tbl-type-source
@@
 As such, we interpret these cherts as local, but possibly from nodules and inland outcrops that are no longer frequently available or visible in the landscape.
-This is especially relevant since the inland outcrops of Ferrel are currently mostly destroyed and can be found in poor preservation conditions, due to recent human intevention and construction work, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].
+This is especially relevant since the inland outcrops of Ferrel are currently mostly destroyed and can be found in poor preservation conditions, due to recent human intervention and construction work, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].
 
@@
 These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from Central Portugal hosted at LusoLit.
-Alike the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matiasRawMaterialSourcing2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%).
+Similar to the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matiasRawMaterialSourcing2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), iron oxides and low porosity (\<1%).
 From these Cretaceous cherts from Central Portugal, only the dark grey nodules (sample RT231, with petrographic descriptions available in our online research compendium) showed identifiable fossils which were rare ostracods.
@@
 Similar cherts have also been identified and described in previous works, describing grey cherts with peloidal and oolitic facies and occasional foraminifera [@rodriguezPrehistoricFlintExploitations2011].
-Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allow us to interpret Type 7 as coming from the Upper Jurassic formations of the Betic Systems, located in the south and southeastern Iberian Peninsula, at least 250 km (as the crow flies) east from Vale Boi (@fig-map).
+Despite the non-existence of thin sections for comparison, the macroscopic similarities between the cherts allow us to interpret Type 7 as coming from the Upper Jurassic formations of the Betic Systems, located in the south and southeastern Iberian Peninsula, at least 250 km (as the crow flies) east from Vale Boi (@fig-map).
 
@@
 Petrographically, this type is a packstone (mudstone texture close to the edges) with a massive microstructure.
-It is composed of 97% micro-cryptocrystalline quartz, macrocrystalline quartz, and fibrous chalcedony, found mostly replacing fossils.
+It is composed of 97% micro-cryptocrystalline quartz, macrocrystalline quartz, and fibrous chalcedony mostly replacing fossils.
 Other minerals include uncommon iron oxides and the porosity is of vug type (3%).
@@
 It corresponds to less than \~2% of the overall cherts in the Terrace and Shelter areas.
-It is opaque, has a heterogeneous structure and is characterized by a variable colour distribution: it shows a horizontal laminated and finely laminated pattern, with grey dark bands intercalated with light grey bands (@fig-distance-types g1).
+It is opaque, has a heterogeneous structure and is characterized by a variable colour distribution: it shows a horizontal, finely laminated pattern, with grey dark bands intercalated with light grey bands (@fig-distance-types g1).
 Whenever present, the cortex is rounded and is generally thin (@tbl-cortex_tab).
@@
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
                               "#996666", "#BDD2B6", "#798777", "#BDC7C9"))
@@
 ggarrange(count_plot, weight_plot, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-```{r}
-#| label: chisq
-#| include: false
-
-library(corrplot)
-
-chisq <- phase_dataset %>% 
-  filter(RMTYPE != "INDET") %>% 
-  filter(RMTYPE != "TYPE 11")
-
-chisq_tab <- table(chisq$RMTYPE, chisq$Level)
-chisq_matrix <- as.table(as.matrix(chisq_tab))
-chisq_rm <- chisq.test(chisq_tab)
-round(chisq_rm$residuals, 3)
-
-corrplot(chisq_rm$residuals, method = "color", is.corr = FALSE)
-
-mosaicplot(chisq_tab, shade = TRUE, las = 2, main = "RMTYPE")
-
-library(vcd)
-vcd_labels <- list(RMTYPE = c("TL", "1", "10", "2", "3", "4",
-                                   "5", "6", "7", "8", "9"))
-mosaic(~ Level + RMTYPE,
-      gp = shading_max,
-      data = chisq, 
-      set_labels = vcd_labels,
-      rot_labels = c(0, 0))
 
@@
   add_overall()
-
-```
-
-```{r}
-#| label: extra-graph
-#| include: false
-
-# Violin plot for RMtype per depth
-phase_dataset %>%
-  filter(Z != "0") %>% 
-  filter(!Z >= "25") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  filter(RMTYPE != "INDET") %>% 
-  ggplot(aes(RMTYPE, Z, fill = RMTYPE)) +
-  geom_violin(trim = FALSE, scale="width", adjust = .5) + 
-  geom_point(position = position_jitter(seed = 1, width = 0.2), alpha = 0.1, color = "darkorange") +
-  geom_hline(yintercept = c(24.5,24.3,24.1,23.8,23.45), colour="#BB0000", linetype="dashed") +
-  labs(title="Groups per depth", 
-       subtitle="Chert groups in the Terrace area by depth",
-       caption="Source: mpg",
-       x="",
-       y="Depth") +
-  stat_summary(fun.data=mean_sdl, geom="errorbar", color="black", width=0.2) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank())
 
@@
   scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
                               "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
@@
 The hypothesis of long-distance procurement of Cretaceous cherts from Central Portugal, situated north of Vale Boi, is corroborated by visual and petrographic evidence [@pereira_territory_2016-1; @bicho_ecodynamics_2013], and this pattern persists throughout the UP.
-Furthermore, the identification of cherts bearing macroscopic resemblance to the peloidal/oolitic Upper Jurassic cherts from the Betic System in Southern Spain confirms a west-east mobility axis, previously inferred from techno-typological studies, spanning also the entirety of the UP period.
+Furthermore, the identification of cherts bearing macroscopic resemblance to the peloidal/oolitic Upper Jurassic cherts from the Betic System in Southern Spain confirms a west-east mobility axis, previously inferred from techno-typological studies, also spanning the entirety of the UP period.
 

### 2024-11-06 analysis/paper/paper.qmd
Small details and grammar corrections

@@
 The Upper Jurassic nodules can be found in sub-primary deposition and are easily accessible, or in secondary deposition.
-These cherts are frequently opaque with yellow, grey, purple or red colours, occasionally with bands or laminations, and frequently with visible sponge spicules.
+The cherts from the previously mentioned formations are are frequently opaque with yellow, grey, purple or red colours, occasionally with bands or laminations, and frequently with visible sponge spicules.
 Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being \~20 cm, and frequently characterized by high levels of dolomitization and fractures.
-Petrographically, they are composed mainly of microcrystalline quartz, with wackestone packstone textures, small percentages of macrocrystalline quartz, fibrous chalcedony and allochems composed of oxides and poorly preserved bioclasts.
-When identifiable, the bioclasts are Echinoderms, Radiolarians, Sponge spicules and bivalve shells.
+Petrographically, they are composed mainly of microcrystalline quartz, with wackestone or packstone textures, small percentages of macrocrystalline quartz, fibrous chalcedony and allochems composed of oxides and poorly preserved bioclasts.
+When identifiable, the bioclasts are Echinoderms, Radiolarians, Sponge spicules and Bivalve shells.
 The Upper Jurassic cherts of Mareta show similar petrographic characteristics, but with abundant Calcispheres which are not always apparent under the stereomicroscope.
@@
 The Middle Jurassic Malhão formation chert outcrops were identified at around 80-100 km east of the site and characterized by chert nodules with similar macroscopic features to those found in Western Algarve.
-Petrographically they are composed mainly by microcrystalline quartz and characterized by wackestone textures, while dolomite, chalcedony and macrocrystalline quartz are also present in small percentages (\<10%).
+Petrographically they are composed mainly of microcrystalline quartz and characterized by wackestone textures, while dolomite, chalcedony and macrocrystalline quartz are also present in small percentages (\<10%).
 Allochems are iron oxides, and albeit poorly preserved, several bioclasts can be identified: Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres and possibly Tentaculites.
 The Upper Jurassic Jordana formation chert outcrops were identified at around 90 km east of Vale Boi and characterized by chert nodules with significantly different macroscopic features to those of other formations, with frequently small sizes (\~5 cm), difficult removal from the parent rock and unidentified secondary deposits.
-Petrographically they are composed mostly of microcrystalline quartz, with wackstone to packstone textures, and small percentages of fibrous chalcedony and dolomite (\<10%).
+Petrographically they are composed mostly of microcrystalline quartz, with wackestone to packstone textures, and small percentages of fibrous chalcedony and dolomite (\<10%).
 They show frequent iron oxides and very frequent bioclasts, albeit often poorly preserved and unidentifiable; identified fossils are Calcispheres, Bivalve shells, Sponge spicules, Ostracods, Echinoderms and Gastropods.
@@
 The term chalcedony is used in its petrographic sense and refers to cherts with a fibrous quartz structure.
-Since this definition of chalcedony implies a destructive method which cannot be applied to all archaeological chalcedony samples, we use the term based on the confirmed correlation between the petrographic characteristics of chalcedony and their macroscopic appearance, both through geological and archaeological thin sections.
+Since this definition of chalcedony implies a destructive method which cannot be applied to all archaeological chalcedony samples, we use the term based on the confirmed correlation between the macroscopic appearance of chalcedony and its petrographic characteristics, obtained through geological and archaeological thin sections.
 
@@
 For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), and given the inexistence of a reference collection for local altered cherts, they were analysed using the individual sample dataset (Online Resource 1) but grouped in a category of indeterminate cherts (INDET).
-This category included key alterations such as heat/fire related alterations which clearly impacted the macroscopic characteristics of the samples, fully changing its color, luster and texture, and covering them in crazings (e.g., @fig-terrace-alterations a).
-Similarly, it also included alteration rinds that extended through most of the sample, impeding the identification of its original color, texture, translucency, inclusions and fossils (e.g., @fig-terrace-alterations b) and the presence of extended pitting.
+This category included key alterations such as heat/fire-related alterations which clearly impacted the macroscopic characteristics of the samples, fully changing its color, luster and texture, and covering them in crazings (e.g., @fig-terrace-alterations a).
+Similarly, it also included alteration rinds that extended through most of the sample, impeding the identification of its original colour, texture, translucency, inclusions and fossils (e.g., @fig-terrace-alterations b) and the presence of extended pitting.
 However, whenever surface alterations were present in only a limited area of the sample leaving recognizable macroscopic features and fossils, or when key elements were distinguishable to make a characterization (e.g., reddening in the Type 2 cherts, Online Resource 2, figure s5), the samples were not collapsed into the INDET category.
@@
   ~"Chert type", ~"Potential source", ~"Distance from source",
-  "Type 1", "Triassic volcanic formations ", " ",
+  "Type 1", "Triassic volcanic formations ", "<10 km",
   "Type 2-5", "Lower Carixian formations, Southwestern Portugal", "<20 km",
@@
 Consequently, we cannot rule out the possibility that some of the cherts we have identified as being sourced within a 20 km radius of Vale Boi may actually be the result of regional procurement and mobility, potentially from areas approximately 80-100 km away, where the Middle Jurassic cherts are located.
-Given the similarity of the raw materials and the absence of geochemical studies which have shown to be key to differentiate between cherts belonging to the same geological formations or cherts formed under similar conditions [@gomezdesolerMultitechniqueApproachCharacterization2023], and considering the studies and model mentioned above, most of these cherts likely originate from local sources, while a minor portion may be attributable to regional movements.
+Given the similarity of the raw materials and the absence of geochemical studies which have shown to be key to differentiating between cherts belonging to the same geological formations or cherts formed under similar conditions [@gomezdesolerMultitechniqueApproachCharacterization2023], and considering the studies and model mentioned above, most of these cherts likely originate from local sources, while a minor portion may be attributable to regional movements.
 

### 2025-05-06 analysis/paper/paper.qmd
Added tables created in code to derived data folder. Replaced pipes for quarto pipes. No code was altered

@@
 # Concatenate UNIT and ID variables in one single variable (e.g. A6-100)
-context <- context %>%
-  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column from Unit to UNIT to use cols_to_concat
-  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
-  select(ID, Level, Spit, Code) %>% # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
+context <- context |>
+  dplyr::rename("UNIT" = "Unit") |> # Renaming first column from Unit to UNIT to use cols_to_concat
+  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) |> # Unite UNIT with ID using cols_to_concat
+  select(ID, Level, Spit, Code) |> # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
   distinct(ID, .keep_all = TRUE)
 
-xyz <- xyz %>%
-  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
-  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
-  select(ID, X, Y, Z) %>% # Selecting important variables
+xyz <- xyz |>
+  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) |> # Unite UNIT with ID using cols_to_concat
+  filter(Suffix == 0) |> # To avoid several entries related to several points of the same artefact
+  select(ID, X, Y, Z) |> # Selecting important variables
   distinct(ID, .keep_all = TRUE)
@@
 
-T_dataset <- T_dataset %>% 
-  filter(!Level %in% c("2", "3")) %>% # Filtering out Upper levels
-  drop_na(Level) %>% # Dropping artifacts without level
+T_dataset <- T_dataset |> 
+  filter(!Level %in% c("2", "3")) |> # Filtering out Upper levels
+  drop_na(Level) |> # Dropping artifacts without level
   mutate(Level = case_when(Level == "7C" ~ "7",
@@
                            Level == "4B" ~ "4", # collapsing samples from 4B to 4 due to small number
-                           TRUE ~ Level)) %>% 
+                           TRUE ~ Level)) |> 
   mutate(RMTYPE = case_when(RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
@@
 # Mutate level 5 to add phases: top of 5 (4E/5) and bottom of 5 (5), following Belmiro et al. (2020).
-phase_dataset <- T_dataset %>% 
+phase_dataset <- T_dataset |> 
   mutate(Level = case_when(
@@
     Level == 5 & Z < 24.1 ~ "5",
-    TRUE ~ Level)) %>% 
+    TRUE ~ Level)) |> 
   mutate(Technocomplex = case_when(
@@
 
-S_RM_data <- S_RM_data %>%
-  mutate_at("ID", str_replace, "VB", "") %>% # Remove "VB" from ID
+S_RM_data <- S_RM_data |>
+  mutate_at("ID", str_replace, "VB", "") |> # Remove "VB" from ID
   mutate(RMTYPE = case_when(
@@
 
-S_layerdata <- S_layerdata %>% 
-  mutate_at("Date", str_replace, "20", "") %>%
-  dplyr::rename("YEAR" = "Date") %>% 
-  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% 
-  select(ID, Layer) %>% 
-  distinct(ID, .keep_all = TRUE) %>%
+S_layerdata <- S_layerdata |> 
+  mutate_at("Date", str_replace, "20", "") |>
+  dplyr::rename("YEAR" = "Date") |> 
+  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) |> 
+  select(ID, Layer) |> 
+  distinct(ID, .keep_all = TRUE) |>
   filter(!Layer %in% c("D", "Z"))
 
-S_field_data <- S_field_data %>%
-  mutate_at("YEAR", str_replace, "20", "") %>%
-  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% # Unite YEAR with Id s«using cols_to_concat
-  select(ID, X, Y, Z, CODE) %>%
-  filter(CODE == "L") %>% # Selecting important variables
+S_field_data <- S_field_data |>
+  mutate_at("YEAR", str_replace, "20", "") |>
+  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) |> # Unite YEAR with Id s«using cols_to_concat
+  select(ID, X, Y, Z, CODE) |>
+  filter(CODE == "L") |> # Selecting important variables
   distinct(ID, .keep_all = TRUE)
@@
 S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID")
-S_dataset <- S_dataset %>% 
-  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
+S_dataset <- S_dataset |> 
+  mutate_at(vars(Layer), ~replace_na(., "ND")) |> 
   filter(!Layer == "ND")
@@
 ## Data preparation
-context_comp <- field_data %>% # Using the field data to obtain lithic info beyond the analysed chert/chalcedony
+context_comp <- field_data |> # Using the field data to obtain lithic info beyond the analysed chert/chalcedony
   mutate(Level = case_when(Level == "4B" ~ "4", # Collapsing samples from level 4B into 4
@@
                            Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7
-                           TRUE ~ Level)) %>% 
+                           TRUE ~ Level)) |> 
   mutate(Level = case_when( # Dividing 5 as previous
@@
     Level == 5 & Z < 24.1 ~ "5",
-    TRUE ~ Level)) %>% 
-  filter(!Level %in% c("3B", "5B", "5C")) %>% 
+    TRUE ~ Level)) |> 
+  filter(!Level %in% c("3B", "5B", "5C")) |> 
   dplyr::filter(Code %in% c("LITHIC", "POINT")) # Field data includes ALL artifacts; filtering only lithics
 
-dataset_comp <- phase_dataset %>% # Using the analysed DB to get chert/chalcedony data
+dataset_comp <- phase_dataset |> # Using the analysed DB to get chert/chalcedony data
   select(ID, RMTYPE) # Selecting only ID (to join dbs and RMTYPE to know whether chert or chalcedont)
@@
 
-T_comp <- T_comp %>% 
+T_comp <- T_comp |> 
  mutate(Rawmaterial = case_when(
@@
     RMTYPE == "TYPE 1" ~ "Chalcedony", # Naming Chalcedony
-    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category (All but Other and TYPE1)
+    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) |> # Collapsing all cherts into one single category (All but Other and TYPE1)
   dplyr::filter(Level %in% c("4", "4C", "4D", "4E", "4E/5","5", "6", "7")) # Choosing only UP levels (excluding 3)
@@
 ## Barplot
-T_comp_bar <- T_comp %>%
-  dplyr::group_by(Level, Rawmaterial) %>% 
-  dplyr::summarise(n = n()) %>%
-  mutate(percent = n/sum(n)) %>% 
+T_comp_bar <- T_comp |>
+  dplyr::group_by(Level, Rawmaterial) |> 
+  dplyr::summarise(n = n()) |>
+  mutate(percent = n/sum(n)) |> 
   ggplot(aes(x=Level, y = percent, fill = Rawmaterial)) +
@@
 S_field_comp <- S_field_data
-S_data_comp <- S_dataset %>% 
+S_data_comp <- S_dataset |> 
   dplyr::select(ID, RMTYPE)
 S_comp <- full_join(S_field_comp, S_data_comp, by = "ID")
-S_layerdata <- S_layerdata %>% 
+S_layerdata <- S_layerdata |> 
   distinct() # Keeping only individual IDs to avoid repetitions
@@
 
-S_comp <- S_comp %>% 
+S_comp <- S_comp |> 
  mutate(Rawmaterial = case_when(
@@
     RMTYPE == "TYPE 1" ~ "Chalcedony",
-    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category
+    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) |> # Collapsing all cherts into one single category
   dplyr::mutate(Layer = case_when(
     Layer == "c" ~ "C",
-    TRUE ~ Layer)) %>% 
+    TRUE ~ Layer)) |> 
   filter(Layer %in% c("A", "B", "C")) # Choosing only Solutrean occupation layers (A, B and C)
@@
 ## Barplot
-S_comp_bar <- S_comp %>%
-  dplyr::group_by(Layer, Rawmaterial) %>% 
-  dplyr::summarise(n = n()) %>%
-  mutate(percent = n/sum(n)) %>% 
+S_comp_bar <- S_comp |>
+  dplyr::group_by(Layer, Rawmaterial) |> 
+  dplyr::summarise(n = n()) |>
+  mutate(percent = n/sum(n)) |> 
   ggplot(aes(x=Layer, y = percent, fill = Rawmaterial)) +
@@
 ## Unused barplot: here to understand how the chert ratio changes when A, B and C are collapsed since "No info" has most of the other rms.
-#S_comp %>% 
-#  dplyr::group_by(Rawmaterial) %>% 
-#  dplyr::summarise(n = n()) %>%
-#  mutate(percent = n/sum(n)) %>% 
+#S_comp |> 
+#  dplyr::group_by(Rawmaterial) |> 
+#  dplyr::summarise(n = n()) |>
+#  mutate(percent = n/sum(n)) |> 
 #  ggplot(aes(x=Rawmaterial, y = percent, fill = Rawmaterial)) +
@@
 
-terrace_indet_table <- phase_dataset %>% 
-  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>%
-  tabyl(Group,Level) %>% 
-  adorn_totals("row") %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
+terrace_indet_table <- phase_dataset |> 
+  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) |>
+  tabyl(Group,Level) |> 
+  adorn_totals("row") |>
+  adorn_percentages("col") |>
+  adorn_pct_formatting(2) |>
   adorn_ns("front")
 
-shelter_indet_table <- S_dataset %>% 
-  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
-  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
-  filter(!Layer %in% c("D", "ND")) %>% 
-  tabyl(Group,Layer) %>% 
-  adorn_totals("row") %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
+shelter_indet_table <- S_dataset |> 
+  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) |> 
+  mutate_at(vars(Layer), ~replace_na(., "ND")) |> 
+  filter(!Layer %in% c("D", "ND")) |> 
+  tabyl(Group,Layer) |> 
+  adorn_totals("row") |>
+  adorn_percentages("col") |>
+  adorn_pct_formatting(2) |>
   adorn_ns("front")
@@
 kable(head(indet_table))
+
+#write.csv(indet_table,"analysis/data/derived_data/indet_tbl.csv", row.names = FALSE)
 
@@
 # Results
-
-```{r}
-#| label: t-table
-
-T_table <- phase_dataset %>%
-  filter(RMTYPE !="INDET") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  tabyl(RMTYPE, Level) %>%
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
-  adorn_ns("front")
-
-```
 
@@
 ## Preparing cortex db Terrace
-cortex_tab <- phase_dataset %>% 
-  select(RMTYPE, CORTEX, CORTEXAPPEARANCE, CORTEXTHICKNESS) %>%
-  filter(CORTEX == "YES") %>% # filtering only samples with cortex
-  filter(RMTYPE != "INDET") %>% 
-  filter(RMTYPE != "TYPE 11") %>% # removing Type 11 since it barely has samples
+cortex_tab <- phase_dataset |> 
+  select(RMTYPE, CORTEX, CORTEXAPPEARANCE, CORTEXTHICKNESS) |>
+  filter(CORTEX == "YES") |> # filtering only samples with cortex
+  filter(RMTYPE != "INDET") |> 
+  filter(RMTYPE != "TYPE 11") |> # removing Type 11 since it barely has samples
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
@@
 
-cortex_tab <- cortex_tab %>% # dropping the caps and organizing value order
+cortex_tab <- cortex_tab |> # dropping the caps and organizing value order
   mutate(CORTEXAPPEARANCE = case_when(
@@
     CORTEXAPPEARANCE == "OTHER" | CORTEXAPPEARANCE == "Other" | CORTEXAPPEARANCE == "INDETERMINATE" ~ "Indeterminate",
-    TRUE ~ CORTEXAPPEARANCE)) %>%
+    TRUE ~ CORTEXAPPEARANCE)) |>
   dplyr::mutate(CORTEXTHICKNESS = case_when(
@@
     CORTEXTHICKNESS == "THIN" ~ "Thin",
-    TRUE ~ CORTEXTHICKNESS)) %>% 
-  mutate(CORTEXAPPEARANCE = factor(CORTEXAPPEARANCE, levels = c("Irregular", "Rounded", "Pebble", "Indeterminate"))) %>%
-  mutate(CORTEXTHICKNESS = factor(CORTEXTHICKNESS, levels = c("Thin", "Medium", "Thick"))) %>% 
+    TRUE ~ CORTEXTHICKNESS)) |> 
+  mutate(CORTEXAPPEARANCE = factor(CORTEXAPPEARANCE, levels = c("Irregular", "Rounded", "Pebble", "Indeterminate"))) |>
+  mutate(CORTEXTHICKNESS = factor(CORTEXTHICKNESS, levels = c("Thin", "Medium", "Thick"))) |> 
   drop_na() # removing missing values that may be cortex=NO instead passed as YES
@@
 ## gt summary table
-cortex_tab %>%
-  select(RMTYPE, CORTEXAPPEARANCE, CORTEXTHICKNESS) %>% 
-  tbl_summary(by=RMTYPE, label = list(CORTEXAPPEARANCE ~ "Cortex type", CORTEXTHICKNESS ~ "Thickness")) %>% 
-  modify_header(label ~ "**Variable**") %>% 
-  bold_labels()
+cortex_tbl <- cortex_tab |>
+  select(RMTYPE, CORTEXAPPEARANCE, CORTEXTHICKNESS) |> 
+  tbl_summary(by=RMTYPE, label = list(CORTEXAPPEARANCE ~ "Cortex type", CORTEXTHICKNESS ~ "Thickness")) |> 
+  modify_header(label ~ "**Variable**") |> 
+  bold_labels() |> 
+  as_tibble()
+
+#write.csv(cortex_tbl,"analysis/data/derived_data/cortex_tbl.csv", row.names = FALSE)
 
@@
 
-count_table <- phase_dataset %>%
-  filter(RMTYPE != "INDET") %>% 
+count_table <- phase_dataset |>
+  filter(RMTYPE != "INDET") |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  dplyr::group_by(Level, RMTYPE) %>% 
-  dplyr::summarise(n = n()) %>%
-  mutate(percent = n/sum(n)) %>% 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
+  dplyr::group_by(Level, RMTYPE) |> 
+  dplyr::summarise(n = n()) |>
+  mutate(percent = n/sum(n)) |> 
   mutate(Technocomplex = case_when(
@@
     Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
-    Level == 6 | Level == 7 ~ "Gravettian")) %>% 
+    Level == 6 | Level == 7 ~ "Gravettian")) |> 
   mutate(Technocomplex = factor(Technocomplex, levels = c("Solutrean", "Proto-Solutrean", "Gravettian")))
@@
 
-weight_table <- phase_dataset %>%
-  filter(RMTYPE != "INDET" & WEIGHT != "0") %>% 
+weight_table <- phase_dataset |>
+  filter(RMTYPE != "INDET" & WEIGHT != "0") |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  dplyr::group_by(Level, RMTYPE) %>% 
-  dplyr::summarise(wsum = sum(WEIGHT)) %>% 
-  mutate(percent = wsum/sum(wsum)) %>% 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
+  dplyr::group_by(Level, RMTYPE) |> 
+  dplyr::summarise(wsum = sum(WEIGHT)) |> 
+  mutate(percent = wsum/sum(wsum)) |> 
   mutate(Technocomplex = case_when(
@@
 # Tables f Terrace and Shelter with n(%) of chert types both overall by area and by level. Perhaps going to SOM (?)
-terrace_table <- phase_dataset %>%
-  filter(RMTYPE != "INDET") %>% 
-  select(RMTYPE, Level) %>% 
+terrace_table <- phase_dataset |>
+  filter(RMTYPE != "INDET") |> 
+  select(RMTYPE, Level) |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1", "TYPE 2", "TYPE 3", "TYPE 4", "TYPE 5", "TYPE 6",
-                                            "TYPE 7", "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11","TL"))) %>%
-  tbl_summary(include = c(RMTYPE), by = Level) %>% 
+                                            "TYPE 7", "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11","TL"))) |>
+  tbl_summary(include = c(RMTYPE), by = Level) |> 
   add_overall()
@@
 S_dataset["Layer"][is.na(S_dataset["Layer"])] <- "No info"
-shelter_table <- S_dataset %>% 
-  filter(RMTYPE != "INDET") %>%
-  filter(Layer != "D") %>% 
-  select(RMTYPE, Layer) %>% 
+shelter_table <- S_dataset |> 
+  filter(RMTYPE != "INDET") |>
+  filter(Layer != "D") |> 
+  select(RMTYPE, Layer) |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1", "TYPE 2", "TYPE 3", "TYPE 4", "TYPE 5", "TYPE 6",
-                                            "TYPE 7", "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11","TL"))) %>%
-  tbl_summary(include = c(RMTYPE), by = Layer) %>% 
+                                            "TYPE 7", "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11","TL"))) |>
+  tbl_summary(include = c(RMTYPE), by = Layer) |> 
   add_overall()
@@
 # Weight boxplots terrace
-phase_dataset %>%
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+phase_dataset |>
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(Technocomplex = factor(Technocomplex, levels = c("Solutrean", "Proto-Solutrean", "Gravettian"))) %>%
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
+  mutate(Technocomplex = factor(Technocomplex, levels = c("Solutrean", "Proto-Solutrean", "Gravettian"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
@@
 ## Preparing db for the barplots
-s_data_plots <- S_dataset %>% # sorting missing values and filtering non-solutrean layers
-  filter(RMTYPE != "INDET" & WEIGHT != "0") %>% 
-  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
-  filter(CODE != "B" & Layer != "D" & Layer != "ND" & Layer != "No info") %>% #removing buckets (B)
+s_data_plots <- S_dataset |> # sorting missing values and filtering non-solutrean layers
+  filter(RMTYPE != "INDET" & WEIGHT != "0") |> 
+  mutate_at(vars(Layer), ~replace_na(., "ND")) |> 
+  filter(CODE != "B" & Layer != "D" & Layer != "ND" & Layer != "No info") |> #removing buckets (B)
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
@@
 ## Count barplot
-shelter_count <- s_data_plots %>%
-  dplyr::group_by(Layer, RMTYPE) %>% 
-  dplyr::summarise(n = n()) %>%
-  mutate(percent = n/sum(n)) %>% 
+shelter_count <- s_data_plots |>
+  dplyr::group_by(Layer, RMTYPE) |> 
+  dplyr::summarise(n = n()) |>
+  mutate(percent = n/sum(n)) |> 
   ggplot(aes(x = Layer, y = percent, fill = RMTYPE)) +
@@
 ## Weight barplot
-shelter_weight <- s_data_plots %>%
-  dplyr::group_by(Layer, RMTYPE) %>% 
-  dplyr::summarise(wsum = sum(WEIGHT)) %>% 
-  mutate(percent = wsum/sum(wsum)) %>% 
+shelter_weight <- s_data_plots |>
+  dplyr::group_by(Layer, RMTYPE) |> 
+  dplyr::summarise(wsum = sum(WEIGHT)) |> 
+  mutate(percent = wsum/sum(wsum)) |> 
   ggplot(aes(x = Layer, y = percent, fill = RMTYPE)) +
@@
 ## Weight boxplot
-shelter_box <- s_data_plots %>% 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","TL") & Layer != "ND") %>%
+shelter_box <- s_data_plots |> 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","TL") & Layer != "ND") |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
